---
title: "Analysis: Evolution of Sovereignty Attitudes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analysis: Evolution of Sovereignty Attitudes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
```

```{r}
library(qesR)
library(dplyr)
library(ggplot2)
library(knitr)
```

This page uses the merged file to estimate sovereignty support over time.

## Load merged data

```{r}
master <- if (file.exists("qes_master.csv")) {
  read.csv("qes_master.csv", stringsAsFactors = FALSE)
} else {
  get_qes_master(assign_global = FALSE, strict = FALSE, quiet = TRUE)
}

master <- master %>%
  mutate(
    qes_year_num = suppressWarnings(as.integer(sub("^([0-9]{4}).*$", "\\1", qes_year)))
  )

# CROP polling waves are useful for other questions, but the sovereignty trend
# below is restricted to election-study files.
master <- master %>%
  filter(qes_code != "qes_crop_2007_2010")

if (!("sovereignty_support" %in% names(master))) {
  master$sovereignty_support <- NA_real_
}
```

## Dedicated sovereignty question

```{r}
master <- master %>%
  mutate(
    sovereignty_final = suppressWarnings(as.numeric(sovereignty_support))
  )

master$sovereignty_final[!(master$sovereignty_final %in% c(0, 1))] <- NA_real_

all_years <- data.frame(
  qes_year_num = sort(unique(master$qes_year_num[!is.na(master$qes_year_num)]))
)

sov <- master %>%
  filter(!is.na(qes_year_num)) %>%
  group_by(qes_year_num) %>%
  summarise(
    respondents = n(),
    n_with_measure = sum(!is.na(sovereignty_final)),
    support_share = ifelse(n_with_measure > 0, mean(sovereignty_final, na.rm = TRUE), NA_real_),
    se = ifelse(
      n_with_measure > 0,
      sqrt(pmax(support_share * (1 - support_share), 0) / n_with_measure),
      NA_real_
    ),
    ci_low = ifelse(!is.na(se), pmax(0, support_share - 1.96 * se), NA_real_),
    ci_high = ifelse(!is.na(se), pmin(1, support_share + 1.96 * se), NA_real_),
    .groups = "drop"
  )

sov <- merge(all_years, sov, by = "qes_year_num", all.x = TRUE, sort = TRUE)
```

### Results table

```{r}
knitr::kable(
  sov %>%
    mutate(
      sovereignty_support_pct = round(100 * support_share, 1),
      ci_low_pct = round(100 * ci_low, 1),
      ci_high_pct = round(100 * ci_high, 1)
    ) %>%
    transmute(
      `Study year` = qes_year_num,
      `N respondents` = respondents,
      `N with sovereignty item` = n_with_measure,
      `Sovereignty support (%)` = sovereignty_support_pct,
      `95% CI (%)` = ifelse(is.na(ci_low_pct), NA_character_, sprintf("%.1f to %.1f", ci_low_pct, ci_high_pct))
    ),
  caption = "Sovereignty support by study year"
)
```

### Trend plot

```{r, fig.alt="Line chart of sovereignty support trend over study years."}
plot_sov <- sov %>%
  filter(!is.na(support_share), !is.na(ci_low), !is.na(ci_high))

ggplot(plot_sov, aes(x = qes_year_num, y = support_share)) +
  geom_ribbon(aes(ymin = ci_low, ymax = ci_high), fill = "#9fb6ce", alpha = 0.35, na.rm = TRUE) +
  geom_line(linewidth = 1, color = "#12355b", na.rm = TRUE) +
  geom_point(size = 2.3, color = "#b5651d", na.rm = TRUE) +
  geom_smooth(
    method = "loess",
    se = TRUE,
    span = 0.9,
    color = "#12355b",
    fill = "#c9d6e4",
    alpha = 0.25,
    linetype = "dashed",
    linewidth = 0.8,
    na.rm = TRUE
  ) +
  scale_y_continuous(labels = function(x) paste0(round(100 * x, 0), "%")) +
  scale_x_continuous(breaks = sort(unique(sov$qes_year_num))) +
  labs(
    title = "Sovereignty support across QES studies",
    x = "Study year",
    y = "Share of respondents"
  ) +
  theme_minimal(base_size = 12)
```

## Notes

- The sovereignty trend uses the dedicated sovereignty item from each study.
- Confidence intervals are binomial 95% intervals using normal approximation.
- `n_with_measure` in the table shows where coverage is thinner.
