---
title: "Citing qesR and the studies"
output: rmarkdown::html_vignette
params:
  lang: "en"
vignette: >
  %\VignetteIndexEntry{Citing qesR and the studies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

*[Version française](fr-citations.html)*

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# messages in the language of the page, whatever the locale of the build
options(qesR.lang = params$lang)
```

When you publish results computed with qesR, cite both the package and each
dataset you used. `qes_cite()` writes these citations from the catalog that
ships with qesR, so this page needs no network access. The last section
gives the licence of each study and the attribution the 2022 study
requires.

```{r load}
library(qesR)
```

## Citing qesR

```{r package}
qes_cite()
```

`citation("qesR")` gives the same reference.

## Citing the datasets

Each study code reads one version of one Dataverse deposit. The citation
gives the authors, the year, the deposit title as published, the DOI, the
repository, the version and the dataset's UNF (a checksum of its data). The
three 1998 surveys share one deposit, so their citations also name the data
file.

```{r table, echo = FALSE, warning = FALSE}
studies <- qes_studies()
title <- if (identical(params$lang, "fr")) studies$title_fr else studies$title_en
table <- data.frame(
  study = paste0("`", studies$study, "`"),
  year = ifelse(
    studies$year_end == studies$year,
    studies$year,
    paste0(studies$year, "-", studies$year_end)
  ),
  title = title,
  licence = studies$licence,
  citation = qes_cite(studies$study, lang = params$lang)[-1]
)
knitr::kable(table, col.names = if (identical(params$lang, "fr")) {
  c("Code", "Année", "Étude", "Licence", "Citation")
} else {
  c("Code", "Year", "Study", "Licence", "Citation")
})
```

## Citing only what you used

Pass study codes, or the data returned by `get_qes()`, to cite exactly the
datasets in an analysis:

```{r subset}
qes_cite(c("qes2018", "qes2022"))
```

```{r loaded, eval = FALSE}
qes2018 <- get_qes("qes2018")
qes_cite(qes2018)
```

For a reference manager, ask for BibTeX:

```{r bibtex}
cat(qes_cite("qes2018", style = "bibtex"), sep = "\n\n")
```

## Licences and attribution

The data are not part of qesR: it downloads them from Borealis and the
Harvard Dataverse. Most studies are released under CC0 1.0 (public domain).
The 2022 study is released under CC BY-NC 4.0, which requires attribution
and rules out commercial use. `qes_studies()$licence` gives the licence of
each study.

The MIT licence of qesR covers the package code only. The metadata of the
2022 study that qesR ships (labels, question text, answer counts, and the
harmonization wording, labels and counts derived from them) are derived
from Mahéo, Bélanger, Stephenson and Harell (2023), *2022 Quebec Election
Study*, Harvard Dataverse, V1.1, <https://doi.org/10.7910/DVN/PAQBDR>, and
keep its licence,
[CC BY-NC 4.0](https://creativecommons.org/licenses/by-nc/4.0/):
attribution, no commercial use. This does not imply that the authors
endorse qesR. Cite the study when you use this metadata, as when you use
its data. A printed codebook of `qes2022` repeats this notice, and
codebooks, `qes_question()` and `qes_search()` results that include
`qes2022` keep it in their attribute `licence_notice`.

The metadata of the other studies are CC0 1.0, and the census counts come
from Statistics Canada (Statistics Canada Open Licence).
`system.file("COPYRIGHTS", package = "qesR")` lists each file of the
package, its source and its licence.
