---
title: "Study Citations"
output: rmarkdown::html_vignette
params:
  lang: "en"
vignette: >
  %\VignetteIndexEntry{Study Citations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

*[Version française](fr-citations.html)*

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# messages in the language of the page, whatever the locale of the build
options(qesR.lang = params$lang)
```

When you publish results computed with qesR, cite both the package and each
dataset you used. `qes_cite()` writes these citations from the catalog that
ships with qesR, so this page needs no network access. The French version of
this page is `vignette("fr-citations", package = "qesR")`.

```{r load}
library(qesR)
```

## Citing qesR

```{r package}
qes_cite()
```

`citation("qesR")` gives the same reference.

## Citing the datasets

Each study code points to one Dataverse deposit, pinned to a dataset version.
The citation gives the authors, the year, the deposit title as published, the
DOI, the repository, the version and the dataset's UNF (a checksum of its
data). The three 1998 surveys share one deposit, so their citations also name
the data file.

```{r table, echo = FALSE, warning = FALSE}
studies <- qes_studies()
title <- if (identical(params$lang, "fr")) studies$title_fr else studies$title_en
table <- data.frame(
  study = paste0("`", studies$study, "`"),
  year = ifelse(
    studies$year_end == studies$year,
    studies$year,
    paste0(studies$year, "-", studies$year_end)
  ),
  title = title,
  licence = studies$licence,
  citation = qes_cite(studies$study, lang = params$lang)[-1]
)
knitr::kable(table, col.names = if (identical(params$lang, "fr")) {
  c("Code", "Année", "Étude", "Licence", "Citation")
} else {
  c("Code", "Year", "Study", "Licence", "Citation")
})
```

The 2022 study is released under CC BY-NC 4.0, which requires attribution and
rules out commercial use; the other studies are CC0 (public domain). The
licence of each study is in `qes_studies()$licence`. The metadata of the 2022
study that qesR ships (its codebook, question text, value labels and counts)
carries the same licence and needs the same citation.

## Citing only what you used

Pass study codes, or the data returned by `get_qes()`, to cite exactly the
datasets in an analysis:

```{r subset}
qes_cite(c("qes2018", "qes2022"))
```

```{r loaded, eval = FALSE}
qes2018 <- get_qes("qes2018")
qes_cite(qes2018)
```

For a reference manager, ask for BibTeX:

```{r bibtex}
cat(qes_cite("qes2018", style = "bibtex"), sep = "\n\n")
```
