---
title: "Getting Started with qesR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with qesR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

*[Version française](fr-demarrage.html)*

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`qesR` helps you access Quebec Election Study data by survey code, plus metadata, codebooks, and a merged harmonized dataset.

## Install

```{r, eval = FALSE}
if (!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")
remotes::install_github("ThomasGareau/qesR")
```

```{r}
library(qesR)
```

## Using the package

Core calls:

- `get_qes(srvy)` download a survey by code
- `qes_studies()` list the available studies (offline catalog)
- `qes_codebook(srvy)` retrieve the codebook (offline)
- `qes_question(srvy, variables)` the exact wording of questions, EN or FR
- `qes_search(pattern)` find variables across studies, EN and FR
- `qes_missing(data)` set "don't know" and "refused" codes to `NA`
- `get_qes_master()` build merged harmonized data
- `qes_download(studies, path)` save the original files, md5-checked
- `qes_provenance(data)` record which file the data came from

Every function returns its result. Nothing is written into your workspace
unless you ask for it with `assign_global = TRUE`, so assign the result
yourself: `qes2022 <- get_qes("qes2022")`.

## List available studies

```{r}
studies <- qes_studies()
studies[, c("study", "year", "title_en", "title_fr")]
```

## Download one study

```{r, eval = FALSE}
qes2022 <- get_qes("qes2022")
names(qes2022)[1:20]
```

## Codebooks, questions and search

```{r, eval = FALSE}
cb <- qes_codebook("qes2014")
cb_long <- qes_codebook("qes2014", layout = "long", variables = c("Q2", "Q19"))
qes_question("qes2014", "Q19", lang = "en")
qes_search("souverain|sovereign")
qes2014 <- qes_missing(get_qes("qes2014"))
```

## Keep the original files and their provenance

`qes_download()` saves a study's original files in a folder you choose: the
data file as its authors deposited it and, with `what = "docs"`, the
codebooks, questionnaires and reports. Each file is checked against the md5
checksum in the qesR catalog before it gets its final name, and nothing is
written outside that folder. `qes_provenance()` records which file a data
frame was read from (DOI, dataset version, file, md5, when and how it was
retrieved); printed, it gives a paragraph for a replication log.

```{r, eval = FALSE}
dir.create("originals")
qes_download("qes2018", path = "originals", what = c("data", "docs"))
qes2018 <- get_qes("qes2018")
qes_provenance(qes2018)
qes_cite(qes2018)
```

## Build the legacy master dataset

```{r, eval = FALSE}
master <- get_qes_master(
  surveys = c("qes2022", "qes2018", "qes2014", "qes2007", "qes1998"),
  strict = FALSE
)

dim(master)
head(master)
```

## Details

- Surveys are loaded in labelled form with available variable/value labels.
- Codebook metadata can be inspected in compact, wide, or long layout.
- The merged data keeps the harmonized columns of qesR 0.4.4, every respondent of every file, and the source of every column (`attr(master, "source_map")`); values verified to be wrong are `NA` (`attr(master, "legacy_na_columns")`).

## Messages and errors

Messages, warnings and errors can be shown in French with
`options(qesR.lang = "fr")`; the data returned is the same in both
languages. Errors have classes, so a script can react to them without
reading their text:

```{r}
tryCatch(
  get_qes("QES 2022"),
  qesR_error_unknown_study = function(e) e$suggestions
)
```

The older helpers (`get_codebook()`, `get_question()`, `get_preview()`, ...)
still work and print a one-time notice naming their replacement
(`?qesR-deprecated`).

## Save the master dataset

```{r, eval = FALSE}
get_qes_master(save_path = file.path(tempdir(), "qes_master.csv"), strict = FALSE)
get_qes_master(save_path = file.path(tempdir(), "qes_master.rds"), strict = FALSE)
```
