---
title: "Getting Started with qesR"
output: rmarkdown::html_vignette
params:
  lang: "en"
vignette: >
  %\VignetteIndexEntry{Getting Started with qesR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

*[Version française](fr-demarrage.html)*

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# messages in the language of the page, whatever the locale of the build
options(qesR.lang = params$lang)
```

qesR loads the Quebec Election Studies and related Quebec election surveys
into R by study code. Each code points to one Dataverse deposit, pinned to a
dataset version and to one original data file that is checked against its
md5 checksum before use. The catalog of studies, their documents, their
citations and the description of every variable of the studies released
under CC0 ship with the package.

This page goes from a study code to a weighted estimate. Every chunk on it
runs without a network connection: it uses the catalog, the shipped metadata
and `qes_demo`, a small synthetic study that ships with qesR. The chunks
that would download a real study are shown but not run.

## Install

```{r install, eval = FALSE}
install.packages("qesR")
# or the development version:
# remotes::install_github("ThomasGareau/qesR")
```

```{r load}
library(qesR)
```

## Which studies are there?

`qes_studies()` lists every study, with its year, family, design, target
population, licence, DOI and the dataset version it is pinned to.

```{r studies}
studies <- qes_studies()
studies[, c("study", "year", "family", "study_design", "licence")]
```

Not every study is a Quebec Election Study (family `qes`): the Durand panels,
the CROP polls and the 1998 polls have their own designs and populations.
`qes_docs()` lists the codebooks, questionnaires and reports of each study.

```{r docs}
qes_docs("qes2014")[, c("role", "lang", "file_name")]
```

## Load a study

`get_qes()` returns the data. It writes nothing into your workspace, so
assign the result yourself. The demonstration study is read from the
package; a real study is downloaded once per session (or kept between
sessions with `options(qesR.cache = "disk")`).

```{r demo}
demo <- get_qes("qes_demo", quiet = TRUE)
dim(demo)
head(demo[, c("QSEXE", "Q2", "Q3", "Q19")])
```

```{r real, eval = FALSE}
qes2014 <- get_qes("qes2014")
```

Columns keep the codes and labels of the file (`haven` labelled columns).
Codes such as "don't know" and "refused" are values until you ask
`qes_missing()` to set them to `NA`; it records what it changed.

```{r missing}
demo_clean <- qes_missing(demo, quiet = TRUE)
attr(demo_clean, "qes_missing_log")
```

## Codebooks, questions and search, offline

The codebook of every study released under CC0 ships with qesR: variable
labels, question text in English and French from the deposited
questionnaires, value labels and missing codes.

```{r codebook}
cb <- qes_codebook("qes2014")
head(cb[, c("variable", "label", "n_value_labels")])
qes_codebook("qes2014", layout = "long", variables = "Q19")
```

`qes_question()` gives the exact wording, in either language:

```{r question}
qes_question("qes2014", "Q19", lang = "en")$question
qes_question("qes2014", "Q19", lang = "fr")$question
```

`qes_search()` searches every study at once, ignoring case and accents:

```{r search}
hits <- qes_search("souverain|sovereign")
head(hits[, c("study", "variable", "label")])
```

The 2022 study is licensed CC BY-NC 4.0, so qesR ships none of its
metadata: its codebook is built from your own copy of the data file the
first time you ask for it.

## A weighted estimate

What share of Quebec adults would have voted YES in a referendum on
independence, just after the 2014 election? Question Q19 of `qes2014` asks
it (its wording is above). A correct estimate states four things, and the
steps above give each of them.

1. **The population.** The catalog gives each study's design and target
   population: `qes2014` is a post-election survey that targets the Quebec
   adult population, nothing wider. Its respondents come from an opt-in
   online panel, not from a probability sample.
2. **The weight.** `POND` is the weight of `qes2014`. Its technical report
   (`qes_docs("qes2014")`) says it adjusts the sample to the latest census
   by sex, age, region and language. The estimate describes the adult
   population only to the extent that this adjustment corrects the panel.

```{r population}
studies[studies$study == "qes2014",
        c("study", "study_design", "target_population_en", "target_population_fr")]
qes_codebook("qes2014", variables = "POND")[, c("variable", "label")]
```

3. **The question and its codes.** Q19 codes YES as 1, NO as 2, "don't
   know" as 8 and "refused" as 9 (the long codebook above).
4. **The denominator.** `qes_missing()` set codes 8 and 9 to `NA`. The share
   below is among respondents who answered YES or NO, and says so.

On the demonstration study, which copies these names and codes:

```{r estimate}
answered <- demo_clean$Q19 %in% c(1, 2)
yes <- demo_clean$Q19[answered] == 1
weight <- demo_clean$POND[answered]
c(
  respondents = sum(answered),
  unweighted = mean(yes),
  weighted = sum(weight * yes) / sum(weight)
)
```

The demonstration data are synthetic, so these numbers describe no
population; they show the computation. On the real file the same lines give
34.8% YES among the 1,353 respondents who answered YES or NO (34.1%
unweighted), weighted to the Quebec adult population of 2014.

```{r estimate-real, eval = FALSE}
qes2014 <- qes_missing(get_qes("qes2014"))
answered <- qes2014$Q19 %in% c(1, 2)
yes <- qes2014$Q19[answered] == 1
weight <- qes2014$POND[answered]
sum(weight * yes) / sum(weight)
```

This is a point estimate. `qes2014` is a non-probability online panel: its
technical report (`qes_docs("qes2014")`) gives no margin of error, and a
design-based standard error, from the `survey` package for instance, would
not have its usual meaning. Compute each estimate within one study: studies differ in design,
population and wording, and their weights are on their own scales.

## Harmonized data across studies (experimental)

`qes_harmonize()` builds one data frame from several studies, one column
per harmonized variable ("target"), from a reviewed specification only.
`qes_spec()` shows which question of each study feeds a target and how
comparable it is to the target's anchor question:

```{r spec}
xw <- qes_spec("crosswalk", targets = "sov_indep", lang = params$lang)
xw[, c("study", "wave", "source_var", "grade", "weight_var")]
```

The rows of this first specification are checked against the original
files but not yet signed off by a reviewer, so they are applied only with
`include_draft = TRUE`. On the demonstration study, which stands in for
`qes2014`:

```{r harmonize}
h <- qes_harmonize("qes_demo", targets = c("sov_indep", "vote_prov_recall"),
                   missing = "reasons", include_draft = TRUE, quiet = TRUE,
                   lang = params$lang)
h[1:4, c("study", "eligible_voter", "sov_indep", "sov_indep__na", "weight_post")]
attr(h, "qes_weight_guide")[, c("target", "study", "weight_column", "weight_var")]
```

Every missing value has a reason (`sov_indep__na`). `eligible_voter` says
whether the respondent could vote in the election, and each wave's
recommended weight has mean 1: these questions were asked after the
election, so `weight_post` is their weight, as the weight guide says.
`qes_design()` hands the data to the `survey` package:

```{r design}
if (requireNamespace("survey", quietly = TRUE)) {
  d <- qes_design(h, weight = "weight_post")
  survey::svymean(~sov_indep, d, na.rm = TRUE)
}
```

The standard error treats the weighted sample as a probability sample; for
an opt-in panel such as qes2014, read it as a rough guide only (see
`?qes_design`).

A pre-election question (vote intention) takes `weight_pre` instead, and a
study's two waves can have different weights: `qes_design()` asks you to
choose when the targets need different ones. The reference of every target,
generated from the specification, is
`vignette("harmonization-reference", package = "qesR")`.

## Provenance and citations

Data returned by `get_qes()` records which file it was read from.
`qes_provenance()` returns that record, and `qes_cite()` cites qesR and the
datasets you used (see `vignette("citations", package = "qesR")`).

```{r provenance}
qes_provenance(demo)
qes_cite("qes2014")
```

`qes_download()` saves the original files themselves (data file and
documents), md5-checked, in a folder you choose:

```{r download, eval = FALSE}
dir <- file.path(tempdir(), "qes_originals")
dir.create(dir)
qes_download("qes2014", path = dir, what = c("data", "docs"))
```

## Messages and errors

Messages, warnings and errors can be shown in French with
`options(qesR.lang = "fr")`; the data returned is the same in both
languages. Errors have classes, so a script can react to them without
reading their text:

```{r errors}
tryCatch(
  get_qes("QES 2014"),
  qesR_error_unknown_study = function(e) e$suggestions
)
```

## Coming from qesR 0.4.4

`get_qes_master()` and the other functions of qesR 0.4.4 keep working. What
changed, and how to keep results reproducible, is in
`vignette("migrating-0.5", package = "qesR")`.
