---
title: "Getting Started with qesR"
output: rmarkdown::html_vignette
params:
  lang: "en"
vignette: >
  %\VignetteIndexEntry{Getting Started with qesR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

*[Version française](fr-demarrage.html)*

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# messages in the language of the page, whatever the locale of the build
options(qesR.lang = params$lang)
```

qesR loads the Quebec Election Studies and related Quebec election surveys
into R by study code. Each code points to one Dataverse deposit, pinned to a
dataset version and to one original data file that is checked against its
md5 checksum before use. The catalog of studies, their documents, their
citations and the description of every variable of the studies released
under CC0 ship with the package.

Every chunk on this page runs without a network connection: it uses the
catalog, the shipped metadata and `qes_demo`, a small synthetic study that
ships with qesR. The chunks that would download a real study are shown but
not run.

## Install

```{r install, eval = FALSE}
install.packages("qesR")
# or the development version:
# remotes::install_github("ThomasGareau/qesR")
```

```{r load}
library(qesR)
```

## Which studies are there?

`qes_studies()` lists every study, with its year, family, design, target
population, licence, DOI and the dataset version it is pinned to.

```{r studies}
studies <- qes_studies()
studies[, c("study", "year", "family", "study_design", "licence")]
```

Not every study is a Quebec Election Study (family `qes`): the Durand panels,
the CROP polls and the 1998 polls have their own designs and populations.
`qes_docs()` lists the codebooks, questionnaires and reports of each study.

```{r docs}
qes_docs("qes2014")[, c("role", "lang", "file_name")]
```

## Load a study

`get_qes()` returns the data. It writes nothing into your workspace, so
assign the result yourself. The demonstration study is read from the
package; a real study is downloaded once per session (or kept between
sessions with `options(qesR.cache = "disk")`).

```{r demo}
demo <- get_qes("qes_demo", quiet = TRUE)
dim(demo)
head(demo[, c("QSEXE", "Q2", "Q3", "Q19")])
```

```{r real, eval = FALSE}
qes2014 <- get_qes("qes2014")
```

Columns keep the codes and labels of the file (`haven` labelled columns).
Codes such as "don't know" and "refused" are values until you ask
`qes_missing()` to set them to `NA`; it records what it changed.

```{r missing}
demo_clean <- qes_missing(demo, quiet = TRUE)
attr(demo_clean, "qes_missing_log")
```

## Codebooks, questions and search, offline

The codebook of every study released under CC0 ships with qesR: variable
labels, question text in English and French from the deposited
questionnaires, value labels and missing codes.

```{r codebook}
cb <- qes_codebook("qes2014")
head(cb[, c("variable", "label", "n_value_labels")])
qes_codebook("qes2014", layout = "long", variables = "Q19")
```

`qes_question()` gives the exact wording, in either language:

```{r question}
qes_question("qes2014", "Q19", lang = "en")$question
qes_question("qes2014", "Q19", lang = "fr")$question
```

`qes_search()` searches every study at once, ignoring case and accents:

```{r search}
hits <- qes_search("souverain|sovereign")
head(hits[, c("study", "variable", "label")])
```

The 2022 study is licensed CC BY-NC 4.0, so qesR ships none of its
metadata: its codebook is built from your own copy of the data file the
first time you ask for it.

## Provenance and citations

Data returned by `get_qes()` records which file it was read from.
`qes_provenance()` returns that record, and `qes_cite()` cites qesR and the
datasets you used (see `vignette("citations", package = "qesR")`).

```{r provenance}
qes_provenance(demo)
qes_cite("qes2014")
```

`qes_download()` saves the original files themselves (data file and
documents), md5-checked, in a folder you choose:

```{r download, eval = FALSE}
dir <- file.path(tempdir(), "qes_originals")
dir.create(dir)
qes_download("qes2014", path = dir, what = c("data", "docs"))
```

## Messages and errors

Messages, warnings and errors can be shown in French with
`options(qesR.lang = "fr")`; the data returned is the same in both
languages. Errors have classes, so a script can react to them without
reading their text:

```{r errors}
tryCatch(
  get_qes("QES 2014"),
  qesR_error_unknown_study = function(e) e$suggestions
)
```

## Coming from qesR 0.4.4

`get_qes_master()` and the other functions of qesR 0.4.4 keep working. What
changed, and how to keep results reproducible, is in
`vignette("migrating-0.5", package = "qesR")`.
