---
title: "Getting Started with qesR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with qesR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`qesR` helps you access Quebec Election Study data by survey code, plus metadata, codebooks, and a merged harmonized dataset.

## Install

```{r, eval = FALSE}
if (!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")
remotes::install_github("ThomasGareau/qesR")
```

```{r}
library(qesR)
```

## Using the package

Core calls:

- `get_qes(srvy)` download a survey by code
- `qes_studies()` list the available studies (offline catalog)
- `get_question(data, variable)` retrieve survey question text
- `qes_codebook(srvy)` retrieve codebook
- `get_qes_master()` build merged harmonized data

Every function returns its result. Nothing is written into your workspace
unless you ask for it with `assign_global = TRUE`, so assign the result
yourself: `qes2022 <- get_qes("qes2022")`.

## List available studies

```{r}
studies <- qes_studies()
studies[, c("study", "year", "title_en", "title_fr")]
```

## Download one study

```{r, eval = FALSE}
qes2022 <- get_qes("qes2022")
names(qes2022)[1:20]
```

## Retrieve the codebook in different layouts

```{r, eval = FALSE}
cb_compact <- qes_codebook("qes2022", layout = "compact")
cb_wide <- qes_codebook("qes2022", layout = "wide")
cb_long <- qes_codebook("qes2022", layout = "long")
```

## Get a preview and question text

```{r, eval = FALSE}
head(qes2022, 10)
get_question(qes2022, "cps_age_in_years")
get_question(qes2022, "cps_age_in_years", full = TRUE)
```

## Build a harmonized master dataset

```{r, eval = FALSE}
master <- get_qes_master(
  surveys = c("qes2022", "qes2018", "qes2014", "qes2007", "qes1998"),
  strict = FALSE
)

dim(master)
head(master)
```

## Details

- Surveys are loaded in labelled form with available variable/value labels.
- Codebook metadata can be inspected in compact, wide, or long layout.
- Merged data includes harmonized variables, de-duplication, and source-map tracing.

## Messages and errors

Messages, warnings and errors can be shown in French with
`options(qesR.lang = "fr")`; the data returned is the same in both
languages. Errors have classes, so a script can react to them without
reading their text:

```{r}
tryCatch(
  get_qes("QES 2022"),
  qesR_error_unknown_study = function(e) e$suggestions
)
```

The older helpers `get_codebook()`, `get_qes_codebook()` and `get_preview()`
still work and print a one-time notice naming their replacement
(`?qesR-deprecated`).

## Save the master dataset

```{r, eval = FALSE}
get_qes_master(save_path = file.path(tempdir(), "qes_master.csv"), strict = FALSE)
get_qes_master(save_path = file.path(tempdir(), "qes_master.rds"), strict = FALSE)
```
