---
title: "Merged Dataset Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Merged Dataset Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`qesR` provides a merged harmonized dataset through `get_qes_master()`.

## Build merged data

```{r}
library(qesR)

master <- get_qes_master(strict = FALSE)
dim(master)
head(master)
```

## What it harmonizes

The merged data includes cross-study fields such as:

- `respondent_id`
- `age`, `age_group`, `year_of_birth`
- `gender`, `education`, `language`
- `turnout`, `vote_choice`, `party_lean`
- `survey_weight`

It also records source-variable provenance:

```{r}
source_map <- attr(master, "source_map")
head(source_map)
```

## Quality controls in merge

`get_qes_master()` applies:

- within-year respondent de-duplication for multi-study years
- derived age groups where age or year-of-birth are available
- removal of rows empty across harmonized variables

## Save merged data

```{r}
get_qes_master(save_path = "qes_master.csv", strict = FALSE)
get_qes_master(save_path = "qes_master.rds", strict = FALSE)
```
