---
title: "Merged Dataset Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Merged Dataset Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

*[Version française](fr-donnees-fusionnees.html)*

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`qesR` provides a merged harmonized dataset through `get_qes_master()`.

```{r}
library(qesR)
qes_studies()$study
```

By default, and with `surveys = "all"`, `get_qes_master()` builds the 11 studies of qesR 0.4.4. The 1998 firm files `qes1998_crop` and `qes1998_createc` are not in the master yet (their respondents are already in `qes1998`); read them with `get_qes()`, as `?get_qes_master` explains.

## Build merged data

```{r, eval = FALSE}
master <- get_qes_master(strict = FALSE)
dim(master)
head(master)
```

## What it harmonizes

Harmonized fields include demographics (age, gender, education, income), vote choice, party identification, sovereignty attitudes, leader thermometers, and more. Use `colnames(master)` to see the full list.

`get_qes_master()` also keeps the merge logic explicit:

- uses cross-study variable maps and value normalization for comparability
- derives `age_group` where age or year-of-birth is available
- retains panel and non-panel respondents as separate observations
- renames opaque legacy variables (for example `q10`, `q16`, `voteprec`) to readable merged names
- stores source-variable provenance in `attr(master, "source_map")`
- stores the list of added cross-study variables in `attr(master, "crossstudy_variables_added")`
- stores old-to-new variable renaming in `attr(master, "variable_name_map")`

It also records source-variable provenance:

```{r, eval = FALSE}
source_map <- attr(master, "source_map")
head(source_map)
```

```{r, eval = FALSE}
name_map <- attr(master, "variable_name_map")
head(name_map)
```

## Quality controls in merge

`get_qes_master()` applies:

- respondent de-duplication within the same survey code
- derived age groups where age or year-of-birth are available
- removal of rows empty across harmonized variables

## Save merged data

```{r, eval = FALSE}
get_qes_master(save_path = file.path(tempdir(), "qes_master.csv"), strict = FALSE)
get_qes_master(save_path = file.path(tempdir(), "qes_master.rds"), strict = FALSE)
```
