---
title: "Merged Dataset Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Merged Dataset Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`qesR` provides a merged harmonized dataset through `get_qes_master()`.

## Build merged data

```{r}
library(qesR)

master <- get_qes_master(strict = FALSE)
dim(master)
head(master)
```

## What it harmonizes

Current harmonized columns in `qes_master`:

`qes_code`, `qes_year`, `qes_name_en`, `respondent_id`, `interview_start`, `interview_end`, `interview_recorded`, `language`, `citizenship`, `year_of_birth`, `age`, `age_group`, `gender`, `province_territory`, `education`, `income`, `religion`, `born_canada`, `political_interest`, `ideology`, `turnout`, `vote_choice`, `vote_choice_text`, `party_best`, `party_lean`, `sovereignty_support`, `sovereignty`, `federal_pid`, `provincial_pid`, `survey_weight`, `definibin`, `ethn1`, `influperso`, `luentend`, `occup`, `patron`, `pondam1`, `q10`, `q16`, `q17`, `q18`, `q18a`, `q18b`, `q20`, `q20b`, `q22`, `q23`, `q24`, `q3`, `q31`, `q33`, `q35`, `q37`, `q38`, `q39`, `q4`, `q40`, `q41`, `q42`, `q43`, `q44`, `q45`, `q46`, `q47`, `q48`, `q49`, `q50`, `q51`, `q53`, `q54`, `q55`, `q56`, `q58`, `q59`, `q60`, `q61b`, `q61d`, `q62a`, `q62b`, `q64`, `q65`, `q66`, `q68`, `q7`, `q70`, `q72`, `q73`, `q74`, `q79`, `q8`, `q80`, `q81`, `q9`, `raison1`, `raison2`, `s_jse`, `satisf`, `sefie`, `sondbons`, `voteprec`

`get_qes_master()` also keeps the merge logic explicit:

- uses cross-study variable maps and value normalization for comparability
- derives `age_group` where age or year-of-birth is available
- retains panel and non-panel respondents as separate observations
- stores source-variable provenance in `attr(master, "source_map")`
- stores the list of added cross-study variables in `attr(master, "crossstudy_variables_added")`

It also records source-variable provenance:

```{r}
source_map <- attr(master, "source_map")
head(source_map)
```

## Quality controls in merge

`get_qes_master()` applies:

- respondent de-duplication within the same survey code
- derived age groups where age or year-of-birth are available
- removal of rows empty across harmonized variables

## Save merged data

```{r}
get_qes_master(save_path = "qes_master.csv", strict = FALSE)
get_qes_master(save_path = "qes_master.rds", strict = FALSE)
```
