---
title: "Moving from qesR 0.4.4 to 0.7.0"
output: rmarkdown::html_vignette
params:
  lang: "en"
vignette: >
  %\VignetteIndexEntry{Moving from qesR 0.4.4 to 0.7.0}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

*[Version française](fr-migrer-0.7.html)*

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# messages in the language of the page, whatever the locale of the build
options(qesR.lang = params$lang)
```

This guide is for code written for qesR 0.4.4, such as the replication
scripts of a published paper. Every function of 0.4.4 keeps its name and
its arguments, but one change can stop that code: `get_qes()` no longer
creates the object for you (first row below). What else changes is what
some of it produces, which depends on what the code uses:

| Your code uses | Under 0.7.0 | What to do |
|---|---|---|
| `get_qes("qes2018")` alone, then the object `qes2018` | error "object 'qes2018' not found" | write `qes2018 <- get_qes("qes2018")` ([below](#assign-the-result-yourself)) |
| raw codes from `get_qes()` (`as.numeric()`, ranges of valid codes) | same numbers | nothing |
| label text from `get_qes()` (`haven::as_factor()`, text columns) | some text changes | [check the labels you use](#get_qes-the-same-codes-read-from-the-original-files) |
| `get_qes_master()` | values change | [read the column table](#get_qes_master-what-changed), or keep 0.4.4 to reproduce |
| `get_decon()` | values change; one-time notice | as for the master |
| the other helpers of 0.4.4 | same results as their replacements; one-time notice | nothing, or [rename them](#old-names-new-names) |

Every chunk on this page runs offline, on the synthetic study `qes_demo`.
The complete list of changes is in the package NEWS
(`news(package = "qesR")`), versions 0.5.0 and 0.7.0.

```{r load}
library(qesR)
```

## Before you upgrade: record the version

A published result should say which qesR produced it. Before installing
0.7.0, record the version you have, and keep it with the results:

```{r record, eval = FALSE}
packageVersion("qesR")
packageDescription("qesR")$RemoteSha   # the commit, when installed from GitHub
```

To reproduce a 0.4.4 result exactly, install 0.4.4, preferably in a
separate library or an renv project so that new work can use 0.7.0:

```{r pin, eval = FALSE}
remotes::install_github("ThomasGareau/qesR", ref = "v0.4.4")
```

0.5.0 and 0.6.0 were development versions and were never released. A
result computed with one of them is reproducible by installing the commit
it was built from, which `packageDescription("qesR")$RemoteSha` records for
an installation from GitHub
(`remotes::install_github("ThomasGareau/qesR", ref = "<sha>")`).

## Assign the result yourself

In 0.4.4, `get_qes("qes2018")` created an object `qes2018` in your
workspace. Since 0.5.0 it returns the data and writes nothing: assign it.

```{r assign}
qes_demo <- get_qes("qes_demo", quiet = TRUE)
dim(qes_demo)
```

A script that calls `get_qes("qes2018")` on its own line and then uses
`qes2018` stops with "object 'qes2018' not found". Look also for such calls
inside `tryCatch()` or `try()`: there the error is caught, and the part of
the script that used the object is skipped without a visible error.

`assign_global = TRUE` still assigns, into the environment the function is
called from: the global environment at top level, the function's own frame
inside a function. It still assigns the codebook too, as `<code>_codebook`.

```{r assign-global}
f <- function() {
  get_qes("qes_demo", assign_global = TRUE, quiet = TRUE)
  c(data = exists("qes_demo", inherits = FALSE),
    codebook = exists("qes_demo_codebook", inherits = FALSE))
}
f()
```

## `get_qes()`: the same codes, read from the original files

`get_qes()` reads each study's original SPSS or Stata file, checked against
the md5 in the catalog. Column names, codes and `is.na()` counts are those
of 0.4.4 for all 11 studies, so code that works on codes (`as.numeric()`,
ranges of valid codes, `%in%`) gives the same numbers and the same N. Text
changes where the old conversion damaged it or took labels from elsewhere:

- accents are repaired (1,805 cells in `qes2018`, 2,517 in `qes2022`);
- `qes2012` labels keep their original case and length (0.4.4 lowercased
  them and cut them at 80 characters);
- the `qes2022` interview dates are date-times, not text;
- codes are stored as doubles instead of integers.

`haven::as_factor()` and anything else built from label text can therefore
give different text. The data record which file they came from:

```{r provenance}
qes_provenance(qes_demo)
```

## `get_qes_master()`: what changed

The master keeps its 30 documented columns, in the same order and with the
same types; new columns are appended after them and are never removed.
Its values changed twice.

**In 0.5.0**, mostly by deletion:

- every respondent of every file is kept: 40,987 rows. 0.4.4 dropped 380
  `qes2007_panel` respondents and 1 CROP respondent as duplicates, and, in
  a session with French messages, lost 1,208 `qes2007_panel` and 616
  `qes2018_panel` rows while reading them;
- the 70 columns 0.4.4 appended by stacking variables that share a name
  across studies are removed, because they held different questions in
  different studies (`attr(, "removed_columns")` lists them);
- `qes2018` `turnout` is 0 rather than `NA` for the 336 respondents who
  said they did not vote (`q5` codes 1 and 3): its valid N goes from 2,303
  to 2,639 and its mean from 0.96 to 0.84;
- `qes_name_en` uses each study's own title for the Durand panels, the
  CROP polls and the 1998 polls (for example "Panel Survey on the 2018
  Quebec Election"): select studies with `qes_code`, not with the name;
- label text is fixed: `qes2012` `religion` keeps its original case, CROP
  `income` and `province_territory` read their accented letters correctly
  ("20 000 $ à 39 999 $", "Quebec"), and the `qes2022` `interview_*`
  columns lose their trailing `.000`.

**In 0.7.0**, the master is rendered from the harmonization engine
(`qes_harmonize()`) for all 11 studies: each column is built from the
harmonized variables ("targets") it needs, and a code the specification
does not map is `NA`. The columns that change estimates:

| Column | Since 0.7.0 | Changed in |
|---|---|---|
| `vote_choice` | The vote reported after the election, in every study that asked it (0.4.4 used the campaign intention for `qes2022` and `qes1998`). Non-voters, spoiled ballots, "don't know" and refusals are `NA`: the categories "Did not vote / None" and "Don't know / Refused" are gone, and `turnout` says who voted. The CROP polls asked only intentions: `NA`, with the intention in the appended `vote_intent`. | every study |
| `turnout` | Reported turnout, 1 or 0 (`qes2022` from its post-election wave; the 2018 panel from its turnout question). | `qes2022`, `qes2018`, `qes2018_panel`, `qes2007_panel` |
| `sovereignty_support`, `sovereignty` | The referendum on Quebec becoming an independent country only. Other wordings are `NA`; the 1995 sovereignty-partnership question is in the appended `sov_partnership_1995`. | the studies without that question |
| `ideology` | 0-10 with its end points: 0.4.4 lost the 0 and 10 answers of `qes2014` (142 respondents) and found no source in `qes2018`. | `qes2014`, `qes2018` |
| `political_interest` | The four-point questions as 10, 7, 3 and 0; 0.4.4 left the 2018 codes 1 to 4 unconverted and reversed. Mixes a 4-point and a 0-10 question across studies. | `qes2018`, `qes2012_panel` |
| `language` | The mother tongue; 0.4.4 used the interview language in `qes2014` and the interface language in `qes2022`. Two first languages are `NA`. | `qes2014`, `qes2022`, `qes2007`, `qes2007_panel`, `qes2018_panel` |
| `income`, `religion` | Each study's own categories; `-99` and raw codes are `NA`; `qes2012` income is filled. | `income`: 9 studies; `religion`: `qes2012`, `qes2014`, `qes2018`, `qes2022` |
| `age_group`, `education`, `born_canada` | Categories corrected: six age bands in the 2007 and 2012 panels, *maîtrise* as University, people born elsewhere in Canada as born in Canada, and similar cases listed in NEWS. | a few studies each |
| `provincial_pid` | Filled where the study asked it (0.4.4 found a source only in `qes2022`). | `qes2007`, `qes2008`, `qes2012`, `qes2014`, `qes2018` |
| `party_best`, `party_lean` | `NA` everywhere: each study's source was another question. | `party_best`: `qes2022`, `qes2018`, `qes2014`, `qes2012`, `qes2007`; `party_lean`: every study but `qes1998` |
| `survey_weight` | Unchanged: each study's own weight, on its own scale. Do not pool weighted estimates across studies with it. | none |

Three consequences for pooled analyses: an estimate that pooled
`vote_choice` across studies mixed intentions and reported votes in 0.4.4,
and one that pooled `ideology` across 2012, 2014 and 2022 lost the extreme
answers of 2014. Estimates of that kind move most. Third, a filter on
non-missing values now keeps studies that 0.4.4 left empty: columns that
were `NA` for a whole study are filled, `ideology` for `qes2018`,
`provincial_pid` for `qes2007`, `qes2008`, `qes2012`, `qes2014` and
`qes2018`, `income` for `qes2012` and `year_of_birth` for `qes2022`. A
complete-case sample such as
`subset(master, !is.na(vote_choice) & !is.na(ideology))` now contains
`qes2018`; if the script also reads `qes2018` on its own, those respondents
are counted twice. Select studies explicitly, with
`qes_code %in% c(...)`, and compare `table(sample$qes_code)` for the
analysis sample with the counts you published.

The attributes of the master say what each column holds, which studies'
values changed and why a column is empty in a study:

```{r master}
master <- get_qes_master(surveys = "qes_demo", quiet = TRUE)
map <- attr(master, "legacy_column_map")
map[map$column %in% c("vote_choice", "ideology", "sovereignty_support"),
    c("column", "target", "flag")]
attr(master, "legacy_na_columns")[1:5, c("column", "study", "reason")]
unique(master[, c("qes_code", "vote_choice_timing", "sovereignty_item")])
```

`map$studies_changed` names, for each column, the studies whose values
differ from 0.4.4, and `attr(master, "source_map")` the question each
column read in each study. The NEWS of 0.7.0 counts the changed cells,
column by column and study by study.

`get_decon()` is soft-deprecated since 0.7.0, rendered from the same
targets: its categorical columns are factors with English levels, and for
`qes2022` its `turnout` and `votechoice` stay the campaign-period
likelihood and intention (`attr(, "timing")` says `"pre"`).

## From the master to `qes_harmonize()`

For new work, `qes_harmonize()` gives what the master cannot: one column
per question stimulus (a reported vote and an intention are different
targets, and so are the sovereignty wordings), a comparability grade for
each study's question, a reason for every missing value and each wave's
weight. The `target` column of the map above names the target of each
master column. The master's `ideology` and `vote_choice` become:

```{r harmonize}
h <- qes_harmonize("qes_demo", targets = c("lr_self", "vote_prov_recall"),
                   missing = "reasons", include_draft = TRUE, quiet = TRUE,
                   lang = params$lang)
table(h$lr_self__na, useNA = "ifany")[c("dk", "refused")]
attr(h, "qes_weight_guide")[, c("target", "study", "weight_column")]
```

Compute each estimate within one study, with the weight of the wave that
asked the question. The mean left-right position of each party's voters in
the demonstration study:

```{r estimate}
ok <- !is.na(h$lr_self) & !is.na(h$vote_prov_recall)
d <- h[ok, ]
means <- sapply(split(d, d$vote_prov_recall, drop = TRUE), function(p) {
  weighted.mean(p$lr_self, p$weight_post)
})
round(means, 2)
```

The rows of the specification are checked against the original files but
not yet signed off by a reviewer, so `include_draft = TRUE` is needed for
now. `qes_spec()` shows which study has which target and how comparable its
question is, and `vignette("harmonization-reference", package = "qesR")`
gives each target's questions, wordings and grades.

## Old names, new names

The eleven helpers of 0.4.4 below keep working and will not be removed.
Each prints a one-time notice naming its replacement
(`options(qesR.quiet_deprecated = TRUE)` hides the notices).

| 0.4.4 | Replacement | Note |
|---|---|---|
| `get_codebook()`, `get_qes_codebook()` | `qes_codebook()` | same arguments; offline; `refresh` is ignored |
| `format_codebook(cb, layout)` | `qes_codebook(cb, layout = )` | a plain data frame (for example a codebook read back from CSV) is an error |
| `get_value_labels(cb, variable)` | `qes_codebook(layout = "long")` | an unknown variable is an error, not an empty list |
| `get_question(data, q)` | `qes_question(study, variables)` | exact names only; `full = FALSE` no longer changes the result |
| `get_codebook_files()`, `get_qes_codebook_files()` | `qes_docs()` | lists every document, offline |
| `download_codebook()` | `qes_download(what = "docs")` | `file` now selects documents by name; files are md5-checked |
| `get_preview(srvy, obs)` | `head(get_qes(srvy), obs)` | |
| `get_qescodes()` | `qes_studies()` | the 1998 firm files are added after the 11 old codes |
| `get_decon()` | `qes_harmonize(srvy, targets = "decon", include_draft = TRUE)` | soft-deprecated since 0.7.0; values changed (above) |

`get_qes()`, `get_qes_master()` and `qes_codebook()` keep their names. A
legacy function returns what its replacement returns. The first call of
`get_codebook()` in a new session also prints this notice (it is not shown
below, because it appears only once per session):

> `get_codebook()` is soft-deprecated; use `qes_codebook()`. It keeps
> working and will not be removed.

```{r deprecated, message = FALSE}
cb <- get_codebook("qes_demo")
identical(cb, qes_codebook("qes_demo"))
```

## Keeping new results reproducible

Record what each result was computed from, and cite it:

```{r reproducible}
qes_provenance(h, level = "spec")[, c("spec_version", "spec_hash", "qesR_version")]
qes_cite("qes2014")
```

`qes_provenance(x)` gives the file of each study (DOI, dataset version,
md5), and `level = "spec"` the version and content hash of the
harmonization specification that produced harmonized data. A result is
reproduced exactly by the same qesR version, which fixes both.
