---
title: "Moving from qesR 0.4.4 to 0.5.0"
output: rmarkdown::html_vignette
params:
  lang: "en"
vignette: >
  %\VignetteIndexEntry{Moving from qesR 0.4.4 to 0.5.0}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

*[Version française](fr-migrer-0.5.html)*

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# messages in the language of the page, whatever the locale of the build
options(qesR.lang = params$lang)
```

Code written for qesR 0.4.4 keeps running under 0.5.0: every function keeps
its name and its arguments. Three things change what that code produces:

1. functions no longer write into your workspace unless you ask;
2. data are read from the original files, which fixes damaged text;
3. `get_qes_master()` and `get_decon()` keep every respondent and set values
   verified to be wrong to `NA`, so estimates computed from them change.

The complete list is in the package NEWS (`news(package = "qesR")`). Every
chunk on this page runs offline, on the synthetic study `qes_demo`.

```{r load}
library(qesR)
```

## Assign the result yourself

In 0.4.4, `get_qes("qes2018")` created an object `qes2018` in your
workspace. In 0.5.0 it returns the data and writes nothing: assign it.

```{r assign}
qes_demo <- get_qes("qes_demo", quiet = TRUE)
dim(qes_demo)
```

`assign_global = TRUE` still assigns, into the environment the function is
called from: the global environment at top level, the function's own frame
inside a function. It still assigns the codebook too, as `<code>_codebook`.

```{r assign-global}
f <- function() {
  get_qes("qes_demo", assign_global = TRUE, quiet = TRUE)
  c(data = exists("qes_demo", inherits = FALSE),
    codebook = exists("qes_demo_codebook", inherits = FALSE))
}
f()
```

The first call in a session that leaves `assign_global` unset prints a
one-time note; passing the argument, or assigning the result, is enough.

## The same data, read from the original files

`get_qes()` now reads each study's original SPSS or Stata file, checked
against the md5 in the catalog. Column names, codes and `is.na()` counts are
those of 0.4.4 for all 11 studies, except where the old text conversion
damaged the data: for example 1,805 cells with garbled accents in `qes2018`
and 2,517 in `qes2022` are repaired, and the `qes2022` interview dates are
now date-times. The data record which file they came from:

```{r provenance}
qes_provenance(qes_demo)
```

## Old names, new names

The eleven helpers of 0.4.4 below keep working and will not be removed.
Each prints a one-time notice naming its replacement
(`options(qesR.quiet_deprecated = TRUE)` hides the notices).

| 0.4.4 | 0.5.0 | Note |
|---|---|---|
| `get_codebook()`, `get_qes_codebook()` | `qes_codebook()` | same arguments; offline; `refresh` is ignored |
| `format_codebook(cb, layout)` | `qes_codebook(cb, layout = )` | a plain data frame (for example a codebook read back from CSV) is an error |
| `get_value_labels(cb, variable)` | `qes_codebook(layout = "long")` | an unknown variable is an error, not an empty list |
| `get_question(data, q)` | `qes_question(study, variables)` | exact names only; `full = FALSE` no longer changes the result |
| `get_codebook_files()`, `get_qes_codebook_files()` | `qes_docs()` | lists every document, offline |
| `download_codebook()` | `qes_download(what = "docs")` | `file` now selects documents by name; files are md5-checked |
| `get_preview(srvy, obs)` | `head(get_qes(srvy), obs)` | |
| `get_qescodes()` | `qes_studies()` | the 1998 firm files are added after the 11 old codes |
| `get_decon()` | kept as it is until qesR 0.7.0 | values changed (below) |

`get_qes()`, `get_qes_master()` and `qes_codebook()` keep their names. A
legacy function returns what its replacement returns. The first call of
`get_codebook()` in a new session also prints this notice (it is not shown
below, because it appears only once per session):

> `get_codebook()` is soft-deprecated; use `qes_codebook()`. It keeps
> working and will not be removed.

```{r deprecated, message = FALSE}
cb <- get_codebook("qes_demo")
identical(cb, qes_codebook("qes_demo"))
```

## `get_qes_master()` and `get_decon()`

The master keeps its 30 columns, in the same order and with the same types.
Its values change, by deletion only:

- every respondent of every file is kept (0.4.4 dropped 380 `qes2007_panel`
  respondents and 1 CROP respondent as duplicates);
- the 70 columns 0.4.4 appended by stacking variables that share a name
  across studies are removed, because they held different questions in
  different studies;
- values verified to be wrong are `NA`: vote intentions stored as reported
  votes, sovereignty questions other than the referendum on an independent
  country, raw codes left on converted scales, and similar cases.

Attributes say what was removed and why:

```{r master}
master <- get_qes_master(surveys = "qes_demo", quiet = TRUE)
head(attr(master, "removed_columns"))
attr(master, "legacy_na_columns")[, c("column", "study", "reason", "n_cells", "cause")]
unique(master[, c("qes_code", "vote_choice_timing", "sovereignty_item")])
```

`attr(master, "legacy_column_map")` says what each column means and which
studies' values changed. `get_decon()` changes the same way.

## Keeping results reproducible

Estimates computed from `get_qes_master()` or `get_decon()` under 0.4.4 will
differ under 0.5.0. To reproduce a 0.4.4 result exactly, install that
version in a separate library:

```{r pin, eval = FALSE}
remotes::install_github("ThomasGareau/qesR", ref = "v0.4.4")
```

Results computed from `get_qes()` data keep their values, apart from the
repaired text and dates listed above. For new work, record what you used
with `qes_provenance()` and cite it with `qes_cite()`.
