---
title: "Example: respondents by study"
params:
  lang: "en"
---

*[Version française](fr-analyse-descriptive.html)*

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
# messages in the language of the page, whatever the locale of the build
options(qesR.lang = params$lang)
```

This page describes the respondents of the studies in the legacy merged file,
`get_qes_master()`, study by study. It is built when the website is built:
the data files are downloaded from their Dataverse deposits through qesR's
cache, so the numbers are those of the full files, not of a sample.

For estimates with the harmonization engine (experimental), with a
comparability grade for each study's question and the weight of the wave
that asked it, see [Harmonizing across studies](harmonization.html).

The studies were not designed alike: the Quebec Election Studies (family
`qes`), the Durand panels (`durand_panel`), the CROP polls (`crop_polls`)
and the 1998 polls (`polls_1998`) differ in mode, target population and
questions. Every estimate below is computed within one study, with that
study's own weight (`survey_weight`), and never pooled across studies. The
estimates are point estimates: they ignore each survey's design, so no
confidence interval is shown.

```{r packages, message = FALSE, warning = FALSE}
library(qesR)
library(ggplot2)
```

```{r data, message = FALSE}
tr <- function(en, fr) if (identical(params$lang, "fr")) fr else en

master <- get_qes_master(quiet = TRUE)
studies <- qes_studies()
master$family <- studies$family[match(master$qes_code, studies$study)]
master$year <- as.integer(master$qes_year)

# Weighted share of each value of x, within one study (or one CROP year)
wshare <- function(x, w) {
  ok <- !is.na(x) & !is.na(w)
  if (!any(ok)) return(NULL)
  tapply(w[ok], x[ok], sum) / sum(w[ok])
}
cells <- split(master, list(master$qes_code, master$year), drop = TRUE)
```

## Table 1: respondents by study

Every respondent of every data file is kept. The CROP polls are monthly
polls from 2007 to 2010, shown by year.

```{r sizes}
sizes <- do.call(rbind, lapply(cells, function(d) {
  data.frame(study = d$qes_code[1], year = d$year[1], family = d$family[1], n = nrow(d))
}))
sizes <- sizes[order(sizes$year, sizes$study), ]
knitr::kable(
  sizes, row.names = FALSE,
  col.names = tr(c("Study", "Year", "Family", "Respondents"),
                 c("Étude", "Année", "Famille", "Répondants"))
)
```

## Table 2: age groups

The studies record age in different bands: six bands in most studies, three
(18-34, 35-54, 55+) in the 2018 Durand panel (`qes2018_panel`). The six bands nest in the three,
so the table uses three. Target populations differ too: `qes2018` covers
people aged 16 and over, the 1998 polls francophones only (see
`qes_studies()$target_population_en`).

Shares are among respondents with a known age group, counted in the "N"
column. `qes2018`'s respondents aged 16 or 17 have no age group in the
legacy file and are left out, as are the few respondents of other studies
whose age is missing.

```{r age}
band3 <- c("18-24" = "18-34", "25-34" = "18-34", "18-34" = "18-34",
           "35-44" = "35-54", "45-54" = "35-54", "35-54" = "35-54",
           "55-64" = "55+", "65+" = "55+", "55+" = "55+")
master$age3 <- unname(band3[master$age_group])
cells <- split(master, list(master$qes_code, master$year), drop = TRUE)

age <- do.call(rbind, lapply(cells, function(d) {
  s <- wshare(d$age3, d$survey_weight)
  if (is.null(s)) return(NULL)
  data.frame(study = d$qes_code[1], year = d$year[1], family = d$family[1],
             n = sum(!is.na(d$age3) & !is.na(d$survey_weight)),
             age3 = names(s), pct = round(100 * as.numeric(s), 1))
}))
age_wide <- reshape(age, idvar = c("study", "year", "family", "n"), timevar = "age3",
                    direction = "wide")
age_wide <- age_wide[order(age_wide$year, age_wide$study), ]
knitr::kable(
  age_wide, row.names = FALSE, format.args = list(decimal.mark = tr(".", ",")),
  col.names = c(tr(c("Study", "Year", "Family", "N with an age group"),
                   c("Étude", "Année", "Famille", "N avec un groupe d'âge")),
                "18-34 (%)", "35-54 (%)", "55+ (%)")
)
```

```{r age-plot, fig.alt = tr("Weighted share of each age group in each study, by year of the study.", "Part pondérée de chaque groupe d'âge dans chaque étude, par année de l'étude.")}
ggplot(age, aes(x = year, y = pct, colour = age3, shape = family)) +
  geom_point(size = 2.5) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(
    x = tr("Year of the study", "Année de l'étude"),
    y = tr("Weighted share (%)", "Part pondérée (%)"),
    colour = tr("Age group", "Groupe d'âge"),
    shape = tr("Family", "Famille")
  ) +
  theme_minimal(base_size = 12)
```

## Table 3: reported turnout

`turnout` is the turnout reported after the election (1 voted, 0 did not),
in every study that asked it; `qes2022` answers in its post-election wave.
It is `NA` in the CROP polls, which asked only about intentions (see
`attr(master, "legacy_na_columns")`). Surveys overestimate turnout: people
who vote are more likely to answer, and some who did not vote say they did.

```{r turnout}
turnout <- do.call(rbind, lapply(cells, function(d) {
  s <- wshare(d$turnout, d$survey_weight)
  if (is.null(s) || !"1" %in% names(s)) return(NULL)
  data.frame(study = d$qes_code[1], year = d$year[1], family = d$family[1],
             n = sum(!is.na(d$turnout)), pct = round(100 * s[["1"]], 1))
}))
turnout <- turnout[order(turnout$year, turnout$study), ]
knitr::kable(
  turnout, row.names = FALSE, format.args = list(decimal.mark = tr(".", ",")),
  col.names = tr(c("Study", "Year", "Family", "N answering", "Reported turnout (%)"),
                 c("Étude", "Année", "Famille", "N répondants",
                   "Participation déclarée (%)"))
)
```

```{r turnout-plot, fig.alt = tr("Weighted reported turnout in each study, by year of the study.", "Participation déclarée pondérée dans chaque étude, par année de l'étude.")}
ggplot(turnout, aes(x = year, y = pct, colour = family, shape = family)) +
  geom_point(size = 2.5) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(
    x = tr("Year of the study", "Année de l'étude"),
    y = tr("Reported turnout, weighted (%)", "Participation déclarée, pondérée (%)"),
    colour = tr("Family", "Famille"),
    shape = tr("Family", "Famille")
  ) +
  theme_minimal(base_size = 12)
```

## Notes

- `survey_weight` is the weight variable qesR 0.4.4 read for each study, on
  that study's own scale; `attr(master, "legacy_column_map")` describes it.
- The legacy merged file is kept for code written for qesR 0.4.4; since
  0.7.0 it is rendered from the harmonization engine. What changed since
  0.4.4 is in `vignette("migrating-0.7", package = "qesR")`.
