---
title: "Example: reported vote"
params:
  lang: "en"
---

*[Version française](fr-analyse-choix-vote.html)*

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
# messages in the language of the page, whatever the locale of the build
options(qesR.lang = params$lang)
```

This page shows the reported provincial vote in the studies of the legacy
merged file, `get_qes_master()`. It is built when the website is built: the
data files are downloaded from their Dataverse deposits through qesR's
cache.

For estimates with the harmonization engine (experimental), with a
comparability grade for each study's question and the weight of the wave
that asked it, see [Harmonizing across studies](harmonization.html).

Since qesR 0.7.0, `vote_choice` is rendered from the harmonization engine:
it holds the vote reported after the election in every study that asked it
(`vote_choice_timing` is `"post"`), `qes2022` and `qes1998` included, and is
`NA` for respondents who did not vote, did not know or refused to say. The
CROP polls asked only vote intentions, so they are not shown. Each estimate
is computed within one study, with that study's own weight
(`survey_weight`); it is a point estimate that ignores the survey's design.
The Quebec Election Studies, the Durand panels and the 1998 polls are shown
apart, because their designs differ. The 1998 polls sampled francophones
only, and the firms over-selected the undecided and those who refused to
say; their weight (`survey_weight`, as read by 0.4.4) is not reviewed, so
their shares are not comparable with the other studies or with the
official results.

```{r packages, message = FALSE, warning = FALSE}
library(qesR)
library(ggplot2)
```

```{r data, message = FALSE}
tr <- function(en, fr) if (identical(params$lang, "fr")) fr else en

master <- get_qes_master(quiet = TRUE)
studies <- qes_studies()
master$family <- studies$family[match(master$qes_code, studies$study)]
master$year <- as.integer(master$qes_year)
```

## Table 1: reported vote by study

Shares are among respondents who reported voting for a party, including
"Other party". Respondents who did not vote, did not know or refused to say
are `NA` and left out of the denominator.

Two kinds of cell are structural zeros, not estimates, and are not shown as
0:

- **—** the party did not run in that election, according to the official
  results of Élections Québec that ship with qesR
  (`inst/extdata/validation/official_results.csv`). The ADQ merged into the
  CAQ in 2012, and Option nationale ran only in 2012 and 2014.
- **n.l.** the party ran, but the study's question did not list it as an
  answer, so its voters could only answer "Other party". The answers each
  question offered come from the harmonization specification
  (`qes_spec("crosswalk", targets = "vote_prov_recall")`): the 2012, 2014
  and 2018 studies omit the Conservative Party (PCQ), and the 2018 and 2022
  studies the Green Party (PVQ).

A 0.0 is a party that was listed but that no respondent of the study
reported.

```{r vote}
parties <- c("PLQ", "PQ", "ADQ", "CAQ", "QS", "PVQ", "ON", "PCQ", "Other party")
# Parties with candidates at each general election: the official results of
# Elections Quebec that ship with qesR. A party that did not run cannot be
# reported.
official <- read.csv(system.file("extdata", "validation", "official_results.csv",
                                 package = "qesR"))
ran <- split(official$party, official$election_id)
# The answers each study's question offered, from the harmonization spec
# (the levels of the reported-vote target); "other" is "Other party"
xw <- qes_spec("crosswalk", targets = "vote_prov_recall")
offered <- lapply(setNames(strsplit(xw$levels_offered, ";"), xw$study),
                  function(x) sub("^other$", "Other party", x))

voters <- master[master$vote_choice %in% parties & !is.na(master$survey_weight), ]
vote <- do.call(rbind, lapply(split(voters, voters$qes_code), function(d) {
  w <- tapply(d$survey_weight, factor(d$vote_choice, levels = parties), sum)
  w[is.na(w)] <- 0
  election <- studies$election_id[match(d$qes_code[1], studies$study)]
  status <- ifelse(!parties %in% c(ran[[election]], "Other party"), "not_run",
                   ifelse(parties %in% offered[[d$qes_code[1]]], "offered", "not_listed"))
  data.frame(study = d$qes_code[1], year = d$year[1], family = d$family[1],
             n = nrow(d), party = parties, pct = round(100 * as.numeric(w) / sum(w), 1),
             status = status)
}))
vote$cell <- ifelse(vote$status == "not_run", "\u2014",
                    ifelse(vote$status == "not_listed", tr("n.l.", "n.p."),
                           formatC(vote$pct, format = "f", digits = 1,
                                   decimal.mark = tr(".", ","))))
vote_wide <- reshape(vote[, c("study", "year", "family", "n", "party", "cell")],
                     idvar = c("study", "year", "family", "n"),
                     timevar = "party", direction = "wide")
names(vote_wide) <- sub("^cell\\.", "", names(vote_wide))
vote_wide <- vote_wide[order(vote_wide$family, vote_wide$year), ]
knitr::kable(
  vote_wide, row.names = FALSE, align = c("l", "r", "l", rep("r", 1 + length(parties))),
  col.names = c(tr(c("Study", "Year", "Family", "N voters"),
                   c("Étude", "Année", "Famille", "N votants")),
                tr(parties, c(parties[-length(parties)], "Autre parti")))
)
```

## Figure 1: the main parties

```{r vote-plot, fig.alt = tr("Weighted reported vote share of the PLQ, PQ, ADQ, CAQ and QS in each study, by year, with the Quebec Election Studies, the Durand panels and the 1998 polls in separate panels.", "Part pondérée du vote déclaré pour le PLQ, le PQ, l'ADQ, la CAQ et QS dans chaque étude, par année, les Études électorales québécoises, les panels Durand et les sondages de 1998 dans des panneaux distincts."), fig.height = 7}
main <- vote[vote$status == "offered" & vote$party %in% c("PLQ", "PQ", "ADQ", "CAQ", "QS"), ]
families <- c(qes = tr("Quebec Election Studies", "Études électorales québécoises"),
              durand_panel = tr("Durand panels", "Panels Durand"),
              polls_1998 = tr("1998 polls", "Sondages de 1998"))
main$family <- factor(families[main$family], levels = families)
ggplot(main, aes(x = year, y = pct, colour = party, shape = party)) +
  geom_line() +
  geom_point(size = 2.4) +
  facet_wrap(~family, ncol = 1) +
  scale_colour_manual(values = c(PLQ = "#d71920", PQ = "#004c9d", ADQ = "#6d9eeb",
                                 CAQ = "#00a3c7", QS = "#ff6f00")) +
  scale_x_continuous(breaks = sort(unique(main$year))) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(
    x = tr("Year of the study", "Année de l'étude"),
    y = tr("Share of reported votes, weighted (%)", "Part des votes déclarés, pondérée (%)"),
    colour = tr("Party", "Parti"),
    shape = tr("Party", "Parti")
  ) +
  theme_minimal(base_size = 12)
```

## Notes

- A line joins studies of the same family only; the ADQ and the CAQ are
  separate parties, and their lines are not joined.
- qesR 0.4.4 coded "Did not vote / None" and "Don't know / Refused" as
  categories of `vote_choice`, and used the campaign-period vote intention
  for `qes2022` and `qes1998`; since 0.7.0 these respondents are `NA` and
  both studies give the reported vote
  (`vignette("migrating-0.7", package = "qesR")`).
- Reported votes differ from official results: compare them with the
  results published by Élections Québec, not with each other across
  designs.
