---
title: "Example: reported vote"
params:
  lang: "en"
---

*[Version française](fr-analyse-choix-vote.html)*

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
# messages in the language of the page, whatever the locale of the build
options(qesR.lang = params$lang)
```

This page shows the reported provincial vote in the studies of the legacy
merged file, `get_qes_master()`. It is built when the website is built: the
data files are downloaded from their Dataverse deposits through qesR's
cache.

For estimates with the harmonization engine (experimental), with a
comparability grade for each study's question and the weight of the wave
that asked it, see [Harmonizing across studies](harmonization.html).

Since qesR 0.5.0, `vote_choice` holds the vote reported after the election
only (`vote_choice_timing` is `"post"`). It is `NA` in the studies whose
source was a vote intention: `qes2022`, the CROP polls and `qes1998`. Each
estimate is computed within one study, with that study's own weight
(`survey_weight`); it is a point estimate that ignores the survey's design.
The Quebec Election Studies and the Durand panels are shown apart, because
their designs differ.

```{r packages, message = FALSE, warning = FALSE}
library(qesR)
library(ggplot2)
```

```{r data, message = FALSE}
tr <- function(en, fr) if (identical(params$lang, "fr")) fr else en

master <- get_qes_master(quiet = TRUE)
studies <- qes_studies()
master$family <- studies$family[match(master$qes_code, studies$study)]
master$year <- as.integer(master$qes_year)
```

## Table 1: reported vote by study

Shares are among respondents who reported voting for a party, including
"Other party". Respondents who did not vote, did not know or refused to say
are left out of the denominator.

Two kinds of cell are structural zeros, not estimates, and are not shown as
0:

- **—** the party did not run in that election (official results of
  Élections Québec). The ADQ merged into the CAQ in 2012, and Option
  nationale ran only in 2012 and 2014.
- **n.l.** the party ran, but the study's question did not list it as an
  answer, so its voters could only answer "Other party". Both 2018 studies
  omit the Green Party (PVQ): see
  `qes_codebook("qes2018", layout = "long", variables = "q6")`.

A 0.0 is a party that was listed but that no respondent of the study
reported.

```{r vote}
parties <- c("PLQ", "PQ", "ADQ", "CAQ", "QS", "PVQ", "ON", "Other party")
# Parties with candidates at each general election (official results of
# Elections Quebec). A party that did not run cannot be reported.
ran <- list(
  QC2007 = c("PLQ", "PQ", "ADQ", "QS", "PVQ"),
  QC2008 = c("PLQ", "PQ", "ADQ", "QS", "PVQ"),
  QC2012 = c("PLQ", "PQ", "CAQ", "QS", "PVQ", "ON"),
  QC2014 = c("PLQ", "PQ", "CAQ", "QS", "PVQ", "ON"),
  QC2018 = c("PLQ", "PQ", "CAQ", "QS", "PVQ"),
  QC2022 = c("PLQ", "PQ", "CAQ", "QS", "PVQ")
)
# Parties that ran but that the study's question did not list as an answer
# (value labels of the source variable, from qes_codebook())
not_listed <- list(qes2018 = "PVQ", qes2018_panel = "PVQ")

voters <- master[master$vote_choice %in% parties & !is.na(master$survey_weight), ]
vote <- do.call(rbind, lapply(split(voters, voters$qes_code), function(d) {
  w <- tapply(d$survey_weight, factor(d$vote_choice, levels = parties), sum)
  w[is.na(w)] <- 0
  election <- studies$election_id[match(d$qes_code[1], studies$study)]
  status <- ifelse(!parties %in% c(ran[[election]], "Other party"), "not_run",
                   ifelse(parties %in% not_listed[[d$qes_code[1]]], "not_listed", "offered"))
  data.frame(study = d$qes_code[1], year = d$year[1], family = d$family[1],
             n = nrow(d), party = parties, pct = round(100 * as.numeric(w) / sum(w), 1),
             status = status)
}))
vote$cell <- ifelse(vote$status == "not_run", "\u2014",
                    ifelse(vote$status == "not_listed", tr("n.l.", "n.p."),
                           formatC(vote$pct, format = "f", digits = 1,
                                   decimal.mark = tr(".", ","))))
vote_wide <- reshape(vote[, c("study", "year", "family", "n", "party", "cell")],
                     idvar = c("study", "year", "family", "n"),
                     timevar = "party", direction = "wide")
names(vote_wide) <- sub("^cell\\.", "", names(vote_wide))
vote_wide <- vote_wide[order(vote_wide$family, vote_wide$year), ]
knitr::kable(
  vote_wide, row.names = FALSE, align = c("l", "r", "l", rep("r", 1 + length(parties))),
  col.names = c(tr(c("Study", "Year", "Family", "N voters"),
                   c("Étude", "Année", "Famille", "N votants")),
                tr(parties, c(parties[-length(parties)], "Autre parti")))
)
```

## Figure 1: the main parties

```{r vote-plot, fig.alt = tr("Weighted reported vote share of the PLQ, PQ, ADQ, CAQ and QS in each study, by year, with Quebec Election Studies and Durand panels in separate panels.", "Part pondérée du vote déclaré pour le PLQ, le PQ, l'ADQ, la CAQ et QS dans chaque étude, par année, les Études électorales québécoises et les panels Durand dans des panneaux distincts.")}
main <- vote[vote$status == "offered" & vote$party %in% c("PLQ", "PQ", "ADQ", "CAQ", "QS"), ]
main$family <- ifelse(main$family == "qes",
                      tr("Quebec Election Studies", "Études électorales québécoises"),
                      tr("Durand panels", "Panels Durand"))
ggplot(main, aes(x = year, y = pct, colour = party, shape = party)) +
  geom_line() +
  geom_point(size = 2.4) +
  facet_wrap(~family, ncol = 1) +
  scale_colour_manual(values = c(PLQ = "#d71920", PQ = "#004c9d", ADQ = "#6d9eeb",
                                 CAQ = "#00a3c7", QS = "#ff6f00")) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(
    x = tr("Year of the study", "Année de l'étude"),
    y = tr("Share of reported votes, weighted (%)", "Part des votes déclarés, pondérée (%)"),
    colour = tr("Party", "Parti"),
    shape = tr("Party", "Parti")
  ) +
  theme_minimal(base_size = 12)
```

## Notes

- A line joins studies of the same family only; the ADQ and the CAQ are
  separate parties, and their lines are not joined.
- "Don't know / Refused" is a category qesR 0.4.4 coded on purpose, kept as
  it is in `vote_choice` until the harmonized variables of qesR 0.7.0.
- Reported votes differ from official results: compare them with the
  results published by Élections Québec, not with each other across
  designs.
