---
title: "Use case example: Support for Sovereignty"
params:
  lang: "en"
---

*[Version française](fr-analyse-souverainete.html)*

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
```

This page estimates support for Quebec independence in the studies of the
legacy merged file, `get_qes_master()`. It is built when the website is
built: the data files are downloaded from their Dataverse deposits through
qesR's cache.

A sovereignty question is only comparable across studies when it asks the
same thing. Since qesR 0.5.0, `sovereignty_support` holds one question only:
how the respondent would vote in a referendum on Quebec becoming an
independent country. Studies that asked something else (the 1995 question
on sovereignty with a partnership offer, a "sovereign country", follow-up
questions put to undecided respondents, or a favourable/opposed scale) have
`NA` in that column. Each estimate is computed within one study, with that
study's own weight (`survey_weight`); it is a point estimate that ignores the
survey's design.

```{r data}
library(qesR)
library(ggplot2)

tr <- function(en, fr) if (identical(params$lang, "fr")) fr else en

master <- get_qes_master(quiet = TRUE)
studies <- qes_studies()
master$family <- studies$family[match(master$qes_code, studies$study)]
master$year <- as.integer(master$qes_year)

# Weighted share of each value of x, within one study
wshare <- function(x, w) {
  ok <- !is.na(x) & !is.na(w)
  if (!any(ok)) return(NULL)
  tapply(w[ok], x[ok], sum) / sum(w[ok])
}
```

## Table 1: which studies ask the independence question

`attr(master, "legacy_na_columns")` says why a column is `NA` in a study;
`attr(master, "source_map")` names the variable each study's column reads.

```{r items}
sources <- attr(master, "source_map")
sources <- sources[sources$harmonized_variable == "sovereignty_support",
                   c("qes_code", "source_variable")]
blanked <- attr(master, "legacy_na_columns")
blanked <- blanked[blanked$column == "sovereignty_support", c("study", "cause")]
items <- merge(sources, blanked, by.x = "qes_code", by.y = "study", all.x = TRUE)
items$year <- studies$year[match(items$qes_code, studies$study)]
items$in_column <- ifelse(is.na(items$cause), tr("yes", "oui"), tr("no", "non"))
items <- items[order(items$year, items$qes_code), c("qes_code", "year", "source_variable", "in_column")]
knitr::kable(
  items, row.names = FALSE,
  col.names = tr(c("Study", "Year", "Source variable", "Independence question"),
                 c("Étude", "Année", "Variable source", "Question sur l'indépendance"))
)
```

The wording in the four studies that ask it:

```{r wording}
asked <- items$qes_code[items$in_column == tr("yes", "oui")]
wording <- do.call(rbind, lapply(asked, function(s) {
  q <- qes_question(s, items$source_variable[items$qes_code == s], lang = params$lang)
  q[, c("study", "variable", "question", "truncated")]
}))
knitr::kable(wording, row.names = FALSE)
```

## Table 2: support for independence

The share who would vote YES among respondents who answered YES or NO.
"Don't know" and refusals are left out.

```{r support}
sov <- do.call(rbind, lapply(split(master, master$qes_code), function(d) {
  s <- wshare(d$sovereignty_support, d$survey_weight)
  if (is.null(s) || !"1" %in% names(s)) return(NULL)
  data.frame(study = d$qes_code[1], year = d$year[1],
             n = sum(!is.na(d$sovereignty_support)), pct = round(100 * s[["1"]], 1))
}))
sov <- sov[order(sov$year), ]
knitr::kable(
  sov, row.names = FALSE,
  col.names = tr(c("Study", "Year", "N answering", "YES (%)"),
                 c("Étude", "Année", "N répondants", "OUI (%)"))
)
```

```{r support-plot, fig.alt = "Weighted share who would vote YES in a referendum on independence, in the 2012, 2014, 2018 and 2022 Quebec Election Studies."}
ggplot(sov, aes(x = year, y = pct)) +
  geom_line(colour = "#12355b") +
  geom_point(size = 2.5, colour = "#12355b") +
  scale_x_continuous(breaks = sov$year) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(
    x = tr("Year of the study", "Année de l'étude"),
    y = tr("Would vote YES, weighted (%)", "Voterait OUI, pondéré (%)")
  ) +
  theme_minimal(base_size = 12)
```

## Notes

- The four studies are Quebec Election Studies, but they differ in mode and
  population: `qes2018` covers people aged 16 and over, and `qes2022` is an
  online panel of citizens aged 18 and over (`qes_studies()`).
- `qes2022`'s question text is its variable label, which its data file cut
  at 80 characters (`truncated`); the full wording is in its questionnaire
  (`qes_docs("qes2022")`).
- For the other studies, read their own question with `get_qes()` and
  `qes_question()`; comparing it with the independence question needs care,
  because the wording differs.
