---
title: "Use case example: Reported Vote Over Time"
params:
  lang: "en"
---

*[Version française](fr-analyse-choix-vote.html)*

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
```

This page shows the reported provincial vote in the studies of the legacy
merged file, `get_qes_master()`. It is built when the website is built: the
data files are downloaded from their Dataverse deposits through qesR's
cache.

Since qesR 0.5.0, `vote_choice` holds the vote reported after the election
only (`vote_choice_timing` is `"post"`). It is `NA` in the studies whose
source was a vote intention: `qes2022`, the CROP polls and `qes1998`. Each
estimate is computed within one study, with that study's own weight
(`survey_weight`); it is a point estimate that ignores the survey's design.
The Quebec Election Studies and the Durand panels are shown apart, because
their designs differ.

```{r data}
library(qesR)
library(ggplot2)

tr <- function(en, fr) if (identical(params$lang, "fr")) fr else en

master <- get_qes_master(quiet = TRUE)
studies <- qes_studies()
master$family <- studies$family[match(master$qes_code, studies$study)]
master$year <- as.integer(master$qes_year)

# Weighted share of each value of x, within one study
wshare <- function(x, w) {
  ok <- !is.na(x) & !is.na(w)
  if (!any(ok)) return(NULL)
  tapply(w[ok], x[ok], sum) / sum(w[ok])
}
```

## Table 1: reported vote by study

Shares are among respondents who reported voting for a party, including
"Other party". Respondents who did not vote, did not know or refused to say
are left out of the denominator. A blank cell means that no respondent of
that study reported that party; the ADQ merged into the CAQ in 2012, so the
ADQ has no voters after 2008 and the CAQ none before 2012.

```{r vote}
options(knitr.kable.NA = "")
parties <- c("PLQ", "PQ", "ADQ", "CAQ", "QS", "PVQ", "ON", "Other party")
voters <- master[master$vote_choice %in% parties, ]
vote <- do.call(rbind, lapply(split(voters, voters$qes_code), function(d) {
  s <- wshare(d$vote_choice, d$survey_weight)
  data.frame(study = d$qes_code[1], year = d$year[1], family = d$family[1],
             n = nrow(d), party = names(s), pct = round(100 * as.numeric(s), 1))
}))
vote_wide <- reshape(vote, idvar = c("study", "year", "family", "n"),
                     timevar = "party", direction = "wide")
names(vote_wide) <- sub("^pct\\.", "", names(vote_wide))
vote_wide <- vote_wide[order(vote_wide$family, vote_wide$year),
                       c("study", "year", "family", "n", intersect(parties, names(vote_wide)))]
knitr::kable(
  vote_wide, row.names = FALSE,
  col.names = c(tr(c("Study", "Year", "Family", "N voters"),
                   c("Étude", "Année", "Famille", "N votants")),
                intersect(parties, names(vote_wide)))
)
```

## Figure 1: the main parties

```{r vote-plot, fig.alt = "Weighted reported vote share of the PLQ, PQ, ADQ, CAQ and QS in each study, by year, with Quebec Election Studies and Durand panels in separate panels."}
main <- vote[vote$party %in% c("PLQ", "PQ", "ADQ", "CAQ", "QS"), ]
main$family <- ifelse(main$family == "qes",
                      tr("Quebec Election Studies", "Études électorales québécoises"),
                      tr("Durand panels", "Panels Durand"))
ggplot(main, aes(x = year, y = pct, colour = party)) +
  geom_line() +
  geom_point(size = 2.2) +
  facet_wrap(~family, ncol = 1) +
  scale_colour_manual(values = c(PLQ = "#d71920", PQ = "#004c9d", ADQ = "#6d9eeb",
                                 CAQ = "#00a3c7", QS = "#ff6f00")) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(
    x = tr("Year of the study", "Année de l'étude"),
    y = tr("Share of reported votes, weighted (%)", "Part des votes déclarés, pondérée (%)"),
    colour = tr("Party", "Parti")
  ) +
  theme_minimal(base_size = 12)
```

## Notes

- A line joins studies of the same family only; the ADQ and the CAQ are
  separate parties, and their lines are not joined.
- "Don't know / Refused" is a category qesR 0.4.4 coded on purpose, kept as
  it is in `vote_choice` until the harmonized variables of qesR 0.7.0.
- Reported votes differ from official results: compare them with the
  results published by Élections Québec, not with each other across
  designs.
