---
title: "Coverage by study"
params:
  lang: "en"
---

*[Version française](fr-couverture.html)*

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# messages in the language of the page, whatever the locale of the build
options(qesR.lang = params$lang)
```

Which harmonized variable ("target") of `qes_harmonize()` each study has,
and how comparable its question is. The grid and the table of studies below
are generated from the rules that ship with qesR.

```{r coverage, echo = FALSE, results = "asis"}
cat(qesR:::.spec_coverage_md(lang = params$lang, site = TRUE))
```

## Relaxed columns of qes_decon()

`qes_decon()` returns one column per concept for every study, relaxed
where the questions differ ([One file for every study](decon.html)). The
grid says where each study's values come from: *strict* where the column
is built from a target or a pooled variable of the grid above, *relaxed*
where a relaxed mapping of the study's own question gives them, and
*relaxed (awaiting sign-off)* where that mapping is not yet signed off, so that
`qes_decon()` leaves the column missing there for now. Relaxed columns
carry no grade, and a mapping leaves a study's column missing where its
categories straddle the column's.

```{r relaxed, echo = FALSE, results = "asis"}
fr <- identical(params$lang, "fr")
rx <- qesR::qes_spec("relaxed", lang = params$lang)
studies <- intersect(names(rx), qesR::qes_studies()$study)
word <- c(strict = if (fr) "stricte" else "strict", relaxed = if (fr) "souple" else "relaxed")
ref <- if (fr) "fr-reference-harmonisation.html" else "harmonization-reference.html"
rows <- lapply(seq_len(nrow(rx)), function(j) {
  v <- unlist(rx[j, studies])
  # a relaxed mapping not yet signed off: "relaxed (review)"
  held <- !is.na(v) & startsWith(v, "relaxed (")
  v[held] <- paste0(word[["relaxed"]], if (fr) " (en attente d\u2019approbation)" else " (awaiting sign-off)")
  cells <- ifelse(is.na(v), "\u2014", ifelse(v %in% names(word), word[v], v))
  c(sprintf("[`%s`](%s#relaxed-%s)<br>%s", rx$column[j], ref, rx$column[j], rx$label[j]), cells)
})
cat(qesR:::.qes_md_table(c(if (fr) "Colonne" else "Column", studies), rows), sep = "\n")
```

## The same grid in R

`qes_spec()` returns the grid as a data frame, one row per target and one
column per study, each cell the grade of that study's question:

```{r spec, eval = FALSE}
library(qesR)
qes_spec(lang = "en")
```

A grade compares a study's question with the target's anchor question.
[How harmonization works](harmonization.html) shows how grades, missing
values and weights carry into an estimate.
