---
title: "Coverage by study"
params:
  lang: "en"
---

*[Version française](fr-couverture.html)*

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# messages in the language of the page, whatever the locale of the build
options(qesR.lang = params$lang)
```

Which harmonized variable ("target") of `qes_harmonize()` each study has,
and how comparable its question is. The grid and the table of studies below
are generated from the specification that ships with qesR when the site is
built; none of it is written by hand.

```{r coverage, echo = FALSE, results = "asis"}
cat(qesR:::.spec_coverage_md(lang = params$lang))
```

## The same grid in R

`qes_spec()` returns the grid as a data frame, one row per target and one
column per study, each cell the grade of that study's question:

```{r spec, eval = FALSE}
library(qesR)
qes_spec(lang = params$lang)
```

A grade compares a study's question with the target's anchor question.
[Harmonizing across studies](harmonization.html) shows how grades, missing
values and weights carry into an estimate.
