---
title: "Coverage by study"
params:
  lang: "en"
---

*[Version française](fr-couverture.html)*

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# messages in the language of the page, whatever the locale of the build
options(qesR.lang = params$lang)
```

Which harmonized variable ("target") of `qes_harmonize()` each study has,
and how comparable its question is. The grid and the table of studies below
are generated from the rules that ship with qesR.

```{r coverage, echo = FALSE, results = "asis"}
cat(qesR:::.spec_coverage_md(lang = params$lang, site = TRUE))
```

## The same grid in R

`qes_spec()` returns the grid as a data frame, one row per target and one
column per study, each cell the grade of that study's question:

```{r spec, eval = FALSE}
library(qesR)
qes_spec(lang = params$lang)
```

A grade compares a study's question with the target's anchor question.
[How harmonization works](harmonization.html) shows how grades, missing
values and weights carry into an estimate.
