---
title: "One file for every study: qes_decon()"
description: >
  One data frame for the 11 harmonized studies, one column per concept
  even where the questions differ: what qes_decon() gives, how each
  column was relaxed, and when the strict qes_harmonize() is the better
  tool.
params:
  lang: "en"
---

*[Version française](fr-decon.html)*

```{r setup, include = FALSE}
source("_setup.R", encoding = "UTF-8")
```

```{r data, include = FALSE}
d <- qes_decon(lang = params$lang, quiet = TRUE)

lead_cols <- c("study", "year", "wave", "qes_id", "weight", "weight_var")
rx_cols <- setdiff(names(d), lead_cols)
n_rows <- nrow(d)
n_resp <- length(unique(d$qes_id))
n_studies <- length(unique(d$study))
big <- function(x) formatC(x, format = "d", big.mark = tr(",", " "))
# "a, b and c" / "a, b et c"
and_list <- function(x) if (length(x) < 2) as.character(x) else
  paste0(paste(head(x, -1), collapse = ", "), tr(" and ", " et "), tail(x, 1))

# the weight of each study and wave, and why it is missing where it is
w <- attr(d, "weights")
w_ok <- tapply(w$reason, w$study, function(r) all(is.na(r)))
weighted_studies <- intersect(qz_studies, names(w_ok)[w_ok])
no_weight <- intersect(qz_studies, unique(w$study[w$reason %in% "no_recommended_weight"]))
review_weight <- intersect(qz_studies, unique(w$study[w$reason %in% "needs_review"]))

# the studies in which each column has at least one value
has_value <- sapply(rx_cols, function(v) tapply(!is.na(d[[v]]), d$study, any))
covered <- lapply(rx_cols, function(v) intersect(qz_studies, rownames(has_value)[has_value[, v]]))
names(covered) <- rx_cols
n_cov <- lengths(covered)
src <- attr(d, "decon_sources")
```

```{r strings, include = FALSE}
# French typography in the figure and table text: a non-breaking space
# before : ; ? ! % and inside « » (a "%s" placeholder is left alone)
trs <- function(en, fr) tr(en, gsub("« ", "« ", gsub(" (?=[:;?!»]|%(?!s))", " ", fr, perl = TRUE)))
short_study <- c(qes1998 = "1998", qes2007 = "2007", qes2007_panel = tr("2007 panel", "panel 2007"),
                 qes2008 = "2008", qes_crop_2007_2010 = "CROP", qes2012 = "2012",
                 qes2012_panel = tr("2012 panel", "panel 2012"), qes2014 = "2014", qes2018 = "2018",
                 qes2018_panel = tr("2018 panel", "panel 2018"), qes2022 = "2022")
reference <- tr("harmonization-reference.html", "fr-reference-harmonisation.html")
```

For some years now, the Canadian Election Study has had a simple point of
entry. The [cesR](https://hodgettsp.github.io/cesR/) package's
`get_decon()` returns one data frame of the 2019 study, with 21 variables
under plain names (`education`, `income_cat`, `lr`, `econ_retro`...) and
readable labels. The Quebec Election Studies had no such file. Their
harmonization in qesR, `qes_harmonize()`, is strict by design: one target
per question, a comparability grade for each study's question and a
reason for each missing value. It is the right tool when the difference
between two questions matters. However, many uses do not call for that
much: a descriptive table, a control variable, a first look at 25 years of
data. `qes_decon()` answers this need. It returns one data frame for the
`r n_studies` harmonized studies, with `r big(n_rows)` rows (one per
respondent and wave) and `r length(rx_cols)` columns of data, one per
concept, even where the wording or the answer options differ from one
study to the next. The price is exactness. The categories are coarse, and
no column claims that two studies asked the same question.

## What the file holds

```{r load, eval = FALSE}
d <- qes_decon(lang = params$lang)
```

`lang = "fr"` gives the labels in French (`params$lang` is the language of
this page); the column names stay the same. The first rows:

```{r peek}
d[1:4, c("study", "wave", "weight", "gender", "education", "vote_choice", "vote_type")]
```

The first columns identify the row: `study`, `year`, `wave` and `qes_id`,
the respondent's identifier within the study. The file has
`r big(n_resp)` respondents, and a respondent of a panel has one row per
wave. The socio-demographic columns (education, income, language,
religion, region...) repeat on every row of a respondent, while the vote,
turnout and attitudes sit on the row of the wave that asked them: the vote
intention on the campaign wave, the reported vote on the post-election
wave. `weight` is the recommended weight of the study and wave, rescaled
to a mean of 1 within each, and `weight_var` names its source. Of the
`r n_studies` studies, `r length(weighted_studies)` have such a weight on
every row: `r and_list(qz_study_label(weighted_studies))`. The others have
`NA`. The `r and_list(qz_study_label(no_weight))` has no recommended
weight, since its weights are calibrated on the reported vote, and the
weights of the following studies are still under review:
`r and_list(qz_study_label(review_weight))`. `attr(d, "weights")` gives
the detail, wave by wave.

## The columns

The table below is generated from the relaxed specification that ships
with qesR (`qes_spec("relaxed")`). A study counts as covered when the
column has at least one value there. Each name links to the column's
section of the [variable reference](harmonization-reference.html), which
gives each study's source variable and recoding rule.

```{r columns, echo = FALSE}
rx <- as.data.frame(qes_spec("relaxed", lang = params$lang))
rx <- rx[match(rx_cols, rx$column), ]
cats <- vapply(seq_len(nrow(rx)), function(j) {
  v <- d[[rx$column[j]]]
  if (!is.factor(v)) return(sprintf(tr("number, %s", "nombre, %s"), rx$levels[j]))
  lv <- levels(v)
  if (length(lv) > 6) return(sprintf(tr("%d categories (%s...)", "%d catégories (%s...)"), length(lv),
                                     paste(head(lv, 2), collapse = ", ")))
  paste(lv, collapse = " · ")
}, character(1))
studies_txt <- vapply(rx$column, function(v) {
  s <- covered[[v]]
  if (length(s) == length(qz_studies)) return(tr("all 11", "les 11"))
  paste0(length(s), " (", paste(short_study[s], collapse = ", "), ")")
}, character(1))
tab <- data.frame(
  sprintf("[`%s`](%s#relaxed-%s)", rx$column, reference, rx$column),
  rx$label, cats, studies_txt, rx$relaxed
)
names(tab) <- c(tr("Column", "Colonne"), tr("Concept", "Concept"), tr("Categories", "Catégories"),
                tr("Studies", "Études"), tr("How it was relaxed", "Comment elle a été assouplie"))
knitr::kable(tab, row.names = FALSE, escape = FALSE)
```

`r sum(n_cov == length(qz_studies))` columns cover all 11 studies, and
`r sum(n_cov >= 5)` cover at least five. The other `r sum(n_cov < 5)`
were asked in `r min(n_cov)` to `r max(n_cov[n_cov < 5])` studies, and are
kept for analyses within those studies. Two columns are companions:
`vote_type` says whether `vote_choice` comes from the reported vote or
from a vote intention, and `sovereignty_type` which referendum question
`sovereignty` comes from. The rule of thumb is simple. Within one study, a
relaxed column loses detail but not meaning; across studies, it is as
comparable as the sentence in the last column says, and no more.

## How each column was relaxed

Each column carries its relaxation rule and its sources:

```{r education}
attr(d$education, "relaxed")
head(attr(d$education, "sources")[, c("study", "wave", "source_var", "recode")], 3)
```

```{r education-facts, include = FALSE}
edu_src <- src[src$column == "education", ]
edu_empty <- intersect(qz_studies, edu_src$study[edu_src$n_value == 0])
```

Education is a good example of what relaxing means. The studies asked
about schooling in different ways, from detailed lists of diplomas to
bands of years of schooling, and `qes_decon()` groups every answer into
four categories: no high school diploma, high school, college (CEGEP,
technical or trade school) and university. The grouping is coarse on
purpose; it is the finest one every study supports. Where a study's
categories straddle two groups, the column stays missing rather than
guess: this is the case of the `r and_list(qz_study_label(edu_empty))`,
whose question cannot be split into the four groups.
`attr(d, "decon_sources")` gathers the sources of every column, with the
number of values and the reasons for missing ones;
`qes_spec("relaxed_maps")` lists the relaxed mappings themselves, with
their review status.

## A weighted estimate in one study

An estimate is made within one study and one wave, with that wave's
weight. Here, the reported vote for the Coalition avenir Québec (CAQ) by
education in 2022, from the post-election wave (`pes`):

```{r one-study}
d22 <- subset(d, study == "qes2022" & wave == "pes" & !is.na(vote_choice) & !is.na(education))
d22$caq <- as.numeric(d22$vote_choice == "CAQ")
des <- survey::svydesign(ids = ~1, weights = ~weight, data = d22)
survey::svyby(~caq, ~education, des, survey::svymean)
```

```{r one-study-facts, include = FALSE}
caq22 <- survey::svyby(~caq, ~education, des, survey::svymean)
caq_edu <- setNames(100 * caq22$caq, as.integer(caq22$education))
```

The education gradient is steep. In 2022, `r qz_pct(caq_edu[["1"]], 0)`
of respondents without a high school diploma report voting for the CAQ,
against `r qz_pct(caq_edu[["4"]], 0)` of university graduates. Before
reading too much into it, note that education is related to mother tongue,
and that non-francophones voted for the CAQ far less than francophones
did. The next example holds language constant.

## Across elections: a diploma divide?

It is often said that the CAQ, like the Action démocratique du Québec
(ADQ) before it, draws its support from voters without a university
degree. The relaxed columns allow a quick test over six elections, among
francophones only. The ADQ ran in 2007 and 2008, the CAQ from 2012 on;
the two parties stay separate in `vote_choice`, and each election has one
of them.

```{r elections}
elections <- c("qes2007", "qes2008", "qes2012", "qes2014", "qes2018", "qes2022")
d$edu3 <- cut(as.integer(d$education), c(0, 2, 3, 4), labels = c("hs", "college", "university"))
est <- do.call(rbind, lapply(elections, function(s) {
  x <- subset(d, study == s & wave %in% c("post", "pes") & language %in% c("French", "Français") &
                !is.na(vote_choice) & !is.na(edu3))
  weighted <- !anyNA(x$weight)
  if (!weighted) x$weight <- 1
  x$adq_caq <- as.numeric(x$vote_choice %in% c("ADQ", "CAQ"))
  des <- survey::svydesign(ids = ~1, weights = ~weight, data = x)
  e <- survey::svyby(~adq_caq, ~edu3, des, survey::svymean)
  data.frame(study = s, edu3 = e$edu3, pct = 100 * e$adq_caq, se = 100 * e$se,
             n = as.vector(table(x$edu3)), weighted = weighted)
}))
```

```{r elections-facts, include = FALSE}
est$year <- qz_year(est$study)
ci <- qz_logit_ci(est$pct / 100, est$se / 100)
est$lo <- 100 * ci[, "lo"]
est$hi <- 100 * ci[, "hi"]
ev <- function(y, g) est$pct[est$year == y & est$edu3 == g]
gap <- function(y) ev(y, "hs") - ev(y, "university")
gaps <- sapply(qz_year(elections), gap)
names(gaps) <- qz_year(elections)
# the text below reads these patterns off the estimates
stopifnot(which.max(gaps) == length(gaps), abs(gaps[["2008"]]) < 2, all(abs(gaps[c("2012", "2014")]) < 7),
          gaps[["2007"]] > 9, gaps[["2018"]] > 12)
off <- qz_official()
adq_off <- setNames(off$share_valid[off$party == "ADQ" & off$year %in% c(2007, 2008)],
                    off$year[off$party == "ADQ" & off$year %in% c(2007, 2008)])
```

```{r fig-divide, echo = FALSE}
edu_lab <- c(hs = tr("High school or less", "Secondaire ou moins"),
             college = tr("College or CEGEP", "Collégial ou cégep"),
             university = tr("University", "Universitaire"))
fd <- est
fd$edu3 <- factor(as.character(fd$edu3), levels = names(edu_lab))
fd$party <- ifelse(fd$year < 2012, "ADQ", "CAQ")
fd$x <- fd$year + c(hs = -0.2, college = 0, university = 0.2)[as.character(fd$edu3)]
fd$line <- paste(fd$party, fd$edu3)
fd$label <- edu_lab[as.character(fd$edu3)]
tags <- data.frame(x = c(2007.5, 2017), y = 62, label = c("ADQ", "CAQ"))

fig_divide <- function(mode) {
  k <- qz_tokens[[mode]]
  cols <- setNames(k$ord3, names(edu_lab))
  ggplot(fd, aes(x, pct, colour = edu3, group = line)) +
    annotate("segment", x = 2010, xend = 2010, y = 0, yend = 64, colour = k$axis, linewidth = 0.35,
             linetype = "22") +
    geom_text(data = tags, aes(x, y, label = label), inherit.aes = FALSE, size = 3.3, fontface = "bold",
              colour = k$ink2, family = qz_font()) +
    qz_ci(aes(ymin = lo, ymax = hi), show.legend = FALSE) +
    qz_line(show.legend = FALSE) +
    qz_dots(mode, fd, aes(x, pct), colour_aes = "edu3", key = FALSE) +
    qz_end_labels(fd, "x", "pct", "label", "edu3", mode, min_gap = 4, nudge_x = 1.2) +
    scale_colour_manual(values = cols, guide = "none") +
    scale_fill_manual(values = cols, breaks = names(edu_lab), labels = edu_lab, guide = "none") +
    scale_x_continuous(breaks = qz_year(elections), labels = function(x) ifelse(x == 2008, "’08", x),
                       limits = c(2006.4, 2028.5), expand = expansion(0)) +
    scale_y_continuous(breaks = seq(0, 60, 20), labels = qz_pct_axis) +
    coord_cartesian(ylim = c(0, 65), clip = "off") +
    labs(x = NULL, y = tr("Reported vote for the ADQ or the CAQ", "Vote déclaré pour l'ADQ ou la CAQ")) +
    theme_qes(mode)
}

tab <- fd[order(fd$year, fd$edu3), ]
tab_divide <- setNames(data.frame(
  tab$year, tab$party, tab$label, qz_cell(tab$pct, tab$lo, tab$hi, tab$n), tab$n,
  ifelse(tab$weighted, tr("weighted", "pondéré"), tr("unweighted (no recommended weight)", "non pondéré (aucune pondération recommandée)")),
  row.names = NULL
), c(tr("Election", "Élection"), tr("Party", "Parti"), tr("Education", "Scolarité"),
     trs("Reported vote, % [95% CI]", "Vote déclaré, % [IC à 95 %]"), "n", tr("Weighting", "Pondération")))
alt <- sprintf(trs(
  "Line chart of the share of francophone respondents who report voting for the ADQ (2007, 2008) or the CAQ (2012 to 2022), for three education groups, with 95%% confidence intervals. The gap between those with high school or less and university graduates is %s points in 2007, %s in 2012, %s in 2018 and %s in 2022. Values in the table view.",
  "Graphique linéaire de la part des répondants francophones qui déclarent avoir voté pour l'ADQ (2007, 2008) ou la CAQ (2012 à 2022), pour trois groupes de scolarité, avec intervalles de confiance à 95 %%. L'écart entre les personnes qui ont au plus un diplôme d'études secondaires et les diplômés universitaires est de %s points en 2007, de %s en 2012, de %s en 2018 et de %s en 2022. Valeurs dans la vue en tableau."),
  qz_num(gap(2007), 0), qz_num(gap(2012), 0), qz_num(gap(2018), 0), qz_num(gap(2022), 0))
qz_figure(fig_divide, "divide",
  title = sprintf(trs("Among francophones, the education gap in the CAQ vote went from %s points in 2014 to %s in 2022",
                      "Chez les francophones, l'écart de scolarité dans le vote caquiste passe de %s points en 2014 à %s en 2022"),
                  qz_num(gap(2014), 0), qz_num(gap(2022), 0)),
  subtitle = trs("Share of francophone respondents who report voting for the ADQ (2007, 2008) or the CAQ (2012 to 2022), by education, with 95% confidence intervals",
                 "Part des répondants francophones qui déclarent avoir voté pour l'ADQ (2007, 2008) ou la CAQ (2012 à 2022), selon la scolarité, avec intervalles de confiance à 95 %"),
  alt = alt, table = tab_divide, height = 4.0,
  note = trs("Source: qes_decon(), relaxed columns vote_choice (reported vote, post-election wave), education (in three groups: no diploma and high school joined) and language (mother tongue French). One Quebec Election Study per election. The ADQ and the CAQ are separate parties and stay separate in the data; the dashed line marks the change of party. Weighted with each study's post-election weight; hollow dots (2008): unweighted, as the 2008 study has no recommended weight.",
             "Source : qes_decon(), colonnes souples vote_choice (vote déclaré, vague postélectorale), education (en trois groupes : sans diplôme et secondaire réunis) et language (langue maternelle française). Une Étude électorale québécoise par élection. L'ADQ et la CAQ sont des partis distincts et le restent dans les données ; le trait pointillé marque le changement de parti. Pondéré avec la pondération postélectorale de chaque étude ; points creux (2008) : non pondéré, l'étude de 2008 n'ayant pas de pondération recommandée."))
```

The conventional wisdom finds mixed support. In 2007, francophones with
high school or less were `r qz_num(gap(2007), 0)` points more likely than
university graduates to report voting for the ADQ. The gap then all but
disappeared. In 2008, when the ADQ's share of the vote fell from
`r qz_pct(adq_off[["2007"]], 0)` to `r qz_pct(adq_off[["2008"]], 0)`,
graduates and the least educated voted for it in the same proportion, and
the CAQ's gap was
`r if (qz_num(gap(2012), 0) == qz_num(gap(2014), 0)) paste(qz_num(gap(2012), 0), "points in both 2012 and 2014") else paste(qz_num(gap(2012), 0), "points in 2012 and", qz_num(gap(2014), 0), "in 2014")`.
It opened again when the CAQ won: `r qz_num(gap(2018), 0)` points in 2018
and `r qz_num(gap(2022), 0)` in 2022, when `r qz_pct(ev(2022, "hs"), 0)`
of francophones with high school or less voted for the CAQ, against
`r qz_pct(ev(2022, "university"), 0)` of university graduates. In other
words, the diploma divide is not a constant of these parties' vote. It is
wide in the elections where they did well (2007, 2018 and 2022) and narrow
in the others.

Two cautions apply, and both come from the relaxation. First, the
education groups are not drawn from the same question in every study, so
a small change in a gap from one election to the next may come from the
grouping rather than from the voters; the changes of 2018 and 2022 are
well beyond that. Second, the reported vote of the 2008 study is
unweighted. Neither caution affects the comparison within an election,
which is where the relaxed columns stand on their firmest ground.

## qes_decon() or qes_harmonize()?

The two functions answer different needs, and neither replaces the other.
`qes_decon()` fits descriptive tables, control variables, teaching and a
first look across studies: one call, plain names, every study. Its columns
carry no grade, and its categories are broad enough for every study to
fit them. `qes_harmonize()` fits analyses where the difference between two
questions matters. It keeps one target per question, grades each study's
question against an anchor, records a reason for each missing value on
every row, and works with `qes_design()` for survey estimates.

Three cases call for other functions. First, a trend in a level: the
share who would vote Yes in a referendum moves with the question asked (in
the Quebec Election Studies, the 1995 question in 2007 and 2008, an
independent country from 2012 on), and `sovereignty` pools them all.
`sovereignty_type` says which question each row comes from, and
`qes_harmonize(targets = "sov_indep")` keeps only one wording. Second, an
analysis that should drop the less comparable studies: the grades and
`min_grade` do this, and the relaxed columns have neither. Third, code
written for qesR 0.4.4: `get_decon()` still returns its columns for one
study, unchanged, and `get_qes_master()` the merged file; `qes_decon()` is
a new function, not their replacement. `?qes_decon` documents every
column.
