---
title: "Two dimensions of competition: sovereignty and left-right"
description: >
  Where each party's voters stand on sovereignty and on the left-right
  scale, and where sovereigntists and federalists cast their vote.
params:
  lang: "en"
---

*[Version française](fr-dimensions.html)*

```{r setup, include = FALSE}
source("_setup.R", encoding = "UTF-8")
```

```{r harmonize, echo = FALSE}
h <- qes_harmonize(
  studies = qz_studies,
  targets = c("vote_choice", "sov_support", "lr_self"),
  types = list(vote_choice = "recall"),
  missing = "reasons", quiet = TRUE
)
```

```{r estimates, echo = FALSE}
h$party <- factor(qz_party_code(h$vote_choice), levels = qz_parties)
h$yes <- factor(ifelse(h$sov_support %in% "Yes", "yes",
                       ifelse(h$sov_support %in% "No", "no", NA)), levels = c("yes", "no"))
dim_studies <- c("qes2012", "qes2014", "qes2018", "qes2022")
bar_studies <- c("qes2007", "qes2008", dim_studies)

electorates <- do.call(rbind, lapply(dim_studies, function(s) {
  d <- qz_design(h, s, "weight_post")
  lr <- qz_mean(d, "lr_self", by = "party")
  yes <- qz_share(d, "yes", by = "party", levels = "yes")
  out <- merge(setNames(lr, c("party", "lr", "lr_lo", "lr_hi", "lr_n", "weighted")),
               setNames(yes[, c("group", "pct", "lo", "hi", "n")], c("party", "yes", "yes_lo", "yes_hi", "yes_n")))
  qz_tag(out, study = s)
}))
electorates$year <- qz_year(electorates$study)
electorates$status <- qz_party_status(electorates$study, electorates$party)
electorates <- electorates[electorates$status == "offered" & electorates$party != "other", ]

where <- do.call(rbind, lapply(bar_studies, function(s) {
  d <- qz_design(h, s, "weight_post")
  qz_tag(qz_share(d, "party", by = "yes"), study = s)
}))
where$year <- qz_year(where$study)
where$status <- qz_party_status(where$study, where$level)
ev <- function(s, p, col = "yes") electorates[[col]][electorates$study == s & electorates$party == p]
wv <- function(s, g, p) where$pct[where$study == s & where$group == g & where$level == p]
```

For forty years the PQ and the PLQ split Quebec on the national question.
The Quebec Election Studies show that this axis no longer sorts the
parties the way it did. In 2012, `r qz_pct(ev("qes2012", "PQ"), 0)` of PQ
voters would have voted Yes to an independent country, against
`r qz_pct(ev("qes2012", "PLQ"), 0)` of PLQ voters. By 2022 the Yes vote was
split: the PQ took `r qz_pct(wv("qes2022", "yes", "PQ"), 0)` of it, Québec
solidaire `r qz_pct(wv("qes2022", "yes", "QS"), 0)` and the CAQ
`r qz_pct(wv("qes2022", "yes", "CAQ"), 0)`, while in 2012 the PQ alone took
`r qz_pct(wv("qes2012", "yes", "PQ"), 0)`. On the left-right scale, Québec
solidaire's voters are the furthest left
(`r qz_num(ev("qes2022", "QS", "lr"))` out of 10 in 2022) and the
Conservatives' the furthest right (`r qz_num(ev("qes2022", "PCQ", "lr"))`).

## The data

Three harmonized variables: the pooled reported vote, the pooled
referendum vote (`sov_support`) and the self-placement on the left-right
scale (`lr_self`, 0 to 10):

```{r show-harmonize, ref.label = "harmonize", eval = FALSE}
```

`lr_self` was asked in the Quebec Election Studies of 2012 to 2022; in
2022 it and the referendum question were asked during the campaign, the
reported vote after the election, so the page keeps the respondents of
both waves and weights them with the post-election weight. For one party
in one study:

```{r example}
d22 <- qes_design(h[h$study == "qes2022", ], weight = "weight_post")
svyby(~lr_self, ~vote_choice, subset(d22, vote_choice %in% c("CAQ", "QS")),
      svymean, na.rm = TRUE)
```

## Party electorates on two axes, 2012 to 2022

```{r fig-paths, echo = FALSE}
path_parties <- c("PLQ", "PQ", "CAQ", "QS", "PCQ")
e <- electorates[electorates$party %in% path_parties & electorates$lr_n >= qz_min_n & electorates$yes_n >= qz_min_n, ]
measure_lab <- c(yes = tr("Would vote Yes (among Yes and No)", "Voteraient Oui (parmi Oui et Non)"),
                 lr = tr("Left-right, mean (0 = left, 10 = right)", "Gauche-droite (0 = gauche, 10 = droite)"))
d <- rbind(data.frame(e[, c("party", "year", "weighted")], measure = "yes", value = e$yes, lo = e$yes_lo, hi = e$yes_hi),
           data.frame(e[, c("party", "year", "weighted")], measure = "lr", value = e$lr, lo = e$lr_lo, hi = e$lr_hi))
d$party <- factor(d$party, levels = path_parties)
d$panel <- factor(measure_lab[d$measure], levels = measure_lab)
# the parties side by side at each election, so their intervals do not overlap
d$x <- d$year + c(PLQ = -0.36, PQ = -0.18, CAQ = 0, QS = 0.18, PCQ = 0.36)[as.character(d$party)]
d$label <- qz_party_label(d$party)
# the two panels have their own scales: a share and a mean on 0-10
blank <- data.frame(panel = factor(measure_lab, levels = measure_lab), x = 2012, value = c(0, 2.5), top = c(100, 7.4))

fig_paths <- function(mode) {
  k <- qz_tokens[[mode]]
  ggplot(d, aes(x, value, colour = party, group = party)) +
    geom_blank(data = blank, aes(x, value), inherit.aes = FALSE) +
    geom_blank(data = blank, aes(x, top), inherit.aes = FALSE) +
    geom_linerange(aes(ymin = lo, ymax = hi), linewidth = 0.4, alpha = 0.45) +
    qz_line() +
    qz_dots(mode, d, aes(x, value), colour_aes = "party", size = 2.4) +
    qz_end_labels(d[d$measure == "yes", ], "x", "value", "label", "party", mode, min_gap = 5.5, nudge_x = 1.4, size = 3) +
    qz_end_labels(d[d$measure == "lr", ], "x", "value", "label", "party", mode, min_gap = 0.24, nudge_x = 1.4, size = 3) +
    facet_wrap(~panel, nrow = 1, scales = "free_y") +
    scale_party(mode) +
    scale_x_continuous(breaks = c(2012, 2014, 2018, 2022), limits = c(2011.3, 2026.2), expand = expansion(0)) +
    # the share panel starts at 0, the left-right panel at 2.5: 0 and above 10 are percentages
    scale_y_continuous(labels = function(x) ifelse(x > 10 | x == 0, qz_pct_axis(x), qz_num(x, 0)), expand = expansion(0)) +
    labs(x = NULL, y = NULL) +
    theme_qes(mode) +
    theme(panel.spacing.x = unit(30, "pt"), strip.text = element_text(size = 9.5))
}

tab <- electorates[order(match(electorates$party, qz_parties), electorates$year), ]
tab_paths <- setNames(data.frame(
  qz_party_label(tab$party), tab$year, qz_study_label(tab$study),
  qz_cell(tab$lr, tab$lr_lo, tab$lr_hi, tab$lr_n, 2), tab$lr_n,
  qz_cell(tab$yes, tab$yes_lo, tab$yes_hi, tab$yes_n), tab$yes_n, qz_weight_text(tab$weighted),
  row.names = NULL
), c(tr("Party", "Parti"), tr("Election", "Élection"), tr("Study", "Étude"),
     tr("Left-right, mean [95% CI]", "Gauche-droite, moyenne [IC à 95 %]"), tr("n (left-right)", "n (gauche-droite)"),
     tr("Yes, % [95% CI]", "Oui, % [IC à 95 %]"), tr("n (Yes/No)", "n (Oui/Non)"), tr("Weighting", "Pondération")))
alt <- sprintf(tr(
  "Two line charts, 2012 to 2022, one line per party (PLQ, PQ, CAQ, QS; the PCQ in 2022 only): on the left the share of the party's voters who would vote Yes to independence, on the right their mean left-right self-placement, with confidence intervals. PQ voters go from %s to %s Yes; CAQ voters from %s to %s; QS voters from %s to %s. PLQ voters move from %s to %s on the left-right scale; QS voters stay the furthest left, at %s in 2022. Values in the table view.",
  "Deux graphiques linéaires, de 2012 à 2022, une ligne par parti (PLQ, PQ, CAQ, QS ; le PCQ en 2022 seulement) : à gauche la part des électeurs du parti qui voteraient Oui à l'indépendance, à droite leur autopositionnement gauche-droite moyen, avec intervalles de confiance. Les électeurs du PQ passent de %s à %s de Oui ; ceux de la CAQ, de %s à %s ; ceux de QS, de %s à %s. Les électeurs du PLQ passent de %s à %s sur l'axe gauche-droite ; ceux de QS restent les plus à gauche, à %s en 2022. Valeurs dans la vue en tableau."),
  qz_pct(ev("qes2012", "PQ"), 0), qz_pct(ev("qes2022", "PQ"), 0),
  qz_pct(ev("qes2012", "CAQ"), 0), qz_pct(ev("qes2022", "CAQ"), 0),
  qz_pct(ev("qes2012", "QS"), 0), qz_pct(ev("qes2022", "QS"), 0),
  qz_num(ev("qes2012", "PLQ", "lr")), qz_num(ev("qes2022", "PLQ", "lr")), qz_num(ev("qes2022", "QS", "lr")))
qz_figure(fig_paths, "paths",
  title = sprintf(tr("CAQ and QS voters converged on sovereignty (%s and %s Yes in 2022), while PLQ voters moved left",
                     "Les électeurs de la CAQ et de QS se sont rapprochés sur la souveraineté (%s et %s de Oui en 2022), et ceux du PLQ se sont déplacés vers la gauche"),
                  qz_pct(ev("qes2022", "CAQ"), 0), qz_pct(ev("qes2022", "QS"), 0)),
  subtitle = tr("Where each party's voters stand on sovereignty and on the left-right scale, 2012 to 2022, with 95% confidence intervals",
                "Où se situent les électeurs de chaque parti sur la souveraineté et sur l'axe gauche-droite, de 2012 à 2022, avec intervalles de confiance à 95 %"),
  alt = alt, table = tab_paths, height = 4.2,
  note = tr("Source: qesR, pooled vote_choice (reported vote) and sov_support (independence), lr_self; QES 2012, 2014, 2018 and 2022, weighted with each study's post-election weight. The parties are set side by side at each election. Lines: 95% confidence intervals (logit for the share, Wald for the mean). The two panels have their own scales. The PCQ was listed in 2022 only.",
            "Source : qesR, variables regroupées vote_choice (vote déclaré) et sov_support (indépendance), lr_self ; EEQ 2012, 2014, 2018 et 2022, pondérées avec la pondération postélectorale de chaque étude. Les partis sont côte à côte à chaque élection. Traits : intervalles de confiance à 95 % (logit pour la part, Wald pour la moyenne). Les deux panneaux ont leur propre échelle. Le PCQ n'était proposé qu'en 2022."))
```

What to notice:

- PQ and QS voters are the sovereigntist ones; PLQ voters are almost
  all federalists at every election (`r qz_pct(ev("qes2022", "PLQ"), 0)` Yes in
  2022).
- The two newer parties moved toward each other on sovereignty: CAQ
  voters from `r qz_pct(ev("qes2012", "CAQ"), 0)` Yes in 2012 to
  `r qz_pct(ev("qes2022", "CAQ"), 0)` in 2022, QS voters from
  `r qz_pct(ev("qes2012", "QS"), 0)` to `r qz_pct(ev("qes2022", "QS"), 0)`.
- PLQ voters moved left, from `r qz_num(ev("qes2012", "PLQ", "lr"))` to
  `r qz_num(ev("qes2022", "PLQ", "lr"))`; the voters right of centre are now
  those of the CAQ and the PCQ.
- PQ and QS voters are close on left-right
  (`r qz_num(ev("qes2012", "PQ", "lr") - ev("qes2012", "QS", "lr"))` points
  apart in 2012, `r qz_num(ev("qes2022", "PQ", "lr") - ev("qes2022", "QS", "lr"))`
  in 2022), but their gap on sovereignty grew from
  `r qz_num(ev("qes2012", "PQ") - ev("qes2012", "QS"), 0)` to
  `r qz_num(ev("qes2022", "PQ") - ev("qes2022", "QS"), 0)` points: the
  national question now separates them more than the left-right scale.

## Where sovereigntists and federalists vote

```{r fig-where, echo = FALSE}
line_parties <- c("PLQ", "PQ", "ADQ", "QS", "CAQ", "PCQ")
d <- where[where$group %in% c("yes", "no") & where$level %in% line_parties & where$status == "offered" &
             where$n >= qz_min_n, ]
d$party <- factor(d$level, levels = line_parties)
d$x <- d$year
# the question changes between 2008 (the 1995 question) and 2012 (independence): no line across
d$era <- ifelse(d$year < 2010, "q1995", "indep")
side_lab <- c(yes = tr("Would vote Yes", "Voteraient Oui"), no = tr("Would vote No", "Voteraient Non"))
d$facet <- factor(side_lab[d$group], levels = side_lab)
lab <- d[d$year == 2022, ]
# the QES 2022 under-reports the CAQ (see the survey-vs-official page)
caq_gap <- qz_official()$share_valid[qz_official()$year == 2022 & qz_official()$party == "CAQ"] -
  qz_share(qz_design(h, "qes2022", "weight_post"), "party", levels = "CAQ")$pct
lab$label <- qz_party_label(lab$party)

fig_where <- function(mode) {
  k <- qz_tokens[[mode]]
  ggplot(d, aes(x, pct, colour = party, group = interaction(party, era))) +
    annotate("rect", xmin = 2008.6, xmax = 2011.4, ymin = -Inf, ymax = Inf, fill = k$wash) +
    annotate("text", x = 2010, y = 78, label = tr("the question\nchanges", "la question\nchange"),
             size = 2.7, colour = k$ink2, family = qz_font(), lineheight = 0.9) +
    geom_linerange(aes(ymin = lo, ymax = hi), linewidth = 0.4, alpha = 0.45, show.legend = FALSE) +
    qz_line() +
    geom_point(aes(fill = party), shape = 21, size = 2.4, stroke = 0.8,
               colour = k$surface, data = d[d$year != 2008, ], show.legend = FALSE) +
    geom_point(aes(colour = party), shape = 21, size = 2.2, stroke = 1.1, fill = k$surface, data = d[d$year == 2008, ],
               show.legend = FALSE) +
    qz_end_labels(lab[lab$group == "yes", ], "x", "pct", "label", "party", mode, min_gap = 4.2, nudge_x = 1.6, size = 3) +
    qz_end_labels(lab[lab$group == "no", ], "x", "pct", "label", "party", mode, min_gap = 4.2, nudge_x = 1.6, size = 3) +
    facet_wrap(~facet, nrow = 1, axes = "all_x") +
    scale_party(mode, aesthetics = "colour", guide = guide_legend(order = 1, nrow = 1)) +
    scale_party(mode, aesthetics = "fill") +
    scale_shape_manual(values = c(weighted = 21, unweighted = 21), limits = c("weighted", "unweighted"),
                       breaks = "unweighted", labels = tr("hollow: unweighted (2008, weight under review)", "creux : non pondéré (2008, pondération en révision)"),
                       guide = guide_legend(order = 2, override.aes = list(fill = k$surface, colour = k$ink2, stroke = 1.1, size = 2.6,
                                                                          alpha = 1, linetype = 0))) +
    geom_point(data = d[1, ], aes(shape = "unweighted"), alpha = 0, colour = NA, show.legend = c(colour = FALSE, shape = TRUE)) +
    scale_x_continuous(breaks = c(2007, 2012, 2014, 2018, 2022), labels = c("2007", "2012", "’14", "2018", "2022"),
                       limits = c(2006.3, 2026.6), expand = expansion(0)) +
    scale_y_continuous(breaks = seq(0, 80, 20), labels = qz_pct_axis) +
    coord_cartesian(ylim = c(0, 84), expand = FALSE, clip = "off") +
    labs(x = NULL, y = tr("Share of the side's reported vote", "Part du vote déclaré du camp")) +
    theme_qes(mode) +
    theme(panel.spacing.x = unit(26, "pt"))
}

tab <- where[where$group %in% c("yes", "no") & where$level %in% c(line_parties, "other"), ]
tab <- tab[order(tab$year, tab$group == "no", match(tab$level, qz_parties)), ]
tab_where <- setNames(data.frame(
  tab$year, qz_study_label(tab$study),
  ifelse(tab$group == "yes", tr("Yes", "Oui"), tr("No", "Non")),
  qz_party_label(tab$level),
  ifelse(tab$status == "not_run", "—", ifelse(tab$status == "not_listed", tr("n.l.", "n.p."),
                                                  qz_cell(tab$pct, tab$lo, tab$hi, tab$n))),
  tab$n, qz_weight_text(tab$weighted), row.names = NULL
), c(tr("Election", "Élection"), tr("Study", "Étude"), tr("Referendum vote", "Vote référendaire"),
     tr("Party", "Parti"), tr("Share of reported vote [95% CI]", "Part du vote déclaré [IC à 95 %]"),
     "n", tr("Weighting", "Pondération")))
alt <- sprintf(tr(
  "Line charts in two panels, the respondents who would vote Yes (left) and No (right) in a referendum, from 2007 to 2022: the share of each side's reported vote that went to each party. The PQ's share of the Yes vote goes from %s in 2012 to %s in 2022, as QS rises to %s and the CAQ to %s; the PLQ's share of the No vote goes from %s in 2012 to %s in 2022. Values in the table view.",
  "Graphiques linéaires en deux panneaux, les répondants qui voteraient Oui (à gauche) et Non (à droite) à un référendum, de 2007 à 2022 : la part du vote déclaré de chaque camp allée à chaque parti. La part du PQ chez les électeurs du Oui passe de %s en 2012 à %s en 2022, tandis que QS monte à %s et la CAQ à %s ; celle du PLQ chez les électeurs du Non passe de %s en 2012 à %s en 2022. Valeurs dans la vue en tableau."),
  qz_pct(wv("qes2012", "yes", "PQ"), 0), qz_pct(wv("qes2022", "yes", "PQ"), 0),
  qz_pct(wv("qes2022", "yes", "QS"), 0), qz_pct(wv("qes2022", "yes", "CAQ"), 0),
  qz_pct(wv("qes2012", "no", "PLQ"), 0), qz_pct(wv("qes2022", "no", "PLQ"), 0))
qz_figure(fig_where, "where",
  title = sprintf(tr("The PQ's share of the Yes vote fell from %s to %s as QS and the CAQ took it; the No vote fragmented",
                     "La part du PQ chez les électeurs du Oui est passée de %s à %s au profit de QS et de la CAQ ; le vote pour le Non s'est fragmenté"),
                  qz_pct(wv("qes2012", "yes", "PQ"), 0), qz_pct(wv("qes2022", "yes", "PQ"), 0)),
  subtitle = tr("How the respondents who would vote Yes and No in a referendum voted: each party's share of the side's reported vote, with 95% confidence intervals",
                "Comment ont voté les répondants qui voteraient Oui et Non à un référendum : la part de chaque parti dans le vote déclaré du camp, avec intervalles de confiance à 95 %"),
  alt = alt, table = tab_where, height = 4.2,
  note = tr("Source: qesR, pooled vote_choice (reported vote) and sov_support; QES 2007 to 2022, weighted with each study's post-election weight except 2008 (hollow: unweighted, weight under review). 2007 and 2008 asked the 1995 question, 2012 to 2022 about an independent country, so the lines break at the change. Lines: 95% confidence intervals (logit). The other parties (PV, ON and others) are in the table view.",
            "Source : qesR, variables regroupées vote_choice (vote déclaré) et sov_support ; EEQ 2007 à 2022, pondérées avec la pondération postélectorale de chaque étude sauf 2008 (creux : non pondérée, pondération en révision). 2007 et 2008 ont posé la question de 1995, 2012 à 2022 la question sur un pays indépendant : les lignes s'interrompent au changement. Traits : intervalles de confiance à 95 % (logit). Les autres partis (PV, ON et autres) sont dans la vue en tableau."))
```

What to notice:

- In 2007 and 2012 the PQ held most of the Yes vote; by 2018 it had less
  than half of it, as Québec solidaire and the CAQ took their shares.
- The No vote fragmented: the PLQ's share fell from
  `r qz_pct(wv("qes2012", "no", "PLQ"), 0)` in 2012 to
  `r qz_pct(wv("qes2022", "no", "PLQ"), 0)` in 2022; the CAQ took
  `r qz_pct(wv("qes2012", "no", "CAQ"), 0)` of it in 2012,
  `r qz_pct(wv("qes2018", "no", "CAQ"), 0)` in 2018 and
  `r qz_pct(wv("qes2022", "no", "CAQ"), 0)` in 2022, and the PCQ
  `r qz_pct(wv("qes2022", "no", "PCQ"), 0)` in 2022. In 2007 the ADQ already
  took `r qz_pct(wv("qes2007", "no", "ADQ"), 0)` of it. The QES 2022
  under-reports the CAQ by about
  `r qz_num(caq_gap, 0)`
  points (see [Survey vs official results](survey-vs-official.html)),
  which affects the 2022 shares.
- The CAQ is the one party with a large share of both sides.

## About the data

- **Studies.** The Quebec Election Studies of 2012 to 2022 for the
  left-right placement, and of 2007 to 2022 for the referendum vote; the
  Durand panels are left out, since they asked other referendum wordings.
- **Variables.** `vote_choice` restricted to the reported vote,
  `sov_support` as the share of Yes among Yes and No, and `lr_self`, the
  respondent's own placement from 0 (left) to 10 (right). The 2007 and 2008
  studies asked the 1995 question, the others the question on an
  independent country: the rows of the bar chart are not one series.
- **Timing.** In 2022 the referendum and left-right questions were asked
  during the campaign and the reported vote after the election; the
  estimates keep the respondents of both waves, weighted with the
  post-election weight.
- **Weights.** Each study's recommended post-election weight, through
  `qes_design()`; 2008 is unweighted (its weights are calibrated on the
  vote and are under review).
