---
title: "Two dimensions of competition: sovereignty and left-right"
description: >
  Has the left-right scale replaced the national question in Quebec
  elections? Where each party's voters stand on both axes, where the Yes
  and No camps vote, and how much each axis still structures the vote.
params:
  lang: "en"
---

*[Version française](fr-dimensions.html)*

```{r setup, include = FALSE}
source("_setup.R", encoding = "UTF-8")
dl <- qz_download(c("qes2007", "qes2008", "qes2012", "qes2014", "qes2018", "qes2022"))
```

::: {.qesr-download}
**This page downloads `r dl$n` studies** (about `r qz_num(dl$mb, 0)` MB) the
first time it runs. `options(qesR.cache = "disk")` keeps them on disk for
later sessions.
:::

```{r harmonize, echo = FALSE}
h <- qes_harmonize(
  studies = c("qes2007", "qes2008", "qes2012", "qes2014", "qes2018", "qes2022"),
  targets = c("vote_choice", "sov_support", "lr_self", "lang_mother"),
  types = list(vote_choice = "recall"),
  missing = "reasons", quiet = TRUE
)
```

```{r estimates, echo = FALSE}
h$party <- factor(qz_party_code(h$vote_choice), levels = qz_parties)
h$yes <- factor(ifelse(h$sov_support %in% "Yes", "yes",
                       ifelse(h$sov_support %in% "No", "no", NA)), levels = c("yes", "no"))
dim_studies <- c("qes2012", "qes2014", "qes2018", "qes2022")
bar_studies <- c("qes2007", "qes2008", dim_studies)

electorates <- do.call(rbind, lapply(dim_studies, function(s) {
  d <- qz_design(h, s, "weight_post")
  lr <- qz_mean(d, "lr_self", by = "party")
  yes <- qz_share(d, "yes", by = "party", levels = "yes")
  out <- merge(setNames(lr, c("party", "lr", "lr_lo", "lr_hi", "lr_n", "weighted")),
               setNames(yes[, c("group", "pct", "lo", "hi", "n")], c("party", "yes", "yes_lo", "yes_hi", "yes_n")))
  qz_tag(out, study = s)
}))
electorates$year <- qz_year(electorates$study)
electorates$status <- qz_party_status(electorates$study, electorates$party)
electorates <- electorates[electorates$status == "offered" & electorates$party != "other", ]

where <- do.call(rbind, lapply(bar_studies, function(s) {
  d <- qz_design(h, s, "weight_post")
  qz_tag(qz_share(d, "party", by = "yes"), study = s)
}))
where$year <- qz_year(where$study)
where$status <- qz_party_status(where$study, where$level)

# How much of the variation in each attitude lies between the parties'
# electorates: the weighted share of the variance accounted for by the
# vote (eta squared), for the referendum vote (Yes = 1, No = 0) and for
# the left-right placement, among the same respondents (those who answered
# both and voted for a listed party). The 95% intervals come from 500
# bootstrap replicates of each study's design.
eta2 <- function(y, g, w) {
  m <- sum(w * y) / sum(w)
  wg <- tapply(w, g, sum)
  mg <- tapply(w * y, g, sum) / wg
  sum(wg * (mg - m)^2, na.rm = TRUE) / sum(w * (y - m)^2)
}
sorting <- do.call(rbind, lapply(dim_studies, function(s) {
  x <- h[h$study == s & h$party %in% setdiff(qz_parties, "other") & !is.na(h$yes) & !is.na(h$lr_self), ]
  x <- x[qz_party_status(s, as.character(x$party)) == "offered", ]
  x$y_sov <- as.numeric(x$yes == "yes")
  groups <- list(all = rep(TRUE, nrow(x)), fr = x$lang_mother %in% "French")
  do.call(rbind, lapply(names(groups), function(g) {
    do.call(rbind, lapply(c(FALSE, TRUE), function(no_pcq) {
      xx <- x[groups[[g]] & !(no_pcq & x$party %in% "PCQ"), ]
      xx$party <- droplevels(xx$party)
      set.seed(2012)
      r <- survey::as.svrepdesign(qz_design(xx, s, "weight_post"), type = "subbootstrap", replicates = 500)
      est <- survey::withReplicates(r, function(w, data) {
        c(eta2(data$y_sov, data$party, w), eta2(data$lr_self, data$party, w))
      })
      se <- sqrt(diag(as.matrix(attr(est, "var"))))
      data.frame(study = s, group = g, no_pcq = no_pcq, measure = c("sov", "lr"),
                 est = 100 * as.numeric(est), lo = 100 * (as.numeric(est) - 1.96 * se),
                 hi = 100 * (as.numeric(est) + 1.96 * se), n = nrow(xx),
                 weighted = qz_weighted(xx, s, "weight_post"))
    }))
  }))
}))
sorting$year <- qz_year(sorting$study)

ev <- function(s, p, col = "yes") electorates[[col]][electorates$study == s & electorates$party == p]
wv <- function(s, g, p) where$pct[where$study == s & where$group == g & where$level == p]
sv <- function(s, m, g = "all", no_pcq = FALSE) {
  sorting$est[sorting$study == s & sorting$measure == m & sorting$group == g & sorting$no_pcq == no_pcq]
}
# for the prose: a gap between two values rounded as the text shows them, so
# that the gap agrees with the figures cited next to it
rgap <- function(a, b, digits = 0) round(a, digits) - round(b, digits)
# the share of each study's respondents who placed themselves on the
# left-right scale (the 2022 item offered no "don't know")
lr_answered <- vapply(dim_studies, function(s) 100 * mean(!is.na(h$lr_self[h$study == s])), numeric(1))
# official results, share of valid votes
off <- function(y, p) qz_official()$share_valid[qz_official()$year == y & qz_official()$party == p]
```

For four decades, the national question organized party competition in
Quebec. The Parti Québécois (PQ) stood for sovereignty, the Parti libéral
du Québec (PLQ) for federalism, and a voter's position on independence
went a long way toward predicting their vote (Bélanger et al. 2018). The
rise of the Coalition avenir Québec (CAQ) and of Québec solidaire (QS),
and the CAQ's victory in 2018, have changed that picture. The
conventional wisdom, reflected in recent work on the new party system
(Bélanger and Mahéo 2020; Bélanger et al. 2022), is now that the
sovereignty axis has faded and that a left-right axis has taken its
place, with the PQ squeezed between a nationalist party that set the
referendum aside and a sovereigntist party on the left. On this page, we
examine whether the Quebec Election Studies bear this out: has the
national question stopped structuring the Quebec vote, and has the
left-right scale taken over?

Overall, we find mixed support for the conventional wisdom. The first half
holds. In 2012, party choice accounted for
`r qz_pct(sv("qes2012", "sov"), 0)` of the variation in the referendum
vote (knowing which party a respondent voted for told us
`r qz_pct(sv("qes2012", "sov"), 0)` of what there was to know about their
referendum vote); in 2022, it accounted for
`r qz_pct(sv("qes2022", "sov"), 0)`. The second half does not hold. The
left-right scale did not fill the gap: party choice accounted for
`r qz_pct(sv("qes2012", "lr"), 0)` of the variation in left-right
placement in 2012 and `r qz_pct(sv("qes2022", "lr"), 0)` in 2022. Leaving
aside the voters of the Parti conservateur du Québec (PCQ), which went
from `r qz_pct(off(2018, "PCQ"), 1)` of the vote in 2018 to
`r qz_pct(off(2022, "PCQ"), 1)` in 2022, the figure falls to
`r qz_pct(sv("qes2022", "lr", no_pcq = TRUE), 0)`. In other words, the
national question lost ground without being replaced. It still separates
the two historic parties as sharply as ever: in 2022,
`r qz_pct(ev("qes2022", "PQ"), 0)` of PQ voters would have voted Yes,
against `r qz_pct(ev("qes2022", "PLQ"), 0)` of PLQ voters.

We use three questions asked, in similar form, in every Quebec Election
Study from 2012 to 2022: the vote reported after the election, the vote in
a referendum on independence, and the respondent's own placement on a
left-right scale from 0 (left) to 10 (right). Mother tongue lets us look at
francophones on their own. In the Quebec Election Studies used here, the
referendum question goes back to 2007, so the second figure starts then.
The end of the page shows how to reproduce the figures.

## The poles held; the newcomers converged

```{r fig-paths, echo = FALSE}
path_parties <- c("PLQ", "PQ", "CAQ", "QS", "PCQ")
e <- electorates[electorates$party %in% path_parties & electorates$lr_n >= qz_min_n & electorates$yes_n >= qz_min_n, ]
measure_lab <- c(yes = tr("Would vote Yes (among Yes and No)", "Voteraient Oui (parmi Oui et Non)"),
                 lr = tr("Left-right, mean (0 = left, 10 = right)", "Gauche-droite (0 = gauche, 10 = droite)"))
d <- rbind(data.frame(e[, c("party", "year", "weighted")], measure = "yes", value = e$yes, lo = e$yes_lo, hi = e$yes_hi),
           data.frame(e[, c("party", "year", "weighted")], measure = "lr", value = e$lr, lo = e$lr_lo, hi = e$lr_hi))
d$party <- factor(d$party, levels = path_parties)
d$panel <- factor(measure_lab[d$measure], levels = measure_lab)
# the parties side by side at each election, so their intervals do not overlap
d$x <- d$year + c(PLQ = -0.36, PQ = -0.18, CAQ = 0, QS = 0.18, PCQ = 0.36)[as.character(d$party)]
d$label <- qz_party_label(d$party)
# the two panels have their own scales: a share and a mean on 0-10
blank <- data.frame(panel = factor(measure_lab, levels = measure_lab), x = 2012, value = c(0, 2.5), top = c(100, 7.4))

fig_paths <- function(mode) {
  k <- qz_tokens[[mode]]
  ggplot(d, aes(x, value, colour = party, group = party)) +
    geom_blank(data = blank, aes(x, value), inherit.aes = FALSE) +
    geom_blank(data = blank, aes(x, top), inherit.aes = FALSE) +
    geom_linerange(aes(ymin = lo, ymax = hi), linewidth = 0.4, alpha = 0.45) +
    qz_line() +
    qz_dots(mode, d, aes(x, value), colour_aes = "party", size = 2.4) +
    qz_end_labels(d[d$measure == "yes", ], "x", "value", "label", "party", mode, min_gap = 5.5, nudge_x = 1.4, size = 3) +
    qz_end_labels(d[d$measure == "lr", ], "x", "value", "label", "party", mode, min_gap = 0.24, nudge_x = 1.4, size = 3) +
    facet_wrap(~panel, nrow = 1, scales = "free_y") +
    scale_party(mode) +
    scale_x_continuous(breaks = c(2012, 2014, 2018, 2022), limits = c(2011.3, 2026.2), expand = expansion(0)) +
    # the share panel starts at 0, the left-right panel at 2.5: 0 and above 10 are percentages
    scale_y_continuous(labels = function(x) ifelse(x > 10 | x == 0, qz_pct_axis(x), qz_num(x, 0)), expand = expansion(0)) +
    labs(x = NULL, y = NULL) +
    theme_qes(mode) +
    theme(panel.spacing.x = unit(30, "pt"), strip.text = element_text(size = 9.5))
}

tab <- electorates[order(match(electorates$party, qz_parties), electorates$year), ]
tab_paths <- setNames(data.frame(
  qz_party_label(tab$party), tab$year, qz_study_label(tab$study),
  qz_cell(tab$lr, tab$lr_lo, tab$lr_hi, tab$lr_n, 2), tab$lr_n,
  qz_cell(tab$yes, tab$yes_lo, tab$yes_hi, tab$yes_n), tab$yes_n, qz_weight_text(tab$weighted),
  row.names = NULL
), c(tr("Party", "Parti"), tr("Election", "Élection"), tr("Study", "Étude"),
     tr("Left-right, mean [95% CI]", "Gauche-droite, moyenne [IC à 95 %]"), tr("n (left-right)", "n (gauche-droite)"),
     tr("Yes, % [95% CI]", "Oui, % [IC à 95 %]"), tr("n (Yes/No)", "n (Oui/Non)"), tr("Weighting", "Pondération")))
alt <- sprintf(tr(
  "Two line charts, 2012 to 2022, one line per party (PLQ, PQ, CAQ, QS; the PCQ in 2022 only): on the left the share of the party's voters who would vote Yes to independence, on the right their mean left-right self-placement, with confidence intervals. PQ voters go from %s to %s Yes; CAQ voters from %s to %s; QS voters from %s to %s. PLQ voters move from %s to %s on the left-right scale; QS voters stay the furthest left, at %s in 2022. Values in the table view.",
  "Deux graphiques linéaires, de 2012 à 2022, une ligne par parti (PLQ, PQ, CAQ, QS ; le PCQ en 2022 seulement) : à gauche la part des électeurs du parti qui voteraient Oui à l'indépendance, à droite leur autopositionnement gauche-droite moyen, avec intervalles de confiance. Les électeurs du PQ passent de %s à %s de Oui ; ceux de la CAQ, de %s à %s ; ceux de QS, de %s à %s. Les électeurs du PLQ passent de %s à %s sur l'axe gauche-droite ; ceux de QS restent les plus à gauche, à %s en 2022. Valeurs dans la vue en tableau."),
  qz_pct(ev("qes2012", "PQ"), 0), qz_pct(ev("qes2022", "PQ"), 0),
  qz_pct(ev("qes2012", "CAQ"), 0), qz_pct(ev("qes2022", "CAQ"), 0),
  qz_pct(ev("qes2012", "QS"), 0), qz_pct(ev("qes2022", "QS"), 0),
  qz_num(ev("qes2012", "PLQ", "lr")), qz_num(ev("qes2022", "PLQ", "lr")), qz_num(ev("qes2022", "QS", "lr")))
qz_figure(fig_paths, "paths",
  title = sprintf(tr("CAQ and QS voters converged on sovereignty (%s and %s Yes in 2022), while PLQ voters moved left",
                     "Les électeurs de la CAQ et de QS se sont rapprochés sur la souveraineté (%s et %s de Oui en 2022), et ceux du PLQ se sont déplacés vers la gauche"),
                  qz_pct(ev("qes2022", "CAQ"), 0), qz_pct(ev("qes2022", "QS"), 0)),
  subtitle = tr("Where each party's voters stand on sovereignty and on the left-right scale, 2012 to 2022, with 95% confidence intervals",
                "Où se situent les électeurs de chaque parti sur la souveraineté et sur l'axe gauche-droite, de 2012 à 2022, avec intervalles de confiance à 95 %"),
  alt = alt, table = tab_paths, height = 4.2,
  note = tr("Source: Quebec Election Studies 2012, 2014, 2018 and 2022, weighted. Reported vote after the election; referendum vote among respondents who chose Yes or No. The parties are set side by side at each election, and the two panels have their own scales. The PCQ was offered as an answer in 2022 only.",
            "Source : Études électorales québécoises de 2012, 2014, 2018 et 2022, pondérées. Vote déclaré après l'élection ; vote référendaire parmi les répondants qui ont choisi Oui ou Non. Les partis sont côte à côte à chaque élection, et les deux panneaux ont leur propre échelle. Le PCQ n'était proposé comme réponse qu'en 2022."))
```

Two results stand out. First, the two historic poles did not move on the
national question. PQ voters remain overwhelmingly sovereigntist
(`r qz_pct(ev("qes2012", "PQ"), 0)` would vote Yes in 2012,
`r qz_pct(ev("qes2022", "PQ"), 0)` in 2022) and PLQ voters almost
unanimously federalist (`r qz_pct(ev("qes2012", "PLQ"), 0)` and
`r qz_pct(ev("qes2022", "PLQ"), 0)`). Second, the two newer parties moved
toward each other. CAQ voters went from `r qz_pct(ev("qes2012", "CAQ"), 0)`
Yes in 2012 to `r qz_pct(ev("qes2022", "CAQ"), 0)` in 2022, and QS voters
from `r qz_pct(ev("qes2012", "QS"), 0)` to `r qz_pct(ev("qes2022", "QS"), 0)`.
The CAQ now draws a good share of its vote from sovereigntists. Contrary to
what one might expect of a sovereigntist party, a majority of QS voters
(`r qz_pct(rgap(100, ev("qes2022", "QS")), 0)`) would now vote No. The 2012
estimate rests on only `r ev("qes2012", "QS", "yes_n")` respondents,
however, and its interval is wide
(`r qz_pct(ev("qes2012", "QS", "yes_lo"), 0)` to
`r qz_pct(ev("qes2012", "QS", "yes_hi"), 0)`).

Movements on the left-right scale are smaller, with one exception. QS
voters are the furthest left at every election
(`r qz_num(ev("qes2022", "QS", "lr"))` out of 10 in 2022) and PCQ voters
the furthest right (`r qz_num(ev("qes2022", "PCQ", "lr"))`). The exception
is the PLQ: its voters went from `r qz_num(ev("qes2012", "PLQ", "lr"))` in
2012 to `r qz_num(ev("qes2022", "PLQ", "lr"))` in 2022. In 2022, they stood
to the left of CAQ voters (`r qz_num(ev("qes2022", "CAQ", "lr"))`), close
to PQ voters (`r qz_num(ev("qes2022", "PQ", "lr"))`). This estimate rests
on `r ev("qes2022", "PLQ", "lr_n")` respondents, and the 2022 question was
worded differently (see [About the data](#about-the-data)), so the size of
the shift should be read with caution. The right of the scale now belongs
to the CAQ and the PCQ. Note, however, what happened between the two
sovereigntist parties. PQ and QS voters are close on left-right
(`r qz_num(rgap(ev("qes2012", "PQ", "lr"), ev("qes2012", "QS", "lr"), 1))`
points apart on the 0-10 scale in 2012,
`r qz_num(rgap(ev("qes2022", "PQ", "lr"), ev("qes2022", "QS", "lr"), 1))`
in 2022), while their gap on sovereignty grew from
`r qz_num(rgap(ev("qes2012", "PQ"), ev("qes2012", "QS")), 0)` to
`r qz_num(rgap(ev("qes2022", "PQ"), ev("qes2022", "QS")), 0)` percentage
points. What separates their electorates today is the national question
more than the left-right scale.

## Yes voters now split among three parties

```{r fig-where, echo = FALSE}
line_parties <- c("PLQ", "PQ", "ADQ", "QS", "CAQ", "PCQ")
d <- where[where$group %in% c("yes", "no") & where$level %in% line_parties & where$status == "offered" &
             where$n >= qz_min_n, ]
d$party <- factor(d$level, levels = line_parties)
d$x <- d$year
# the question changes between 2008 (the 1995 question) and 2012 (independence): no line across
d$era <- ifelse(d$year < 2010, "q1995", "indep")
side_lab <- c(yes = tr("Would vote Yes", "Voteraient Oui"), no = tr("Would vote No", "Voteraient Non"))
d$facet <- factor(side_lab[d$group], levels = side_lab)
lab <- d[d$year == 2022, ]
# the QES 2022 under-reports the CAQ (see the survey-vs-official page)
caq_gap <- qz_official()$share_valid[qz_official()$year == 2022 & qz_official()$party == "CAQ"] -
  qz_share(qz_design(h, "qes2022", "weight_post"), "party", levels = "CAQ")$pct
lab$label <- qz_party_label(lab$party)

fig_where <- function(mode) {
  k <- qz_tokens[[mode]]
  ggplot(d, aes(x, pct, colour = party, group = interaction(party, era))) +
    annotate("rect", xmin = 2008.6, xmax = 2011.4, ymin = -Inf, ymax = Inf, fill = k$wash) +
    annotate("text", x = 2010, y = 78, label = tr("the question\nchanges", "la question\nchange"),
             size = 2.7, colour = k$ink2, family = qz_font(), lineheight = 0.9) +
    geom_linerange(aes(ymin = lo, ymax = hi), linewidth = 0.4, alpha = 0.45, show.legend = FALSE) +
    qz_line() +
    geom_point(aes(fill = party), shape = 21, size = 2.4, stroke = 0.8,
               colour = k$surface, data = d[d$year != 2008, ], show.legend = FALSE) +
    geom_point(aes(colour = party), shape = 21, size = 2.2, stroke = 1.1, fill = k$surface, data = d[d$year == 2008, ],
               show.legend = FALSE) +
    qz_end_labels(lab[lab$group == "yes", ], "x", "pct", "label", "party", mode, min_gap = 4.2, nudge_x = 1.6, size = 3) +
    qz_end_labels(lab[lab$group == "no", ], "x", "pct", "label", "party", mode, min_gap = 4.2, nudge_x = 1.6, size = 3) +
    facet_wrap(~facet, nrow = 1, axes = "all_x") +
    scale_party(mode, aesthetics = "colour", guide = guide_legend(order = 1, nrow = 1)) +
    scale_party(mode, aesthetics = "fill") +
    scale_shape_manual(values = c(weighted = 21, unweighted = 21), limits = c("weighted", "unweighted"),
                       breaks = "unweighted", labels = tr("hollow: 2008, unweighted, no validated weight", "creux : 2008, non pondéré, aucune pondération validée"),
                       guide = guide_legend(order = 2, override.aes = list(fill = k$surface, colour = k$ink2, stroke = 1.1, size = 2.6,
                                                                          alpha = 1, linetype = 0))) +
    geom_point(data = d[1, ], aes(shape = "unweighted"), alpha = 0, colour = NA, show.legend = c(colour = FALSE, shape = TRUE)) +
    scale_x_continuous(breaks = c(2007, 2012, 2014, 2018, 2022), labels = c("2007", "2012", "’14", "2018", "2022"),
                       limits = c(2006.3, 2026.6), expand = expansion(0)) +
    scale_y_continuous(breaks = seq(0, 80, 20), labels = qz_pct_axis) +
    coord_cartesian(ylim = c(0, 84), expand = FALSE, clip = "off") +
    labs(x = NULL, y = tr("Share of the side's reported vote", "Part du vote déclaré du camp")) +
    theme_qes(mode) +
    theme(panel.spacing.x = unit(26, "pt"))
}

tab <- where[where$group %in% c("yes", "no") & where$level %in% c(line_parties, "other"), ]
tab <- tab[order(tab$year, tab$group == "no", match(tab$level, qz_parties)), ]
tab_where <- setNames(data.frame(
  tab$year, qz_study_label(tab$study),
  ifelse(tab$group == "yes", tr("Yes", "Oui"), tr("No", "Non")),
  qz_party_label(tab$level),
  ifelse(tab$status == "not_run", "—", ifelse(tab$status == "not_listed", tr("n.l.", "n.p."),
                                                  qz_cell(tab$pct, tab$lo, tab$hi, tab$n))),
  tab$n, qz_weight_text(tab$weighted), row.names = NULL
), c(tr("Election", "Élection"), tr("Study", "Étude"), tr("Referendum vote", "Vote référendaire"),
     tr("Party", "Parti"), tr("Share of reported vote [95% CI]", "Part du vote déclaré [IC à 95 %]"),
     "n", tr("Weighting", "Pondération")))
alt <- sprintf(tr(
  "Line charts in two panels, the respondents who would vote Yes (left) and No (right) in a referendum, from 2007 to 2022: the share of each side's reported vote that went to each party. The PQ's share of the Yes vote goes from %s in 2012 to %s in 2022, as QS rises to %s and the CAQ to %s; the PLQ's share of the No vote goes from %s in 2012 to %s in 2022. Values in the table view.",
  "Graphiques linéaires en deux panneaux, les répondants qui voteraient Oui (à gauche) et Non (à droite) à un référendum, de 2007 à 2022 : la part du vote déclaré de chaque camp allée à chaque parti. La part du PQ chez les électeurs du Oui passe de %s en 2012 à %s en 2022, tandis que QS monte à %s et la CAQ à %s ; celle du PLQ chez les électeurs du Non passe de %s en 2012 à %s en 2022. Valeurs dans la vue en tableau."),
  qz_pct(wv("qes2012", "yes", "PQ"), 0), qz_pct(wv("qes2022", "yes", "PQ"), 0),
  qz_pct(wv("qes2022", "yes", "QS"), 0), qz_pct(wv("qes2022", "yes", "CAQ"), 0),
  qz_pct(wv("qes2012", "no", "PLQ"), 0), qz_pct(wv("qes2022", "no", "PLQ"), 0))
qz_figure(fig_where, "where",
  title = sprintf(tr("The PQ's share of the Yes vote fell from %s to %s as QS and the CAQ took it; No voters spread across more parties",
                     "La part du PQ chez les électeurs du Oui est passée de %s à %s au profit de QS et de la CAQ ; les électeurs du Non se sont dispersés entre davantage de partis"),
                  qz_pct(wv("qes2012", "yes", "PQ"), 0), qz_pct(wv("qes2022", "yes", "PQ"), 0)),
  subtitle = tr("How the respondents who would vote Yes and No in a referendum voted: each party's share of the side's reported vote, with 95% confidence intervals",
                "Comment ont voté les répondants qui voteraient Oui et Non à un référendum : la part de chaque parti dans le vote déclaré du camp, avec intervalles de confiance à 95 %"),
  alt = alt, table = tab_where, height = 4.2,
  note = tr("Source: Quebec Election Studies 2007 to 2022, weighted, except 2008 (hollow: unweighted, no validated weight). 2007 and 2008 asked the 1995 referendum question, 2012 to 2022 a question on an independent country, so the lines break at the change. The other parties (Green Party, Option nationale and others) are in the table view.",
            "Source : Études électorales québécoises de 2007 à 2022, pondérées, sauf 2008 (creux : non pondérée, aucune pondération validée). 2007 et 2008 ont posé la question référendaire de 1995, 2012 à 2022 une question sur un pays indépendant : les lignes s'interrompent au changement. Les autres partis (Parti vert, Option nationale et autres) sont dans la vue en tableau."))
```

We now turn from where each party's voters stand to where each camp's
voters go. In 2012, the PQ took `r qz_pct(wv("qes2012", "yes", "PQ"), 0)`
of the reported vote of respondents who would vote Yes; in 2022, it took
`r qz_pct(wv("qes2022", "yes", "PQ"), 0)`, about as much as the CAQ
(`r qz_pct(wv("qes2022", "yes", "CAQ"), 0)`), with QS at
`r qz_pct(wv("qes2022", "yes", "QS"), 0)`. In other words, the PQ now
holds about a third of the Yes vote. The party vote of No voters spread
out as well. The PLQ's share fell from
`r qz_pct(wv("qes2012", "no", "PLQ"), 0)` in 2012 to
`r qz_pct(wv("qes2022", "no", "PLQ"), 0)` in 2022, while the CAQ held a
large share throughout (`r qz_pct(wv("qes2012", "no", "CAQ"), 0)` in 2012,
`r qz_pct(wv("qes2022", "no", "CAQ"), 0)` in 2022) and the PCQ took
`r qz_pct(wv("qes2022", "no", "PCQ"), 0)` in 2022. Unsurprisingly for a
party that promised not to hold a referendum, the CAQ is the only party
that took about a third of each camp in 2022
(`r qz_pct(wv("qes2022", "yes", "CAQ"), 0)` of Yes voters and
`r qz_pct(wv("qes2022", "no", "CAQ"), 0)` of No voters).

Two caveats apply. First, the 2007 and 2008 studies asked the 1995
referendum question and the later ones a question on an independent
country, and the wording of the question is known to move support for
sovereignty (Yale and Durand 2011); the lines break at the change for that
reason. Second, the 2022 study under-reports the CAQ vote by about
`r qz_num(caq_gap, 0)` points compared to the official result (see
[Survey vs official results](survey-vs-official.html)), which lowers the
CAQ's share of both camps that year. The near-tie between the PQ and the
CAQ among Yes voters is therefore, if anything, favourable to the PQ.

## Sovereignty divides the vote much less; left-right did not take over

The first two figures show positions. They do not tell us how much each
axis separates the electorates, which is the heart of the conventional
wisdom. To answer that, we apply the same measure to each study: of all
the variation in respondents' referendum vote, or in their left-right
placement, how much lies between the parties' electorates rather than
within them? This is the share of the variance accounted for by party
choice (an eta squared). At 100%, knowing someone's vote would tell us
their answer exactly; at 0%, it would tell us nothing. Both shares are
computed on the same respondents, those who answered both questions and
voted for one of the parties shown above.

```{r fig-sorting, echo = FALSE}
d <- sorting[!sorting$no_pcq, ]
group_lab <- c(all = tr("All voters", "Tous les électeurs"),
               fr = tr("Francophone voters", "Électeurs francophones"))
axis_lab <- c(sov = tr("Referendum vote (Yes or No)", "Vote référendaire (Oui ou Non)"),
              lr = tr("Left-right placement", "Positionnement gauche-droite"))
d$panel <- factor(group_lab[d$group], levels = group_lab)
d$measure <- factor(d$measure, levels = names(axis_lab))
# the two measures side by side at each election
d$x <- d$year + c(sov = -0.18, lr = 0.18)[as.character(d$measure)]
p <- sorting[sorting$no_pcq & sorting$measure == "lr" & sorting$year == 2022, ]
p$panel <- factor(group_lab[p$group], levels = group_lab)
p$x <- p$year + 0.9
p$from <- vapply(p$group, function(g) sv("qes2022", "lr", g), numeric(1))

fig_sorting <- function(mode) {
  k <- qz_tokens[[mode]]
  cols <- c(sov = k$grp[["fr"]], lr = k$grp[["other"]])
  ggplot(d, aes(x, est, colour = measure, group = measure)) +
    qz_ci(aes(ymin = lo, ymax = hi), show.legend = FALSE) +
    qz_line(show.legend = FALSE) +
    qz_dots(mode, d, aes(x, est), colour_aes = "measure") +
    geom_segment(data = p, aes(x = 2022.18, xend = x - 0.12, y = from, yend = est),
                 colour = k$muted, linewidth = 0.3, inherit.aes = FALSE) +
    geom_point(data = p, aes(x, est), shape = 23, size = 2.3, stroke = 0.8, colour = k$surface,
               fill = cols[["lr"]], inherit.aes = FALSE) +
    geom_text(data = p, aes(x + 0.35, est, label = tr("without\nthe PCQ", "sans\nle PCQ")), hjust = 0,
              size = 2.7, colour = k$ink2, family = qz_font(), lineheight = 0.9, inherit.aes = FALSE) +
    facet_wrap(~panel, nrow = 1) +
    scale_colour_manual(values = cols, guide = "none") +
    scale_fill_manual(values = cols, breaks = names(axis_lab), labels = axis_lab, guide = guide_legend(order = 1)) +
    scale_x_continuous(breaks = c(2012, 2014, 2018, 2022), limits = c(2011.3, 2026.4), expand = expansion(0)) +
    scale_y_continuous(breaks = seq(0, 60, 20), labels = qz_pct_axis) +
    coord_cartesian(ylim = c(0, 66), expand = FALSE, clip = "off") +
    labs(x = NULL, y = tr("Share of the variance accounted for by party choice", "Part de la variance liée au choix partisan")) +
    theme_qes(mode) +
    theme(panel.spacing.x = unit(26, "pt"))
}

tab <- sorting[order(sorting$group, sorting$no_pcq, sorting$measure != "sov", sorting$year), ]
tab_sorting <- setNames(data.frame(
  ifelse(tab$group == "all", tr("All voters", "Tous les électeurs"), tr("Francophones", "Francophones")),
  ifelse(tab$no_pcq, tr("without the PCQ", "sans le PCQ"), tr("all parties shown", "tous les partis montrés")),
  unname(axis_lab[tab$measure]), tab$year, qz_study_label(tab$study),
  qz_cell(tab$est, tab$lo, tab$hi, tab$n), tab$n, qz_weight_text(tab$weighted), row.names = NULL
), c(tr("Voters", "Électeurs"), tr("Parties", "Partis"), tr("Question", "Question"), tr("Election", "Élection"),
     tr("Study", "Étude"), tr("Share of variance, % [95% CI]", "Part de la variance, % [IC à 95 %]"), "n",
     tr("Weighting", "Pondération")))
alt <- sprintf(tr(
  "Line charts in two panels, all voters (left) and francophone voters (right), 2012 to 2022: the share of the variance in the referendum vote and in left-right placement accounted for by party choice, with confidence intervals. For the referendum vote it goes from %s in 2012 to %s in 2022 among all voters, and from %s to %s among francophones; for left-right placement from %s to %s among all voters, and %s in 2022 without the PCQ. Values in the table view.",
  "Graphiques linéaires en deux panneaux, tous les électeurs (à gauche) et les électeurs francophones (à droite), de 2012 à 2022 : la part de la variance du vote référendaire et du positionnement gauche-droite liée au choix partisan, avec intervalles de confiance. Pour le vote référendaire, elle passe de %s en 2012 à %s en 2022 chez l'ensemble des électeurs, et de %s à %s chez les francophones ; pour le positionnement gauche-droite, de %s à %s chez l'ensemble des électeurs, et %s en 2022 sans le PCQ. Valeurs dans la vue en tableau."),
  qz_pct(sv("qes2012", "sov"), 0), qz_pct(sv("qes2022", "sov"), 0),
  qz_pct(sv("qes2012", "sov", "fr"), 0), qz_pct(sv("qes2022", "sov", "fr"), 0),
  qz_pct(sv("qes2012", "lr"), 0), qz_pct(sv("qes2022", "lr"), 0), qz_pct(sv("qes2022", "lr", no_pcq = TRUE), 0))
qz_figure(fig_sorting, "sorting",
  title = sprintf(tr("Party choice accounts for %s of the variance in the referendum vote in 2022, down from %s in 2012; left-right did not gain",
                     "Le choix partisan est lié à %s de la variance du vote référendaire en 2022, contre %s en 2012 ; l'axe gauche-droite n'a pas progressé"),
                  qz_pct(sv("qes2022", "sov"), 0), qz_pct(sv("qes2012", "sov"), 0)),
  subtitle = tr("How much of the variation in each question lies between the parties' electorates, 2012 to 2022, with 95% confidence intervals",
                "Quelle part de la variation de chaque question se situe entre les électorats des partis, de 2012 à 2022, avec intervalles de confiance à 95 %"),
  alt = alt, table = tab_sorting, height = 4.2,
  note = tr("Source: Quebec Election Studies 2012, 2014, 2018 and 2022, weighted. Respondents who answered both questions and reported a vote for the PLQ, the PQ, the CAQ, QS or (2022) the PCQ; francophones are respondents whose mother tongue is French. Diamond: the 2022 left-right share without the PCQ's voters. Intervals from 500 bootstrap replicates of each study.",
            "Source : Études électorales québécoises de 2012, 2014, 2018 et 2022, pondérées. Répondants qui ont répondu aux deux questions et déclaré un vote pour le PLQ, le PQ, la CAQ, QS ou (2022) le PCQ ; les francophones sont les répondants de langue maternelle française. Losange : la part gauche-droite de 2022 sans les électeurs du PCQ. Intervalles tirés de 500 répliques bootstrap de chaque étude."))
```

The result is clear on sovereignty. In 2012 and 2014, party choice
accounted for more than half of the variation in the referendum vote
(`r qz_pct(sv("qes2012", "sov"), 0)` and `r qz_pct(sv("qes2014", "sov"), 0)`),
against `r qz_pct(sv("qes2012", "lr"), 0)` and
`r qz_pct(sv("qes2014", "lr"), 0)` for left-right placement. The share for
sovereignty then fell to `r qz_pct(sv("qes2018", "sov"), 0)` in 2018 and
to `r qz_pct(sv("qes2022", "sov"), 0)` in 2022. The decline is steeper
among francophones: from `r qz_pct(sv("qes2012", "sov", "fr"), 0)` in 2012
to `r qz_pct(sv("qes2022", "sov", "fr"), 0)` in 2022. Among them, the two
axes are now about equally tied to party choice
(`r qz_pct(sv("qes2022", "lr", "fr"), 0)` for left-right,
`r qz_pct(sv("qes2022", "sov", "fr"), 0)` for sovereignty; the intervals
overlap), and the left-right figure owes much to the PCQ's voters: without
them, it is `r qz_pct(sv("qes2022", "lr", "fr", no_pcq = TRUE), 0)`.

Contrary to the conventional wisdom, the left-right axis did not rise to
take the place of sovereignty. Its share went from
`r qz_pct(sv("qes2012", "lr"), 0)` in 2012 to
`r qz_pct(sv("qes2018", "lr"), 0)` in 2018, and came back to
`r qz_pct(sv("qes2022", "lr"), 0)` in 2022 only with the arrival of the
PCQ, whose voters stand furthest to the right. Without them, the 2022
share is `r qz_pct(sv("qes2022", "lr", no_pcq = TRUE), 0)`. It is the
PCQ's base, which crossed the language divide and owed little to
sovereignty (Bélanger et al. 2025), that accounts for the rise in 2022. In
short, the 2022 vote is much less structured by the national question,
without being more structured by the left-right scale.

These shares have limits. First, a Yes or No answer and a placement on a
scale from 0 to 10 are not measured the same way, so the comparison is
about trends and orders of magnitude, not about the exact gap between the
two lines in a given year. Second, a share of variance also depends on the
number of parties and on their size: the arrival of the PCQ in 2022 shows
how much one party at the edge of the scale can move it. Third, the 2022
left-right question differs from the earlier ones (see
[About the data](#about-the-data)), so the 2022 left-right figures, on
which the main finding rests, compare only approximately with earlier
years. Finally, these are cross-sections. They show how the electorates
divide, not why voters chose their party.

## What this means

All in all, the national question no longer organizes Quebec's party
system on its own, but no other single axis has replaced it. It still
divides the PQ and the PLQ as sharply as it did in 2012. What changed is
that the sovereigntist vote is now shared by three parties and that the
CAQ draws from both camps, so that knowing a voter's position on
independence tells us much less about their vote than it did a decade
ago. This differs from what Daoust and Gareau-Paquette (2024) found, with a
different measure, for federal elections, where the weight of sovereignty
in the Bloc québécois vote declined between 2004 and 2006 and then
remained remarkably stable from 2006 to 2021. At the provincial level, the
decline began between 2014 and 2018, during the CAQ's rise, and had not
levelled off by 2022. Of course, things might evolve differently as
younger cohorts, less inclined to vote PQ (Mahéo and Bélanger 2018), make
up a larger share of the electorate. The page on
[support for sovereignty across generations](sovereignty-generations.html)
takes up that question.

## About the data

- **Studies.** The Quebec Election Studies of 2012 to 2022 for the
  left-right placement, and of 2007 to 2022 for the referendum vote. The
  panels run by Claire Durand and colleagues at the same elections are
  left out, so that the series has one study per election; in 2012 and
  2018 they also asked other wordings of the referendum question. The 1998
  polls are left out as well: they have no validated weight, and only one
  of the two asked the referendum question.
- **Questions.** The vote reported after the election; the referendum
  vote as the share of Yes among respondents who chose Yes or No (those
  who remain undecided are set aside); the respondent's own placement
  from 0 (left) to 10 (right). Francophones are respondents whose mother
  tongue is French. The 2007 and 2008 studies asked the 1995 referendum
  question, with a follow-up that asked the undecided which way they
  leaned; the others asked a question on an independent country. In 2022
  the left-right question was worded differently, was asked on its own
  rather than after the respondent placed the parties, and offered no
  "don't know"; almost every respondent answered it
  (`r qz_pct(lr_answered[["qes2022"]], 0)`, against about
  `r qz_pct(mean(lr_answered[c("qes2012", "qes2014", "qes2018")]), 0)`
  before), so 2022 left-right figures compare only approximately with
  earlier years.
- **Timing.** In 2022 the referendum and left-right questions were asked
  during the campaign and the vote after the election; the estimates use
  the respondents who took part in both waves, with the post-election
  weight.
- **Weights.** Each study's post-election weight. The 2008 study is shown
  unweighted: its only weights were calibrated on the reported vote, which
  makes them unsuited to a study of the vote (see the
  [weights table](studies.html#weights)).
- **Share of variance.** Computed within each study on the respondents
  who answered both questions and voted for one of the parties shown, with
  95% intervals from 500 bootstrap replicates of the study's design.

## Reproduce this page

The data come from one call to `qes_harmonize()`, which pools the reported
vote (`vote_choice`), the referendum vote (`sov_support`), left-right
placement (`lr_self`) and mother tongue (`lang_mother`) across the
studies:

```{r show-harmonize, ref.label = "harmonize", eval = FALSE}
```

A weighted mean by party, within one study:

```{r example}
library(survey)
d22 <- qes_design(h[h$study == "qes2022", ], weight = "weight_post")
svyby(~lr_self, ~vote_choice, subset(d22, vote_choice %in% c("CAQ", "QS")),
      svymean, na.rm = TRUE)
```

## Going further

- Bélanger, Éric, Jean-François Daoust, Valérie-Anne Mahéo and Richard
  Nadeau. 2022. *Le nouvel électeur québécois*. Montréal: Presses de
  l'Université de Montréal. <https://doi.org/10.1515/9782760645929>
- Bélanger, Éric, and Valérie-Anne Mahéo. 2020. “Le Parti québécois dans
  le nouveau système multipartite.” *Politique et Sociétés* 39 (3):
  19–55. <https://doi.org/10.7202/1072084ar>
- Bélanger, Éric, Philippe Mongrain, Thomas Gareau-Paquette and
  Valérie-Anne Mahéo. 2025. "A Party that Went Viral? The Drivers of
  Support for the Parti Conservateur du Québec in the 2022 Election."
  *Canadian Journal of Political Science* 58 (2): 277–296.
  <https://doi.org/10.1017/S0008423924000829>
- Bélanger, Éric, Richard Nadeau, Ailsa Henderson and Eve Hepburn. 2018.
  *The National Question and Electoral Politics in Quebec and Scotland*.
  Montreal and Kingston: McGill-Queen's University Press.
  <https://doi.org/10.1515/9780773554139>
- Daoust, Jean-François, and Thomas Gareau-Paquette. 2024. "Is Quebec
  independence still key in making sense of Canadian elections? A
  longitudinal analysis (2000–2021)." *Regional & Federal Studies* 34 (5):
  781–806. <https://doi.org/10.1080/13597566.2023.2233422>
- Mahéo, Valérie-Anne, and Éric Bélanger. 2018. "Is the Parti Québécois
  Bound to Disappear? A Study of the Current Generational Dynamics of
  Electoral Behaviour in Quebec." *Canadian Journal of Political Science*
  51 (2): 335–356. <https://doi.org/10.1017/S0008423917001147>
- Yale, François, and Claire Durand. 2011. "What did Quebeckers Want?
  Impact of Question Wording, Constitutional Proposal and Context on
  Support for Sovereignty, 1976–2008." *American Review of Canadian
  Studies* 41 (3): 242–258. <https://doi.org/10.1080/02722011.2011.594517>
