---
title: "Harmonizing across studies"
params:
  lang: "en"
---

*[Version française](fr-harmonisation.html)*

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# messages in the language of the page, whatever the locale of the build
options(qesR.lang = params$lang, width = 90)
```

This page goes from `qes_harmonize()` to a weighted estimate, study by
study, and shows what qesR records on the way: the comparability grade of
each study's question, the reason for every missing value, the weight that
fits each question, and the provenance of each cell. It is built when the
website is built, from the full data files, which qesR downloads from their
Dataverse deposits through its cache.

The harmonization engine is **experimental**. Its specification says, for
each study and harmonized variable ("target"), which question feeds the
target and how each of its codes maps to the target's levels; nothing is
matched by name. The rows of this first specification were checked against
the original files and documents but are not yet signed off by a reviewer,
so this page applies them with `include_draft = TRUE`.

```{r packages, message = FALSE}
library(qesR)
tr <- function(en, fr) if (identical(params$lang, "fr")) fr else en
# percentages with one decimal, in the page's style
pct <- function(x) formatC(100 * x, format = "f", digits = 1, decimal.mark = tr(".", ","))
```

## Which studies have the question

`qes_spec()` gives one row per target and one column per study, each cell
the grade of that study's question against the target's anchor question
(the full grid is on the [coverage page](coverage.html)). Two targets here:
the referendum question on Quebec becoming an independent country, and the
vote reported after the election.

```{r spec}
grid <- qes_spec(targets = c("sov_indep", "vote_prov_recall"), lang = params$lang)
knitr::kable(grid[, c("target", "label", "qes2022", "qes2018", "qes2014", "qes2012")])
```

The crosswalk view says which question each grade is about, and why:

```{r crosswalk}
xw <- qes_spec("crosswalk", targets = "sov_indep", lang = params$lang)
knitr::kable(xw[, c("study", "wave", "source_var", "grade", "grade_reason")])
```

`identical` means the same question, options and universe as the anchor;
`comparable` the same stimulus with differences that should not move the
shares; `approximate` a difference that can. The
[harmonization reference](harmonization-reference.html) gives the wording
and levels of every study's question.

## Harmonize

`studies = NULL`, the default, takes the Quebec Election Studies that have
a question for at least one of the targets; the Durand panels, the CROP
polls and the 1998 panel are left out unless you name them. The weights of
`qes2007` and `qes2008` are not documented yet, so their rows have no
weight and `qes_design()` leaves them out below, with a message.
`layout = "long"` gives one row
per respondent and wave, so each answer sits on the row of the wave that
asked it, next to that wave's weight. `missing = "reasons"` adds a column
`<target>__na` with the reason of every missing value.

```{r harmonize}
h <- qes_harmonize(targets = c("sov_indep", "vote_prov_recall"), layout = "long",
                   missing = "reasons", include_draft = TRUE, lang = params$lang)
table(h$study, h$wave)
```

The messages are part of the result. The first lists the *structural
zeros*: levels a study's question did not offer, such as a party missing
from its list. Their share in that study is 0 because nobody could choose
them, not because nobody supported them. `qes2022` asked the referendum
question in its campaign-period wave (`cps`) and the reported vote in its
post-election wave (`pes`), which have different weights.

## Why values are missing

Every missing value has a reason. Counts of the valid answers and of each
reason, for the referendum question:

```{r reasons}
reasons <- table(h$study, h$sov_indep__na)
reasons <- reasons[, colSums(reasons) > 0, drop = FALSE]
knitr::kable(cbind(valid = tapply(!is.na(h$sov_indep), h$study, sum), reasons[, , drop = FALSE]))
```

`dk` is "don't know" and `refused` a refusal; `not_in_wave` marks the
`qes2022` post-election rows, since the question was asked in the other
wave. `?qes_harmonize` lists every reason.

## Grades

`min_grade` keeps only the cells at or above a grade and sets the others to
`NA`, with reason `below_grade`. With `min_grade = "identical"`, only the
studies that asked the anchor question itself remain:

```{r min-grade}
strict <- qes_harmonize(targets = "sov_indep", layout = "long", missing = "reasons",
                        min_grade = "identical", include_draft = TRUE, quiet = TRUE,
                        lang = params$lang)
table(strict$study, strict$sov_indep__na)[, c("dk", "refused", "below_grade")]
```

`qes2018` and `qes2022` added "today" to the question (and `qes2022`
offered no "prefer not to answer"), so both are graded `comparable` and are
dropped here. Whether a `comparable` cell belongs in your analysis is a judgement
the grade reason helps you make; qesR never makes it for you.

## Weights

Each wave has at most one recommended weight, normalized to mean 1 in each
study and wave. `attr(, "qes_weight_guide")` says which wave each target
came from and which weight fits it:

```{r weights}
guide <- attr(h, "qes_weight_guide")
knitr::kable(guide[, c("target", "study", "wave", "weight_column", "weight_var")])
```

In the long layout the `weight` column holds, on each row, the weight of
that row's wave, so one design serves both targets.

## A weighted estimate

`qes_design()` hands the data to the `survey` package: each study is a
stratum and, in the long layout, the respondent is the sampling unit.
Estimates are computed within each study with `svyby()`; the table marks
the levels a study did not offer instead of printing a 0.

```{r design, message = FALSE}
d <- qes_design(h)
cells <- qes_provenance(h, level = "cell")
spec <- qes_spec("spec")

# Weighted percentage of each level of `target` by study, with its
# standard error; levels the study's question did not offer are marked.
share_table <- function(design, target) {
  est <- survey::svyby(stats::as.formula(paste0("~", target)), ~study, design,
                       survey::svymean, na.rm = TRUE)
  lv <- levels(h[[target]])
  tg <- spec$tables$targets
  names_lv <- spec$tables$levels$name[spec$tables$levels$levels_id ==
                                        tg$levels_id[tg$target == target]]
  out <- sapply(lv, function(l) {
    paste0(pct(est[[paste0(target, l)]]), " (", pct(est[[paste0("se.", target, l)]]), ")")
  })
  out <- matrix(out, nrow = nrow(est), dimnames = list(est$study, lv))
  for (s in rownames(out)) {
    off <- cells$levels_not_offered[cells$study == s & cells$target == target]
    gone <- lv[names_lv %in% strsplit(off, ";", fixed = TRUE)[[1]]]
    out[s, gone] <- tr("not offered", "non offert")
  }
  out
}
knitr::kable(share_table(d, "sov_indep"))
```

Each cell is a weighted percentage with its standard error in parentheses.
No study offered "would not vote" as an answer to this question, so that
level is a structural zero everywhere. The standard errors treat each
weighted sample as a probability sample. The four studies are web surveys
whose weights adjust to census margins, so read the standard errors as a
rough guide only (see `?qes_design`).

Reported vote describes the electorate, so keep the respondents who could
vote: `eligible_voter` is `TRUE` for those aged 18 or older on election day
and, where asked, Canadian citizens. `qes2018` sampled people aged 16 and
over:

```{r eligible}
table(h$study[h$wave %in% c("post", "pes")],
      h$eligible_voter[h$wave %in% c("post", "pes")], useNA = "ifany")
voters <- subset(d, eligible_voter %in% TRUE)
knitr::kable(share_table(voters, "vote_prov_recall"))
```

Keeping eligible voters does not recalibrate the weights, which target the
population each study sampled. A party marked "not offered" was not on that
study's list of answers, so its share there cannot be compared with the
other studies.

Pooling studies into one estimate is possible (`qes_design(pool = "equal")`
gives each study the same total), but the studies differ in population,
mode and wording; compare them side by side, as above, before pooling.

## Provenance

`qes_provenance(level = "cell")` gives, for each study and target, the
crosswalk row applied, its grade and review status, the weight, the levels
not offered and the count of valid values and of each missing reason:

```{r provenance}
knitr::kable(cells[, c("study", "target", "source_var", "grade", "status", "weight_var",
                       "levels_not_offered", "n_valid", "n_dk", "n_not_in_wave")])
```

These counts are per respondent of the study, as in the respondent layout,
so `n_valid` and the counts of the reasons add up to the study's
respondents; the table in *Why values are missing* counts rows of the long
layout instead. Every `qes2022` respondent took part in the campaign wave,
so none is `not_in_wave` for `sov_indep`, while 301 did not take part in
the post-election wave and are `not_in_wave` for `vote_prov_recall`.

`level = "study"` names the pinned file of each study, and `level = "spec"`
the specification version and its content hash; the same specification,
the same files and the same qesR version give the same values. `qes_cite()`
cites qesR with the specification, then each dataset:

```{r cite}
spec_record <- qes_provenance(h, level = "spec")
spec_record[, c("spec_version", "spec_hash", "qesR_version")]
cat(qes_cite(h, lang = params$lang), sep = "\n\n")
```

## Four sovereignty questions, never pooled

A target is one question stimulus. Other studies asked about sovereignty
with other wordings, so they feed other targets, which are never merged
with `sov_indep`:

```{r sovereignty}
sov <- qes_spec("crosswalk", targets = "sovereignty", lang = params$lang)
sov <- sov[sov$rule != "none", ]
rownames(sov) <- NULL
knitr::kable(sov[, c("target", "study", "source_var", "grade", "weight_var")])
```

`sov_sovereign_country` asks about a "sovereign country" (`qes2012_panel`),
`sov_favour` is a four-point scale of favour (`qes2018_panel`), and
`sov_partnership_1995` repeats the question of the 1995 referendum,
sovereignty with an offer of partnership to the rest of Canada (`qes2007`,
`qes2008`, `qes2007_panel` and the CROP respondents of `qes1998`). Their
weights are not documented yet, so `qes_harmonize()` returns `NA` for them
and `qes_design()` leaves these respondents out, with a message, until the
weights are reviewed. Put the targets side by side and read each one
against its own question.
