---
title: "How harmonization works"
params:
  lang: "en"
---

*[Version française](fr-harmonisation.html)*

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE)
# messages in the language of the page, whatever the locale of the build
options(qesR.lang = params$lang, width = 90)
# decimal comma on the French page (formatC() and print() follow OutDec)
if (identical(params$lang, "fr")) options(OutDec = ",")
library(qesR)
# what the page downloads (the note below): the studies qes_harmonize()
# reads by default for these targets, and the size of their data files
dl <- local({
  grid <- qes_spec(targets = c("sov_indep", "vote_prov_recall"))
  st <- as.data.frame(qes_studies())
  st <- st[st$default_member & st$study %in% names(grid), , drop = FALSE]
  st <- st[vapply(st$study, function(s) any(!is.na(grid[[s]])), logical(1)), , drop = FALSE]
  files <- utils::read.csv(system.file("extdata", "catalog", "files.csv", package = "qesR"),
                           colClasses = "character")
  ids <- paste(st$study, c(st$data_file_id, st$label_file_id))
  list(n = nrow(st), mb = sum(as.numeric(files$bytes[paste(files$study, files$file_id) %in% ids])) / 1e6)
})
```

::: {.qesr-download}
**This page downloads `r dl$n` studies** (about `r round(dl$mb)` MB) the first time it runs. `options(qesR.cache = "disk")` keeps them on disk for later sessions.
:::

This page goes from `qes_harmonize()` to a weighted estimate, study by
study, and shows what qesR records on the way: the comparability grade of
each study's question, the reason for every missing value, the weight that
fits each question, and the provenance of each cell. It is built on the
full data files.

The harmonization is **experimental**: targets, grades and mappings may
still change from one version of qesR to the next. For each study and
harmonized variable ("target"), the rules that ship with qesR say which
question feeds the target and how each of its codes maps to the target's
levels; nothing is matched by name. Some studies have no validated weight,
so their weight columns are `NA`; their answers are harmonized like the
others' (the [weights table](studies.html#weights) gives each study's
weight).

```{r packages, message = FALSE}
library(qesR)
```

## Which studies have the question

`qes_spec()` gives one row per target and one column per study, each cell
the grade of that study's question against the target's anchor question
(the full grid is on the [coverage page](coverage.html)). Two targets here:
the referendum question on Quebec becoming an independent country, and the
vote reported after the election.

```{r spec}
grid <- qes_spec(targets = c("sov_indep", "vote_prov_recall"), lang = "en")
knitr::kable(grid[, c("target", "label", "qes2022", "qes2018", "qes2014", "qes2012")])
```

The crosswalk view says which question each grade is about, and why:

```{r crosswalk}
xw <- qes_spec("crosswalk", targets = "sov_indep", lang = "en")
knitr::kable(xw[, c("study", "wave", "source_var", "grade", "grade_reason")])
```

`identical` means the same question, options and universe as the anchor;
`comparable` the same stimulus with differences that should not move the
shares; `approximate` a difference that can. The
[variable reference](harmonization-reference.html) gives the wording
and levels of every study's question.

## Harmonize

`studies = NULL`, the default, takes the Quebec Election Studies that have
a question for at least one of the targets; the Durand panels, the CROP
polls and the 1998 panel are left out unless you name them. `qes2008` has
no recommended weight (both of its weights are calibrated on the vote or
on turnout), so its rows have no weight and `qes_design()` leaves them out
below, with a message.
`layout = "long"` gives one row
per respondent and wave, so each answer sits on the row of the wave that
asked it, next to that wave's weight. `missing = "reasons"` adds a column
`<target>__na` with the reason of every missing value.

```{r harmonize}
h <- qes_harmonize(targets = c("sov_indep", "vote_prov_recall"), layout = "long",
                   missing = "reasons", lang = "en")
table(h$study, h$wave)
```

The messages are part of the result. The first lists the *structural
zeros*: levels a study's question did not offer, such as a party missing
from its list. Their share in that study is 0 because nobody could choose
them, not because nobody supported them. `qes2022` asked the referendum
question in its campaign-period wave (`cps`) and the reported vote in its
post-election wave (`pes`), which have different weights.

## Why values are missing

Every missing value has a reason. Counts of the valid answers and of each
reason, for the referendum question:

```{r reasons}
reasons <- table(h$study, h$sov_indep__na)
reasons <- reasons[, colSums(reasons) > 0, drop = FALSE]
knitr::kable(cbind(valid = tapply(!is.na(h$sov_indep), h$study, sum), reasons[, , drop = FALSE]))
```

`dk` is "don't know" and `refused` a refusal; `not_in_wave` marks the
`qes2022` post-election rows, since the question was asked in the other
wave. `?qes_harmonize` lists every reason.

## Grades

`min_grade` keeps only the cells at or above a grade and sets the others to
`NA`, with reason `below_grade`. With `min_grade = "identical"`, only the
studies that asked the anchor question itself remain:

```{r min-grade}
strict <- qes_harmonize(targets = "sov_indep", layout = "long", missing = "reasons",
                        min_grade = "identical", quiet = TRUE,
                        lang = "en")
table(strict$study, strict$sov_indep__na)[, c("dk", "refused", "below_grade")]
```

`qes2018` and `qes2022` added "today" to the question (and `qes2022`
offered no "prefer not to answer"), so both are graded `comparable` and are
dropped here. Whether a `comparable` cell belongs in your analysis is a judgement
the grade reason helps you make; qesR never makes it for you.

## Weights

Each wave has at most one recommended weight, normalized to mean 1 in each
study and wave. `attr(, "qes_weight_guide")` says which wave each target
came from and which weight fits it:

```{r weights}
guide <- attr(h, "qes_weight_guide")
knitr::kable(guide[, c("target", "study", "wave", "weight_column", "weight_var")])
```

In the long layout the `weight` column holds, on each row, the weight of
that row's wave, so one design serves both targets.

## A weighted estimate

`qes_design()` hands the data to the `survey` package: each study is a
stratum and, in the long layout, the respondent is the sampling unit.
Estimates are computed within each study with `svyby()`; a dash marks the
levels a study did not offer, instead of a 0.

```{r design}
d <- qes_design(h)
```

```{r design-table, message = FALSE}
cells <- qes_provenance(h, level = "cell")
spec <- qes_spec("spec")

# Weighted percentage of each level of `target` by study, with its
# standard error; a dash marks the levels the study's question did not offer.
pct <- function(x) formatC(100 * x, format = "f", digits = 1)
share_table <- function(design, target) {
  est <- survey::svyby(stats::as.formula(paste0("~", target)), ~study, design,
                       survey::svymean, na.rm = TRUE)
  lv <- levels(h[[target]])
  tg <- spec$tables$targets
  names_lv <- spec$tables$levels$name[spec$tables$levels$levels_id ==
                                        tg$levels_id[tg$target == target]]
  out <- sapply(lv, function(l) {
    paste0(pct(est[[paste0(target, l)]]), " (", pct(est[[paste0("se.", target, l)]]), ")")
  })
  out <- matrix(out, nrow = nrow(est), dimnames = list(est$study, lv))
  for (s in rownames(out)) {
    off <- cells$levels_not_offered[cells$study == s & cells$target == target]
    gone <- lv[names_lv %in% strsplit(off, ";", fixed = TRUE)[[1]]]
    out[s, gone] <- "\u2014"
  }
  out
}
knitr::kable(share_table(d, "sov_indep"))
```

Each cell is a weighted percentage with its standard error in parentheses.
No study offered "would not vote" as an answer to this question, so that
level is a structural zero everywhere. The standard errors treat each
weighted sample as a probability sample. These studies are web surveys
whose weights adjust to census margins, so read the standard errors as a
rough guide only (see `?qes_design`).

Reported vote describes the electorate, so keep the respondents who could
vote: `eligible_voter` is `TRUE` for those aged 18 or older on election day
and, where asked, Canadian citizens. `qes2018` sampled people aged 16 and
over:

```{r eligible}
table(h$study[h$wave %in% c("post", "pes")],
      h$eligible_voter[h$wave %in% c("post", "pes")], useNA = "ifany")
voters <- subset(d, eligible_voter %in% TRUE)
knitr::kable(share_table(voters, "vote_prov_recall"))
```

Keeping eligible voters does not recalibrate the weights, which target the
population each study sampled. A party marked with a dash was not on that
study's list of answers, so its share there cannot be compared with the
other studies.

Pooling studies into one estimate is possible (`qes_design(pool = "equal")`
gives each study the same total), but the studies differ in population,
mode and wording; compare them side by side, as above, before pooling.

## Provenance

`qes_provenance(level = "cell")` gives, for each study and target, the
question applied, its grade, the weight, the levels not offered and the
count of valid values and of each missing reason:

```{r provenance}
knitr::kable(cells[, c("study", "target", "source_var", "grade", "weight_var",
                       "levels_not_offered", "n_valid", "n_dk", "n_not_in_wave")])
```

These counts are per respondent of the study, as in the respondent layout,
so `n_valid` and the counts of the reasons add up to the study's
respondents; the table in *Why values are missing* counts rows of the long
layout instead. Every `qes2022` respondent took part in the campaign wave,
so none is `not_in_wave` for `sov_indep`, while
`r cells$n_not_in_wave[cells$study == "qes2022" & cells$target == "vote_prov_recall"]`
did not take part in the post-election wave and are `not_in_wave` for `vote_prov_recall`.

qesR records which files and which version of the rules produced every
cell: `level = "study"` names the data file of each study, and
`level = "spec"` the version of the rules. The same files and the same qesR
version give the same values. `qes_cite()` cites qesR with the rules, then
each dataset:

```{r cite}
spec_record <- qes_provenance(h, level = "spec")
spec_record[, c("spec_version", "qesR_version")]
cat(qes_cite(h, lang = "en"), sep = "\n\n")
```

## Five sovereignty targets, one pooled variable

A target is one question stimulus. Other studies asked about sovereignty
with other wordings, so they feed other targets, never merged into
`sov_indep`:

```{r sovereignty}
sov <- qes_spec("crosswalk", targets = "sovereignty", lang = "en")
sov <- sov[sov$rule != "none", ]
rownames(sov) <- NULL
knitr::kable(sov[, c("target", "study", "source_var", "grade", "weight_var")])
```

`sov_sovereign_country` asks about a "sovereign country" (`qes2012_panel`),
`sov_favour` is a four-point scale of favour (`qes2018_panel`), and
`sov_partnership_1995` repeats the question of the 1995 referendum,
sovereignty with an offer of partnership to the rest of Canada (`qes2007`,
`qes2008`, `qes2007_panel` and the CROP respondents of `qes1998`). Only
`qes2007` and the 2018 panel have a usable weight for these questions: both
weights of `qes2008` are calibrated on the vote or on turnout, and the
others' weights are not validated (see the
[weights table](studies.html#weights)). The other studies'
rows are left out of weighted estimates, with a message. Put the targets
side by side and read each one against its own question.

To follow support across every study anyway, the pooled variable
`sov_support` takes each study's referendum question, whatever its
wording, and says which one each value comes from (`sov_support__type`),
with the grade of that question (`sov_support__grade`); the favourable or
opposed scale of `qes2018_panel` is collapsed to yes or no and graded
approximate. `vote_choice` does the same for the vote (the reported vote,
else the vote intention), and `pol_interest` for interest in politics:

```{r pooled}
pooled <- qes_spec("pooled", targets = "sov_support", lang = "en")
knitr::kable(pooled[, c("type_name", "member", "precedence", "default", "grade_cap")])
```

```{r reset, include = FALSE}
# the decimal mark of the French page is for this page only
options(OutDec = ".")
```
