---
title: "Merged Dataset Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Merged Dataset Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

*[Version française](fr-donnees-fusionnees.html)*

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`get_qes_master()` stacks the Quebec Election Studies of qesR 0.4.4 in one data frame, with the 30 harmonized columns of 0.4.4: same names, order and types. It is the fixed legacy schema, kept stable for code written for 0.4.4.

```{r}
library(qesR)
qes_studies()$study
```

By default, and with `surveys = "all"`, `get_qes_master()` builds the 11 studies of qesR 0.4.4. The 1998 firm files `qes1998_crop` and `qes1998_createc` are not in the master yet (their respondents are already in `qes1998`); read them with `get_qes()`, as `?get_qes_master` explains.

## Build merged data

The synthetic demonstration study runs offline:

```{r}
demo_master <- get_qes_master(surveys = "qes_demo", quiet = TRUE)
dim(demo_master)
head(demo_master[, c("qes_code", "age_group", "gender", "turnout", "vote_choice")])
```

The real studies are read from their pinned original files:

```{r, eval = FALSE}
master <- get_qes_master(strict = FALSE)
dim(master)
table(master$qes_code)
```

## How it is built

- Each column reads the variable qesR 0.4.4 read for that study, and converts it as 0.4.4 did. The sources are frozen: nothing is chosen by name when you run it. `attr(master, "source_map")` gives the source of every column.
- Every respondent of every file is kept: there is no de-duplication and no removal of empty rows. Panel and cross-section respondents stay separate rows.
- Values verified to be wrong are `NA`: for example vote intentions in `vote_choice` (`qes2022`, CROP, `qes1998`), sovereignty questions other than the referendum on an independent country, and `party_best` and `party_lean` everywhere. `attr(master, "legacy_na_columns")` lists each column and study that is `NA` by design, with the reason.
- `vote_choice_timing` and `sovereignty_item` say what `vote_choice` and `sovereignty_support` hold in each study.
- qesR 0.4.4 also appended 70 columns by stacking variables that share a name across studies; they mixed different questions and are no longer built (`attr(master, "removed_columns")`). Read those items from each study with `get_qes()`.

```{r}
head(attr(demo_master, "source_map")[, c("qes_code", "harmonized_variable", "source_variable")])
attr(demo_master, "legacy_na_columns")[, c("column", "study", "reason", "cause")]
```

`attr(master, "legacy_column_map")` says what each column means and flags the columns that mix instruments (`political_interest`, `age_group`, `education`). `survey_weight` is each study's own weight on its own scale: do not pool weighted estimates across studies without rescaling.

Results from qesR 0.4.4 are reproducible only with that version (`remotes::install_github("ThomasGareau/qesR", ref = "v0.4.4")`).

## Save merged data

`save_path` writes a UTF-8 CSV file or an RDS file, and next to it `<stem>_provenance.csv`, the record of the file read for each study.

```{r, eval = FALSE}
get_qes_master(save_path = file.path(tempdir(), "qes_master.csv"), strict = FALSE)
get_qes_master(save_path = file.path(tempdir(), "qes_master.rds"), strict = FALSE)
```
