---
title: "The legacy merged file"
params:
  lang: "en"
---

*[Version française](fr-donnees-fusionnees.html)*

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# messages in the language of the page, whatever the locale of the build
options(qesR.lang = params$lang)
```

`get_qes_master()` stacks the Quebec Election Studies of qesR 0.4.4 in one data frame, with the 30 harmonized columns of 0.4.4: same names, order and types. It is the fixed legacy schema, kept stable for code written for 0.4.4. [Upgrading from qesR 0.4.4](migrating-0.7.html) says what changed in its values.

```{r}
library(qesR)
qes_studies()$study
```

By default, and with `surveys = "all"`, `get_qes_master()` builds the 11 studies of qesR 0.4.4. The 1998 firm files `qes1998_crop` and `qes1998_createc` are not in the master (their respondents are already in `qes1998`); read them with `get_qes()`, as `?get_qes_master` explains.

## Build merged data

The synthetic demonstration study runs offline:

```{r}
demo_master <- get_qes_master(surveys = "qes_demo", quiet = TRUE)
dim(demo_master)
head(demo_master[, c("qes_code", "age_group", "gender", "turnout", "vote_choice")])
```

The real studies are read from their original files:

```{r real}
master <- get_qes_master(quiet = TRUE)
dim(master)
table(master$qes_code)
```

## How it is built

- Each study is harmonized by `qes_harmonize()`, and each column is built from the harmonized variables ("targets") it needs. A code the harmonization rules do not map is `NA`, never passed through. `attr(master, "source_map")` gives, for every column of every study, the question read, its target and its comparability grade.
- Every respondent of every file is kept: there is no de-duplication and no removal of empty rows. Panel and cross-section respondents stay separate rows.
- `vote_choice` and `turnout` are the reported vote and turnout in every study; the CROP polls asked only vote intentions (in `vote_intent`). `sovereignty_support` is the referendum on an independent country only, and `party_best` and `party_lean` are `NA` everywhere. `attr(master, "legacy_na_columns")` lists each column and study that is `NA` throughout, with the reason and, in `basis`, why in words (`?get_qes_master` lists the causes).
- `vote_choice_timing` and `sovereignty_item` say what `vote_choice` and `sovereignty_support` hold in each study.
- qesR 0.4.4 also appended 70 columns by stacking variables that share a name across studies; they mixed different questions and are no longer built (`attr(master, "removed_columns")`). Read those items from each study with `get_qes()`.

```{r}
head(attr(demo_master, "source_map")[, c("qes_code", "harmonized_variable", "source_variable", "target", "grade")])
attr(demo_master, "legacy_na_columns")[, c("column", "study", "reason", "cause")]
```

`attr(master, "legacy_column_map")` says what each column means and flags the columns that mix instruments (`political_interest`, `age_group`, `education`). `survey_weight` is each study's own weight on its own scale: do not pool weighted estimates across studies without rescaling.

To reproduce a 0.4.4 result, install 0.4.4 (`remotes::install_github("ThomasGareau/qesR", ref = "v0.4.4")`).

## Save merged data

`save_path` writes a UTF-8 CSV file or an RDS file, and next to it `<stem>_provenance.csv`, the record of the file read for each study.

```{r, eval = FALSE}
get_qes_master(save_path = file.path(tempdir(), "qes_master.csv"), strict = FALSE)
get_qes_master(save_path = file.path(tempdir(), "qes_master.rds"), strict = FALSE)
```
