---
title: "From two parties to four: Quebec's party system, 1998-2022"
description: >
  Who moved when the PQ-PLQ duopoly became a four-party system:
  francophones and the others, young and old voters.
params:
  lang: "en"
---

*[Version française](fr-realignement.html)*

```{r setup, include = FALSE}
source("_setup.R", encoding = "UTF-8")
```

```{r harmonize, echo = FALSE}
h <- qes_harmonize(
  studies = qz_studies,
  targets = c("vote_choice", "lang_mother", "age_group3"),
  types = list(vote_choice = "recall"),
  missing = "reasons", quiet = TRUE
)
```

```{r estimates, echo = FALSE}
h$party <- factor(qz_party_code(h$vote_choice), levels = qz_parties)
# the 1998 polls interviewed francophones only
h$language <- ifelse(h$study == "qes1998" | h$lang_mother %in% "French", "fr",
                     ifelse(is.na(h$lang_mother), NA, "other"))
recall_studies <- setdiff(qz_studies, "qes_crop_2007_2010")

shares <- do.call(rbind, lapply(recall_studies, function(s) {
  d <- qz_design(h, s, "weight_post")
  rbind(qz_tag(qz_share(d, "party"), study = s),
        qz_tag(qz_share(d, "party", by = "language"), study = s),
        qz_tag(qz_share(d, "party", by = "age_group3"), study = s))
}))
shares$year <- qz_year(shares$study)
shares$status <- qz_party_status(shares$study, shares$level)
shares <- shares[shares$status == "offered", ]

enp <- do.call(rbind, lapply(recall_studies, function(s) {
  d <- qz_design(h, s, "weight_post")
  lang <- d$variables$language
  rbind(qz_tag(qz_enp(d, "party"), study = s, group = "all"),
        qz_tag(qz_enp(d, "party", lang %in% "fr"), study = s, group = "fr"),
        qz_tag(qz_enp(d, "party", lang %in% "other"), study = s, group = "other"),
        make.row.names = FALSE)
}))
enp$year <- qz_year(enp$study)
enp$weighted <- vapply(enp$study, function(s) qz_weighted(h, s, "weight_post"), logical(1))
official <- qz_official()
enp_official <- qz_enp_official(official)

grade <- setNames(qz_first(h, "vote_choice__grade", recall_studies), recall_studies)
item <- setNames(qz_first(h, "vote_choice__item", recall_studies), recall_studies)
e <- function(s, g) enp$enp[enp$study == s & enp$group == g]
p <- function(s, g, party) shares$pct[shares$study == s & shares$group == g & shares$level == party]
```

In 1998 the PQ and the PLQ took 86% of the valid votes. In 2022 five
parties took more than 12% each, and the effective number of parties
(official results) rose from
`r qz_num(enp_official$share_valid[enp_official$year == 1998])` to
`r qz_num(enp_official$share_valid[enp_official$year == 2022])`. The
surveys show where this happened: among francophones. Their effective
number of parties went from `r qz_num(e("qes1998", "fr"))` in 1998 (the
CROP and CREATEC polls, which interviewed francophones only and are
unweighted) to `r qz_num(e("qes2022", "fr"))` in the 2022 Quebec Election
Study, as their vote split among the PQ, the CAQ, Québec solidaire and
the PLQ. Non-francophones kept voting PLQ: `r qz_pct(p("qes2018", "other", "PLQ"), 0)`
of them in 2018, against `r qz_pct(p("qes2018", "fr", "PLQ"), 0)` of
francophones. The CAQ won that year with `r qz_pct(p("qes2018", "fr", "CAQ"), 0)`
of the francophone vote and `r qz_pct(p("qes2018", "other", "CAQ"), 0)` of the
other voters.

## The data

Every figure on this page uses one pooled variable, `vote_choice`,
restricted to the vote reported after the election (`types`):

```{r show-harmonize, ref.label = "harmonize", eval = FALSE}
```

A weighted share with its confidence interval, within one study, is a
`qes_design()` and a call to the survey package:

```{r example}
d18 <- qes_design(h[h$study == "qes2018", ], weight = "weight_post")
d18 <- subset(d18, !is.na(vote_choice) & !is.na(lang_mother))
svyby(~I(vote_choice == "CAQ"), ~lang_mother, d18, svyciprop,
      method = "logit", vartype = "ci")
```

The figures repeat this for every study, with the helpers of
[`_data.R`](https://github.com/ThomasGareau/qesR/blob/main/vignettes/articles/_data.R).
The estimates of each study are made within that study and never pooled
across studies.

## Francophones fragmented; non-francophones stayed with the PLQ

```{r fig-enp, echo = FALSE}
d <- enp[enp$study %in% qz_primary & enp$group != "all" & enp$n >= qz_min_n, ]
d$series <- d$group
o <- data.frame(study = NA, group = "official", enp = enp_official$share_valid, lo = NA, hi = NA, n = NA,
                year = enp_official$year, weighted = TRUE, series = "official")
o <- o[o$year %in% as.integer(names(qz_primary)), ]
d <- rbind(d[, names(o)], o)
d$x <- qz_ex(d$year)
d$era <- qz_era(d$year)
d$series <- factor(d$series, levels = c("official", "fr", "other"))
series_lab <- c(official = tr("Official result, all voters", "Résultat officiel, tous les électeurs"),
                fr = tr("Francophones", "Francophones"),
                other = tr("Non-francophones", "Non-francophones"))
d$label <- c(official = tr("Official", "Officiel"), fr = tr("Francophones", "Francophones"),
             other = tr("Non-francophones", "Non-francophones"))[as.character(d$series)]
surveyed <- d[d$series != "official", ]

fig_enp <- function(mode) {
  k <- qz_tokens[[mode]]
  cols <- c(official = k$ink, fr = k$grp[["fr"]], other = k$grp[["other"]])
  ggplot(d, aes(x, enp, colour = series, group = interaction(series, era))) +
    qz_xbreak(mode) +
    qz_band(data = surveyed, aes(ymin = lo, ymax = hi, fill = series), mode = mode) +
    qz_ci(data = surveyed[surveyed$era == "before", ], aes(ymin = lo, ymax = hi), show.legend = FALSE) +
    qz_line(show.legend = FALSE) +
    qz_dots(mode, d, aes(x, enp), colour_aes = "series") +
    qz_end_labels(d, "x", "enp", "label", "series", mode, min_gap = 0.14, nudge_x = 1.1) +
    scale_colour_manual(values = cols, guide = "none") +
    scale_fill_manual(values = cols, breaks = names(series_lab), labels = series_lab,
                      guide = guide_legend(order = 1)) +
    scale_x_election(add = c(0.9, 5.2)) +
    scale_y_continuous(breaks = 1:4) +
    coord_cartesian(ylim = c(1, 4.6), expand = FALSE, clip = "off") +
    labs(x = NULL, y = tr("Effective number of parties", "Nombre effectif de partis")) +
    theme_qes(mode) +
    theme(axis.line.x = element_blank())
}

tab <- enp[enp$group != "all", ]
tab <- tab[order(tab$year, tab$study, tab$group), ]
tab_enp <- setNames(data.frame(
  tab$year,
  qz_study_label(tab$study),
  series_lab[tab$group],
  qz_cell(tab$enp, tab$lo, tab$hi, tab$n, 2),
  tab$n,
  qz_num(enp_official$share_valid[match(tab$year, enp_official$year)], 2),
  qz_weight_text(tab$weighted)
, row.names = NULL), c(
  tr("Election", "Élection"),
  tr("Study", "Étude"),
  tr("Group", "Groupe"),
  tr("Effective number [95% CI]", "Nombre effectif [IC à 95 %]"),
  "n",
  tr("Official, all voters", "Officiel, tous les électeurs"),
  tr("Weighting", "Pondération")
))
alt <- sprintf(tr(
  "Line chart of the effective number of parties at each Quebec election from 1998 to 2022: the official result, francophone respondents and non-francophone respondents. Among francophones it rises from %s in 1998 to %s in 2022; among non-francophones it stays between %s and %s until 2018 and reaches %s in 2022, with a wide interval. Values in the table view.",
  "Graphique linéaire du nombre effectif de partis à chaque élection québécoise de 1998 à 2022 : le résultat officiel, les répondants francophones et les répondants non francophones. Chez les francophones, il passe de %s en 1998 à %s en 2022 ; chez les non-francophones, il reste entre %s et %s jusqu'en 2018 et atteint %s en 2022, avec un intervalle large. Valeurs dans la vue en tableau."),
  qz_num(e("qes1998", "fr")), qz_num(e("qes2022", "fr")),
  qz_num(min(enp$enp[enp$study %in% qz_primary[2:6] & enp$group == "other"])),
  qz_num(max(enp$enp[enp$study %in% qz_primary[2:6] & enp$group == "other"])), qz_num(e("qes2022", "other")))
qz_figure(fig_enp, "enp",
  title = sprintf(tr("The francophone vote fragmented, from %s to %s effective parties; the non-francophone vote stayed concentrated until 2018",
                     "Le vote francophone s'est fragmenté, de %s à %s partis effectifs ; le vote non francophone est resté concentré jusqu'en 2018"),
                  qz_num(e("qes1998", "fr")), qz_num(e("qes2022", "fr"))),
  subtitle = tr("Effective number of parties at each election: official result and reported vote by mother tongue, with 95% confidence intervals",
                "Nombre effectif de partis à chaque élection : résultat officiel et vote déclaré selon la langue maternelle, avec intervalles de confiance à 95 %"),
  alt = alt, table = tab_enp, height = 4.0,
  note = tr("Source: qesR, pooled vote_choice (reported vote) of one Quebec Election Study per election and the 1998 polls (CROP and CREATEC, francophones only, unweighted; their recontact over-selected undecided voters and refusers); official results of Élections Québec. Effective number of parties = 1 / sum of squared shares, with the PV, ON and other parties counted as one. Bands: 95% confidence intervals. Weighted with each study's post-election weight; hollow dots (1998, 2008): unweighted, weight under review. The x axis is in years, with a break between 1998 and 2007. The Durand panels are in the table view.",
            "Source : qesR, variable regroupée vote_choice (vote déclaré) d'une Étude électorale québécoise par élection et des sondages de 1998 (CROP et CREATEC, francophones seulement, non pondérés ; leur recontact a surreprésenté les indécis et les refus) ; résultats officiels d'Élections Québec. Nombre effectif de partis = 1 / somme des carrés des parts, le PV, ON et les autres partis comptant pour un. Bandes : intervalles de confiance à 95 %. Pondéré avec la pondération postélectorale de chaque étude ; points creux (1998, 2008) : non pondéré, pondération en révision. L'axe horizontal est en années, avec une coupure entre 1998 et 2007. Les panels Durand sont dans la vue en tableau."))
```

What to notice:

- Over the whole period both lines rise, the official one from
  `r qz_num(enp_official$share_valid[enp_official$year == 1998])` to
  `r qz_num(enp_official$share_valid[enp_official$year == 2022])` and the
  francophone one from `r qz_num(e("qes1998", "fr"))` to
  `r qz_num(e("qes2022", "fr"))`, but not election by election: from 2008
  to 2012 the official number rises
  (`r qz_num(enp_official$share_valid[enp_official$year == 2008])` to
  `r qz_num(enp_official$share_valid[enp_official$year == 2012])`) while the
  francophone one stays at `r qz_num(e("qes2012", "fr"))`, and in 2014 the
  official number falls
  (`r qz_num(enp_official$share_valid[enp_official$year == 2014])`) while the
  francophone one rises (`r qz_num(e("qes2014", "fr"))`). The first peak,
  in 2007, is the three-way race of the PQ, the PLQ and the ADQ.
- Among non-francophones the vote stays concentrated, with an effective
  number between `r qz_num(min(enp$enp[enp$study %in% qz_primary[2:6] & enp$group == "other"]))`
  and `r qz_num(max(enp$enp[enp$study %in% qz_primary[2:6] & enp$group == "other"]))`
  from 2007 to 2018. It reaches `r qz_num(e("qes2022", "other"))` in 2022,
  when the PLQ fell to `r qz_pct(p("qes2022", "other", "PLQ"), 0)` of their
  vote, but the interval runs from
  `r qz_num(enp$lo[enp$study == "qes2022" & enp$group == "other"])` to
  `r qz_num(enp$hi[enp$study == "qes2022" & enp$group == "other"])`: only
  `r enp$n[enp$study == "qes2022" & enp$group == "other"]` non-francophone
  voters answered, so the size of the jump is uncertain.
- The survey line runs close to the official one, but not on it: surveys
  over-report some parties (see
  [Survey vs official results](survey-vs-official.html)).

## Each party's vote, francophones and the rest

```{r fig-lang, echo = FALSE}
panel_parties <- c("PLQ", "PQ", "QS", "ADQ", "CAQ")
d <- shares[shares$study %in% qz_primary & shares$group %in% c("fr", "other") &
              shares$level %in% panel_parties & shares$n >= qz_min_n, ]
d$x <- qz_ex(d$year)
d$era <- qz_era(d$year)
d$party <- factor(d$level, levels = panel_parties)
d$series <- d$group
o <- official[official$party %in% panel_parties & official$year %in% as.integer(names(qz_primary)), ]
o$x <- qz_ex(o$year)
o$party <- factor(o$party, levels = panel_parties)
group_lab <- c(fr = tr("Francophones", "Francophones"), other = tr("Non-francophones", "Non-francophones"))

fig_lang <- function(mode) {
  k <- qz_tokens[[mode]]
  cols <- c(fr = k$grp[["fr"]], other = k$grp[["other"]])
  ggplot(d, aes(x, pct, colour = series, group = interaction(series, era))) +
    qz_xbreak(mode) +
    geom_segment(data = o, aes(x = x - 0.45, xend = x + 0.45, y = share_valid, yend = share_valid, linetype = "official"),
                 colour = k$ink, linewidth = 0.7, inherit.aes = FALSE) +
    qz_ci(aes(ymin = lo, ymax = hi), show.legend = FALSE) +
    qz_line(show.legend = FALSE) +
    qz_dots(mode, d, aes(x, pct), colour_aes = "series", size = 2.4, wrap = TRUE) +
    facet_wrap(~party, ncol = 3, axes = "all_x", labeller = as_labeller(function(x) qz_party_label(x))) +
    scale_colour_manual(values = cols, guide = "none") +
    scale_fill_manual(values = cols, breaks = names(group_lab), labels = group_lab,
                      guide = guide_legend(order = 1)) +
    scale_linetype_manual(values = c(official = "solid"),
                          labels = c(official = tr("Official result,\nall voters", "Résultat officiel,\ntous les électeurs")),
                          guide = guide_legend(order = 2, override.aes = list(colour = k$ink, linewidth = 0.7))) +
    scale_x_election(short = TRUE, add = c(1.4, 1.4)) +
    scale_y_continuous(breaks = seq(0, 80, 20), labels = qz_pct_axis) +
    coord_cartesian(ylim = c(0, 92), expand = FALSE, clip = "off") +
    labs(x = NULL, y = NULL) +
    theme_qes(mode) +
    theme(axis.line.x = element_blank(), panel.spacing.x = unit(22, "pt"),
          legend.position = "inside", legend.position.inside = c(0.695, 0.25),
          legend.justification.inside = c(0, 0.5), legend.direction = "vertical",
          legend.box = "vertical", legend.box.just = "left", legend.spacing.y = unit(4, "pt"),
          legend.key.spacing.y = unit(4, "pt"))
}

tab <- shares[shares$group %in% c("fr", "other") & shares$level %in% c(panel_parties, "PCQ", "other"), ]
tab <- tab[order(match(tab$level, qz_parties), tab$year, tab$study, tab$group), ]
tab_lang <- setNames(data.frame(
  qz_party_label(tab$level),
  tab$year,
  qz_study_label(tab$study),
  group_lab[tab$group],
  qz_cell(tab$pct, tab$lo, tab$hi, tab$n),
  tab$n,
  qz_num(official$share_valid[match(paste(tab$year, tab$level), paste(official$year, official$party))]),
  qz_weight_text(tab$weighted),
  qz_grade_text(grade[tab$study])
, row.names = NULL), c(
  tr("Party", "Parti"),
  tr("Election", "Élection"),
  tr("Study", "Étude"),
  tr("Group", "Groupe"),
  tr("Share of reported vote [95% CI]", "Part du vote déclaré [IC à 95 %]"),
  "n",
  tr("Official, all voters", "Officiel, tous les électeurs"),
  tr("Weighting", "Pondération"),
  tr("Grade", "Niveau")
))
alt <- sprintf(tr(
  "Five small line charts, one per party (PLQ, PQ, QS, ADQ, CAQ), of the share of the reported vote among francophones and among non-francophones at each election from 1998 to 2022, with the official result as a tick. The PLQ takes %s of the non-francophone vote in 2018 against %s of the francophone vote; the CAQ takes %s of francophones and %s of the others. Values in the table view.",
  "Cinq petits graphiques linéaires, un par parti (PLQ, PQ, QS, ADQ, CAQ), de la part du vote déclaré chez les francophones et chez les non-francophones à chaque élection de 1998 à 2022, avec le résultat officiel en trait. Le PLQ obtient %s du vote non francophone en 2018 contre %s du vote francophone ; la CAQ obtient %s des francophones et %s des autres. Valeurs dans la vue en tableau."),
  qz_pct(p("qes2018", "other", "PLQ"), 0), qz_pct(p("qes2018", "fr", "PLQ"), 0),
  qz_pct(p("qes2018", "fr", "CAQ"), 0), qz_pct(p("qes2018", "other", "CAQ"), 0))
qz_figure(fig_lang, "lang",
  title = sprintf(tr("Non-francophones vote PLQ (%s in 2018); the CAQ's wins are francophone wins (%s of francophones, %s of the others)",
                     "Les non-francophones votent PLQ (%s en 2018) ; les victoires de la CAQ sont francophones (%s des francophones, %s des autres)"),
                  qz_pct(p("qes2018", "other", "PLQ"), 0), qz_pct(p("qes2018", "fr", "CAQ"), 0), qz_pct(p("qes2018", "other", "CAQ"), 0)),
  subtitle = tr("Share of the reported vote among francophones and non-francophones, by party, with 95% confidence intervals and the official result",
                "Part du vote déclaré chez les francophones et les non-francophones, selon le parti, avec intervalles de confiance à 95 % et résultat officiel"),
  alt = alt, table = tab_lang, height = 5.2,
  note = tr("Source: qesR, pooled vote_choice (reported vote), one Quebec Election Study per election and the 1998 polls (francophones only). Tick: official share of valid votes, all voters. Bars: 95% confidence intervals (logit). Weighted; hollow dots (1998, 2008): unweighted, weight under review. Cells of fewer than 30 respondents are not drawn. The x axis is in years, with a break between 1998 and 2007. The ADQ merged into the CAQ in 2012; the two stay separate parties here. The PCQ, the Durand panels and the other parties are in the table view.",
            "Source : qesR, variable regroupée vote_choice (vote déclaré), une Étude électorale québécoise par élection et les sondages de 1998 (francophones seulement). Trait : part officielle des votes valides, tous les électeurs. Barres : intervalles de confiance à 95 % (logit). Pondéré ; points creux (1998, 2008) : non pondéré, pondération en révision. Les cellules de moins de 30 répondants ne sont pas tracées. L'axe horizontal est en années, avec une coupure entre 1998 et 2007. L'ADQ a fusionné avec la CAQ en 2012 ; les deux restent des partis distincts ici. Le PCQ, les panels Durand et les autres partis sont dans la vue en tableau."))
```

What to notice:

- The PLQ remained the party of non-francophones:
  `r qz_pct(p("qes2007", "other", "PLQ"), 0)` of their vote in 2007,
  `r qz_pct(p("qes2014", "other", "PLQ"), 0)` in 2014, while its francophone
  vote fell from `r qz_pct(p("qes1998", "fr", "PLQ"), 0)` in 1998 (the 1998
  polls) to
  `r qz_pct(p("qes2022", "fr", "PLQ"), 0)` in 2022, far below its official share.
- The PQ lost the few non-francophone voters it had
  (`r qz_pct(p("qes2007", "other", "PQ"), 0)` in 2007,
  `r qz_pct(p("qes2022", "other", "PQ"), 0)` in 2022). Québec solidaire grew
  among francophones first, and among the others in 2022.
- The CAQ's wins are francophone wins; like the ADQ before it, it barely
  reaches the other voters (`r qz_pct(p("qes2022", "other", "CAQ"), 0)` in 2022).

## The age gap, party by party

```{r fig-age, echo = FALSE}
age_parties <- c("PLQ", "PQ", "ADQ", "QS", "CAQ", "PCQ")
agegap <- do.call(rbind, lapply(recall_studies, function(s) {
  d <- qz_design(h, s, "weight_post")
  do.call(rbind, lapply(age_parties, function(pp) {
    qz_tag(qz_gap(d, "party", pp, "age_group3", "18-34", "55 and over"),
           study = s, party = pp, weighted = isTRUE(attr(d, "qz_weighted")))
  }))
}))
agegap$year <- qz_year(agegap$study)
agegap$status <- qz_party_status(agegap$study, agegap$party)
agegap <- agegap[agegap$status == "offered", ]
d <- agegap[agegap$study %in% qz_primary & !is.na(agegap$gap), ]
d$x <- qz_ex(d$year)
d$era <- qz_era(d$year)
d$party <- factor(d$party, levels = age_parties)
ext <- d[d$year == 2022 & d$party %in% c("QS", "CAQ"), ]
ext$label <- qz_pts(ext$gap)
g <- function(y, party) agegap$gap[agegap$study == qz_primary[as.character(y)] & agegap$party == party]
age_lab <- c(`18-34` = tr("18 to 34", "18 à 34 ans"), `35-54` = tr("35 to 54", "35 à 54 ans"),
             `55 and over` = tr("55 and over", "55 ans et plus"))

fig_age <- function(mode) {
  k <- qz_tokens[[mode]]
  ggplot(d, aes(x, gap, colour = party, group = era)) +
    geom_hline(yintercept = 0, colour = k$axis, linewidth = 0.35) +
    qz_ci(aes(ymin = lo, ymax = hi)) +
    qz_line() +
    qz_dots(mode, d, aes(x, gap), colour_aes = "party", size = 2.4) +
    geom_text(data = ext, aes(x - 0.4, ifelse(gap > 0, hi + 2, lo - 2), label = label,
                              vjust = ifelse(gap > 0, 0, 1)), hjust = 1, size = 3.2, colour = k$ink,
              family = qz_font(), fontface = "bold") +
    qz_xbreak(mode) +
    facet_wrap(~party, ncol = 3, axes = "all_x", labeller = as_labeller(function(x) qz_party_label(x))) +
    scale_party(mode) +
    scale_x_election(short = TRUE, add = c(1.4, 1.4)) +
    scale_y_continuous(breaks = seq(-40, 40, 20), labels = function(x) qz_pts(x)) +
    coord_cartesian(ylim = c(-50, 46), expand = FALSE, clip = "off") +
    labs(x = NULL, y = tr("Share among the 18-34 minus share among the 55+", "Part chez les 18-34 ans moins part chez les 55 ans et plus")) +
    theme_qes(mode) +
    theme(axis.line.x = element_blank(), panel.spacing.x = unit(22, "pt"))
}

tab <- merge(agegap, reshape(shares[shares$group %in% c("18-34", "55 and over"), c("study", "level", "group", "pct", "n")],
                             idvar = c("study", "level"), timevar = "group", direction = "wide"),
             by.x = c("study", "party"), by.y = c("study", "level"))
tab <- tab[order(match(tab$party, qz_parties), tab$year, tab$study), ]
tab_age <- setNames(data.frame(
  qz_party_label(tab$party),
  tab$year,
  qz_study_label(tab$study),
  qz_cell(tab[["pct.18-34"]], NA, NA, tab[["n.18-34"]]),
  qz_cell(tab[["pct.55 and over"]], NA, NA, tab[["n.55 and over"]]),
  ifelse(is.na(tab$gap), tr("n < 30", "n < 30"),
         paste0(qz_pts(tab$gap, 1), " [", qz_num(tab$lo), tr(", ", " ; "), qz_num(tab$hi), "]")),
  qz_weight_text(tab$weighted)
, row.names = NULL), c(
  tr("Party", "Parti"),
  tr("Election", "Élection"),
  tr("Study", "Étude"),
  tr("18 to 34, %", "18 à 34 ans, %"),
  tr("55 and over, %", "55 ans et plus, %"),
  tr("Gap, 18-34 minus 55+ [95% CI]", "Écart, 18-34 moins 55+ [IC à 95 %]"),
  tr("Weighting", "Pondération")
))
alt <- sprintf(tr(
  "Six small charts, one per party (PLQ, PQ, ADQ, QS, CAQ, PCQ): the share of the party among voters aged 18 to 34 minus its share among voters aged 55 and over, in points, at each election from 1998 to 2022, with confidence intervals. The PLQ is below zero at every election. In 2022 Québec solidaire is at %s and the CAQ at %s. Values in the table view.",
  "Six petits graphiques, un par parti (PLQ, PQ, ADQ, QS, CAQ, PCQ) : la part du parti chez les 18 à 34 ans moins sa part chez les 55 ans et plus, en points, à chaque élection de 1998 à 2022, avec intervalles de confiance. Le PLQ est sous zéro à chaque élection. En 2022, Québec solidaire est à %s et la CAQ à %s. Valeurs dans la vue en tableau."),
  qz_pts(g(2022, "QS")), qz_pts(g(2022, "CAQ")))
qz_figure(fig_age, "age",
  title = sprintf(tr("In 2022 Québec solidaire won the young and the CAQ the old: age gaps of %s and %s",
                     "En 2022, Québec solidaire a gagné les jeunes et la CAQ les aînés : des écarts d'âge de %s et %s"),
                  qz_pts(g(2022, "QS")), qz_pts(g(2022, "CAQ"))),
  subtitle = tr("Each party's share of the reported vote among the 18 to 34 minus its share among the 55 and over, in points, with 95% confidence intervals",
                "Part du vote déclaré de chaque parti chez les 18 à 34 ans moins sa part chez les 55 ans et plus, en points, avec intervalles de confiance à 95 %"),
  alt = alt, table = tab_age, height = 4.6,
  note = tr("Source: qesR, pooled vote_choice (reported vote) and age_group3; one Quebec Election Study per election and the 1998 polls (francophones only). Above zero: the party does better among the young. Bars: 95% confidence intervals of the difference, from the covariance of the two age groups' estimates. Weighted with each study's post-election weight; hollow (1998, 2008): unweighted, weight under review. The x axis is in years, with a break between 1998 and 2007. The shares of each age group and the Durand panels are in the table view.",
            "Source : qesR, variables regroupées vote_choice (vote déclaré) et age_group3 ; une Étude électorale québécoise par élection et les sondages de 1998 (francophones seulement). Au-dessus de zéro : le parti fait mieux chez les jeunes. Barres : intervalles de confiance à 95 % de la différence, à partir de la covariance des estimations des deux groupes d'âge. Pondéré avec la pondération postélectorale de chaque étude ; creux (1998, 2008) : non pondéré, pondération en révision. L'axe horizontal est en années, avec une coupure entre 1998 et 2007. Les parts de chaque groupe d'âge et les panels Durand sont dans la vue en tableau."))
```

What to notice:

- In 2007 the young voted PQ (`r qz_pct(p("qes2007", "18-34", "PQ"), 0)`) and
  ADQ (`r qz_pct(p("qes2007", "18-34", "ADQ"), 0)`); in 2018 and 2022 their
  first choice was Québec solidaire (`r qz_pct(p("qes2022", "18-34", "QS"), 0)`
  in 2022), which draws little from voters aged 55 and over
  (`r qz_pct(p("qes2022", "55 and over", "QS"), 0)`).
- The CAQ became the party of the old: `r qz_pct(p("qes2022", "55 and over", "CAQ"), 0)`
  of the vote of the 55 and over in 2022, against
  `r qz_pct(p("qes2022", "18-34", "CAQ"), 0)` of the 18 to 34.
- Until 2012 the PQ did as well or better among the young (by
  `r qz_pts(g(1998, "PQ"))` in 1998 and `r qz_pts(g(2007, "PQ"))` in 2007);
  from 2014 on it does better among the old (`r qz_pts(g(2014, "PQ"))` in
  2014). The PLQ does better among the old at every election.

## About the data

- **Studies.** One Quebec Election Study per election (2007, 2008, 2012,
  2014, 2018, 2022) and the 1998 polls, which interviewed francophones
  only. The Durand panels of 2007, 2012 and 2018 are in the table views.
  The CROP polls asked no reported vote.
- **Variable.** `vote_choice` with `types = list(vote_choice = "recall")`:
  the vote reported after the election, in every study. The source item
  of each study is `vote_choice__item` (for example `r item[["qes2018"]]`).
  Respondents who did not vote, spoiled their ballot, did not know or
  refused are `NA` with a reason in `vote_choice__na`, and are left out of
  the denominator.
- **Grades.** The reported-vote question is graded `identical` in 2012,
  `approximate` in the 2007 and 2012 panels, `comparable` elsewhere: the
  lists of parties differ (`qes_spec("crosswalk", targets = "vote_prov_recall")`).
  A party a study did not list, or that did not run, is not drawn.
- **Mother tongue.** `lang_mother`, "French" against every other answer.
  In 2022 it comes from a select-all question (the languages learned in
  childhood and still understood), graded `approximate`: the 146
  respondents who ticked two or more languages (117 of them French and
  English) have no mother tongue there and are left out of the language
  split.
- **Weights.** Each study's recommended post-election weight, through
  `qes_design()`. The 1998 polls, the 2008 study and the 2007 and 2012
  Durand panels are unweighted (weights under review) and drawn hollow;
  the 2018 panel uses its reviewed post-election weight. The recontact of
  the 1998 polls also over-selected undecided voters and refusers.
