---
title: "From two parties to four: Quebec's party system, 1998-2022"
description: >
  Among whom the Quebec vote fragmented from 1998 to 2022, and whether the
  sovereignty question still organizes it.
params:
  lang: "en"
---

*[Version française](fr-realignement.html)*

```{r setup, include = FALSE}
source("_setup.R", encoding = "UTF-8")
dl <- qz_download(qz_studies)
```

::: {.qesr-download}
**This page downloads `r dl$n` studies** (about `r qz_num(dl$mb, 0)` MB) the
first time it runs. `options(qesR.cache = "disk")` keeps them on disk for
later sessions.
:::

```{r harmonize, echo = FALSE}
h <- qes_harmonize(
  studies = c("qes1998", "qes2007", "qes2007_panel", "qes2008", "qes_crop_2007_2010",
              "qes2012", "qes2012_panel", "qes2014", "qes2018", "qes2018_panel", "qes2022"),
  targets = c("vote_choice", "lang_mother", "sov_support"),
  types = list(vote_choice = "recall"),
  missing = "reasons", quiet = TRUE
)
```

```{r estimates, echo = FALSE}
h$party <- factor(qz_party_code(h$vote_choice), levels = qz_parties)
# the 1998 polls interviewed francophones only
h$language <- ifelse(h$study == "qes1998" | h$lang_mother %in% "French", "fr",
                     ifelse(is.na(h$lang_mother), NA, "other"))
# the side a respondent would take in a referendum, and whether their
# vote went to the party of that side (Yes: PQ; No: PLQ)
h$side <- factor(ifelse(h$sov_support %in% "Yes", "yes",
                        ifelse(h$sov_support %in% "No", "no", NA)), levels = c("yes", "no"))
h$aligned <- factor(ifelse(is.na(h$side) | is.na(h$party), NA,
                           ifelse((h$side %in% "yes" & h$party %in% "PQ") |
                                    (h$side %in% "no" & h$party %in% "PLQ"), "aligned", "not")),
                    levels = c("aligned", "not"))
recall_studies <- setdiff(qz_studies, "qes_crop_2007_2010")

shares <- do.call(rbind, lapply(recall_studies, function(s) {
  d <- qz_design(h, s, "weight_post")
  rbind(qz_tag(qz_share(d, "party"), study = s),
        qz_tag(qz_share(d, "party", by = "language"), study = s))
}))
shares$year <- qz_year(shares$study)
shares$status <- qz_party_status(shares$study, shares$level)
shares <- shares[shares$status == "offered", ]

enp <- do.call(rbind, lapply(recall_studies, function(s) {
  d <- qz_design(h, s, "weight_post")
  lang <- d$variables$language
  rbind(qz_tag(qz_enp(d, "party"), study = s, group = "all"),
        qz_tag(qz_enp(d, "party", lang %in% "fr"), study = s, group = "fr"),
        qz_tag(qz_enp(d, "party", lang %in% "other"), study = s, group = "other"),
        make.row.names = FALSE)
}))
enp$year <- qz_year(enp$study)
enp$weighted <- vapply(enp$study, function(s) qz_weighted(h, s, "weight_post"), logical(1))
official <- qz_official()
enp_official <- qz_enp_official(official)

# the francophone vote within each sovereignty camp
sides <- do.call(rbind, lapply(recall_studies, function(s) {
  d <- subset(qz_design(h, s, "weight_post"), language %in% "fr" & !is.na(side))
  rbind(qz_tag(qz_share(d, "party", by = "side"), study = s, what = "party"),
        qz_tag(qz_share(d, "aligned"), study = s, what = "aligned"))
}))
sides$year <- qz_year(sides$study)
sides$status <- ifelse(sides$what == "aligned", "offered", qz_party_status(sides$study, sides$level))
sides <- sides[sides$status == "offered", ]
sov_type <- setNames(qz_first(h, "sov_support__type", recall_studies), recall_studies)

e <- function(s, g) enp$enp[enp$study == s & enp$group == g]
eo <- function(y) enp_official$share_valid[enp_official$year == y]
off <- function(y, party) official$share_valid[official$year == y & official$party == party]
p <- function(s, g, party) shares$pct[shares$study == s & shares$group == g & shares$level == party]
ps <- function(s, side, party) sides$pct[sides$what == "party" & sides$study == s & sides$group == side & sides$level == party]
al <- function(s) sides$pct[sides$what == "aligned" & sides$study == s & sides$level == "aligned"]
late <- qz_primary[as.integer(names(qz_primary)) >= 2012]
qz_word <- function(n) tr(c("one", "two", "three", "four", "five", "six", "seven"),
                          c("un", "deux", "trois", "quatre", "cinq", "six", "sept"))[n]
big22 <- official[official$year == 2022 & official$party != "other", ]
big_cut <- 12
n_big <- sum(big22$share_valid > big_cut)
other_n <- enp$n[enp$group == "other" & enp$study %in% qz_primary & enp$n > 0]
other_range <- function(i) range(enp$enp[enp$study %in% qz_primary[i] & enp$group == "other"])
# the Yes side among francophones who reported a vote and chose a side
yes <- do.call(rbind, lapply(recall_studies, function(s) {
  d <- subset(qz_design(h, s, "weight_post"), language %in% "fr" & !is.na(side) & !is.na(party))
  qz_tag(qz_share(d, "side"), study = s)
}))
yes_pct <- function(s) yes$pct[yes$study == s & yes$level == "yes"]
```

For about three decades, Quebec's party system rested on the sovereignty
question. From the 1970s on, the Parti Québécois (PQ) and the Parti
libéral du Québec (PLQ) alternated in power, and where Quebecers stood on
independence went a long way toward explaining how they voted. In 1998 the
two parties still took `r qz_pct(off(1998, "PQ") + off(1998, "PLQ"), 0)` of
the valid votes. However, after the breakthrough of the Action démocratique
du Québec (ADQ) in 2007, Bélanger and Nadeau (2009) asked whether the
recent change in voters' behaviour was a passing incident or a lasting
transformation of Quebec's political landscape. The rise of Québec
solidaire (QS) and of the Coalition avenir Québec (CAQ) has since settled
part of that question. In 2022 `r qz_word(n_big)` parties each took more
than `r qz_pct(big_cut, 0)` of the vote, and the effective number of parties
(Laakso and Taagepera 1979) rose from `r qz_num(eo(1998))` in 1998 to
`r qz_num(eo(2022))`, the equivalent of `r qz_word(round(eo(2022)))` parties
of equal size. It is now commonplace to say that the system has
fragmented, and the 2018 election, which ended the alternation of the PQ
and the PLQ in power, is the starting point of recent work on the "new"
Quebec voter (Bélanger et al. 2022). The official results tell us how much
the system fragmented. They do not tell us among whom, nor whether the
sovereignty question still organizes the vote.

The Quebec Election Studies give a fairly clear answer to both questions.
First, the fragmentation is concentrated among francophones. From the
first Quebec Election Study, in 2007, to 2022, their effective number of
parties went from `r qz_num(e("qes2007", "fr"))` to
`r qz_num(e("qes2022", "fr"))`. The 1998 polls put it lower, at
`r qz_num(e("qes1998", "fr"))`, but they surveyed francophones only and
have no validated weight, so that point is a rough guide. Non-francophones, by contrast, kept voting
Liberal until 2018: `r qz_pct(p("qes2018", "other", "PLQ"), 0)` of them
supported the PLQ that year, compared to
`r qz_pct(p("qes2018", "fr", "PLQ"), 0)` of francophones. Second, the
sovereignty question still divides the francophone vote, but it no longer
channels it into two parties. Since 2012 sovereigntists have almost never
voted PLQ and federalists have rarely voted PQ, yet the share of
francophones who voted for the party of their own camp fell from
`r qz_pct(al("qes2012"), 0)` in 2012 to `r qz_pct(al("qes2022"), 0)` in 2022.
In other words, knowing where a francophone voter stands on independence
now tells us more about the party they will not choose than about the one
they will. The idea that the sovereignty question has faded from Quebec
elections therefore finds only mixed support.

## The francophone vote fragmented first

```{r fig-enp, echo = FALSE}
d <- enp[enp$study %in% qz_primary & enp$group != "all" & enp$n >= qz_min_n, ]
d$series <- d$group
o <- data.frame(study = NA, group = "official", enp = enp_official$share_valid, lo = NA, hi = NA, n = NA,
                year = enp_official$year, weighted = TRUE, series = "official")
o <- o[o$year %in% as.integer(names(qz_primary)), ]
d <- rbind(d[, names(o)], o)
d$x <- qz_ex(d$year)
d$era <- qz_era(d$year)
d$series <- factor(d$series, levels = c("official", "fr", "other"))
series_lab <- c(official = tr("Official result, all voters", "Résultat officiel, tous les électeurs"),
                fr = tr("Francophones", "Francophones"),
                other = tr("Non-francophones", "Non-francophones"))
d$label <- c(official = tr("Official", "Officiel"), fr = tr("Francophones", "Francophones"),
             other = tr("Non-francophones", "Non-francophones"))[as.character(d$series)]
surveyed <- d[d$series != "official", ]

fig_enp <- function(mode) {
  k <- qz_tokens[[mode]]
  cols <- c(official = k$ink, fr = k$grp[["fr"]], other = k$grp[["other"]])
  ggplot(d, aes(x, enp, colour = series, group = interaction(series, era))) +
    qz_xbreak(mode) +
    qz_band(data = surveyed, aes(ymin = lo, ymax = hi, fill = series), mode = mode) +
    qz_ci(data = surveyed[surveyed$era == "before", ], aes(ymin = lo, ymax = hi), show.legend = FALSE) +
    qz_line(show.legend = FALSE) +
    qz_dots(mode, d, aes(x, enp), colour_aes = "series") +
    qz_end_labels(d, "x", "enp", "label", "series", mode, min_gap = 0.14, nudge_x = 1.1) +
    scale_colour_manual(values = cols, guide = "none") +
    scale_fill_manual(values = cols, breaks = names(series_lab), labels = series_lab,
                      guide = guide_legend(order = 1)) +
    scale_x_election(add = c(0.9, 5.2)) +
    scale_y_continuous(breaks = 1:4) +
    coord_cartesian(ylim = c(1, 4.6), expand = FALSE, clip = "off") +
    labs(x = NULL, y = tr("Effective number of parties", "Nombre effectif de partis")) +
    theme_qes(mode) +
    theme(axis.line.x = element_blank())
}

tab <- enp[enp$group != "all" & enp$n > 0, ]
tab <- tab[order(tab$year, tab$study, tab$group), ]
tab_enp <- setNames(data.frame(
  tab$year,
  qz_study_label(tab$study),
  series_lab[tab$group],
  qz_cell(tab$enp, tab$lo, tab$hi, tab$n, 2),
  tab$n,
  qz_num(enp_official$share_valid[match(tab$year, enp_official$year)], 2),
  qz_weight_text(tab$weighted)
, row.names = NULL), c(
  tr("Election", "Élection"),
  tr("Study", "Étude"),
  tr("Group", "Groupe"),
  tr("Effective number [95% CI]", "Nombre effectif [IC à 95 %]"),
  "n",
  tr("Official, all voters", "Officiel, tous les électeurs"),
  tr("Weighting", "Pondération")
))
alt <- sprintf(tr(
  "Line chart of the effective number of parties at each Quebec election from 1998 to 2022: the official result, francophone respondents and non-francophone respondents. Among francophones it rises from %s in 1998 to %s in 2022; among non-francophones it stays between %s and %s until 2018 and reaches %s in 2022, with a wide interval. Values in the table view.",
  "Graphique linéaire du nombre effectif de partis à chaque élection québécoise de 1998 à 2022 : le résultat officiel, les répondants francophones et les répondants non francophones. Chez les francophones, il passe de %s en 1998 à %s en 2022 ; chez les non-francophones, il reste entre %s et %s jusqu'en 2018 et atteint %s en 2022, avec un intervalle large. Valeurs dans la vue en tableau."),
  qz_num(e("qes1998", "fr")), qz_num(e("qes2022", "fr")),
  qz_num(min(enp$enp[enp$study %in% qz_primary[2:6] & enp$group == "other"])),
  qz_num(max(enp$enp[enp$study %in% qz_primary[2:6] & enp$group == "other"])), qz_num(e("qes2022", "other")))
qz_figure(fig_enp, "enp",
  title = sprintf(tr("The francophone vote fragmented, from %s to %s effective parties; the non-francophone vote stayed far more concentrated",
                     "Le vote francophone s'est fragmenté, de %s à %s partis effectifs ; le vote non francophone est resté beaucoup plus concentré"),
                  qz_num(e("qes2007", "fr")), qz_num(e("qes2022", "fr"))),
  subtitle = tr("Effective number of parties at each election: official result and reported vote by mother tongue, with 95% confidence intervals",
                "Nombre effectif de partis à chaque élection : résultat officiel et vote déclaré selon la langue maternelle, avec intervalles de confiance à 95 %"),
  alt = alt, table = tab_enp, height = 4.0,
  note = tr("Source: Quebec Election Studies (one per election) and, for 1998, the CROP and CREATEC polls, which interviewed francophones only, each by its own definition (mother tongue for CREATEC, the language spoken most often at home for CROP); official results of Élections Québec. The effective number of parties (Laakso and Taagepera 1979) is 1 divided by the sum of the squared vote shares; the smaller parties are grouped into a single category, other parties. Bands: 95% confidence intervals. Hollow dots (1998, 2008): unweighted, no validated weight. The x axis is in years, with a break between 1998 and 2007.",
            "Source : Études électorales québécoises (une par élection) et, pour 1998, les sondages CROP et CREATEC, qui n'ont interrogé que des francophones, chacun selon sa propre définition (la langue maternelle pour CREATEC, la langue parlée le plus souvent à la maison pour CROP) ; résultats officiels d'Élections Québec. Le nombre effectif de partis (Laakso et Taagepera, 1979) est 1 divisé par la somme des carrés des parts de vote ; les petits partis sont regroupés en une seule catégorie, « autres ». Bandes : intervalles de confiance à 95 %. Points creux (1998, 2008) : non pondéré, aucune pondération validée. L'axe horizontal est en années, avec une coupure entre 1998 et 2007."))
```

The two survey lines do not follow the same path. Among francophones the
effective number of parties rose over the period, though not steadily. It
reached a first peak in 2007 (`r qz_num(e("qes2007", "fr"))`), the year of
the ADQ's breakthrough, fell back to `r qz_num(e("qes2012", "fr"))` in 2012,
and has been no lower than
`r qz_num(min(e("qes2014", "fr"), e("qes2018", "fr"), e("qes2022", "fr")))`
since 2014. Among non-francophones it was `r qz_num(e("qes2007", "other"))`
in 2007, when the PLQ took `r qz_pct(p("qes2007", "other", "PLQ"), 0)` of
their vote, and from 2008 to 2018 it stayed between
`r qz_num(other_range(3:6)[1])` and `r qz_num(other_range(3:6)[2])`, a vote
largely dominated by the PLQ. It reached `r qz_num(e("qes2022", "other"))`
in 2022, when the PLQ fell to `r qz_pct(p("qes2022", "other", "PLQ"), 0)` of
their vote. That said, only
`r enp$n[enp$study == "qes2022" & enp$group == "other"]` non-francophone
voters answered that year, and the interval runs from
`r qz_num(enp$lo[enp$study == "qes2022" & enp$group == "other"])` to
`r qz_num(enp$hi[enp$study == "qes2022" & enp$group == "other"])`, wide
enough to include the 2007 value. The rise is clear against the 2008 to
2018 elections; its size is not.

The survey lines run close to the official one, but not on it. Surveys
over-report some parties and under-report others (see
[Survey vs official results](survey-vs-official.html)). In 2022, for
instance, `r qz_pct(p("qes2022", "all", "CAQ"), 0)` of respondents reported a
vote for the CAQ, against `r qz_pct(off(2022, "CAQ"), 0)` of the ballots,
so the survey makes the 2022 vote look more fragmented than it was
(`r qz_num(e("qes2022", "all"))` effective parties among all respondents,
`r qz_num(eo(2022))` officially). The survey lines therefore say more about
the gap between the two groups, and how it changed, than about the level
of either.

## Party by party, the language divide held

```{r fig-lang, echo = FALSE}
panel_parties <- c("PLQ", "PQ", "QS", "ADQ", "CAQ")
d <- shares[shares$study %in% qz_primary & shares$group %in% c("fr", "other") &
              shares$level %in% panel_parties & shares$n >= qz_min_n, ]
d$x <- qz_ex(d$year)
d$era <- qz_era(d$year)
d$party <- factor(d$level, levels = panel_parties)
d$series <- d$group
o <- official[official$party %in% panel_parties & official$year %in% as.integer(names(qz_primary)), ]
o$x <- qz_ex(o$year)
o$party <- factor(o$party, levels = panel_parties)
group_lab <- c(fr = tr("Francophones", "Francophones"), other = tr("Non-francophones", "Non-francophones"))

fig_lang <- function(mode) {
  k <- qz_tokens[[mode]]
  cols <- c(fr = k$grp[["fr"]], other = k$grp[["other"]])
  ggplot(d, aes(x, pct, colour = series, group = interaction(series, era))) +
    qz_xbreak(mode) +
    geom_segment(data = o, aes(x = x - 0.45, xend = x + 0.45, y = share_valid, yend = share_valid, linetype = "official"),
                 colour = k$ink, linewidth = 0.7, inherit.aes = FALSE) +
    qz_ci(aes(ymin = lo, ymax = hi), show.legend = FALSE) +
    qz_line(show.legend = FALSE) +
    qz_dots(mode, d, aes(x, pct), colour_aes = "series", size = 2.4, wrap = TRUE) +
    facet_wrap(~party, ncol = 3, axes = "all_x", labeller = as_labeller(function(x) qz_party_label(x))) +
    scale_colour_manual(values = cols, guide = "none") +
    scale_fill_manual(values = cols, breaks = names(group_lab), labels = group_lab,
                      guide = guide_legend(order = 1)) +
    scale_linetype_manual(values = c(official = "solid"),
                          labels = c(official = tr("Official result,\nall voters", "Résultat officiel,\ntous les électeurs")),
                          guide = guide_legend(order = 2, override.aes = list(colour = k$ink, linewidth = 0.7))) +
    scale_x_election(short = TRUE, add = c(1.4, 1.4)) +
    scale_y_continuous(breaks = seq(0, 80, 20), labels = qz_pct_axis) +
    coord_cartesian(ylim = c(0, 92), expand = FALSE, clip = "off") +
    labs(x = NULL, y = NULL) +
    theme_qes(mode) +
    theme(axis.line.x = element_blank(), panel.spacing.x = unit(22, "pt"),
          legend.position = "inside", legend.position.inside = c(0.695, 0.25),
          legend.justification.inside = c(0, 0.5), legend.direction = "vertical",
          legend.box = "vertical", legend.box.just = "left", legend.spacing.y = unit(4, "pt"),
          legend.key.spacing.y = unit(4, "pt"))
}

tab <- shares[shares$group %in% c("fr", "other") & shares$level %in% c(panel_parties, "PCQ", "other") & shares$n > 0, ]
tab <- tab[order(match(tab$level, qz_parties), tab$year, tab$study, tab$group), ]
tab_lang <- setNames(data.frame(
  qz_party_label(tab$level),
  tab$year,
  qz_study_label(tab$study),
  group_lab[tab$group],
  qz_cell(tab$pct, tab$lo, tab$hi, tab$n),
  tab$n,
  qz_num(official$share_valid[match(paste(tab$year, tab$level), paste(official$year, official$party))]),
  qz_weight_text(tab$weighted)
, row.names = NULL), c(
  tr("Party", "Parti"),
  tr("Election", "Élection"),
  tr("Study", "Étude"),
  tr("Group", "Groupe"),
  tr("Share of reported vote [95% CI]", "Part du vote déclaré [IC à 95 %]"),
  "n",
  tr("Official, all voters", "Officiel, tous les électeurs"),
  tr("Weighting", "Pondération")
))
alt <- sprintf(tr(
  "Five small line charts, one per party (PLQ, PQ, QS, ADQ, CAQ), of the share of the reported vote among francophones and among non-francophones at each election from 1998 to 2022, with the official result as a tick. The PLQ takes %s of the non-francophone vote in 2018 against %s of the francophone vote; the CAQ takes %s of francophones and %s of the others. Values in the table view.",
  "Cinq petits graphiques linéaires, un par parti (PLQ, PQ, QS, ADQ, CAQ), de la part du vote déclaré chez les francophones et chez les non-francophones à chaque élection de 1998 à 2022, avec le résultat officiel en trait. Le PLQ obtient %s du vote non francophone en 2018 contre %s du vote francophone ; la CAQ obtient %s des francophones et %s des autres. Valeurs dans la vue en tableau."),
  qz_pct(p("qes2018", "other", "PLQ"), 0), qz_pct(p("qes2018", "fr", "PLQ"), 0),
  qz_pct(p("qes2018", "fr", "CAQ"), 0), qz_pct(p("qes2018", "other", "CAQ"), 0))
qz_figure(fig_lang, "lang",
  title = sprintf(tr("The CAQ's wins are francophone wins: in 2018 it took %s of francophones and %s of other voters",
                     "Les victoires de la CAQ sont francophones : en 2018, elle a obtenu %s des francophones et %s des autres électeurs"),
                  qz_pct(p("qes2018", "fr", "CAQ"), 0), qz_pct(p("qes2018", "other", "CAQ"), 0)),
  subtitle = tr("Share of the reported vote among francophones and non-francophones, by party, with 95% confidence intervals and the official result",
                "Part du vote déclaré chez les francophones et les non-francophones, selon le parti, avec intervalles de confiance à 95 % et résultat officiel"),
  alt = alt, table = tab_lang, height = 5.2,
  note = tr("Source: Quebec Election Studies (one per election) and, for 1998, the CROP and CREATEC polls (francophones only). Tick: official share of valid votes, all voters. Bars: 95% confidence intervals. Hollow dots (1998, 2008): unweighted, no validated weight. Groups of fewer than 30 respondents are not drawn. The ADQ merged into the CAQ in 2012; the two are kept apart here. The PCQ and the smaller parties are in the table view.",
            "Source : Études électorales québécoises (une par élection) et, pour 1998, les sondages CROP et CREATEC (francophones seulement). Trait : part officielle des votes valides, tous les électeurs. Barres : intervalles de confiance à 95 %. Points creux (1998, 2008) : non pondéré, aucune pondération validée. Les groupes de moins de 30 répondants ne sont pas tracés. L'ADQ a fusionné avec la CAQ en 2012 ; les deux restent distincts ici. Le PCQ et les petits partis sont dans la vue en tableau."))
```

The PLQ remained the party of non-francophones. It took
`r qz_pct(p("qes2007", "other", "PLQ"), 0)` of their vote in 2007,
`r qz_pct(p("qes2014", "other", "PLQ"), 0)` in 2014 and still
`r qz_pct(p("qes2022", "other", "PLQ"), 0)` in 2022, its lowest share over
the period. Among francophones, however, its share fell from
`r qz_pct(p("qes1998", "fr", "PLQ"), 0)` in 1998 to
`r qz_pct(p("qes2022", "fr", "PLQ"), 0)` in 2022, well below its official
result of `r qz_pct(off(2022, "PLQ"), 0)` that year. The PQ never had many
non-francophone voters (`r qz_pct(p("qes2007", "other", "PQ"), 0)` in 2007)
and kept almost none (`r qz_pct(p("qes2022", "other", "PQ"), 0)` in 2022).

The CAQ's victories, for their part, are francophone victories. In 2018 it
won `r qz_pct(p("qes2018", "fr", "CAQ"), 0)` of the francophone vote and
`r qz_pct(p("qes2018", "other", "CAQ"), 0)` of the others, and in 2022
`r qz_pct(p("qes2022", "fr", "CAQ"), 0)` and
`r qz_pct(p("qes2022", "other", "CAQ"), 0)`. Like the ADQ before it, it
barely reaches non-francophones. The two parties that came closest to
crossing the language divide in 2022 are the two whose breakthrough is
most recent. Québec solidaire, which grew among francophones first
(`r qz_pct(p("qes2018", "fr", "QS"), 0)` in 2018), took
`r qz_pct(p("qes2022", "fr", "QS"), 0)` of francophones and
`r qz_pct(p("qes2022", "other", "QS"), 0)` of the others; the Parti
conservateur du Québec (PCQ) took `r qz_pct(p("qes2022", "fr", "PCQ"), 0)` and
`r qz_pct(p("qes2022", "other", "PCQ"), 0)`. With so few non-francophone
respondents these gaps are uncertain. Using the same 2022 study, Bélanger
et al. (2025) find that, unlike most other parties, the PCQ drew as much
from non-francophones as from francophones once other factors are held
constant; the raw shares shown here do not hold those factors constant.
The new parties thus divided the francophone vote first. Among
non-francophones the break came only in 2022, and it rests on
`r enp$n[enp$study == "qes2022" & enp$group == "other"]` respondents.

## Sovereignty still divides the francophone vote, but no longer channels it

```{r fig-sov, echo = FALSE}
side_lab <- c(yes = tr("Would vote Yes (sovereigntists)", "Voteraient Oui (souverainistes)"),
              no = tr("Would vote No (federalists)", "Voteraient Non (fédéralistes)"))
type_lab <- c(partnership_1995_push = tr("The 1995 question (partnership)", "La question de 1995 (partenariat)"),
              independence = tr("An independent country", "Un pays indépendant"),
              sovereign_country = tr("A sovereign country", "Un pays souverain"),
              favour = tr("Favourable to independence", "Favorable à l'indépendance"))
d <- sides[sides$what == "party" & sides$study %in% qz_primary & sides$n >= qz_min_n & sides$pct > 0, ]
d$level <- factor(d$level, levels = qz_parties)
d$side <- factor(d$group, levels = c("yes", "no"))
d <- d[order(d$study, d$side, d$level), ]
d$xmax <- ave(d$pct, d$study, d$side, FUN = cumsum)
d$xmin <- d$xmax - d$pct
# one row per election, top to bottom, with a gap where the question changes
row_y <- setNames(seq_along(qz_primary) + ifelse(as.integer(names(qz_primary)) >= 2012, 0.8, 0), qz_primary)
d$y <- -row_y[d$study]
d$label <- ifelse(d$pct >= 10, qz_num(d$pct, 0), "")
change <- data.frame(side = factor("yes", levels = c("yes", "no")), x = 0,
                     y = -mean(row_y[c("qes2008", "qes2012")]),
                     label = tr("From 2012: the question on an independent country",
                                "Dès 2012 : la question sur un pays indépendant"))

fig_sov <- function(mode) {
  k <- qz_tokens[[mode]]
  d$ink <- qz_label_ink(k$party[as.character(d$level)], mode)
  half <- 0.36
  ggplot(d) +
    geom_rect(aes(xmin = xmin, xmax = xmax, ymin = y - half, ymax = y + half, fill = level),
              colour = k$surface, linewidth = 0.6) +
    geom_text(aes(x = (xmin + xmax) / 2, y = y, label = label, colour = ink), size = 2.8, family = qz_font()) +
    geom_text(data = change, aes(x, y, label = label), hjust = 0, size = 2.8, colour = k$ink2,
              family = qz_font(), fontface = "italic") +
    facet_wrap(~side, ncol = 2, labeller = as_labeller(side_lab)) +
    scale_fill_manual(values = k$party, breaks = qz_parties, labels = qz_party_label(qz_parties),
                      guide = guide_legend(nrow = 1)) +
    scale_colour_identity() +
    scale_y_continuous(breaks = -row_y, labels = names(qz_primary), expand = expansion(add = 0.5)) +
    scale_x_continuous(limits = c(0, 100.5), breaks = c(0, 50, 100), labels = qz_pct_axis, expand = expansion(0)) +
    coord_cartesian(clip = "off") +
    labs(x = tr("Share of the reported vote", "Part du vote déclaré"), y = NULL) +
    theme_qes(mode, grid = "none") +
    theme(legend.key.width = unit(10, "pt"), legend.key.height = unit(10, "pt"),
          panel.spacing.x = unit(36, "pt"))
}

tab <- sides[sides$what == "party" & sides$n > 0, ]
tab <- tab[order(tab$year, tab$study, match(tab$group, c("yes", "no")), match(tab$level, qz_parties)), ]
tab_sov <- setNames(data.frame(
  tab$year,
  qz_study_label(tab$study),
  unname(type_lab[sov_type[tab$study]]),
  unname(side_lab[tab$group]),
  qz_party_label(tab$level),
  qz_cell(tab$pct, tab$lo, tab$hi, tab$n),
  tab$n,
  qz_weight_text(tab$weighted)
, row.names = NULL), c(
  tr("Election", "Élection"),
  tr("Study", "Étude"),
  tr("Question", "Question"),
  tr("Side", "Camp"),
  tr("Party", "Parti"),
  tr("Share of reported vote [95% CI]", "Part du vote déclaré [IC à 95 %]"),
  "n",
  tr("Weighting", "Pondération")
))
alt <- sprintf(tr(
  "Two panels of stacked horizontal bars, one bar per election from 1998 to 2022: how francophones who would vote Yes and those who would vote No in a referendum split their reported vote among the parties. Among those who would vote Yes, the PQ takes %s in 2012 and %s in 2022; among those who would vote No, the PLQ takes %s in 2012 and %s in 2022, and the CAQ %s in 2022. Values in the table view.",
  "Deux panneaux de barres horizontales empilées, une barre par élection de 1998 à 2022 : comment les francophones qui voteraient Oui et ceux qui voteraient Non à un référendum répartissent leur vote déclaré entre les partis. Chez ceux qui voteraient Oui, le PQ obtient %s en 2012 et %s en 2022 ; chez ceux qui voteraient Non, le PLQ obtient %s en 2012 et %s en 2022, et la CAQ %s en 2022. Valeurs dans la vue en tableau."),
  qz_pct(ps("qes2012", "yes", "PQ"), 0), qz_pct(ps("qes2022", "yes", "PQ"), 0),
  qz_pct(ps("qes2012", "no", "PLQ"), 0), qz_pct(ps("qes2022", "no", "PLQ"), 0),
  qz_pct(ps("qes2022", "no", "CAQ"), 0))
qz_figure(fig_sov, "sov",
  title = sprintf(tr("Each camp's old party lost its hold: from 2012 to 2022 the PQ fell from %s to %s of francophone sovereigntists",
                     "Chaque camp a délaissé son parti traditionnel : de 2012 à 2022, le PQ est passé de %s à %s chez les souverainistes francophones"),
                  qz_pct(ps("qes2012", "yes", "PQ"), 0), qz_pct(ps("qes2022", "yes", "PQ"), 0)),
  subtitle = sprintf(tr("Reported vote of francophones, by the side they would take in a referendum on sovereignty. Among federalists, the PLQ went from %s in 2012 to %s in 2022",
                        "Vote déclaré des francophones, selon le camp qu'ils prendraient à un référendum sur la souveraineté. Chez les fédéralistes, le PLQ est passé de %s en 2012 à %s en 2022"),
                     qz_pct(ps("qes2012", "no", "PLQ"), 0), qz_pct(ps("qes2022", "no", "PLQ"), 0)),
  alt = alt, table = tab_sov, height = 3.9,
  note = sprintf(tr("Source: Quebec Election Studies (one per election) and, for 1998, the CROP poll; francophone respondents who reported a vote and chose a side. From 1998 to 2008 the question is the one of the 1995 referendum, which offered a partnership with Canada, with the undecided asked which way they lean; from 2012 on, it asks about an independent country. The Yes side is about as large among these francophones in 2008 (%s) as in 2012 (%s), but the two questions may not draw the same voters, so compare the rows within each period. Labels show shares of 10%% or more; the 95%% confidence intervals and the Durand panels are in the table view. 1998 and 2008: unweighted, no validated weight.",
                    "Source : Études électorales québécoises (une par élection) et, pour 1998, le sondage CROP ; répondants francophones qui ont déclaré un vote et choisi un camp. De 1998 à 2008, la question est celle du référendum de 1995, qui proposait un partenariat avec le Canada, et les indécis sont relancés ; à partir de 2012, elle porte sur un pays indépendant. Le camp du Oui est à peu près aussi grand chez ces francophones en 2008 (%s) qu'en 2012 (%s), mais les deux questions n'attirent pas forcément les mêmes électeurs : il faut comparer les rangées à l'intérieur de chaque période. Les étiquettes indiquent les parts de 10 %% ou plus ; les intervalles de confiance à 95 %% et les panels Durand sont dans la vue en tableau. 1998 et 2008 : non pondéré, aucune pondération validée."),
                 qz_pct(yes_pct("qes2008"), 0), qz_pct(yes_pct("qes2012"), 0)))
```

Two things stand out. First, the wall between the two old camps holds.
From 2012 on, the PLQ never took more than
`r qz_pct(max(vapply(late, ps, numeric(1), side = "yes", party = "PLQ")), 0)`
of the francophone sovereigntist vote, and the PQ never more than
`r qz_pct(max(vapply(late, ps, numeric(1), side = "no", party = "PQ")), 0)`
of the federalist vote. Second, within each camp the old party lost its
hold, and not at the same pace. Federalist francophones were the first to
leave, though not in a straight line. In 2007 the ADQ already took about as
many of them as the PLQ (`r qz_pct(ps("qes2007", "no", "ADQ"), 0)` against
`r qz_pct(ps("qes2007", "no", "PLQ"), 0)`), and after a brief return to the
PLQ in 2008 (`r qz_pct(ps("qes2008", "no", "PLQ"), 0)`), the CAQ took
`r qz_pct(ps("qes2012", "no", "CAQ"), 0)` of them in 2012. From 2012 to
2022 the PLQ's share of federalist francophones fell from
`r qz_pct(ps("qes2012", "no", "PLQ"), 0)` to
`r qz_pct(ps("qes2022", "no", "PLQ"), 0)`, behind the CAQ, the PCQ and even
QS. Sovereigntist francophones stayed with the PQ longer. Its share of
their vote went from `r qz_pct(ps("qes2012", "yes", "PQ"), 0)` in 2012 to
`r qz_pct(ps("qes2022", "yes", "PQ"), 0)` in 2022, and most of that loss
came in 2018, when Québec solidaire and the CAQ divided them. As a result,
the share of francophones who voted for the party of their camp fell from
`r qz_pct(al("qes2012"), 0)` in 2012 to `r qz_pct(al("qes2022"), 0)` in 2022
(the table view gives every election).

The party that gained most is the one that sits on neither side of the
question. In 2022 the CAQ drew `r qz_pct(ps("qes2022", "no", "CAQ"), 0)` of
the federalist francophone vote and
`r qz_pct(ps("qes2022", "yes", "CAQ"), 0)` of the sovereigntist one. The
sovereignty question has therefore not disappeared from Quebec elections;
what disappeared is the two-party system that translated it into votes.
This echoes Daoust and Gareau-Paquette (2024), who find that in federal
elections support for independence remains a key predictor of the vote,
its weight stable since 2006 after a fall from the 2000 and 2004
elections. Here, by contrast, what changed is how that division maps onto
parties.

## What this means

Overall, the Quebec party system did not fragment evenly. The new parties
divided the francophone vote, and they did so within each sovereignty
camp: the wall between the camps held while the PQ and the PLQ lost their
hold on their own side. Two limits deserve mention. First, these are reported votes in
separate surveys: they show how groups of voters split at each election,
not whether the same people changed parties (the
[Transitions](transitions.html) page follows the same respondents through
a campaign). Second, the non-francophone samples are small, between
`r min(other_n)` and `r max(other_n)` voters per election, so their
estimates move a lot from one election to the next, and the 2022 break
among them remains to be confirmed. The next question is whether the
francophone vote will settle again around new poles, or whether
generational replacement will keep it divided (the
[sovereignty page](sovereignty-generations.html) follows the birth cohorts).

## How the figures are made

Every figure on this page uses one harmonized variable, `vote_choice`,
restricted to the vote reported after the election, with mother tongue
and the referendum question:

```{r show-harmonize, ref.label = "harmonize", eval = FALSE}
```

Within one study, a weighted share and its confidence interval take one
`qes_design()` and one call to the survey package:

```{r example}
library(survey)
d18 <- qes_design(h[h$study == "qes2018", ], weight = "weight_post")
d18 <- subset(d18, !is.na(vote_choice) & !is.na(lang_mother))
svyby(~I(vote_choice == "CAQ"), ~lang_mother, d18, svyciprop,
      method = "logit", vartype = "ci")
```

The figures repeat this for every study. Each estimate is made within one
study and never pooled across studies.

## Going further

- Bélanger, Éric, and Richard Nadeau. 2009. *Le comportement électoral des
  Québécois*. Montréal: Presses de l'Université de Montréal.
  <https://doi.org/10.4000/books.pum.9772>
- Bélanger, Éric, Jean-François Daoust, Valérie-Anne Mahéo, and Richard
  Nadeau. 2022. *Le nouvel électeur québécois*. Montréal: Presses de
  l'Université de Montréal. <https://doi.org/10.1515/9782760645929>
- Bélanger, Éric, Philippe Mongrain, Thomas Gareau-Paquette, and
  Valérie-Anne Mahéo. 2025. "A Party that Went Viral? The Drivers of
  Support for the Parti Conservateur du Québec in the 2022 Election."
  *Canadian Journal of Political Science* 58 (2): 277-296.
  <https://doi.org/10.1017/S0008423924000829>
- Daoust, Jean-François, and Thomas Gareau-Paquette. 2024. "Is Quebec
  Independence Still Key in Making Sense of Canadian Elections? A
  Longitudinal Analysis (2000-2021)." *Regional & Federal Studies* 34 (5):
  781-806. <https://doi.org/10.1080/13597566.2023.2233422>
- Laakso, Markku, and Rein Taagepera. 1979. "'Effective' Number of
  Parties: A Measure with Application to West Europe." *Comparative
  Political Studies* 12 (1): 3-27.
  <https://doi.org/10.1177/001041407901200101>
- Mahéo, Valérie-Anne, and Éric Bélanger. 2018. "Is the Parti Québécois
  Bound to Disappear? A Study of the Current Generational Dynamics of
  Electoral Behaviour in Quebec." *Canadian Journal of Political Science*
  51 (2): 335-356. <https://doi.org/10.1017/S0008423917001147>

## About the data

- **Studies.** One Quebec Election Study per election (2007, 2008, 2012,
  2014, 2018, 2022) and, for 1998, the CROP and CREATEC polls, which
  surveyed francophones only (CREATEC by mother tongue, CROP by the
  language of use); qesR has no mother-tongue column for 1998, and this
  page counts every 1998 respondent as francophone. The table views also include the Durand
  panel surveys of 2007, 2012 and 2018, which interviewed the same
  respondents during the campaign and after the election; they are not
  drawn. The CROP polls of 2007 to 2010 asked no reported vote and are not
  used.
- **Vote.** The party respondents said they voted for, asked after the
  election. Those who did not vote, spoiled their ballot, did not know or
  refused are left out. A party that a study did not list, or that did not
  run, is not drawn. The ADQ and the CAQ are kept apart, although they
  merged in 2012.
- **Language.** Mother tongue, French against every other answer. In 2022
  the question asked for all the languages learned in childhood, and the
  `r sum(is.na(h$language[h$study == "qes2022"]))` respondents who named more
  than one are left out of the language split, as are
  `r sum(is.na(h$language[h$study == "qes2014"]))` respondents in 2014 whose
  answer could not be classed.
- **Sovereignty.** The side a respondent would take in a referendum, among
  those who chose one. From 1998 to 2008 the studies asked the 1995
  referendum question, and those who hesitated were asked which way they
  leaned; from 2012 on, the question is about an independent country. In
  1998 only the CROP poll asked it. The Durand panels in the table views
  use their own wordings (see the Question column).
- **Weights.** Each study's post-election weight, where it has a validated
  one (the [weights table](studies.html#weights) gives each study's); the
  others, 1998 and 2008 among the drawn studies, are unweighted and drawn
  as hollow dots. The 1998 respondents had
  first answered pre-election polls and were interviewed again after the
  election, and that second round over-selected undecided voters and
  those who would not say how they would vote.
