---
title: "Recipes: the language divide in the Liberal vote, step by step"
description: >
  Copyable code for the pooled harmonized variables, worked through one
  question: did the language divide in the Liberal vote narrow as the
  sovereignty question receded?
params:
  lang: "en"
---

*[Version française](fr-recettes.html)*

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
options(qesR.lang = params$lang)
# the site's helpers (numbers in the page's language, the chart system,
# the official results); the recipes themselves do not need them
source("_setup.R", encoding = "UTF-8")
```

```{r early, include = FALSE, ref.label = c("packages", "pooled", "recall", "loop", "composition", "pool")}
# runs the recipes once before the text, so that the opening paragraph can
# quote their results; each recipe runs again, with its output, below
```

```{r numbers, include = FALSE}
gap_of <- function(y) plq$pct[plq$year == y & plq$group == "other"] - plq$pct[plq$year == y & plq$group == "francophone"]
fr_of <- function(y) plq$pct[plq$year == y & plq$group == "francophone"]
other_of <- function(y) plq$pct[plq$year == y & plq$group == "other"]
comp_of <- function(y) comp$pct[comp$year == y]
first <- min(plq$year)
last <- max(plq$year)
b1 <- 100 * coef(m1)[["francophoneother"]]
b2 <- 100 * coef(m2)[["francophoneother"]]
ci1 <- 100 * confint(m1)["francophoneother", ]
p_equal <- as.numeric(regTermTest(m3, ~ francophone:study)$p)
n_four <- if (identical(params$lang, "fr")) c("une", "deux", "trois", "quatre", "cinq", "six")[length(four)] else
  xfun::numbers_to_words(length(four))
stopifnot(p_equal > 0.05)  # "the evidence that the gap changed is weak"

# the weighted share of non-francophones among the voters of each study, and
# the share of non-francophones among Liberal voters had it stayed at its
# first level (each group's Liberal vote as estimated in that study)
voters_other <- setNames(sapply(qes, function(s) {
  ds <- subset(study_design(s), !is.na(vote_choice) & !is.na(francophone))
  100 * as.numeric(coef(svymean(~as.numeric(francophone == "other"), ds)))
}), plq$year[match(qes, plq$study)])
comp_fixed <- function(y) {
  s <- voters_other[[as.character(first)]] / 100
  100 * s * other_of(y) / (s * other_of(y) + (1 - s) * fr_of(y))
}
low_share <- voters_other[c("2007", "2014")]
high_share <- voters_other[c("2012", "2018", "2022")]
stopifnot(abs(diff(low_share)) < 1, min(high_share) > max(low_share) + 4,
          comp_fixed(last) > 50, comp_fixed(last) < comp_of(last))

# French typography in the figure strings (the prose has it already): a
# no-break space before ":" and "%", a narrow one before ";"
typo <- function(x) if (identical(params$lang, "fr")) gsub(" ;", " ;", gsub(" ([:%])", " \\1", x)) else x
typo_table <- function(d) {
  names(d) <- typo(names(d))
  d[] <- lapply(d, function(v) if (is.character(v)) typo(v) else v)
  d
}
# the weighting column of the table views, and the weighting key of the charts
weight_text <- function(w) ifelse(w, tr("weighted", "pondéré"), tr("unweighted", "non pondéré"))
weight_key <- function(mode) {
  s <- scale_weight_key(mode)
  s$labels <- tr("hollow point: unweighted", "point vide : non pondéré")
  s
}
```

Since the first provincial election studies, the Quebec vote has been read
through language and the national question (Hamilton and Pinard 1976;
Pinard and Hamilton 1978; Bélanger et al. 2018). The Parti libéral du
Québec (PLQ) wins most of the anglophone and allophone vote, and its
fortunes among francophones decide elections. A common reading holds that
this language divide is the national question under another name:
non-francophones vote Liberal because they are federalists. If that reading
is right, the divide should have narrowed after 2012, as the Coalition
avenir Québec (CAQ), a nationalist party that sets independence aside,
courted francophone federalists and the sovereignty question was said to
lose ground. That decline is itself less certain than it seems: in federal
elections, the weight of support for independence in the vote held
remarkably steady from 2006 to 2021 (Daoust and Gareau-Paquette 2024).

Overall, the Quebec Election Studies give this reading only limited
support. The gap in the Liberal vote between non-francophones and
francophones was about `r qz_num(gap_of(first), 0)` percentage points in
`r first` and `r qz_num(gap_of(2012), 0)` in 2012; it widened in 2014 and
2018, and it was still `r qz_num(gap_of(last), 0)` points in `r last`.
Support for independence accounts for part of it, but only part: in the
`r n_four` studies from 2012 to 2022, once support for independence is held
constant, the gap falls from `r qz_num(b1, 0)` to `r qz_num(b2, 0)` points.
What did change is the Liberal electorate itself. Francophones left it
(`r qz_pct(fr_of(first), 0)` of them voted PLQ in `r first`,
`r qz_pct(fr_of(last), 0)` in `r last`), so that non-francophones went from
`r qz_pct(comp_of(first), 0)` of Liberal voters to
`r qz_pct(comp_of(last), 0)` (`r qz_pct(comp_fixed(last), 0)` had their
share of the voters in each sample stayed at its `r first` level).

The recipes below build that answer step by step; each runs as is.

```{r packages}
library(qesR)
library(survey)
library(ggplot2)
```

## 1. One vote-choice variable for every study

`vote_choice` pools three questions into one column: the vote reported
after the election, the intention with the undecided pushed toward a party,
and the plain intention. `vote_choice__type` says which question each row
answers.

```{r pooled}
studies <- c("qes1998", "qes2007", "qes2007_panel", "qes2008", "qes_crop_2007_2010",
             "qes2012", "qes2012_panel", "qes2014", "qes2018", "qes2018_panel", "qes2022")
h <- qes_harmonize(studies = studies, targets = c("vote_choice", "lang_mother"),
                   layout = "long", quiet = TRUE)
table(h$study, h$vote_choice__type)
```

In this long format, a study with a wave before and a wave after the
election has one row per wave, so the same respondent appears once with an
intention and once with a reported vote. The CROP polls, fielded between
elections, only have intentions.

## 2. Keep the reported vote only

Setting an intention in one study against a reported vote in another would
compare two different questions. For the language divide we keep the
reported vote, which every Quebec Election Study asked after the election,
and read each study in the default format, one row per respondent. `types`
keeps only some of the questions a pooled variable brings together:

```{r recall}
r <- qes_harmonize(studies = studies, targets = c("vote_choice", "lang_mother", "sov_support"),
                   types = list(vote_choice = "recall"), quiet = TRUE)
table(r$study, r$vote_choice__type, useNA = "ifany")
```

`vote_choice` is now missing for the CROP polls, and for respondents not
reached after the vote. When a study asked several of these questions and
no type is set, the reported vote comes first, then the pushed intention,
then the plain one (`qes_spec("pooled")` lists that order).

## 3. Where each value comes from

```{r sources}
prov <- qes_provenance(r, level = "pooled")
prov[prov$pooled == "vote_choice", c("study", "type", "wave", "grade", "n_value")]
```

```{r sources-text, include = FALSE}
approx <- prov$study[prov$pooled == "vote_choice" & prov$grade %in% "approximate"]
```

Each study's question has a grade. It is `identical` when the question, the
options and the format are the same, `comparable` when the differences
(the order of the options, whether "don't know" is offered) should not move
the shares, and `approximate` when a format or a filter may move them. Only
`r knitr::combine_words(paste("the", qz_study_label(approx)))` are approximate here.
The charts below therefore leave the panels aside and keep one Quebec
Election Study per election.

## 4. A weighted share with its confidence interval, in one study

```{r one-study}
d <- qes_design(r[r$study == "qes2018", ], weight = "weight_post")
d <- subset(d, !is.na(vote_choice))
caq18 <- svyciprop(~I(vote_choice == "CAQ"), d, method = "logit")
caq18
```

```{r one-study-text, include = FALSE}
off <- qz_official()
caq18_ci <- 100 * attr(caq18, "ci")
caq18_off <- off$share_valid[off$year == 2018 & off$party == "CAQ"]
caq18_in <- caq18_off >= caq18_ci[[1]] & caq18_off <= caq18_ci[[2]]
```

The CAQ received `r qz_pct(caq18_off)` of the valid votes in 2018, and
`r qz_pct(100 * as.numeric(caq18))` of respondents to the 2018 study
reported voting for it (95% confidence interval, `r qz_num(caq18_ci[[1]])`
to `r qz_num(caq18_ci[[2]])`). The official result thus falls
`r if (caq18_in) "inside" else "outside"` that interval.

## 5. By group

```{r by-group}
plq18 <- svyby(~I(vote_choice == "PLQ"), ~lang_mother, subset(d, !is.na(lang_mother)),
               svyciprop, method = "logit", vartype = "ci")
plq18
```

```{r by-group-text, include = FALSE}
plq18_pct <- setNames(100 * plq18[[2]], plq18$lang_mother)
n_allo18 <- sum(d$variables$lang_mother %in% "Other")
```

In 2018, `r qz_pct(plq18_pct[["English"]], 0)` of anglophones and
`r qz_pct(plq18_pct[["Other"]], 0)` of allophones voted Liberal, against
`r qz_pct(plq18_pct[["French"]], 0)` of francophones. The allophone
estimate rests on `r n_allo18` respondents, hence its wider interval. The
next recipes group anglophones and allophones as non-francophones, since
the allophone samples are too small to follow on their own from one
election to the next.

## 6. Every study, one at a time

Studies differ in population, mode, wording and weighting, so each estimate
is made within one study before any comparison. The 1998 study does not
measure mother tongue, so the series starts in 2007. The weight of the 2008
study is still under review: its estimates are unweighted, and the charts
draw them as hollow points.

```{r loop}
qes <- c("qes2007", "qes2008", "qes2012", "qes2014", "qes2018", "qes2022")
r$francophone <- factor(ifelse(r$lang_mother == "French", "francophone", "other"))

# one design per study; a study without a post-election weight (2008) gets
# a unit weight, so its estimates are unweighted
is_weighted <- function(s, data = r) !all(is.na(data$weight_post[data$study == s]))
study_design <- function(s, data = r) {
  x <- data[data$study == s, ]
  if (!is_weighted(s, data)) x$weight_post <- 1
  qes_design(x, weight = "weight_post")
}

plq <- do.call(rbind, lapply(qes, function(s) {
  ds <- subset(study_design(s), !is.na(vote_choice) & !is.na(francophone))
  est <- svyby(~I(vote_choice == "PLQ"), ~francophone, ds, svyciprop, method = "logit", vartype = "ci")
  data.frame(study = s, year = as.integer(substr(ds$variables$year[1], 1, 4)),
             weighted = is_weighted(s), group = est$francophone,
             pct = 100 * est[[2]], lo = 100 * est$ci_l, hi = 100 * est$ci_u)
}))
print(plq, digits = 3, row.names = FALSE)
```

```{r loop-site, echo = FALSE}
# the gap and its interval (from the covariance of the two estimates) and
# the sample sizes, for the figure's table view
n_lang <- with(r[!is.na(r$vote_choice) & r$study %in% qes, ], table(study, francophone))
gaps <- do.call(rbind, lapply(qes, function(s) {
  ds <- subset(study_design(s), !is.na(vote_choice) & !is.na(francophone))
  qz_tag(qz_gap(ds, "vote_choice", "PLQ", "francophone", "other", "francophone"),
         study = s, year = plq$year[plq$study == s][1], weighted = is_weighted(s))
}))
gap_rng <- range(gaps$gap)
gap_lo_year <- gaps$year[which.min(gaps$gap)]
gap_hi_year <- gaps$year[which.max(gaps$gap)]
gap_last <- gaps[gaps$year == last, ]
n_other <- n_lang[qes, "other"]
other_rng <- range(plq$pct[plq$group == "other"])
# the claims of the text below
stopifnot(gap_of(2014) > gap_of(2012), gap_of(2018) > gap_of(2012),
          gap_last$lo < gap_of(2012), gap_last$hi > gap_of(2012),
          other_of(last) < other_of(first),
          other_of(2018) - other_of(last) > fr_of(2018) - fr_of(last),
          which.min(n_other) == length(qes))

dl <- plq
dl$series <- ifelse(dl$group == "francophone", "fr", "other")
dl$x <- qz_ex(dl$year)
series_lab <- c(fr = tr("Francophones", "Francophones"), other = tr("Non-francophones", "Non-francophones"))
dl$label <- series_lab[dl$series]
years <- sort(unique(dl$year))

qz_figure(function(mode) {
  k <- qz_tokens[[mode]]
  cols <- c(fr = k$grp[["fr"]], other = k$grp[["other"]])
  ggplot(dl, aes(x, pct, colour = series, group = series)) +
    qz_band(aes(ymin = lo, ymax = hi, fill = series), mode = mode) +
    qz_line(show.legend = FALSE) +
    qz_dots(mode, dl, aes(x, pct), colour_aes = "series") +
    weight_key(mode) +
    qz_end_labels(dl, "x", "pct", "label", "series", mode, min_gap = 7, nudge_x = 1.1) +
    scale_colour_manual(values = cols, guide = "none") +
    scale_fill_manual(values = cols, guide = "none") +
    scale_x_election(years = years, add = c(0.9, 5.4)) +
    scale_y_continuous(breaks = seq(0, 100, 25), labels = qz_pct_axis) +
    coord_cartesian(ylim = c(0, 100), expand = FALSE, clip = "off") +
    labs(x = NULL, y = tr("Reported vote for the PLQ", "Vote déclaré pour le PLQ")) +
    theme_qes(mode)
}, "plq-language",
  title = typo(sprintf(tr("At every election since %s, non-francophones were %s to %s points more likely than francophones to vote Liberal",
                          "À chaque élection depuis %s, le vote libéral est de %s à %s points de pourcentage plus élevé chez les non-francophones que chez les francophones"),
                       first, qz_num(gap_rng[1], 0), qz_num(gap_rng[2], 0))),
  subtitle = typo(tr("Share of the reported vote for the PLQ among francophones and non-francophones (mother tongue), one Quebec Election Study per election, with 95% confidence intervals",
                     "Part du vote déclaré pour le PLQ chez les francophones et les non-francophones (langue maternelle), une Étude électorale québécoise par élection, avec intervalles de confiance à 95 %")),
  alt = typo(sprintf(tr("Line chart of the share of the reported vote for the PLQ at each Quebec election from %s to %s, among francophones and among non-francophones. Among francophones it goes from %s to %s; among non-francophones it stays between %s and %s. Values in the table view.",
                        "Graphique linéaire de la part du vote déclaré pour le PLQ à chaque élection québécoise de %s à %s, chez les francophones et chez les non-francophones. Chez les francophones, elle passe de %s à %s ; chez les non-francophones, elle reste entre %s et %s. Valeurs dans la vue en tableau."),
                     first, last, qz_pct(fr_of(first), 0), qz_pct(fr_of(last), 0),
                     qz_pct(other_rng[1], 0), qz_pct(other_rng[2], 0))),
  table = typo_table(setNames(data.frame(
    gaps$year, qz_study_label(gaps$study),
    qz_cell(plq$pct[plq$group == "francophone"], plq$lo[plq$group == "francophone"],
            plq$hi[plq$group == "francophone"], n_lang[gaps$study, "francophone"]),
    qz_cell(plq$pct[plq$group == "other"], plq$lo[plq$group == "other"],
            plq$hi[plq$group == "other"], n_lang[gaps$study, "other"]),
    paste0(qz_pts(gaps$gap, 1), " [", qz_num(gaps$lo), tr(", ", " ; "), qz_num(gaps$hi), "]"),
    as.integer(n_lang[gaps$study, "francophone"]), as.integer(n_lang[gaps$study, "other"]),
    weight_text(gaps$weighted), row.names = NULL),
    c(tr("Election", "Élection"), tr("Study", "Étude"),
      tr("Francophones, % PLQ [95% CI]", "Francophones, % PLQ [IC à 95 %]"),
      tr("Non-francophones, % PLQ [95% CI]", "Non-francophones, % PLQ [IC à 95 %]"),
      tr("Gap [95% CI]", "Écart [IC à 95 %]"),
      tr("n, francophones", "n, francophones"), tr("n, non-francophones", "n, non-francophones"),
      tr("Weighting", "Pondération")))),
  note = typo(tr("Source: qesR, pooled vote_choice (reported vote) and lang_mother; non-francophones are respondents whose mother tongue is English or another language. Weighted with each study's post-election weight; hollow point (2008): unweighted.",
                 "Source : qesR, variables regroupées vote_choice (vote déclaré) et lang_mother ; les non-francophones sont les répondants de langue maternelle anglaise ou autre. Pondéré avec la pondération postélectorale de chaque étude ; point vide (2008) : non pondéré.")),
  height = 4)
```

Two results emerge. First, the divide is large at every election, from
`r qz_num(gap_rng[1], 0)` points in `r gap_lo_year` to
`r qz_num(gap_rng[2], 0)` points in `r gap_hi_year`, and it did not clearly
narrow after 2012. Relative to 2012 (about `r qz_num(gap_of(2012), 0)` points),
the gap widened in 2014 and 2018 and was back to about
`r qz_num(gap_of(last), 0)` points in `r last`, an estimate whose interval
(`r qz_num(gap_last$lo, 0)` to `r qz_num(gap_last$hi, 0)`) includes the
2012 gap. Second, the two groups did not move in step. Among francophones,
the Liberal vote fell from `r qz_pct(fr_of(first), 0)` in `r first` to
`r qz_pct(fr_of(last), 0)` in `r last`. Among non-francophones it moved
between `r qz_pct(other_rng[1], 0)` and `r qz_pct(other_rng[2], 0)` with no
clear trend, and in `r last` it was about
`r qz_num(other_of(first) - other_of(last), 0)` points below its `r first`
level. The narrower gap of `r last` comes from the non-francophone side:
from 2018 to `r last`, the Liberal vote fell
`r qz_num(other_of(2018) - other_of(last), 0)` points among non-francophones
and `r qz_num(fr_of(2018) - fr_of(last), 0)` among francophones. The
non-francophone estimate of `r last`, however, rests on only
`r n_other[[length(qes)]]` respondents, the smallest sample of the series
(the largest has `r max(n_other)`), and its interval is wide. This is also
why the page does not split non-francophones into anglophones and
allophones.

## 7. Who votes Liberal

The same data can be read from the other side: not the share of each group
that votes Liberal, but the share of Liberal voters who belong to each
group. Read this way, the change is plain.

```{r composition}
comp <- do.call(rbind, lapply(qes, function(s) {
  ds <- subset(study_design(s), vote_choice %in% "PLQ" & !is.na(francophone))
  est <- svyciprop(~I(francophone == "other"), ds, method = "logit")
  data.frame(study = s, year = plq$year[plq$study == s][1], weighted = is_weighted(s),
             pct = 100 * as.numeric(est), lo = 100 * attr(est, "ci")[[1]], hi = 100 * attr(est, "ci")[[2]])
}))
print(comp, digits = 3, row.names = FALSE)
```

```{r composition-site, echo = FALSE}
n_plq <- with(r[r$vote_choice %in% "PLQ" & !is.na(r$francophone) & r$study %in% qes, ], table(study))
stopifnot(max(n_plq) < 1000)  # "a few hundred Liberal voters at most"
dc <- comp
dc$x <- qz_ex(dc$year)

qz_figure(function(mode) {
  k <- qz_tokens[[mode]]
  col <- k$grp[["other"]]
  ggplot(dc, aes(x, pct)) +
    geom_hline(yintercept = 50, colour = k$axis, linewidth = 0.35, linetype = "22") +
    annotate("text", x = qz_ex(first) - 0.6, y = 51.5, label = tr("half of PLQ voters", "la moitié des électeurs du PLQ"),
             hjust = 0, vjust = 0, size = 3, colour = k$ink2, family = qz_font()) +
    qz_band(aes(ymin = lo, ymax = hi), fill = col, mode = mode) +
    qz_line(colour = col) +
    qz_dots(mode, dc, aes(x, pct), fixed = col) +
    weight_key(mode) +
    scale_x_election(years = years, add = c(0.9, 0.9)) +
    scale_y_continuous(breaks = seq(0, 100, 25), labels = qz_pct_axis) +
    coord_cartesian(ylim = c(0, 100), expand = FALSE, clip = "off") +
    labs(x = NULL, y = tr("Non-francophones among PLQ voters", "Non-francophones parmi les électeurs du PLQ")) +
    theme_qes(mode)
}, "plq-composition",
  title = typo(sprintf(tr("Non-francophones went from %s of Liberal voters in %s to %s in %s",
                          "Les non-francophones passent de %s des électeurs libéraux en %s à %s en %s"),
                       qz_pct(comp_of(first), 0), first, qz_pct(comp_of(last), 0), last)),
  subtitle = typo(tr("Share of respondents whose mother tongue is not French among those who reported a vote for the PLQ, one Quebec Election Study per election, with 95% confidence intervals",
                     "Part des répondants dont la langue maternelle n'est pas le français parmi ceux qui déclarent avoir voté pour le PLQ, une Étude électorale québécoise par élection, avec intervalles de confiance à 95 %")),
  alt = typo(sprintf(tr("Line chart of the share of non-francophones among respondents who reported a PLQ vote, at each Quebec election from %s to %s, with a dashed line at 50%%. It goes from %s in %s to %s in %s. Values in the table view.",
                        "Graphique linéaire de la part des non-francophones parmi les répondants qui déclarent un vote pour le PLQ, à chaque élection québécoise de %s à %s, avec une ligne pointillée à 50 %%. Elle passe de %s en %s à %s en %s. Valeurs dans la vue en tableau."),
                     first, last, qz_pct(comp_of(first), 0), first, qz_pct(comp_of(last), 0), last)),
  table = typo_table(setNames(data.frame(
    comp$year, qz_study_label(comp$study),
    qz_cell(comp$pct, comp$lo, comp$hi, as.integer(n_plq[comp$study])),
    as.integer(n_plq[comp$study]), qz_num(voters_other[as.character(comp$year)]),
    qz_num(sapply(comp$year, comp_fixed)), weight_text(comp$weighted), row.names = NULL),
    c(tr("Election", "Élection"), tr("Study", "Étude"),
      tr("Non-francophones among PLQ voters, % [95% CI]", "Non-francophones parmi les électeurs du PLQ, % [IC à 95 %]"),
      tr("n, PLQ voters", "n, électeurs du PLQ"),
      tr("Non-francophones among all voters, %", "Non-francophones parmi tous les électeurs, %"),
      sprintf(tr("Non-francophones among PLQ voters, at their %s share of all voters, %%", "Non-francophones parmi les électeurs du PLQ, à leur part de %s parmi tous les électeurs, %%"), first),
      tr("Weighting", "Pondération")))),
  note = typo(tr("Source: qesR, pooled vote_choice (reported vote) and lang_mother. Weighted with each study's post-election weight; hollow point (2008): unweighted.",
                 "Source : qesR, variables regroupées vote_choice (vote déclaré) et lang_mother. Pondéré avec la pondération postélectorale de chaque étude ; point vide (2008) : non pondéré.")),
  height = 3.4)
```

Non-francophones made up `r qz_pct(comp_of(first), 0)` of PLQ voters in
`r first` and `r qz_pct(comp_of(last), 0)` in `r last`. The path is not
smooth, and each point rests on a few hundred Liberal voters at most. The
share also depends on something other than the Liberal vote of each group:
the weighted share of non-francophones among the voters of each study,
which swings more than any change in the population would. That share is about
`r qz_pct(mean(low_share), 0)` in 2007 and 2014, and
`r qz_num(min(high_share), 0)` to `r qz_pct(max(high_share), 0)` in 2012,
2018 and 2022; the dip of 2014 lines up with that year's low share. Held
at its `r first` level, the share of non-francophones among Liberal voters
would still reach `r qz_pct(comp_fixed(last), 0)` in `r last`. The rise
holds, then, but its size depends on the composition of each sample. In
other words, the PLQ went from a party whose voters were mostly francophone
to one whose voters mostly are not. This is not because non-francophones
rallied to it: francophones left.

## 8. Sovereignty: one variable, several wordings

`sov_support` pools the referendum questions. The wording changed over
time, and the pooled column records it in `sov_support__type`: shares
should be compared within one wording only.

```{r sovereignty}
table(r$study, r$sov_support__type)
```

The `r n_four` Quebec Election Studies from 2012 to 2022 asked the same question
on independence, which the next recipe uses. The earlier studies asked the
partnership question of the 1995 referendum, and the 2012 and 2018 panels
used wordings of their own.

## 9. Pooling studies in one model

Pooling studies only makes sense in a model that keeps them apart, with
study fixed effects, so that each association is estimated within studies.
`pool = "equal"` gives every study the same total weight among the rows it
receives, so we select the rows of the model first; by default a study
counts in proportion to its number of respondents. We use a linear
probability model, so a coefficient times 100 reads in percentage points.
The first model has only the study fixed effects; the second adds support
for independence; the third lets the language gap differ from one study to
the next.

```{r pool}
four <- c("qes2012", "qes2014", "qes2018", "qes2022")
x <- r[r$study %in% four & !is.na(r$vote_choice) & !is.na(r$francophone) & !is.na(r$sov_support), ]
dp <- qes_design(x, weight = "weight_post", pool = "equal")
m1 <- svyglm(I(vote_choice == "PLQ") ~ francophone + study, dp)
m2 <- svyglm(I(vote_choice == "PLQ") ~ francophone + I(sov_support == "Yes") + study, dp)
m3 <- svyglm(I(vote_choice == "PLQ") ~ francophone * study + I(sov_support == "Yes"), dp)
100 * rbind(study_effects = coef(m1)[2], plus_independence = coef(m2)[2])
regTermTest(m3, ~ francophone:study)
```

Across the `r n_four` elections, and with study fixed effects only,
non-francophones are about `r qz_num(b1, 0)` points more likely than
francophones to vote Liberal (95% confidence interval,
`r qz_num(ci1[[1]], 0)` to `r qz_num(ci1[[2]], 0)`). Among voters who give
the same answer to the independence question, the gap is still
`r qz_num(b2, 0)` points. Support for independence thus accounts for about
`r qz_num(b1 - b2, 0)` of the `r qz_num(b1, 0)` points of the language
divide in the Liberal vote. In other words, most of the divide is not the
national question under another name, at least as one question on
independence measures it. Finally, the evidence that the gap changed
between 2012 and 2022 is weak (a test of equal gaps across the `r n_four`
studies gives p = `r qz_num(p_equal, 2)`).

## 10. The ADQ and the CAQ as one series, against the vote count

The harmonized columns keep the Action démocratique du Québec (ADQ) and
the CAQ apart. `qes_party_lineage()` adds a column that joins them, for time
series; its rows fielded before the merger of 2012 are graded
`approximate`.

```{r lineage}
l <- qes_party_lineage(r, cols = "vote_choice")
table(l$study, l$vote_choice_lineage)[, c("PLQ", "PQ", "ADQ/CAQ", "QS")]
```

Weighted within each study and set against the official result, that
column gives a first check on the reported vote. `p` is a plain ggplot2
chart to adapt (print it with `p`).

```{r plot}
lineage <- do.call(rbind, lapply(qes, function(s) {
  ds <- subset(study_design(s, l), !is.na(vote_choice_lineage))
  est <- svyciprop(~I(vote_choice_lineage == "ADQ/CAQ"), ds, method = "logit")
  ci <- attr(est, "ci")
  data.frame(study = s, year = plq$year[plq$study == s][1], weighted = is_weighted(s, l),
             pct = 100 * as.numeric(est), lo = 100 * ci[[1]], hi = 100 * ci[[2]])
}))
# the official share of valid votes of the ADQ (2007, 2008) and of the CAQ
# (2012 on), from Élections Québec
official <- data.frame(year = c(2007, 2008, 2012, 2014, 2018, 2022),
                       pct = c(30.8, 16.4, 27.1, 23.1, 37.4, 41.0))
fr <- identical(params$lang, "fr")
lab_series <- if (fr) c(lineage = "ADQ (2007, 2008), puis CAQ : vote déclaré", official = "Résultat officiel") else
  c(lineage = "ADQ (2007, 2008), then CAQ: reported vote", official = "Official result")
lab_weight <- if (fr) c(`TRUE` = "pondéré", `FALSE` = "non pondéré") else
  c(`TRUE` = "weighted", `FALSE` = "unweighted")

p <- ggplot(lineage, aes(year, pct)) +
  geom_line(data = official, aes(colour = "official")) +
  geom_point(data = official, aes(colour = "official"), shape = 45, size = 9) +
  geom_linerange(aes(ymin = lo, ymax = hi, colour = "lineage")) +
  geom_point(aes(colour = "lineage", fill = weighted), shape = 21, size = 3, stroke = 1) +
  scale_colour_manual(values = c(lineage = "#1d97b0", official = "grey20"), labels = lab_series) +
  scale_fill_manual(values = c(`TRUE` = "#1d97b0", `FALSE` = "white"), labels = lab_weight) +
  scale_x_continuous(breaks = c(2007, 2012, 2014, 2018, 2022)) +
  scale_y_continuous(limits = c(0, 50)) +
  labs(x = NULL, y = if (fr) "% du vote déclaré" else "% of the reported vote", colour = NULL, fill = NULL)
print(lineage, digits = 3, row.names = FALSE)
```

```{r plot-site, echo = FALSE}
# the same chart in the example pages' chart system (vignettes/articles/_theme.R),
# drawn for the light and the dark theme, with its table view
lineage$official <- official$pct[match(lineage$year, official$year)]
n_lineage <- with(l[!is.na(l$vote_choice_lineage) & l$study %in% qes, ], table(study))
gap <- lineage$pct - lineage$official
before <- max(abs(gap[lineage$year < last]))
before_year <- max(lineage$year[lineage$year < last])
qz_figure(function(mode) {
  k <- qz_tokens[[mode]]
  p +
    scale_colour_manual(values = c(lineage = k$party[["CAQ"]], official = k$ink), labels = lab_series,
                        guide = guide_legend(order = 1, override.aes = list(fill = c(k$party[["CAQ"]], k$ink)))) +
    scale_fill_manual(values = c(`TRUE` = k$party[["CAQ"]], `FALSE` = k$surface), labels = lab_weight,
                      guide = guide_legend(order = 2, override.aes = list(colour = k$ink2, linetype = 0, shape = 21))) +
    theme_qes(mode)
}, "lineage",
  title = typo(sprintf(tr("With the ADQ and the CAQ joined, the reported vote is within %s points of the official result until %s, and %s points short in %s",
                          "L'ADQ et la CAQ réunies, le vote déclaré reste à %s points ou moins du résultat officiel jusqu'en %s, puis se situe %s points en deçà en %s"),
                       qz_num(before, 0), before_year, qz_num(abs(gap[lineage$year == last]), 0), last)),
  subtitle = typo(tr("Share of the reported vote for the ADQ (2007, 2008) and the CAQ (2012 on), one Quebec Election Study per election, with 95% confidence intervals and the official result",
                     "Part du vote déclaré pour l'ADQ (2007, 2008) et la CAQ (à partir de 2012), une Étude électorale québécoise par élection, avec intervalles de confiance à 95 % et résultat officiel")),
  alt = typo(sprintf(tr("Dot chart with confidence intervals of the share of the reported vote for the ADQ in 2007 and 2008 and for the CAQ from 2012 to 2022, in the Quebec Election Studies, with the official result as a line with ticks. In 2022 the study has %s against %s officially. Values in the table view.",
                        "Graphique à points avec intervalles de confiance de la part du vote déclaré pour l'ADQ en 2007 et 2008 et pour la CAQ de 2012 à 2022, dans les Études électorales québécoises, avec le résultat officiel en ligne avec des traits. En 2022, l'étude donne %s contre %s officiellement. Valeurs dans la vue en tableau."),
                     qz_pct(lineage$pct[lineage$year == last], 0), qz_pct(lineage$official[lineage$year == last], 0))),
  table = typo_table(setNames(data.frame(lineage$year, qz_study_label(lineage$study),
                                         ifelse(lineage$year < 2012, "ADQ", "CAQ"),
                                         qz_cell(lineage$pct, lineage$lo, lineage$hi, as.integer(n_lineage[lineage$study])),
                                         qz_num(lineage$official), qz_pts(gap, 1), weight_text(lineage$weighted), row.names = NULL),
                              c(tr("Election", "Élection"), tr("Study", "Étude"), tr("Party", "Parti"),
                                tr("Reported vote, % [95% CI]", "Vote déclaré, % [IC à 95 %]"),
                                tr("Official, % of valid votes", "Officiel, % des votes valides"), tr("Difference", "Écart"),
                                tr("Weighting", "Pondération")))),
  note = typo(tr("Source: qesR, pooled vote_choice (reported vote) with qes_party_lineage(); official results of Élections Québec. The ADQ merged into the CAQ in 2012, and the lineage column joins them for a time series. Weighted with each study's post-election weight; hollow point (2008): unweighted.",
                 "Source : qesR, variable regroupée vote_choice (vote déclaré) avec qes_party_lineage() ; résultats officiels d'Élections Québec. L'ADQ a fusionné avec la CAQ en 2012, et la colonne de lignée les réunit pour une série chronologique. Pondéré avec la pondération postélectorale de chaque étude ; point vide (2008) : non pondéré.")),
  height = 3.6)
```

```{r plot-text, include = FALSE}
plq_last <- svyciprop(~I(vote_choice == "PLQ"), subset(study_design(qes[length(qes)]), !is.na(vote_choice)),
                      method = "logit")
plq_last_off <- off$share_valid[off$year == last & off$party == "PLQ"]
```

Up to `r before_year`, the reported vote stays within
`r qz_num(before, 0)` points of the official result. In `r last`, contrary
to the usual overreporting of the winner after an election, it falls
`r qz_num(abs(gap[lineage$year == last]), 0)` points short, while the same
study puts the PLQ at `r qz_pct(100 * as.numeric(plq_last), 0)` against
`r qz_pct(plq_last_off, 0)` officially. The levels of `r last` call for
some caution. The argument above rests on differences between groups within
each study, which are less exposed than levels to an error of this kind,
though not immune to it.

## What the recipes show

In sum, the Quebec Election Studies give only limited support to the idea
that the language divide is the national question under another name.
Support for independence explains part of the gap in the Liberal vote, and
the gap did not clearly narrow after 2012. What changed is who votes Liberal: as
francophones left, the PLQ became a party whose voters are mostly
non-francophones. Our analysis is not without limits. First, mother tongue
is a coarse measure. Recent work also treats language as an attitude, the
worry that French is under threat, and that worry varies among
francophones themselves (Daoust and Gareau-Paquette 2026). Second, the
estimates are associations within cross-sections, not causes, and the
non-francophone samples are small, and their share of each sample varies
from one study to the next. Of course, the divide does not hold for every party: support for
the Parti conservateur du Québec (PCQ) cut across it in 2022 (Bélanger et
al. 2025). The next question is where the francophone Liberals went, which
the pages on [realignment](realignment.html) and on
[changing minds during the campaign](transitions.html) take up.

## Further reading

Bélanger, Éric, Philippe Mongrain, Thomas Gareau-Paquette, and Valérie-Anne
Mahéo. 2025. "A Party that Went Viral? The Drivers of Support for the Parti
Conservateur du Québec in the 2022 Election." *Canadian Journal of
Political Science* 58 (2): 277–96.
<https://doi.org/10.1017/S0008423924000829>

Bélanger, Éric, Richard Nadeau, Ailsa Henderson, and Eve Hepburn. 2018.
*The National Question and Electoral Politics in Quebec and Scotland*.
Montreal and Kingston: McGill-Queen's University Press.
<https://doi.org/10.1515/9780773554139>

Daoust, Jean-François, and Thomas Gareau-Paquette. 2024. "Is Quebec
Independence Still Key in Making Sense of Canadian Elections? A
Longitudinal Analysis (2000–2021)." *Regional & Federal Studies* 34 (5):
781–806. <https://doi.org/10.1080/13597566.2023.2233422>

Daoust, Jean-François, and Thomas Gareau-Paquette. 2026. « De plus en plus
inquiets : qui sont les Québécois qui considèrent le français comme menacé
et quelles en sont les conséquences électorales? » *Canadian Journal of
Political Science* 59 (1): 61–80.
<https://doi.org/10.1017/S0008423925100942>

Hamilton, Richard, and Maurice Pinard. 1976. "The Bases of Parti Québécois
Support in Recent Quebec Elections." *Canadian Journal of Political
Science* 9 (1): 3–26. <https://doi.org/10.1017/S0008423900043158>

Pinard, Maurice, and Richard Hamilton. 1978. "The Parti Québécois Comes to
Power: An Analysis of the 1976 Quebec Election." *Canadian Journal of
Political Science* 11 (4): 739–76.
<https://doi.org/10.1017/S000842390004659X>
