---
title: "Recipes: pooled variables in five minutes"
description: >
  Short, copyable code for the pooled harmonized variables: one variable
  for all studies, its types and sources, weighted estimates by study.
params:
  lang: "en"
---

*[Version française](fr-recettes.html)*

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
options(qesR.lang = params$lang)
```

Each recipe below runs as is on the full data files, which qesR downloads
once through its cache. The other example pages are built from these
same steps.

```{r packages}
library(qesR)
library(survey)
library(ggplot2)
```

## 1. One vote-choice variable for every study

`vote_choice` pools the reported vote, the pushed intention and the plain
intention of every study into one column; `vote_choice__type` says which
question each row answers.

```{r pooled}
studies <- c("qes1998", "qes2007", "qes2007_panel", "qes2008", "qes_crop_2007_2010",
             "qes2012", "qes2012_panel", "qes2014", "qes2018", "qes2018_panel", "qes2022")
h <- qes_harmonize(studies = studies, targets = c("vote_choice", "lang_mother"),
                   layout = "long", quiet = TRUE)
table(h$study, h$vote_choice__type)
```

In the long layout a pre/post study has one row per wave: the intention
before the election and the reported vote after it. In the default
respondent layout each study's pooled value comes from one wave, so that
one weight fits it.

## 2. Keep one type only

`types` restricts a pooled variable to some of its members. The reported
vote only:

```{r recall}
r <- qes_harmonize(studies = studies, targets = c("vote_choice", "lang_mother"),
                   types = list(vote_choice = "recall"), quiet = TRUE)
table(r$study, r$vote_choice__type, useNA = "ifany")
```

Within a study-wave the members are tried in a fixed order (recall, then
the intention with the undecided pushed, then the plain intention), and
the first one that asked the respondent sets the row. `qes_spec("pooled")`
lists that order.

## 3. Where each value comes from

```{r sources}
unique(r[!is.na(r$vote_choice__item), c("study", "vote_choice__item", "vote_choice__grade")])
qes_provenance(r, level = "pooled")
```

The grade is the grade of the study's question (`identical`,
`comparable`, `approximate`); the harmonization reference gives the
reasons.

## 4. A weighted share with its confidence interval, in one study

```{r one-study}
d <- qes_design(r[r$study == "qes2018", ], weight = "weight_post")
d <- subset(d, !is.na(vote_choice))
svyciprop(~I(vote_choice == "CAQ"), d, method = "logit")
```

## 5. By group

```{r by-group}
svyby(~I(vote_choice == "PLQ"), ~lang_mother, subset(d, !is.na(lang_mother)),
      svyciprop, method = "logit", vartype = "ci")
```

## 6. Every study, without pooling them

Studies differ in population, mode, wording and weighting: estimate
within each study, then compare. Studies whose weight is still under
review have `NA` weights, and `qes_design()` leaves their rows out.

```{r loop}
weighted <- c("qes2007", "qes2012", "qes2014", "qes2018", "qes2018_panel", "qes2022")
# 2007: the CAQ did not yet exist (it was founded in 2011), so the loop starts in 2012
caq <- do.call(rbind, lapply(setdiff(weighted, "qes2007"), function(s) {
  ds <- qes_design(r[r$study == s, ], weight = "weight_post")
  ds <- subset(ds, !is.na(vote_choice))
  est <- svyciprop(~I(vote_choice == "CAQ"), ds, method = "logit")
  ci <- attr(est, "ci")
  data.frame(study = s, pct = 100 * as.numeric(est), lo = 100 * ci[[1]], hi = 100 * ci[[2]])
}))
caq
```

Pooling makes sense for a model that keeps the studies apart, such as a
regression with a term for each study: the coefficient of a French
mother tongue is then estimated within studies (on the logit scale,
common to all of them). (A single
PLQ share pooled over 2007-2022 would mean nothing.) `pool = "equal"`
gives every study the same total weight; by default a study counts in
proportion to its number of respondents:

```{r pool}
dp <- qes_design(r[r$study %in% weighted, ], weight = "weight_post", pool = "equal")
dp <- subset(dp, !is.na(vote_choice) & !is.na(lang_mother))
m <- svyglm(I(vote_choice == "PLQ") ~ I(lang_mother == "French") + study, dp,
            family = quasibinomial())
summary(m)$coefficients[1:2, ]
```

## 7. Sovereignty: one variable, several wordings

`sov_support` pools the referendum questions; compare within one wording
(`sov_support__type`):

```{r sovereignty}
s <- qes_harmonize(studies = studies, targets = "sov_support", quiet = TRUE)
table(s$study, s$sov_support__type)
```

## 8. The ADQ and the CAQ in one time series

The harmonized columns keep the ADQ and the CAQ apart.
`qes_party_lineage()` adds a column that joins them, for time series; the
rows fielded before the merger of 2012 are graded `approximate` there.

```{r lineage}
l <- qes_party_lineage(r, cols = "vote_choice")
table(l$study, l$vote_choice_lineage)[, c("PLQ", "PQ", "ADQ/CAQ", "QS")]
```

## 9. A first figure: the ADQ and the CAQ as one series

Recipe 8's lineage column, weighted within each study, against the
official result. `p` is a plain ggplot2 chart, ready to adapt (print it
with `p`); the site draws the same chart with the theme of its example
pages, below.

```{r plot}
qes <- c("qes2007", "qes2008", "qes2012", "qes2014", "qes2018", "qes2022")
lineage <- do.call(rbind, lapply(qes, function(s) {
  x <- l[l$study == s, ]
  weighted <- any(!is.na(x$weight_post))
  if (!weighted) x$weight_post <- 1   # qes2008: weight under review, so unweighted
  ds <- subset(qes_design(x, weight = "weight_post"), !is.na(vote_choice_lineage))
  est <- svyciprop(~I(vote_choice_lineage == "ADQ/CAQ"), ds, method = "logit")
  ci <- attr(est, "ci")
  data.frame(study = s, year = as.integer(substr(x$year[1], 1, 4)), weighted = weighted,
             pct = 100 * as.numeric(est), lo = 100 * ci[[1]], hi = 100 * ci[[2]])
}))
# the official share of valid votes of the ADQ (2007, 2008) and of the CAQ
# (2012 on), from Élections Québec
official <- data.frame(year = c(2007, 2008, 2012, 2014, 2018, 2022),
                       pct = c(30.8, 16.4, 27.1, 23.1, 37.4, 41.0))
fr <- identical(params$lang, "fr")
lab_series <- if (fr) c(lineage = "ADQ (2007, 2008), puis CAQ : vote déclaré", official = "Résultat officiel") else
  c(lineage = "ADQ (2007, 2008), then CAQ: reported vote", official = "Official result")
lab_weight <- if (fr) c(`TRUE` = "pondéré", `FALSE` = "non pondéré (pondération en révision)") else
  c(`TRUE` = "weighted", `FALSE` = "unweighted (weight under review)")

p <- ggplot(lineage, aes(year, pct)) +
  geom_line(data = official, aes(colour = "official")) +
  geom_point(data = official, aes(colour = "official"), shape = 45, size = 9) +
  geom_linerange(aes(ymin = lo, ymax = hi, colour = "lineage")) +
  geom_point(aes(colour = "lineage", fill = weighted), shape = 21, size = 3, stroke = 1) +
  scale_colour_manual(values = c(lineage = "#1d97b0", official = "grey20"), labels = lab_series) +
  scale_fill_manual(values = c(`TRUE` = "#1d97b0", `FALSE` = "white"), labels = lab_weight) +
  scale_x_continuous(breaks = c(2007, 2012, 2014, 2018, 2022)) +
  scale_y_continuous(limits = c(0, 50)) +
  labs(x = NULL, y = if (fr) "% du vote déclaré" else "% of the reported vote", colour = NULL, fill = NULL)
lineage
```

```{r plot-site, echo = FALSE}
# the same chart in the example pages' chart system (vignettes/articles/_theme.R),
# drawn for the light and the dark theme, with its table view
source("_setup.R", encoding = "UTF-8")
lineage$official <- official$pct[match(lineage$year, official$year)]
gap <- lineage$pct - lineage$official
before <- max(abs(gap[lineage$year < 2022]))
qz_figure(function(mode) {
  k <- qz_tokens[[mode]]
  p +
    scale_colour_manual(values = c(lineage = k$party[["CAQ"]], official = k$ink), labels = lab_series,
                        guide = guide_legend(order = 1, override.aes = list(fill = c(k$party[["CAQ"]], k$ink)))) +
    scale_fill_manual(values = c(`TRUE` = k$party[["CAQ"]], `FALSE` = k$surface), labels = lab_weight,
                      guide = guide_legend(order = 2, override.aes = list(colour = k$ink2, linetype = 0, shape = 21))) +
    theme_qes(mode)
}, "lineage",
  title = sprintf(tr("Read as one lineage, the ADQ/CAQ vote is within %s points of its result until 2018, and %s points short in 2022",
                     "Lu comme une seule lignée, le vote ADQ/CAQ est à %s points ou moins de son résultat jusqu'en 2018, et %s points sous lui en 2022"),
                  qz_num(before, 0), qz_num(abs(gap[lineage$year == 2022]), 0)),
  subtitle = tr("Share of the reported vote for the ADQ (2007, 2008) and the CAQ (2012 on), one Quebec Election Study per election, with 95% confidence intervals and the official result",
                "Part du vote déclaré pour l'ADQ (2007, 2008) et la CAQ (à partir de 2012), une Étude électorale québécoise par élection, avec intervalles de confiance à 95 % et résultat officiel"),
  alt = sprintf(tr("Dot chart with confidence intervals of the share of the reported vote for the ADQ in 2007 and 2008 and for the CAQ from 2012 to 2022, in the Quebec Election Studies, with the official result as a line with ticks. In 2022 the study has %s against %s officially. Values in the table view.",
                   "Graphique à points avec intervalles de confiance de la part du vote déclaré pour l'ADQ en 2007 et 2008 et pour la CAQ de 2012 à 2022, dans les Études électorales québécoises, avec le résultat officiel en ligne avec des traits. En 2022, l'étude donne %s contre %s officiellement. Valeurs dans la vue en tableau."),
                qz_pct(lineage$pct[lineage$year == 2022], 0), qz_pct(lineage$official[lineage$year == 2022], 0)),
  table = setNames(data.frame(lineage$year, qz_study_label(lineage$study),
                              ifelse(lineage$year < 2012, "ADQ", "CAQ"),
                              qz_cell(lineage$pct, lineage$lo, lineage$hi, 100),
                              qz_num(lineage$official), qz_pts(gap, 1), qz_weight_text(lineage$weighted), row.names = NULL),
                   c(tr("Election", "Élection"), tr("Study", "Étude"), tr("Party", "Parti"),
                     tr("Reported vote, % [95% CI]", "Vote déclaré, % [IC à 95 %]"),
                     tr("Official, % of valid votes", "Officiel, % des votes valides"), tr("Difference", "Écart"),
                     tr("Weighting", "Pondération"))),
  note = tr("Source: qesR, pooled vote_choice (reported vote) with qes_party_lineage(); official results of Élections Québec. The ADQ merged into the CAQ in 2012: the lineage column joins them for a time series, and its rows before the merger are graded approximate. Weighted with each study's post-election weight; hollow (2008): unweighted, weight under review.",
            "Source : qesR, variable regroupée vote_choice (vote déclaré) avec qes_party_lineage() ; résultats officiels d'Élections Québec. L'ADQ a fusionné avec la CAQ en 2012 : la colonne de lignée les réunit pour une série chronologique, et ses lignes d'avant la fusion ont le niveau approximatif. Pondéré avec la pondération postélectorale de chaque étude ; creux (2008) : non pondéré, pondération en révision."),
  height = 3.6)
```

## 10. Respondents by study

```{r respondents}
n <- aggregate(qes_id ~ study + year, data = r, FUN = length)
names(n)[3] <- "respondents"
n[order(n$year, n$study), ]
```
