---
title: "Are young Quebecers still the most pro-independence? Support for independence by cohort, 2007-2022"
description: >
  In 2007 the youngest francophones were the most likely to vote Yes;
  in 2022 they were the least likely. Most of the change came from within
  cohorts, not from one generation replacing another.
params:
  lang: "en"
---

*[Version française](fr-souverainete-generations.html)*

```{r setup, include = FALSE}
source("_setup.R", encoding = "UTF-8")
```

```{r harmonize, echo = FALSE}
h <- qes_harmonize(
  studies = qz_studies,
  targets = c("sov_support", "birth_year", "lang_mother"),
  missing = "reasons", quiet = TRUE
)
```

```{r estimates, echo = FALSE}
# Page-local helpers. A value as the page prints it, so that the differences
# given in the text add up with the values shown next to them.
shown <- function(x, digits = 0) as.numeric(formatC(x, format = "f", digits = digits))
# French typography for the strings this page builds: a thin no-break space
# before ; : ? !, a no-break space before % and inside « », the typographic
# apostrophe. English strings pass through unchanged.
typo <- function(x) {
  if (!identical(qz_lang, "fr") || is.null(x)) return(x)
  x <- gsub("'", "’", x, fixed = TRUE)
  x <- gsub(" ([;:?!])", " \\1", x)
  x <- gsub(" %", " %", x, fixed = TRUE)
  x <- gsub("« ", "« ", x, fixed = TRUE)
  gsub(" »", " »", x, fixed = TRUE)
}
qz_pct_page <- qz_pct
qz_pct <- function(x, digits = 1) typo(qz_pct_page(x, digits))
qz_int <- function(x) formatC(round(x), format = "d", big.mark = tr(",", " "))
qz_weight_text <- function(weighted) ifelse(weighted, tr("weighted", "pondéré"), tr("unweighted", "non pondéré"))
qz_figure_page <- qz_figure
qz_figure <- function(plot_fun, id, title, alt, table, note = NULL, subtitle = NULL, ...) {
  table[] <- lapply(table, function(col) if (is.character(col)) typo(col) else col)
  names(table) <- typo(names(table))
  qz_figure_page(plot_fun, id, typo(title), typo(alt), table, typo(note), typo(subtitle), ...)
}
# the hollow-dot key, without the words in parentheses
plain_key <- function(p) {
  tryCatch({
    s <- p$scales$get_scales("shape")
    if (!is.null(s) && is.character(s$labels)) s$labels <- sub(" \\(.*\\)$", "", s$labels)
  }, error = function(e) NULL)
  p
}

# the share of Yes among those who would vote Yes or No
h$yes <- factor(ifelse(h$sov_support %in% "Yes", "yes",
                       ifelse(h$sov_support %in% "No", "no", NA)), levels = c("yes", "no"))
# every answer of those who were asked: Yes, No, or neither (would not vote,
# don't know, refused)
asked <- !is.na(h$sov_support) | h$sov_support__na %in% c("dk", "refused", "dk_refused", "no_answer")
h$answer <- factor(ifelse(!asked, NA, ifelse(h$sov_support %in% "Yes", "yes",
                                             ifelse(h$sov_support %in% "No", "no", "neither"))),
                   levels = c("yes", "no", "neither"))
cohort_breaks <- c(-Inf, 1944, 1959, 1974, 1989, Inf)
cohorts <- c("1944-", "1945-59", "1960-74", "1975-89", "1990+")
h$cohort <- cut(h$birth_year, cohort_breaks, labels = cohorts)
# the cohort that was 18 to 22 in 2012, followed on its own
h$born_90_94 <- h$birth_year >= 1990 & h$birth_year <= 1994
h$francophone <- h$lang_mother %in% "French"

# one Quebec Election Study per election, from the first that asked the year of birth
qes <- qz_primary[c("2007", "2008", "2012", "2014", "2018", "2022")]
design <- lapply(setNames(qes, qes), function(s) qz_design(h, s, qz_weight_col(h, s, "sov_support")))

est <- do.call(rbind, lapply(qes, function(s) {
  d <- design[[s]]
  fr <- subset(d, francophone)
  rbind(qz_tag(qz_share(d, "yes"), study = s, lang = "all"),
        qz_tag(qz_share(fr, "yes"), study = s, lang = "fr"),
        qz_tag(qz_share(fr, "yes", by = "cohort"), study = s, lang = "fr"))
}))
est <- est[est$level == "yes" & est$n > 0, ]
est$year <- qz_year(est$study)
est$type <- qz_first(h, "sov_support__type", est$study)
y <- function(s, g, lang = "fr") est$pct[est$study == s & est$group == g & est$lang == lang]
n_of <- function(s, g, lang = "fr") est$n[est$study == s & est$group == g & est$lang == lang]
# the gap between two cohorts, in points, as the page prints them
gap <- function(s, a, b) shown(y(s, a)) - shown(y(s, b))
type_lab <- c(independence = tr("An independent country", "Un pays indépendant"),
              partnership_1995_push = tr("The 1995 question (partnership)", "La question de 1995 (partenariat)"))
cohort_lab <- c(`1944-` = tr("born 1944 or before", "nés en 1944 ou avant"),
                `1945-59` = tr("born 1945-59", "nés en 1945-1959"),
                `1960-74` = tr("born 1960-74", "nés en 1960-1974"),
                `1975-89` = tr("born 1975-89", "nés en 1975-1989"),
                `1990+` = tr("born 1990+", "nés en 1990 ou après"))

gap07 <- gap("qes2007", "1975-89", "1944-")
gap22 <- gap("qes2022", "1990+", "1945-59")
boomers <- est[est$lang == "fr" & est$group == "1945-59" & est$type == "independence", ]
r4559 <- range(shown(boomers$pct))
# the size of the samples: francophones who answered Yes or No in each study,
# and the largest cohort in a single study
n_fr <- est$n[est$lang == "fr" & est$group == "all"]
n_cell <- max(est$n[est$lang == "fr" & est$group %in% cohorts])
```

```{r turnover, echo = FALSE}
# Page-local helpers. The share of each cohort among francophones who would
# vote Yes or No (w), and the share of Yes in each cohort (p), with their
# covariance matrices, in one study.
cohort_parts <- function(s) {
  d <- subset(design[[s]], francophone & !is.na(yes) & !is.na(cohort))
  d$variables$yes_n <- as.numeric(d$variables$yes == "yes")
  w <- survey::svymean(~cohort, d)
  p <- survey::svyby(~yes_n, ~cohort, d, survey::svymean, covmat = TRUE)
  list(w = unname(coef(w)), vw = unname(vcov(w)), p = unname(coef(p)), vp = unname(vcov(p)),
       n = nrow(d$variables))
}
# Yes if the cohorts had the weights of study A and the levels of study B,
# in percent, with a 95% interval (the two studies are independent samples)
blend <- function(A, B) {
  v <- sum(A$w * B$p)
  se <- sqrt(as.numeric(t(B$p) %*% A$vw %*% B$p + t(A$w) %*% B$vp %*% A$w))
  data.frame(pct = 100 * v, lo = 100 * (v - 1.96 * se), hi = 100 * (v + 1.96 * se))
}
p12 <- cohort_parts("qes2012")
p22 <- cohort_parts("qes2022")
obs <- function(s) {
  x <- qz_share(subset(design[[s]], francophone & !is.na(cohort)), "yes")
  x[x$level == "yes", c("pct", "lo", "hi")]
}
cf <- rbind(obs("qes2012"), blend(p22, p12), blend(p12, p22), obs("qes2022"))
cf$row <- c("obs12", "turnover", "within", "obs22")
cf$kind <- c("observed", "counterfactual", "counterfactual", "observed")
# to one decimal, as the figure and the text print them
v <- setNames(shown(cf$pct, 1), cf$row)
decline <- v[["obs12"]] - v[["obs22"]]
# what replacement alone takes off support with the levels of 2022
repl22 <- v[["within"]] - v[["obs22"]]
# the cohort that was 18 to 22 in 2012, followed on its own
closed <- do.call(rbind, lapply(c("qes2012", "qes2022"), function(s) {
  x <- qz_share(subset(design[[s]], francophone & born_90_94), "yes")
  qz_tag(x[x$level == "yes", ], study = s)
}))
c9094 <- function(s, col = "pct") closed[[col]][closed$study == s]
```

For a long time, the sovereignty project was thought to have time on its
side. On this older view, the young voted Yes in greater numbers than their
elders, and each new generation would add to the Yes camp. Whether that age
gap reflects generations or the life cycle has been studied in its own
right (Dufresne, Tessier and Montigny 2019). A newer view points the other
way. Mahéo and Bélanger (2018) showed that the decline of the Parti
Québécois (PQ) in 2014 was concentrated among millennials, who were less
inclined to see sovereignty as a priority, and Daoust and Gareau-Paquette
(2024) note that the shrinking electoral weight of the baby boomers could
alter the cleavages that structure Quebec politics. On this newer view, the
young have turned away from independence, and generational replacement,
the arrival of new cohorts as older ones leave the electorate, now works
against the Yes side.

The PQ vote does not tell us whether support for independence itself has
followed the same path. Nor does it tell us where the change comes from:
one generation replacing another, or change within the generations
themselves. The Quebec Election Studies allow us to follow francophone
birth cohorts from 2007 to 2022 and to ask both questions. Since each study
interviews new people, we follow cohorts, not individuals.

In 2007, francophones born from 1975 to 1989 were the cohort most likely to
vote Yes: `r qz_pct(y("qes2007", "1975-89"), 0)` would have voted Yes to the
1995 question, compared to `r qz_pct(y("qes2007", "1944-"), 0)` of those
born in 1944 or before. In 2022, francophones born in 1990 or after were the
least likely: `r qz_pct(y("qes2022", "1990+"), 0)` would vote Yes to an
independent country, compared to `r qz_pct(y("qes2022", "1945-59"), 0)` of
those born from 1945 to 1959, roughly the baby boomers. In other words, the
generation gap did not close; it reversed. The newer view is thus right
about the level, but less so about the mechanism. Depending on the year
whose levels we hold fixed, generational replacement accounts for between
none and `r qz_num(repl22, 1)` of the `r qz_num(decline, 1)` points lost
among francophones since 2012. Most of the decline happened within the
cohorts.

## The change of question in 2012

```{r fig-levels, echo = FALSE}
d <- est[est$group == "all" & est$lang == "all", ]
d$weighted <- as.logical(d$weighted)
d$x <- qz_ex(d$year)
d$series <- ifelse(d$type == "independence", "independence", "partnership")
series_lab <- data.frame(
  x = c(qz_ex(2008) + 0.5, qz_ex(2022) + 0.5),
  pct = c(d$pct[d$study == "qes2008"], d$pct[d$study == "qes2022"]),
  label = c(tr("1995\nquestion", "question\nde 1995"), tr("independent\ncountry", "pays\nindépendant")))
r95 <- range(shown(d$pct[d$series == "partnership"]))
rind <- range(shown(d$pct[d$series == "independence"]))

fig_levels <- function(mode) {
  k <- qz_tokens[[mode]]
  plain_key(ggplot(d, aes(x, pct)) +
    annotate("rect", xmin = qz_ex(2008) + 0.6, xmax = qz_ex(2012) - 0.6, ymin = -Inf, ymax = Inf,
             fill = k$wash) +
    annotate("text", x = (qz_ex(2008) + qz_ex(2012)) / 2, y = 64,
             label = tr("the question\nchanges", "la question\nchange"),
             size = 3, colour = k$ink2, family = qz_font(), lineheight = 0.9) +
    geom_hline(yintercept = 49.42, colour = k$ink2, linewidth = 0.35) +
    annotate("text", x = qz_ex(2007) - 0.6, y = 50.6,
             label = tr("1995 referendum, all voters: 49.4% Yes", "Référendum de 1995, tous les électeurs : 49,4 % Oui"),
             hjust = 0, vjust = 0, size = 3, colour = k$ink2, family = qz_font()) +
    qz_line(aes(group = series), colour = k$yes) +
    qz_ci(aes(ymin = lo, ymax = hi), colour = k$yes) +
    qz_dots(mode, d, aes(x, pct), fixed = k$yes) +
    geom_text(data = series_lab, aes(x, pct, label = label), hjust = 0, size = 3, colour = k$ink,
              family = qz_font(), lineheight = 0.9) +
    scale_x_election(years = c("2007", "2008", "2012", "2014", "2018", "2022"), add = c(0.9, 2.6)) +
    scale_y_continuous(breaks = seq(20, 60, 10), labels = qz_pct_axis) +
    coord_cartesian(ylim = c(20, 68), expand = FALSE, clip = "off") +
    labs(x = NULL, y = tr("Would vote Yes (among Yes and No)", "Voterait Oui (parmi Oui et Non)")) +
    theme_qes(mode))
}

tab <- d[order(d$year), ]
tab_levels <- setNames(data.frame(
  tab$year, qz_study_label(tab$study), type_lab[tab$type],
  qz_cell(tab$pct, tab$lo, tab$hi, tab$n), tab$n, qz_weight_text(tab$weighted),
  row.names = NULL
), c(tr("Election", "Élection"), tr("Study", "Étude"), tr("Question", "Question"),
     tr("Yes, % [95% CI]", "Oui, % [IC à 95 %]"), "n", tr("Weighting", "Pondération")))
alt <- sprintf(tr(
  "Line chart with confidence intervals of the share of all respondents who would vote Yes, among those who would vote Yes or No, in the Quebec Election Studies from 2007 to 2022. Support is %s%% to %s%% on the 1995 question (2007, 2008) and %s%% to %s%% on an independent country (2012 to 2022); a grey band marks the change of question. Values in the table view.",
  "Graphique linéaire avec intervalles de confiance de la part de tous les répondants qui voteraient Oui, parmi celles et ceux qui voteraient Oui ou Non, dans les Études électorales québécoises de 2007 à 2022. L'appui est de %s %% à %s %% sur la question de 1995 (2007, 2008) et de %s %% à %s %% sur un pays indépendant (2012 à 2022) ; une bande grise marque le changement de question. Valeurs dans la vue en tableau."),
  r95[1], r95[2], rind[1], rind[2])
qz_figure(fig_levels, "levels",
  title = sprintf(tr("The question changed in 2012: %s-%s%% Yes on the 1995 question, %s-%s%% on an independent country",
                     "La question change en 2012 : de %s à %s %% d'appui au Oui sur la question de 1995, de %s à %s %% sur un pays indépendant"),
                  r95[1], r95[2], rind[1], rind[2]),
  subtitle = tr("Would vote Yes in a referendum, among those who would vote Yes or No, all respondents, with 95% confidence intervals",
                "Voterait Oui à un référendum, parmi celles et ceux qui voteraient Oui ou Non, tous les répondants, avec intervalles de confiance à 95 %"),
  alt = alt, table = tab_levels, height = 3.9,
  note = tr("Source: Quebec Election Studies, 2007 to 2022, all respondents. The 2007 and 2008 studies asked the 1995 referendum question, which offered a partnership with Canada, and asked those who did not know which way they leaned; from 2012 on, the studies asked about an independent country. The 2008 study is unweighted (hollow dot).",
            "Source : Études électorales québécoises, 2007 à 2022, tous les répondants. Les études de 2007 et de 2008 ont posé la question du référendum de 1995, qui offrait un partenariat avec le Canada, et ont relancé les indécis ; à partir de 2012, les études portent sur un pays indépendant. L'étude de 2008 n'est pas pondérée (point creux)."))
```

Before comparing cohorts, a word on the question itself. The 2007 and 2008
studies asked the question of the 1995 referendum, which offered a
partnership with Canada, and asked those who did not know which way they
leaned. From 2012 on, the studies asked about Quebec becoming an
independent country, with no such follow-up. In these studies, the first
wording draws more Yes than the second. Wording has long been known to move
support for sovereignty (Yale and Durand 2011), and here the change of
wording also coincides with the gap between the 2008 and 2012 elections, so
the levels on each side of the grey band cannot be compared. On the
independence question, support among all respondents fell from
`r qz_pct(y("qes2012", "all", "all"), 0)` in 2012 to
`r qz_pct(y("qes2014", "all", "all"), 0)` in 2014, and has barely moved since
(`r qz_pct(y("qes2022", "all", "all"), 0)` in 2022). For this reason, we
compare cohorts within one study and one question, and we read the order of
the cohorts rather than their level.

## The order of the cohorts reverses

```{r fig-gradient, echo = FALSE}
d <- est[est$lang == "fr" & est$group %in% cohorts & est$n >= qz_min_n, ]
d$weighted <- as.logical(d$weighted)
d$cohort <- factor(d$group, levels = rev(cohorts))
q_short <- function(type) ifelse(type == "independence", tr("independence", "indépendance"),
                                 tr("1995 question", "question de 1995"))
d$panel <- factor(paste0(d$year, " · ", q_short(d$type)),
                  levels = unique(paste0(d$year, " · ", q_short(d$type))[order(d$year)]))

fig_gradient <- function(mode) {
  k <- qz_tokens[[mode]]
  plain_key(ggplot(d, aes(pct, cohort)) +
    geom_vline(xintercept = 50, colour = k$axis, linewidth = 0.35) +
    annotate("text", x = 50.8, y = Inf, label = tr("50%", "50 %"), hjust = 0, vjust = 1.2, size = 2.7,
             colour = k$ink2, family = qz_font()) +
    qz_cih(aes(xmin = lo, xmax = hi), colour = k$yes) +
    qz_dots(mode, d, aes(pct, cohort), fixed = k$yes) +
    facet_wrap(~panel, nrow = 2) +
    scale_x_continuous(limits = c(10, 72), breaks = c(20, 40, 60), labels = qz_pct_axis, expand = expansion(0)) +
    scale_y_discrete(labels = function(x) c(`1944-` = tr("1944 or before", "1944 ou avant"), `1945-59` = "1945-1959",
                                            `1960-74` = "1960-1974", `1975-89` = "1975-1989",
                                            `1990+` = tr("1990 or after", "1990 ou après"))[x],
                     expand = expansion(add = c(0.6, 1))) +
    labs(x = tr("Would vote Yes (among Yes and No)", "Voterait Oui (parmi Oui et Non)"),
         y = tr("Born", "Nés")) +
    theme_qes(mode, grid = "x") +
    theme(panel.spacing.x = unit(22, "pt")))
}
alt <- sprintf(tr(
  "Dot charts with confidence intervals in six panels, one per study from 2007 to 2022: the share of francophones who would vote Yes in each birth cohort. In 2007 the youngest cohort (born 1975-1989) is %s from the cohort born in 1944 or before; in 2022 the youngest (born in 1990 or after) is %s from the cohort born 1945-1959. Values in the table view.",
  "Graphiques à points avec intervalles de confiance en six panneaux, un par étude de 2007 à 2022 : la part des francophones qui voteraient Oui dans chaque cohorte de naissance. En 2007, la cohorte la plus jeune (née en 1975-1989) est à %s de celle née en 1944 ou avant ; en 2022, la plus jeune (née en 1990 ou après) est à %s de celle née en 1945-1959. Valeurs dans la vue en tableau."),
  qz_pts(gap07), qz_pts(gap22))
tab <- est[est$lang == "fr" & est$group %in% cohorts, ]
tab <- tab[order(tab$year, match(tab$group, cohorts)), ]
tab_gradient <- setNames(data.frame(
  tab$year, type_lab[tab$type], cohort_lab[tab$group], qz_cell(tab$pct, tab$lo, tab$hi, tab$n), tab$n,
  qz_weight_text(as.logical(tab$weighted)), row.names = NULL
), c(tr("Election", "Élection"), tr("Question", "Question"), tr("Cohort", "Cohorte"),
     tr("Yes, % [95% CI]", "Oui, % [IC à 95 %]"), "n", tr("Weighting", "Pondération")))
qz_figure(fig_gradient, "gradient",
  title = tr("In 2007 the youngest francophones were the cohort most likely to vote Yes; in 2022 they were the least likely",
             "En 2007, les plus jeunes francophones formaient la cohorte la plus souverainiste ; en 2022, la moins souverainiste"),
  subtitle = tr("Francophones who would vote Yes, by birth cohort, in each Quebec Election Study, with 95% confidence intervals",
                "Francophones qui voteraient Oui, selon la cohorte de naissance, dans chaque Étude électorale québécoise, avec intervalles de confiance à 95 %"),
  alt = alt, table = tab_gradient, height = 4.8,
  note = tr("Source: Quebec Election Studies, 2007 to 2022, respondents whose mother tongue is French. The 2007 and 2008 studies asked the 1995 question, the others about an independent country: compare the order of the cohorts within a panel, not the levels across panels. The 2008 study is unweighted (hollow dots). Cohorts of fewer than 30 respondents are not drawn.",
            "Source : Études électorales québécoises, 2007 à 2022, répondants de langue maternelle française. Les études de 2007 et de 2008 ont posé la question de 1995, les autres portent sur un pays indépendant : comparer l'ordre des cohortes dans un panneau, pas les niveaux d'un panneau à l'autre. L'étude de 2008 n'est pas pondérée (points creux). Les cohortes de moins de 30 répondants ne sont pas tracées."))
```

First, in 2007, on the 1995 question, support falls with age: francophones
born from 1975 to 1989 are `r qz_num(gap07, 0)` percentage points above
those born in 1944 or before. This is the pattern the older view has in
mind. The unweighted 2008 study gives a less tidy picture, as it puts those
born from 1945 to 1959 first (`r qz_pct(y("qes2008", "1945-59"), 0)`).
Second, in 2012 the youngest cohort no longer stands out: those born in 1990
or after (`r qz_pct(y("qes2012", "1990+"), 0)` Yes) are level with those
born from 1945 to 1959 (`r qz_pct(y("qes2012", "1945-59"), 0)`), and the
cohort born from 1960 to 1974 is
`r qz_num(gap("qes2012", "1945-59", "1960-74"), 0)` points lower
(`r qz_pct(y("qes2012", "1960-74"), 0)`). The oldest cohort, at
`r qz_pct(y("qes2012", "1944-"), 0)`, is lower still, although it rests on
only `r n_of("qes2012", "1944-")` respondents. Third, from 2018 on the order
is reversed. In 2022, the youngest cohort is the least likely to vote Yes,
`r qz_num(-gap22, 0)` points below those born from 1945 to 1959
(`r qz_pct(y("qes2022", "1945-59"), 0)`, n = `r n_of("qes2022", "1945-59")`).
The oldest cohort, born in 1944 or before, is below the 1945-59 cohort at
every election, but its intervals are among the widest: only
`r n_of("qes2022", "1944-")` francophones of that cohort answered in 2022.

## Which cohorts changed

```{r fig-cohorts, echo = FALSE}
years <- c(2012, 2014, 2018, 2022)
d <- est[est$lang == "fr" & est$group %in% cohorts & est$year %in% years & est$n >= qz_min_n, ]
d$weighted <- as.logical(d$weighted)
d$x <- d$year
d$cohort <- factor(d$group, levels = cohorts)
d$label <- cohort_lab[as.character(d$cohort)]
# the story is two cohorts; the other three are grey context
focus <- c("1945-59", "1990+")
d$series <- factor(ifelse(d$group %in% focus, d$group, "context"), levels = c(focus, "context"))
foc <- d[d$group %in% focus, ]
ctx <- d[!d$group %in% focus, ]
small <- unique(d$year[d$group == "1944-" & d$n < 100])

fig_cohorts <- function(mode) {
  k <- qz_tokens[[mode]]
  # darker = older in both themes
  cols <- if (mode == "light") c(`1945-59` = "#104281", `1990+` = "#5598e7", context = k$muted) else
    c(`1945-59` = "#3987e5", `1990+` = "#b7d3f6", context = k$muted)
  ggplot(d, aes(x, pct, colour = series, group = cohort)) +
    geom_line(data = ctx, linewidth = 0.45, lineend = "round", show.legend = FALSE) +
    qz_band(data = foc, aes(ymin = lo, ymax = hi, fill = series), mode = mode) +
    qz_line(data = foc, show.legend = FALSE) +
    qz_dots(mode, ctx, aes(x, pct), colour_aes = "series", size = 2) +
    qz_dots(mode, foc, aes(x, pct), colour_aes = "series") +
    qz_end_labels(d, "x", "pct", "label", "cohort", mode, min_gap = 2.6, nudge_x = 1.2) +
    scale_colour_manual(values = cols, guide = "none") +
    scale_fill_manual(values = cols, breaks = c(focus, "context"),
                      labels = c(cohort_lab[focus], context = tr("other cohorts", "autres cohortes")),
                      guide = guide_legend(order = 1)) +
    scale_x_continuous(breaks = years, limits = c(2011.3, 2026.4), expand = expansion(0)) +
    scale_y_continuous(breaks = seq(20, 60, 10), labels = qz_pct_axis) +
    coord_cartesian(ylim = c(20, 66), expand = FALSE) +
    labs(x = NULL, y = tr("Would vote Yes (among Yes and No)", "Voterait Oui (parmi Oui et Non)")) +
    theme_qes(mode)
}

tab <- d[order(d$year, match(d$group, cohorts)), ]
tab_cohorts <- setNames(data.frame(
  tab$year, cohort_lab[tab$group], qz_cell(tab$pct, tab$lo, tab$hi, tab$n), tab$n,
  qz_weight_text(tab$weighted), row.names = NULL
), c(tr("Election", "Élection"), tr("Cohort", "Cohorte"), tr("Yes, % [95% CI]", "Oui, % [IC à 95 %]"),
     "n", tr("Weighting", "Pondération")))
alt <- sprintf(tr(
  "Line chart of the share of francophones who would vote Yes to Quebec becoming an independent country, for five birth cohorts, at the elections of 2012, 2014, 2018 and 2022; the cohorts born 1945-1959 and in 1990 or after are highlighted with confidence bands. Those born in 1990 or after go from %s to %s; those born 1945-1959 stay between %s%% and %s%%. Values in the table view.",
  "Graphique linéaire de la part des francophones qui voteraient Oui à ce que le Québec devienne un pays indépendant, pour cinq cohortes de naissance, aux élections de 2012, 2014, 2018 et 2022 ; les cohortes nées en 1945-1959 et en 1990 ou après sont mises en évidence avec des bandes de confiance. Les personnes nées en 1990 ou après passent de %s à %s ; celles nées en 1945-1959 restent entre %s %% et %s %%. Valeurs dans la vue en tableau."),
  qz_pct(y("qes2012", "1990+"), 0), qz_pct(y("qes2022", "1990+"), 0), r4559[1], r4559[2])
qz_figure(fig_cohorts, "cohorts",
  title = sprintf(tr("Francophones born in 1990 or after went from %s to %s Yes; those born 1945-59 stayed between %s%% and %s%%",
                     "Les francophones nés en 1990 ou après sont passés de %s à %s de Oui ; ceux nés de 1945 à 1959 sont restés entre %s et %s %%"),
                  qz_pct(y("qes2012", "1990+"), 0), qz_pct(y("qes2022", "1990+"), 0), r4559[1], r4559[2]),
  subtitle = tr("Francophones who would vote Yes to an independent country, by birth cohort, 2012 to 2022; two cohorts with 95% confidence bands, the others in grey",
                "Francophones qui voteraient Oui à un pays indépendant, selon la cohorte de naissance, de 2012 à 2022 ; deux cohortes avec bandes de confiance à 95 %, les autres en gris"),
  alt = alt, table = tab_cohorts, height = 4.2,
  note = sprintf(tr("Source: Quebec Election Studies 2012, 2014, 2018 and 2022, respondents whose mother tongue is French, weighted. The line of the oldest cohort rests on fewer than 100 respondents in %s. The youngest cohort grows as new voters come of age: born 1990-1994 (aged 18 to 22) in 2012, born 1990-2004 in 2022.",
                    "Source : Études électorales québécoises de 2012, 2014, 2018 et 2022, répondants de langue maternelle française, données pondérées. La ligne de la cohorte la plus âgée repose sur moins de 100 répondants en %s. La cohorte la plus jeune s'agrandit à mesure que de nouveaux électeurs atteignent l'âge de voter : nés en 1990-1994 (18 à 22 ans) en 2012, nés en 1990-2004 en 2022."),
                 knitr::combine_words(small, and = tr(" and ", " et "), oxford_comma = FALSE)))
```

```{r answers, echo = FALSE}
ans <- do.call(rbind, lapply(c("qes2012", "qes2022"), function(s) {
  qz_tag(qz_share(subset(design[[s]], francophone & cohort %in% "1990+"), "answer"), study = s)
}))
a <- function(s, level) ans$pct[ans$study == s & ans$level == level]
```

Comparing the same cohorts from one study to the next shows where the
reversal comes from. Those born from 1945 to 1959, who voted in the
referendums of 1980 and 1995, barely moved: between `r qz_pct(r4559[1], 0)`
and `r qz_pct(r4559[2], 0)` would vote Yes at each election from 2012 to
2022. By contrast, the cohort born from 1975 to 1989 fell from
`r qz_pct(y("qes2012", "1975-89"), 0)` in 2012 to
`r qz_pct(y("qes2014", "1975-89"), 0)` as early as 2014, the election in
which the PQ vote of millennials fell (Mahéo and Bélanger 2018). It stood at
`r qz_pct(y("qes2018", "1975-89"), 0)` in 2018 and
`r qz_pct(y("qes2022", "1975-89"), 0)` in 2022, never close to its 2012
level. The youngest cohort broke later. Still at
`r qz_pct(y("qes2014", "1990+"), 0)` in 2014, it was at
`r qz_pct(y("qes2018", "1990+"), 0)` in 2018 and
`r qz_pct(y("qes2022", "1990+"), 0)` in 2022.

One could object that the young did not turn against independence, but away
from the question, and that the share of Yes among those who answered Yes
or No hides a rise in indecision. The data offer little support for that
reading. Among all francophones born in 1990 or after who were asked the
question, the share who would vote No rose from `r qz_pct(a("qes2012", "no"), 0)`
in 2012 to `r qz_pct(a("qes2022", "no"), 0)` in 2022, while the share who
answered neither Yes nor No went from `r qz_pct(a("qes2012", "neither"), 0)`
to `r qz_pct(a("qes2022", "neither"), 0)`. Indecision grew, but most of the
movement went to the No side.

## Replacement or change within cohorts?

The reversal could come from two places. The first is replacement: the
oldest cohorts shrink, new voters come of age, and the weight of each
generation among those who take a side changes even if no cohort changes its
answer. The second is change within cohorts: a given cohort answers
differently from one election to the next. To separate the two, we use a
simple counterfactual: we apply each cohort's 2012 level of support to the
cohort sizes of 2022, and the reverse.

```{r fig-turnover, echo = FALSE}
row_lab <- c(obs12 = tr("2012, observed", "2012, observé"),
             turnover = tr("2012 levels, 2022 generations\n(replacement only)",
                           "Niveaux de 2012, générations de 2022\n(remplacement seulement)"),
             within = tr("2022 levels, 2012 generations\n(change within cohorts only)",
                         "Niveaux de 2022, générations de 2012\n(changement dans les cohortes seulement)"),
             obs22 = tr("2022, observed", "2022, observé"))
cf$label <- factor(row_lab[cf$row], levels = rev(row_lab))
cf$value <- qz_pct(cf$pct, 1)

fig_turnover <- function(mode) {
  k <- qz_tokens[[mode]]
  ggplot(cf, aes(pct, label)) +
    qz_cih(aes(xmin = lo, xmax = hi), colour = k$yes) +
    geom_point(aes(shape = kind), size = 3.2, stroke = 0.8, fill = k$yes, colour = k$surface) +
    geom_text(aes(label = value), nudge_y = 0.3, size = 3.2, colour = k$ink, family = qz_font()) +
    scale_shape_manual(values = c(observed = 21, counterfactual = 23),
                       labels = c(observed = tr("observed", "observé"),
                                  counterfactual = tr("recombined", "recombiné")),
                       breaks = c("observed", "counterfactual"),
                       guide = guide_legend(override.aes = list(fill = k$ink2, colour = k$surface))) +
    scale_x_continuous(limits = c(34, 60), breaks = seq(35, 55, 5), labels = qz_pct_axis, expand = expansion(0)) +
    labs(x = tr("Would vote Yes (among Yes and No)", "Voterait Oui (parmi Oui et Non)"), y = NULL) +
    theme_qes(mode, grid = "x")
}
tab_turnover <- setNames(data.frame(
  row_lab[cf$row], qz_cell(cf$pct, cf$lo, cf$hi, rep(Inf, nrow(cf))), row.names = NULL
), c(tr("Estimate", "Estimation"), tr("Yes, % [95% CI]", "Oui, % [IC à 95 %]")))
alt <- sprintf(tr(
  "Dot chart with confidence intervals of the share of francophones who would vote Yes to an independent country: %s observed in 2012, %s if each cohort had kept its 2012 level with the cohort sizes of 2022, %s with the 2022 levels and the cohort sizes of 2012, and %s observed in 2022. Values in the table view.",
  "Graphique à points avec intervalles de confiance de la part des francophones qui voteraient Oui à un pays indépendant : %s observé en 2012, %s si chaque cohorte avait gardé son niveau de 2012 avec le poids des cohortes de 2022, %s avec les niveaux de 2022 et le poids des cohortes de 2012, et %s observé en 2022. Valeurs dans la vue en tableau."),
  cf$value[1], cf$value[2], cf$value[3], cf$value[4])
qz_figure(fig_turnover, "turnover",
  title = sprintf(tr("Most of the decline came from within the cohorts: the 2022 levels alone take support from %s to %s",
                     "L'essentiel du recul vient de l'intérieur des cohortes : les niveaux de 2022 suffisent à faire passer l'appui de %s à %s"),
                  qz_pct(v[["obs12"]], 1), qz_pct(v[["within"]], 1)),
  subtitle = tr("Francophones who would vote Yes to an independent country, observed in 2012 and 2022 and recombined from the two studies, with 95% confidence intervals",
                "Francophones qui voteraient Oui à un pays indépendant, observés en 2012 et en 2022 et recombinés à partir des deux études, avec intervalles de confiance à 95 %"),
  alt = alt, table = tab_turnover, height = 3.2,
  note = tr("Source: Quebec Election Studies 2012 and 2022, respondents whose mother tongue is French, weighted. The recombined estimates weight the share of Yes in each of the five birth cohorts of one study by the size of the cohorts in the other study. Cohort sizes are each cohort's share of the francophones who answered Yes or No in that study (weighted). The youngest cohort (born in 1990 or after) takes its 2012 level, that of voters aged 18 to 22, in the replacement-only row, and its 2022 level, which includes voters born from 1995 to 2004, in the row of change within cohorts.",
            "Source : Études électorales québécoises de 2012 et de 2022, répondants de langue maternelle française, données pondérées. Les estimations recombinées pondèrent la part de Oui dans chacune des cinq cohortes de naissance d'une étude par le poids des cohortes dans l'autre étude. Le poids d'une cohorte est sa part des francophones ayant répondu Oui ou Non dans cette étude (données pondérées). La cohorte la plus jeune (née en 1990 ou après) reçoit son niveau de 2012, celui des électeurs de 18 à 22 ans, dans la ligne du remplacement seul, et son niveau de 2022, qui inclut les électeurs nés de 1995 à 2004, dans la ligne du changement dans les cohortes."))
```

Among francophones, support for independence fell by
`r qz_num(decline, 1)` points between 2012 and 2022, from
`r qz_pct(v[["obs12"]], 1)` to `r qz_pct(v[["obs22"]], 1)`. Had each cohort
kept its 2012 level, the new weight of the generations would have left
support at `r qz_pct(v[["turnover"]], 1)`, practically unchanged. This is
because in 2012 the larger cohorts were close to one another, so it made
little difference which of them grew and which shrank. Conversely, the
levels of 2022 applied to the cohort sizes of 2012 already give
`r qz_pct(v[["within"]], 1)`. In other words, most of the decline happened
within the cohorts. The same holds when we follow only francophones born
from 1990 to 1994, who were 18 to 22 years old in 2012, a closed group that
no newcomer joins: `r qz_pct(c9094("qes2012"), 0)` would have voted Yes in
2012 (n = `r c9094("qes2012", "n")`), compared to
`r qz_pct(c9094("qes2022"), 0)` in 2022 (n = `r c9094("qes2022", "n")`). The
interval of the latter is wide, from `r qz_pct(c9094("qes2022", "lo"), 0)`
to `r qz_pct(c9094("qes2022", "hi"), 0)`, but it does not overlap that of
2012 (`r qz_pct(c9094("qes2012", "lo"), 0)` to
`r qz_pct(c9094("qes2012", "hi"), 0)`).

That said, the two mechanisms are not independent. Now that the young are
less likely than their elders to vote Yes, replacement works against the Yes
side, as the caveat of Daoust and Gareau-Paquette (2024) suggested: with the
levels of 2022, the change in the weight of the generations since 2012
lowers support by `r qz_num(repl22, 1)` of the `r qz_num(decline, 1)` points
on its own.

## Conclusion

Overall, each of the two views gets part of the story right. The older view
held in 2007, when the cohort born from 1975 to 1989 was the most likely to
vote Yes on the 1995 question. It no longer held in 2012, when those born in
1990 or after were level with the 1945-59 cohort, and by 2022 the order had
reversed. The newer view is right about the level: support for independence
is now lowest among the young and highest among those born from 1945 to
1959. It is less right about the mechanism. The decline since 2012 came
mostly from net change within cohorts: depending on the year of reference,
replacement accounts for between none and `r qz_num(repl22, 1)` of the
`r qz_num(decline, 1)` points lost, and it began to work against the Yes
side only once the young had changed. Nor do the young form a
single block. Daoust and Gareau-Paquette (2026) find that generation Z
worries more about the French language than millennials do, and the
youngest cohort on this page mixes the two.

The analysis has three limits. First, the samples are modest. Each study
counts between `r qz_int(min(n_fr))` and `r qz_int(max(n_fr))` francophones
who answered Yes or No, so a single cohort in a single year rests on at most
about `r qz_int(round(n_cell, -2))` respondents, and the oldest on far fewer.
Second, the studies do not follow the same people: we observe cohorts, not
individuals, changing. They are independent samples, drawn at different
moments (2022 during the campaign, the others after the election) and in
different ways (by telephone in 2008, online from 2012 on), so change within
a cohort also carries these differences. The youngest cohort of 2022 also
includes voters who were not yet adults in 2012, which is why we followed
those born from 1990 to 1994 on their own. Third, the page describes
cohorts; it does not explain why they changed. The period since 2014 also
saw the rise of the Coalition avenir Québec (CAQ) and Québec solidaire and,
in 2022, of the Parti conservateur du Québec, as well as new issues. A
cohort that turned away from independence in its twenties can come back to
it, and the next Quebec Election Studies will show whether the reversal is
a lasting trait of the younger generations or a moment in their political
life.

## Going further

- Daoust, Jean-François, and Thomas Gareau-Paquette. 2024. "Is Quebec
  independence still key in making sense of Canadian elections? A
  longitudinal analysis (2000–2021)." *Regional & Federal Studies* 34 (5):
  781–806. <https://doi.org/10.1080/13597566.2023.2233422>
- Daoust, Jean-François, and Thomas Gareau-Paquette. 2026. "De plus en plus
  inquiets : qui sont les Québécois qui considèrent le français comme
  menacé et quelles en sont les conséquences électorales?" *Revue
  canadienne de science politique / Canadian Journal of Political Science*
  59 (1): 61–80. <https://doi.org/10.1017/S0008423925100942>
- Dufresne, Yannick, Charles Tessier, and Eric Montigny. 2019.
  "Generational and life-cycle effects on support for Quebec independence."
  *French Politics* 17 (1): 50–63. <https://doi.org/10.1057/s41253-019-00083-9>
- Mahéo, Valérie-Anne, and Éric Bélanger. 2018. "Is the Parti Québécois
  bound to disappear? A study of the current generational dynamics of
  electoral behaviour in Quebec." *Canadian Journal of Political Science*
  51 (2): 335–356. <https://doi.org/10.1017/S0008423917001147>
- Yale, François, and Claire Durand. 2011. "What did Quebeckers want?
  Impact of question wording, constitutional proposal and context on
  support for sovereignty, 1976–2008." *American Review of Canadian
  Studies* 41 (3): 242–258. <https://doi.org/10.1080/02722011.2011.594517>

## About the data

- **Studies.** One Quebec Election Study per election, 2007 to 2022. The
  1998 polls did not ask the year of birth, and the panels that interviewed
  at the same elections are left out, so that each election counts once.
- **Question.** The share of Yes among those who would vote Yes or No, in
  a referendum on Quebec becoming an independent country (2012 to 2022) or
  on the 1995 question, which offered a partnership with Canada (2007,
  2008; those who did not know were asked which way they leaned). Those who
  would not vote, did not know or refused are left out of the share, except
  in the paragraph on indecision.
- **Cohorts and language.** Birth cohorts come from the year of birth.
  Francophones are respondents whose mother tongue is French. In 2007, 2014
  and 2022, those who gave two mother tongues are left out (the 2012 and
  2018 files have no such answer). The 2018 study also interviewed people
  aged 16 and 17; they are kept, in the youngest cohort.
- **Timing and weights.** The 2022 study asked the question during the
  campaign, the others after the election. Each study is weighted, except
  the 2008 study, which is unweighted (hollow dots).
- **Reproducing the page.** The data come from one call:

```{r show-harmonize, ref.label = "harmonize", eval = FALSE}
```
