---
title: "Study catalog"
params:
  lang: "en"
---

*[Version française](fr-etudes.html)*

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE, echo = FALSE)
# messages in the language of the page, whatever the locale of the build
options(qesR.lang = params$lang)
```

Every study qesR can load: its design, population, size, licence and
documents. `qes_studies()` and `qes_docs()` return the same information.

```{r catalog}
library(qesR)
fr <- identical(params$lang, "fr")
pick <- function(en, fr_text) if (fr) fr_text else en
dir <- system.file("extdata", "catalog", package = "qesR", mustWork = TRUE)
read <- function(name) {
  utils::read.csv(file.path(dir, name), colClasses = "character", encoding = "UTF-8", na.strings = "")
}
enums <- read("enums.csv")
label <- function(enum, value) {
  e <- enums[enums$enum == enum, , drop = FALSE]
  e[[pick("label_en", "label_fr")]][match(value, e$value)]
}
files <- read("files.csv")
studies <- qes_studies()
docs <- qes_docs()
data_file <- files[match(paste(studies$study, studies$data_file_id),
                         paste(files$study, files$file_id)), ]
number <- function(x) format(as.integer(x), big.mark = pick(",", "\u00a0"))
years <- ifelse(studies$year_end == studies$year, studies$year,
                paste0(studies$year, "-", studies$year_end))
title <- studies[[pick("title_en", "title_fr")]]
yes_no <- function(x) ifelse(x, pick("yes", "oui"), pick("no", "non"))
colon <- pick(": ", "\u00a0: ")
semi <- pick("; ", "\u00a0; ")
# facts quoted in the prose below, from the catalog as well
enum_list <- function(x) {
  if (length(x) < 2L) return(x)
  paste(paste(x[-length(x)], collapse = ", "), x[length(x)], sep = pick(" and ", " et "))
}
other_families <- enum_list(unique(label("family", studies$family[studies$family != "qes"])))
```

```{r overview}
knitr::kable(data.frame(
  code = sprintf("[`%s`](#%s)", studies$study, studies$study),
  year = years,
  family = label("family", studies$family),
  title = sprintf("[%s](%s)", title, studies$doi_url),
  design = label("study_design", studies$study_design),
  population = studies[[pick("target_population_en", "target_population_fr")]],
  n = number(data_file$n_rows),
  licence = studies$licence,
  codebook = yes_no(as.logical(studies$metadata_shipped))
), col.names = pick(
  c("Code", "Year", "Family", "Study (DOI)", "Design", "Population", "n", "Licence", "Offline codebook"),
  c("Code", "Année", "Famille", "Étude (DOI)", "Devis", "Population", "n", "Licence", "Livre de codes hors ligne")
), align = "llllllrll")
```

`n` is the number of respondents in the data file. Not every study is a
Quebec Election Study: the studies of the families `r other_families` are
listed under their own titles, and their designs and populations differ, so
check both before comparing studies. The licences, and the attribution the
2022 study requires, are on [Citing qesR and the studies](citations.html).

```{r codes}
spec <- qes_spec("spec")$tables
covered <- unique(spec$waves$study)
covered <- studies$study[studies$study %in% covered]
parts_1998 <- setdiff(studies$study[startsWith(studies$study, "qes1998")], "qes1998")
validated <- unique(spec$weights$study[spec$weights$recommended %in% TRUE &
                                       spec$weights$status == "reviewed" &
                                       spec$weights$study %in% covered])
```

## Study codes

`qes_studies()` lists `r nrow(studies)` study codes. `qes1998` is the panel
file that combines the two 1998 polls, which also load on their own
(`r paste0("\x60", parts_1998, "\x60", collapse = " and ")`). The harmonized
data of `qes_harmonize()` and `qes_decon()` cover `r length(covered)` of the
codes: every study but the two 1998 polls taken separately.

The 1998 polls surveyed francophones only, CREATEC by mother tongue and
CROP by language of use. The files have no mother-tongue question that
qesR harmonizes for 1998, so the examples count every 1998 respondent as
francophone.

## Weights {#weights}

Each wave of a harmonized study has one recommended weight, or none. Only a
weight checked against the study's documentation and its file is used:
`r length(validated)` of the `r length(covered)` studies have one. In the
others, the weight columns of `qes_harmonize()` are `NA`, and the examples
draw their estimates as hollow dots, marked "unweighted: no validated
weight". `qes_spec("spec")$tables$weights` gives every weight of every
study, with its source.

```{r weights}
w <- as.data.frame(spec$weights)
w <- w[w$study %in% covered, ]
w$key <- paste(w$study, w$wave)
keys <- unique(w$key)
rows <- lapply(keys, function(k) {
  x <- w[w$key == k, ]
  rec <- x[x$recommended %in% TRUE, ]
  if (nrow(rec)) {
    status <- if (all(rec$status == "reviewed")) {
      pick("weighted: validated, in the weight columns of `qes_harmonize()`",
           "pondéré : validée, dans les colonnes de pondération de `qes_harmonize()`")
    } else {
      pick("unweighted: weight not validated (weight columns `NA`)",
           "non pondéré : pondération non validée (colonnes de pondération à `NA`)")
    }
    data.frame(study = x$study[1], wave = x$wave[1], weight = paste0("`", rec$weight_var, "`", collapse = ", "),
               status = status, calibrated = rec$calibrated_on[1])
  } else {
    data.frame(study = x$study[1], wave = x$wave[1], weight = pick("none recommended", "aucune recommandée"),
               status = pick(paste0("unweighted: the file's weights (", paste0("`", x$weight_var, "`", collapse = ", "),
                                    ") are calibrated on the reported vote"),
                             paste0("non pondéré : les pondérations du fichier (", paste0("`", x$weight_var, "`", collapse = ", "),
                                    ") sont calées sur le vote déclaré")),
               calibrated = x$calibrated_on[1])
  }
})
tab_w <- do.call(rbind, rows)
tab_w$wave[tab_w$wave == "*"] <- pick("every poll", "chaque sondage")
tab_w$study <- sprintf("[`%s`](#%s)", tab_w$study, tab_w$study)
tab_w$calibrated <- gsub(";", "; ", tab_w$calibrated, fixed = TRUE)
knitr::kable(tab_w, col.names = pick(
  c("Study", "Wave", "Recommended weight", "Status", "Calibrated on (as documented)"),
  c("Étude", "Vague", "Pondération recommandée", "Statut", "Calée sur (selon la documentation, en anglais)")
), row.names = FALSE)
```

## By study

The Deposit line links to the deposit's DOI. The document links download the file
from Dataverse; `qes_download()` saves them in a folder you choose.

```{r by-study, results = "asis"}
for (i in seq_len(nrow(studies))) {
  s <- studies[i, ]
  f <- data_file[i, ]
  d <- docs[docs$study == s$study, ]
  cat(sprintf("### `%s`%s%s {#%s}\n\n", s$study, colon, title[i], s$study))
  cat(sprintf("- **%s**%s%s%s%s%s%s\n", pick("Study", "Étude"), colon,
              label("family", s$family), semi, label("study_design", s$study_design), semi,
              s[[pick("target_population_en", "target_population_fr")]]))
  cat(sprintf("- **%s**%s[%s](%s), %s, version %s%slicence [%s](%s)\n", pick("Deposit", "Dépôt"),
              colon, s$doi_url, s$doi_url, s$publisher, s$dataset_version, semi,
              label("licence", s$licence), s$licence_url))
  cat(sprintf("- **%s**%s`%s` (%s), %s %s\n",
              pick("Data file", "Fichier de données"), colon,
              f$original_file_name, label("format", f$format), number(f$n_rows),
              pick("rows", "lignes")))
  cat(sprintf("- **%s**%s%s\n", pick("Offline codebook", "Livre de codes hors ligne"), colon,
              if (as.logical(s$metadata_shipped)) {
                paste0(pick("yes, shipped with qesR (`qes_codebook()`, `qes_search()`)",
                            "oui, livré avec qesR (`qes_codebook()`, `qes_search()`)"),
                       if (s$licence != "CC0 1.0") pick(", under the study's licence", ", sous la licence de l'étude") else "")
              } else {
                pick("no; `qes_codebook()` builds it from the data file",
                     "non ; `qes_codebook()` le construit à partir du fichier de données")
              }))
  note <- s[[pick("notes_en", "notes_fr")]]
  if (!is.na(note) && nzchar(note)) {
    cat(sprintf("- **%s**%s%s\n", "Note", colon, note))
  }
  if (nrow(d)) {
    cat(sprintf("- **%s** (`qes_docs(\"%s\")`)%s\n", "Documents", s$study, colon))
    what <- paste0(label("role", d$role),
                   ifelse(is.na(d$lang), "", paste0(", ", label("lang", d$lang))),
                   ", ", toupper(d$format))
    cat(sprintf("    - [%s](%s)%s%s\n", d$file_name, d$url, colon, what), sep = "")
  }
  cat(sprintf("- **%s**%s`qes_cite(\"%s\")`\n\n", pick("Cite", "Citer"), colon, s$study))
}
```

## Citing

`qes_cite()` gives the citation of each study; see
[Citing qesR and the studies](citations.html).
