---
title: "Study catalog"
params:
  lang: "en"
---

Every study that qesR can load, with the facts qesR pins for it. This page is
generated when the site is built, from the catalog that ships with the
package (`inst/extdata/catalog/`): the same data as `qes_studies()` and
`qes_docs()`, with no hand-written copy. [Version française](fr-etudes.html).

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", echo = FALSE, warning = FALSE)
```

```{r catalog}
library(qesR)
lang <- params$lang
fr <- identical(lang, "fr")
dir <- system.file("extdata", "catalog", package = "qesR", mustWork = TRUE)
read <- function(name) {
  utils::read.csv(file.path(dir, name), colClasses = "character", encoding = "UTF-8", na.strings = "")
}
enums <- read("enums.csv")
label <- function(enum, value) {
  e <- enums[enums$enum == enum, , drop = FALSE]
  e[[if (fr) "label_fr" else "label_en"]][match(value, e$value)]
}
files <- read("files.csv")
studies <- qes_studies()
docs <- qes_docs()
pick <- function(en, fr_text) if (fr) fr_text else en
data_rows <- files$n_rows[match(
  paste(studies$study, studies$data_file_id),
  paste(files$study, files$file_id)
)]
years <- ifelse(studies$year_end == studies$year, studies$year, paste0(studies$year, "-", studies$year_end))
```

## Studies

```{r studies}
knitr::kable(data.frame(
  code = paste0("`", studies$study, "`"),
  year = years,
  title = sprintf("[%s](%s)", studies[[pick("title_en", "title_fr")]], studies$doi_url),
  design = label("study_design", studies$study_design),
  population = studies[[pick("target_population_en", "target_population_fr")]],
  n = format(as.integer(data_rows), big.mark = pick(",", "\u00a0")),
  licence = studies$licence,
  version = studies$dataset_version
), col.names = if (fr) {
  c("Code", "Année", "Étude (DOI)", "Devis", "Population", "n", "Licence", "Version")
} else {
  c("Code", "Year", "Study (DOI)", "Design", "Population", "n", "Licence", "Version")
}, align = "lllllrll")
```

Each title links to the deposit's DOI. `n` is the number of rows in the
pinned data file and *Version* the pinned dataset version. The Durand panels, the
CROP polls and the 1998 polls are not Quebec Election Studies; they are
listed under their own titles. The three 1998 codes share one deposit.

## Notes

```{r notes, results = "asis"}
notes <- studies[[pick("notes_en", "notes_fr")]]
for (i in seq_len(nrow(studies))) {
  if (!is.na(notes[i]) && nzchar(notes[i])) {
    cat(sprintf("- **`%s`**: %s\n", studies$study[i], notes[i]))
  }
}
```

## Documents

Codebooks, questionnaires and reports deposited with each study, as listed by
`qes_docs()`. The links download the file from Dataverse.

```{r docs}
knitr::kable(data.frame(
  code = paste0("`", docs$study, "`"),
  document = sprintf("[%s](%s)", docs$file_name, docs$url),
  role = label("role", docs$role),
  lang = label("lang", docs$lang),
  format = toupper(docs$format)
), col.names = if (fr) {
  c("Code", "Document", "Rôle", "Langue", "Format")
} else {
  c("Code", "Document", "Role", "Language", "Format")
})
```

## Citing

`qes_cite()` gives the citation of each study; see
[Study Citations](study-citations.html).
