---
title: "Do surveys miss the Liberals? The reported vote against the official results"
description: >
  Six Quebec Election Studies and three panels set against the results of
  Élections Québec: the Liberal gap, the over-reported PQ, what language
  and weighting explain, and how far vote intentions move before an
  election.
params:
  lang: "en"
---

*[Version française](fr-enquetes-resultats.html)*

```{r setup, include = FALSE}
source("_setup.R", encoding = "UTF-8")
```

```{r harmonize, echo = FALSE}
h <- qes_harmonize(studies = qz_studies, targets = c("vote_choice", "lang_mother", "lang_home"),
                   missing = "reasons", quiet = TRUE)
```

```{r estimates, echo = FALSE}
official <- qz_official()
xw <- qes_spec("crosswalk", targets = "vote_prov_recall")
offered <- setNames(strsplit(xw$levels_offered, ";", fixed = TRUE), xw$study)
recall_studies <- setdiff(qz_studies, c("qes_crop_2007_2010", "qes1998"))

# a study's parties: those its question listed; the official share of a
# party it did not list goes to "other", the answer its voters had to give
fold <- function(party, s) ifelse(party %in% c(offered[[s]], "other"), party, "other")
h$party <- factor(qz_party_code(h$vote_choice), levels = qz_parties)

recall <- do.call(rbind, lapply(recall_studies, function(s) {
  w <- qz_design(h, s, "weight_post")
  u <- qz_design(h, s, "weight_post", unweighted = TRUE)
  rbind(qz_tag(qz_share(w, "party"), study = s, weighting = if (attr(w, "qz_weighted")) "weighted" else "unweighted"),
        if (attr(w, "qz_weighted")) qz_tag(qz_share(u, "party"), study = s, weighting = "unweighted"))
}))
recall$year <- qz_year(recall$study)
recall$status <- qz_party_status(recall$study, recall$level)
off_study <- do.call(rbind, lapply(recall_studies, function(s) {
  o <- official[official$year == qz_year(s), ]
  o$level <- fold(o$party, s)
  o <- stats::aggregate(share_valid ~ level, o, sum)
  qz_tag(o, study = s)
}))
recall$official <- off_study$share_valid[match(paste(recall$study, recall$level), paste(off_study$study, off_study$level))]
recall$official[is.na(recall$official) & recall$status == "offered"] <- 0

# index of dissimilarity: half the sum of the absolute differences, in points
dissim <- stats::aggregate(cbind(index = abs(pct - official)) ~ study + weighting,
                           recall[recall$status == "offered", ], function(x) sum(x) / 2)
dissim$year <- qz_year(dissim$study)
rv <- function(s, p, w = "weighted") recall$pct[recall$study == s & recall$level == p & recall$weighting == w]
ov <- function(y, p) official$share_valid[official$year == y & official$party == p]
di <- function(s, w) dissim$index[dissim$study == s & dissim$weighting == w]
# the estimate a chart draws: weighted where the study has a reviewed weight, else unweighted
rb <- function(s, p) {
  x <- recall[recall$study == s & recall$level == p, ]
  x$pct[order(x$weighting != "weighted")][1]
}
gap <- function(s, p) rb(s, p) - ov(qz_year(s), p)

# the Quebec Election Studies, one per election, and the claims of the opening
qes_series <- c("qes2007", "qes2008", "qes2012", "qes2014", "qes2018", "qes2022")
early <- recall_studies[qz_year(recall_studies) <= 2014]
plq_early <- vapply(early, gap, numeric(1), p = "PLQ")
pq_qes <- vapply(qes_series, gap, numeric(1), p = "PQ")
winner <- vapply(qes_series, function(s) {
  o <- official[official$year == qz_year(s), ]
  o$party[which.max(o$share_valid)]
}, character(1))
win_gap <- mapply(gap, qes_series, winner)
win_over <- qes_series[win_gap > 0]
stopifnot(all(plq_early < 0), all(pq_qes > 0), length(win_over) == 1)
# the typical number of voters in a study, and the widest 95% margin of error it gives a party
n_typical <- stats::median(vapply(qes_series, function(s) recall$n[recall$study == s][1], numeric(1)))
moe_typical <- 100 * 1.96 * sqrt(0.25 / n_typical)
# a count with the page's thousands separator: 1,278 / 1 278
qz_int <- function(x) formatC(round(x), format = "d", big.mark = tr(",", "\u00a0"))
# the widest 95% margin of error, once weighted, that a Quebec Election Study gives a party
wq <- recall[recall$study %in% qes_series & recall$weighting == "weighted" & recall$status == "offered", ]
moe_weighted <- max((wq$hi - wq$lo) / 2)
# the PQ in every study and panel, not only the Quebec Election Studies
pq_all <- vapply(recall_studies, gap, numeric(1), p = "PQ")

# is the official result inside the 95% CI of the estimate a chart draws?
in_ci <- function(s, p) {
  x <- recall[recall$study == s & recall$level == p, ]
  x <- x[order(x$weighting != "weighted"), ][1, ]
  o <- ov(qz_year(s), p)
  x$lo <= o & o <= x$hi
}
pq_in <- qz_year(qes_series[vapply(qes_series, in_ci, logical(1), p = "PQ")])
plq_in <- qz_year(qes_series[vapply(qes_series, in_ci, logical(1), p = "PLQ")])
win_in <- qz_year(qes_series[mapply(in_ci, qes_series, winner)])
stopifnot(length(pq_in) >= 2, length(plq_in) >= 2, length(win_in) >= 2, !in_ci("qes2022", "CAQ"),
          all(win_gap[qz_year(qes_series) %in% win_in] < 0), moe_weighted > moe_typical,
          mean(pq_all > 0) > 0.5)

# page-local wording: small counts as words, a list of years, and a number of
# points with the page's plural (1 point, 7 points; 1,4 point in French)
num_word <- function(n) tr(c("one", "two", "three", "four", "five", "six", "seven", "eight", "nine")[n],
                           c("un", "deux", "trois", "quatre", "cinq", "six", "sept", "huit", "neuf")[n])
yrs_text <- function(y) {
  y <- sort(unique(y))
  if (length(y) < 2) return(as.character(y))
  paste(paste(head(y, -1), collapse = ", "), tr("and", "et"), tail(y, 1))
}
pt_unit <- function(x, digits = 0) {
  r <- abs(round(x, digits))
  if (identical(qz_lang, "fr")) ifelse(r < 2, "point", "points") else ifelse(r == 1, "point", "points")
}
pt_text <- function(x, digits = 0) paste(qz_num(abs(x), digits), pt_unit(x, digits))
pt_signed <- function(x, digits = 1) sub("pts$", pt_unit(x, digits), qz_pts(x, digits))

# the CROP polls: monthly vote intentions, among respondents who named a party
crop <- h[h$study == "qes_crop_2007_2010", ]
crop$intent <- factor(qz_party_code(crop$vote_choice), levels = qz_parties)
dc <- qz_design(crop, "qes_crop_2007_2010", "weight_pre")
polls <- qz_share(dc, "intent", by = "waves")
polls$date <- as.Date(paste0(sub("^poll_(\\d{4})_(\\d{2})$", "\\1-\\2", polls$group), "-15"))
```

In Quebec, it is almost a reflex to say that surveys underestimate the
Parti libéral du Québec (PLQ). The polls of the 1998 election, which
overestimated the Parti Québécois (PQ) and underestimated the Liberals, are
the textbook case (Durand and Blais 1999). Studies in the United States
point to a second and quite different bias: after the vote, more
respondents say they voted for the winner than the result allows (Atkeson
1999, on the presidential primaries). Both beliefs are about polls and
surveys in general. Individual studies have been checked against the result
(Bélanger et al. 2025 for 2022). However, the Quebec Election Studies,
which ask respondents how they voted once the election is over, have rarely
been compared with the official results across several elections. We do so
for six elections, from 2007 to 2022, with the six studies and the three
panels that Claire Durand and her colleagues ran in 2007, 2012 and 2018.
Overall, the first belief holds only in part and the second not at all. The
PLQ is under-reported in all `r num_word(length(early))` studies and panels
from 2007 to 2014, by `r qz_num(min(abs(plq_early)), 0)` to
`r qz_num(max(abs(plq_early)), 0)` percentage points of the valid vote, and
the PQ is over-reported in every Quebec Election Study, though sometimes by
as little as `r pt_text(min(pq_qes))`. But the Liberal gap narrows in 2018
and is slightly reversed in 2022, within the margin of error. Nor is the
winner over-reported: it falls short in `r num_word(sum(win_gap < 0))` of
the `r num_word(length(qes_series))` elections. In 2022, the Coalition
avenir Québec (CAQ) falls `r pt_text(gap("qes2022", "CAQ"))` short.

## The reported vote, party by party

```{r fig-parties, echo = FALSE}
panel_parties <- c("PLQ", "PQ", "QS", "ADQ", "CAQ")
d <- recall[recall$level %in% panel_parties & recall$status == "offered" &
              (recall$weighting == "weighted" | !recall$study %in% recall$study[recall$weighting == "weighted"]), ]
d$weighted <- d$weighting == "weighted"
# the Quebec Election Study just left of each election, the Durand panel just right
d$family <- ifelse(grepl("_panel$", d$study), "panel", "qes")
d$x <- d$year + ifelse(d$family == "panel", 0.35, ifelse(d$study %in% c("qes2007", "qes2012", "qes2018"), -0.35, 0))
d$party <- factor(d$level, levels = panel_parties)
o <- official[official$party %in% panel_parties & official$year >= 2007, ]
o$party <- factor(o$party, levels = panel_parties)
key_lab <- c(qes = tr("Quebec Election Study", "Étude électorale québécoise"),
             panel = tr("Durand panel (just right)", "Panel Durand (juste à droite)"),
             unweighted = tr("hollow: unweighted", "creux : non pondéré"))

fig_parties <- function(mode) {
  k <- qz_tokens[[mode]]
  ggplot(d, aes(x, pct)) +
    geom_line(data = o, aes(year, share_valid, group = party, linetype = "official"), colour = k$ink, linewidth = 0.6) +
    geom_segment(data = o, aes(x = year - 0.4, xend = year + 0.4, y = share_valid, yend = share_valid),
                 colour = k$ink, linewidth = 0.8) +
    geom_linerange(aes(ymin = lo, ymax = hi, colour = party), linewidth = 0.45, alpha = 0.6) +
    geom_point(data = d[d$weighted, ], aes(shape = family, fill = party), size = 2.9, stroke = 0.8, colour = k$surface) +
    geom_point(data = d[!d$weighted, ], aes(shape = family, colour = party), size = 2.5, stroke = 1.1, fill = k$surface) +
    # an invisible point that trains the "hollow" key of the shape legend
    geom_point(data = d[1, ], aes(shape = "unweighted"), alpha = 0) +
    facet_wrap(~party, ncol = 3, axes = "all_x") +
    scale_party(mode) +
    scale_linetype_manual(values = c(official = "solid"),
                          labels = c(official = tr("Official result (line and tick)", "Résultat officiel (ligne et trait)")),
                          guide = guide_legend(order = 1, override.aes = list(colour = k$ink, linewidth = 0.6))) +
    # a third, unused level keys the hollow encoding
    scale_shape_manual(values = c(qes = 21, panel = 22, unweighted = 21), limits = names(key_lab),
                       breaks = names(key_lab), labels = key_lab,
                       guide = guide_legend(order = 2, override.aes = list(
                         fill = c(k$ink2, k$ink2, k$surface), colour = c(k$surface, k$surface, k$ink2),
                         stroke = c(0.8, 0.8, 1.1), size = 3, alpha = 1))) +
    scale_x_continuous(breaks = c(2007, 2012, 2018, 2022), labels = c("’07", "’12", "’18", "’22"),
                       limits = c(2006, 2023), expand = expansion(0)) +
    scale_y_continuous(limits = c(0, 52), breaks = seq(0, 50, 10), labels = qz_pct_axis, expand = expansion(0)) +
    labs(x = NULL, y = NULL) +
    theme_qes(mode) +
    theme(panel.spacing.x = unit(22, "pt"),
          legend.position = "inside", legend.position.inside = c(0.695, 0.25),
          legend.justification.inside = c(0, 0.5), legend.direction = "vertical",
          legend.key.spacing.y = unit(4, "pt"))
}

tab <- recall[recall$status != "not_run", ]
tab <- tab[order(match(tab$level, qz_parties), tab$year, tab$study, tab$weighting), ]
tab_parties <- setNames(data.frame(
  qz_party_label(tab$level), tab$year, qz_study_label(tab$study),
  ifelse(tab$status == "not_listed", tr("n.l.", "n.p."), qz_cell(tab$pct, tab$lo, tab$hi, tab$n)),
  tab$n, qz_num(tab$official), ifelse(tab$status == "not_listed", "", qz_pts(tab$pct - tab$official, 1)),
  ifelse(tab$weighting == "weighted", tr("weighted", "pondéré"), tr("unweighted", "non pondéré")),
  row.names = NULL
), c(tr("Party", "Parti"), tr("Election", "Élection"), tr("Study", "Étude"),
     tr("Reported vote, % [95% CI]", "Vote déclaré, % [IC à 95 %]"), "n",
     tr("Official, % of valid votes", "Officiel, % des votes valides"), tr("Difference", "Écart"),
     tr("Weighting", "Pondération")))
alt <- sprintf(tr(
  "Five small charts, one per party (PLQ, PQ, QS, ADQ, CAQ): the official share of valid votes at each election from 2007 to 2022 as a line with a tick, and each study's reported vote as a dot with its confidence interval (circles: Quebec Election Studies; squares: Durand panels). The PLQ dots sit below the line from 2007 to 2014 and the PQ dots above it. In 2022 the Quebec Election Study has the CAQ at %s against %s officially. Values in the table view.",
  "Cinq petits graphiques, un par parti (PLQ, PQ, QS, ADQ, CAQ) : la part officielle des votes valides à chaque élection de 2007 à 2022 en ligne avec un trait, et le vote déclaré de chaque étude en point avec son intervalle de confiance (cercles : Études électorales québécoises ; carrés : panels Durand). Les points du PLQ sont sous la ligne de 2007 à 2014 et ceux du PQ au-dessus. L'Étude électorale québécoise de 2022 donne la CAQ à %s contre %s officiellement. Valeurs dans la vue en tableau."),
  qz_pct(rv("qes2022", "CAQ"), 0), qz_pct(ov(2022, "CAQ"), 0))
qz_figure(fig_parties, "parties",
  title = sprintf(tr("The PLQ is under-reported in every study from 2007 to 2014, the PQ over-reported, and the CAQ %s points below its result in 2022",
                     "Le PLQ est sous-estimé dans chaque étude de 2007 à 2014, le PQ surestimé, et la CAQ sous-estimée de %s points en 2022"),
                  qz_num(abs(gap("qes2022", "CAQ")), 0)),
  subtitle = tr("Reported vote in each study (dot, 95% CI) against the official share of valid votes (line and tick), by party",
                "Vote déclaré dans chaque étude (point, IC à 95 %) et part officielle des votes valides (ligne et trait), selon le parti"),
  alt = alt, table = tab_parties, height = 5.2,
  note = tr("Source: Quebec Election Studies (circles, just left of each election) and Durand panels (squares, just right), reported vote among respondents who named a party; official results of Élections Québec. Dots are weighted, with 95% confidence intervals. Hollow dots: unweighted. The table view gives every party, with the unweighted estimates.",
            "Source : Études électorales québécoises (cercles, juste à gauche de chaque élection) et panels Durand (carrés, juste à droite), vote déclaré parmi les répondants qui nomment un parti ; résultats officiels d'Élections Québec. Les points sont pondérés, avec des intervalles de confiance à 95 %. Points creux : non pondéré. La vue en tableau donne tous les partis, avec les estimations non pondérées."))
```

First, the Liberals. From 2007 to 2014 the PLQ is under-reported in all
`r num_word(length(early))` studies and panels, by as much as
`r pt_text(gap("qes2007", "PLQ"))` in 2007 and `r pt_text(gap("qes2014", "PLQ"))`
in 2014, the year it returned to power with `r qz_pct(ov(2014, "PLQ"), 0)`
of the vote. The pattern then fades. The 2018 study puts the PLQ
`r pt_text(gap("qes2018", "PLQ"))` below its result, and the 2022 study puts
it `r pt_text(gap("qes2022", "PLQ"))` above. Second, the PQ is over-reported
in every Quebec Election Study, by as little as `r pt_text(min(pq_qes))` in
`r qz_year(names(which.min(pq_qes)))` and as much as
`r pt_text(max(pq_qes))` in `r qz_year(names(which.max(pq_qes)))`
(`r qz_pct(rb("qes2012", "PQ"), 0)` reported against
`r qz_pct(ov(2012, "PQ"), 0)` official). Taken one at a time, several of
these gaps are within the margin of error: the PQ's in `r yrs_text(pq_in)`,
and the PLQ's in `r yrs_text(plq_in)`.

Third, and contrary to what the literature on post-election surveys would
lead us to expect, the winner is rarely over-reported. The only exception is
`r qz_year(win_over)`, when the `r winner[win_over]` won. In
`r yrs_text(win_in)`, the winner's shortfall is within the margin of error.
Not so in 2022: the CAQ took `r qz_pct(ov(2022, "CAQ"), 0)` of the vote, but
only `r qz_pct(rb("qes2022", "CAQ"), 0)` of the study's respondents say they
voted for it, a shortfall Bélanger et al. (2025) also report with the same
study (33.7% against 41%). Not every 2022 party is off, however: the Parti
conservateur du Québec (PCQ) is close to its result
(`r qz_pct(rb("qes2022", "PCQ"), 0)` against `r qz_pct(ov(2022, "PCQ"), 0)`),
as they also find. What stands out is the direction of the errors. From 2007
to 2014 at least, they are not random noise around the result: they lean
the same way for the same parties.

## Is the Liberal gap a language gap?

```{r language, echo = FALSE}
# mother tongue, or the language spoken at home where it is missing or mixed
lang <- ifelse(is.na(h$lang_mother), as.character(h$lang_home), as.character(h$lang_mother))
h$language <- factor(ifelse(lang %in% "French", "fr", ifelse(is.na(lang), NA, "other")), levels = c("fr", "other"))
lang_studies <- c("qes2007", "qes2012", "qes2014", "qes2018", "qes2022")

# In each weighted study, among respondents whose language is known: the
# reported vote, the vote of each language group, and the reported vote once
# non-francophones among voters get the share they have among all
# respondents (post-stratification of the voters on language)
rebal <- do.call(rbind, lapply(lang_studies, function(s) {
  d <- subset(qz_design(h, s, "weight_post"), !is.na(language))
  all <- qz_share(d, "language")
  v <- subset(d, !is.na(party))
  voters <- qz_share(v, "language")
  groups <- qz_share(v, "party", by = "language")
  pop <- data.frame(language = factor(c("fr", "other"), levels = c("fr", "other")),
                    Freq = all$pct / 100 * sum(weights(v)))
  as_reported <- qz_share(v, "party")
  rebalanced <- qz_share(survey::postStratify(v, ~language, pop), "party")
  o <- official[official$year == qz_year(s), ]
  data.frame(study = s, party = as_reported$level, reported = as_reported$pct, rebalanced = rebalanced$pct,
             lo = rebalanced$lo, hi = rebalanced$hi, n = as_reported$n,
             official = o$share_valid[match(as_reported$level, o$party)],
             fr = groups$pct[groups$group == "fr"], other = groups$pct[groups$group == "other"],
             share_all = all$pct[all$level == "other"], share_voters = voters$pct[voters$level == "other"])
}))
rebal$gap <- rebal$reported - rebal$official
rebal$gap_rb <- rebal$rebalanced - rebal$official
# the share of non-francophones among voters that would match the official
# PLQ result, given the PLQ vote of each language group in the study
rebal$needed <- 100 * (rebal$official - rebal$fr) / (rebal$other - rebal$fr)
lr <- function(s, p, col) rebal[[col]][rebal$study == s & rebal$party == p]
early_lang <- c("qes2007", "qes2012", "qes2014")
closed <- vapply(early_lang, function(s) lr(s, "PLQ", "gap_rb") - lr(s, "PLQ", "gap"), numeric(1))
plq_gap_lang <- vapply(early_lang, function(s) abs(lr(s, "PLQ", "gap")), numeric(1))
needed <- vapply(early_lang, function(s) lr(s, "PLQ", "needed"), numeric(1))
share_all <- vapply(early_lang, function(s) lr(s, "PLQ", "share_all"), numeric(1))
pq_closed <- vapply(early_lang, function(s) lr(s, "PQ", "gap") - lr(s, "PQ", "gap_rb"), numeric(1))
stopifnot(all(closed > 0), all(lr(early_lang, "PLQ", "gap_rb") < 0), all(needed > share_all),
          all(pq_closed > 0), lr("qes2018", "PLQ", "gap") < 0, lr("qes2018", "PLQ", "gap_rb") > 0)
```

The most familiar explanation for the Liberal gap is language.
Non-francophones vote Liberal in very large numbers: in 2014,
`r qz_pct(lr("qes2014", "PLQ", "other"), 0)` of the non-francophones who
reported a vote chose the PLQ, against
`r qz_pct(lr("qes2014", "PLQ", "fr"), 0)` of francophones. A survey that
reaches too few of them will under-report the party. The studies are
weighted, among other things, on language, so non-francophones should make
up roughly their share of the census population among all respondents.
Among those who report a vote, however, they are fewer:
`r qz_pct(lr("qes2014", "PLQ", "share_voters"), 0)` in 2014, against
`r qz_pct(lr("qes2014", "PLQ", "share_all"), 0)` of all respondents. We
therefore examine what happens when non-francophones among voters are given
back the share they have among all respondents. This is a generous test,
since it assumes they name a party as often as francophones do.

```{r fig-lang, echo = FALSE}
d <- rebal[rebal$party %in% c("PLQ", "PQ"), ]
d$row <- factor(qz_study_label(d$study), levels = rev(qz_study_label(lang_studies)))
d$party <- factor(d$party, levels = c("PLQ", "PQ"))
long <- rbind(transform(d, value = gap, kind = "reported"), transform(d, value = gap_rb, kind = "rebalanced"))
long$kind <- factor(long$kind, levels = c("reported", "rebalanced"))
kind_lab <- c(reported = tr("As reported", "Tel que déclaré"),
              rebalanced = tr("Non-francophones given their share of all respondents",
                              "Non-francophones ramenés à leur part de l'ensemble des répondants"))
party_lab <- c(PLQ = tr("PLQ: reported minus official", "PLQ : déclaré moins officiel"),
               PQ = tr("PQ: reported minus official", "PQ : déclaré moins officiel"))

fig_lang <- function(mode) {
  k <- qz_tokens[[mode]]
  ggplot(long, aes(value, row)) +
    geom_vline(xintercept = 0, colour = k$ink, linewidth = 0.5) +
    geom_segment(data = d, aes(x = gap, xend = gap_rb, y = row, yend = row), colour = k$muted, linewidth = 0.6) +
    geom_point(data = long[long$kind == "reported", ], aes(fill = party, shape = kind), size = 3, stroke = 0.8,
               colour = k$surface) +
    geom_point(data = long[long$kind == "rebalanced", ], aes(shape = kind), size = 2.5, stroke = 0.7,
               colour = k$surface, fill = k$ink2) +
    facet_wrap(~party, ncol = 2, labeller = as_labeller(party_lab)) +
    scale_party(mode, aesthetics = "fill") +
    scale_shape_manual(values = c(reported = 21, rebalanced = 23), breaks = names(kind_lab), labels = kind_lab,
                       guide = guide_legend(override.aes = list(fill = c(k$ink2, k$ink2), colour = k$surface,
                                                                size = 3))) +
    scale_x_continuous(limits = c(-9, 9), breaks = seq(-5, 5, 5),
                       labels = function(x) ifelse(x == 0, "0", qz_pts(x)), expand = expansion(0)) +
    labs(x = tr("Points from the official result (0)", "Points d'écart avec le résultat officiel (0)"), y = NULL) +
    theme_qes(mode, grid = "x") +
    theme(panel.spacing.x = unit(26, "pt"))
}

tab <- d[order(d$party, qz_year(d$study)), ]
tab_lang <- setNames(data.frame(
  qz_party_label(tab$party), qz_study_label(tab$study), qz_num(tab$reported), qz_num(tab$rebalanced),
  qz_num(tab$official), qz_pts(tab$gap, 1), qz_pts(tab$gap_rb, 1),
  qz_num(tab$fr), qz_num(tab$other), qz_num(tab$share_voters), qz_num(tab$share_all),
  ifelse(tab$party == "PLQ", qz_num(tab$needed), ""), tab$n, row.names = NULL
), c(tr("Party", "Parti"), tr("Study", "Étude"), tr("Reported, %", "Déclaré, %"),
     tr("Rebalanced, %", "Rééquilibré, %"), tr("Official, %", "Officiel, %"),
     tr("Gap as reported", "Écart déclaré"), tr("Gap rebalanced", "Écart rééquilibré"),
     tr("Among francophones, %", "Chez les francophones, %"), tr("Among non-francophones, %", "Chez les non-francophones, %"),
     tr("Non-francophones among voters, %", "Non-francophones parmi les votants, %"),
     tr("Non-francophones among all respondents, %", "Non-francophones parmi tous les répondants, %"),
     tr("Non-francophone share that would match the PLQ result, %", "Part de non-francophones qui donnerait le résultat du PLQ, %"),
     "n"))
alt <- sprintf(tr(
  "Two dot charts side by side, one row per Quebec Election Study from 2007 to 2022: the PLQ's and the PQ's reported vote minus the official result, as reported (coloured dot) and once non-francophones among voters are given their share of all respondents (grey diamond). For the PLQ in 2014 the gap goes from %s to %s; in 2007 and 2012 the rebalancing also closes little of the gap. Values in the table view.",
  "Deux graphiques à points côte à côte, une rangée par Étude électorale québécoise de 2007 à 2022 : le vote déclaré du PLQ et du PQ moins le résultat officiel, tel que déclaré (point de couleur) et une fois les non-francophones parmi les votants ramenés à leur part de l'ensemble des répondants (losange gris). Pour le PLQ en 2014, l'écart passe de %s à %s ; en 2007 et en 2012, le rééquilibrage comble aussi peu de l'écart. Valeurs dans la vue en tableau."),
  qz_pts(lr("qes2014", "PLQ", "gap"), 1), qz_pts(lr("qes2014", "PLQ", "gap_rb"), 1))
qz_figure(fig_lang, "language",
  title = sprintf(tr("Giving non-francophones their full weight closes %s to %s points of a Liberal gap of %s to %s",
                     "Donner tout leur poids aux non-francophones comble de %s à %s point d'un écart libéral de %s à %s points"),
                  qz_num(min(closed)), qz_num(max(closed)), qz_num(min(plq_gap_lang), 0), qz_num(max(plq_gap_lang), 0)),
  subtitle = tr("Reported vote minus the official result, as reported and with non-francophones among voters given their share of all respondents, in the weighted Quebec Election Studies",
                "Vote déclaré moins le résultat officiel, tel que déclaré et avec les non-francophones parmi les votants ramenés à leur part de l'ensemble des répondants, dans les Études électorales québécoises pondérées"),
  alt = alt, table = tab_lang, height = 3.4,
  note = tr("Source: Quebec Election Studies, weighted, respondents whose mother tongue (or, where it is missing or mixed, home language) is known; official results of Élections Québec. Francophones: French mother tongue. Rebalanced: the respondents who reported a vote are weighted again so that non-francophones make up the same share as among all respondents; the vote of each language group is unchanged. The 2008 study, whose weights are calibrated on the vote, is left out. The table view gives the vote of each group and the share of non-francophones among voters that would have matched the PLQ's result.",
            "Source : Études électorales québécoises, pondérées, répondants dont la langue maternelle (ou, si elle manque ou est multiple, la langue parlée à la maison) est connue ; résultats officiels d'Élections Québec. Francophones : langue maternelle française. Rééquilibré : les répondants qui déclarent un vote sont pondérés de nouveau pour que les non-francophones y aient la même part que parmi l'ensemble des répondants ; le vote de chaque groupe linguistique ne change pas. L'étude de 2008, dont la pondération est calée sur le vote, est laissée de côté. La vue en tableau donne le vote de chaque groupe et la part de non-francophones parmi les votants qui aurait donné le résultat du PLQ."))
```

The answer is fairly clear: language accounts for a small part of the gap.
In the `r num_word(length(early_lang))` weighted Quebec Election Studies of
`r yrs_text(qz_year(early_lang))`, where the Liberal gap is
`r qz_num(min(plq_gap_lang), 0)` to `r qz_num(max(plq_gap_lang), 0)` points,
rebalancing closes `r qz_num(min(closed))` to `r qz_num(max(closed))` points
of it, and trims the PQ's excess by `r qz_num(min(pq_closed))` to
`r qz_num(max(pq_closed))` points. To close the PLQ gap through language
alone, non-francophones would have had to make up
`r qz_pct(min(needed), 0)` to `r qz_pct(max(needed), 0)` of voters, against
`r qz_pct(min(share_all), 0)` to `r qz_pct(max(share_all), 0)` of all
respondents. In other words, the Liberal gap is not only a matter of too few
non-francophones. This suggests that it is also a gap within language
groups, most likely among francophones: in these studies, either Liberal
voters are missing from the samples, or they do not say they voted Liberal.
In 2018, by contrast, the same correction takes the PLQ slightly past its
result (`r pt_signed(lr("qes2018", "PLQ", "gap_rb"), 1)`). The gap was small
that year, and language composition may account for most of it.

## Does weighting help?

```{r fig-dissim, echo = FALSE}
# the studies with both a weighted and an unweighted estimate
both <- intersect(dissim$study[dissim$weighting == "weighted"], dissim$study[dissim$weighting == "unweighted"])
both <- both[order(qz_year(both), both)]
w_index <- dissim$index[dissim$weighting == "weighted"]
stopifnot(sum(w_index > di("qes2008", "unweighted")) > length(w_index) / 2,
          sum(w_index > di("qes2007_panel", "unweighted")) > length(w_index) / 2)
d <- dissim[dissim$study %in% both, ]
d$row <- factor(qz_study_label(d$study), levels = rev(qz_study_label(both)))
wide <- merge(d[d$weighting == "weighted", c("row", "index")], d[d$weighting == "unweighted", c("row", "index")],
              by = "row", suffixes = c("_w", "_u"))
wide$change <- qz_pts(wide$index_w - wide$index_u, 1)
wide$xmax <- pmax(wide$index_w, wide$index_u)
d$weighting <- factor(d$weighting, levels = c("weighted", "unweighted"))

fig_dissim <- function(mode) {
  k <- qz_tokens[[mode]]
  ggplot(d, aes(index, row)) +
    geom_segment(data = wide, aes(x = index_u, xend = index_w, y = row, yend = row),
                 colour = k$muted, linewidth = 0.6) +
    # the unweighted ring is drawn larger, so it stays visible around an equal weighted dot
    geom_point(data = d[d$weighting == "unweighted", ], aes(shape = weighting), size = 3.9, stroke = 1.1,
               colour = k$ink2, fill = k$surface) +
    geom_point(data = d[d$weighting == "weighted", ], aes(shape = weighting), size = 2.5, stroke = 0.8,
               colour = k$surface, fill = k$ink) +
    geom_text(data = wide, aes(x = 12.6, y = row, label = change), hjust = 0, size = 3.1, colour = k$ink,
              family = qz_font()) +
    annotate("text", x = 12.6, y = length(both) + 0.75, label = tr("weighted minus\nunweighted", "pondéré moins\nnon pondéré"),
             hjust = 0, vjust = 0, size = 2.7, colour = k$ink2, family = qz_font(), lineheight = 0.9) +
    scale_shape_manual(values = c(weighted = 21, unweighted = 21), limits = c("weighted", "unweighted"),
                       breaks = c("weighted", "unweighted"),
                       labels = c(tr("filled: weighted", "plein : pondéré"), tr("hollow: unweighted", "creux : non pondéré")),
                       guide = guide_legend(override.aes = list(fill = c(k$ink, k$surface), colour = c(k$surface, k$ink2),
                                                                size = c(2.6, 3.2), stroke = c(0.8, 1.1)))) +
    scale_y_discrete(limits = levels(d$row), expand = expansion(add = c(0.6, 1.2))) +
    scale_x_continuous(limits = c(0, 15.5), breaks = seq(0, 12, 2), labels = function(x) qz_num(x, 0),
                       expand = expansion(0)) +
    coord_cartesian(clip = "off") +
    labs(x = tr("Index of dissimilarity with the official result (points)", "Indice de dissimilarité avec le résultat officiel (points)"),
         y = NULL) +
    theme_qes(mode, grid = "x")
}

tab <- dissim[order(dissim$year, dissim$study, dissim$weighting), ]
tab_dissim <- setNames(data.frame(
  tab$year, qz_study_label(tab$study),
  ifelse(tab$weighting == "weighted", tr("weighted", "pondéré"), tr("unweighted", "non pondéré")),
  qz_num(tab$index), row.names = NULL
), c(tr("Election", "Élection"), tr("Study", "Étude"), tr("Weighting", "Pondération"),
     tr("Index of dissimilarity (points)", "Indice de dissimilarité (points)")))
alt <- sprintf(tr(
  "Dumbbell chart, one row per study with both estimates (2007 to 2022): the index of dissimilarity between the reported and the official vote, unweighted (hollow ring) and weighted (filled dot). Weighting takes 2012 from %s to %s points and 2022 from %s to %s, and moves 2014 from %s to %s. Values in the table view, with the unweighted studies.",
  "Graphique en haltères, une rangée par étude qui a les deux estimations (2007 à 2022) : l'indice de dissimilarité entre le vote déclaré et le vote officiel, non pondéré (anneau creux) et pondéré (point plein). La pondération fait passer 2012 de %s à %s points et 2022 de %s à %s, et 2014 de %s à %s. Valeurs dans la vue en tableau, avec les études non pondérées."),
  qz_num(di("qes2012", "unweighted")), qz_num(di("qes2012", "weighted")),
  qz_num(di("qes2022", "unweighted")), qz_num(di("qes2022", "weighted")),
  qz_num(di("qes2014", "unweighted")), qz_num(di("qes2014", "weighted")))
qz_figure(fig_dissim, "dissim",
  title = sprintf(tr("Weighting cuts the 2012 error from %s to %s points, but does not bring every study closer to the result",
                     "La pondération réduit l'erreur de 2012 de %s à %s points, mais ne rapproche pas chaque étude du résultat"),
                  qz_num(di("qes2012", "unweighted"), 0), qz_num(di("qes2012", "weighted"), 0)),
  subtitle = tr("Index of dissimilarity between the reported and the official vote, unweighted and weighted",
                "Indice de dissimilarité entre le vote déclaré et le vote officiel, non pondéré et pondéré"),
  alt = alt, table = tab_dissim, height = 3.2,
  note = tr("Source: Quebec Election Studies and Durand panels, reported vote; official results of Élections Québec. Index of dissimilarity: half the sum, over parties, of the absolute difference between the reported and the official share, in points. It is the share of respondents who would have to change party for the survey to match the result (0: identical). A party a study did not list counts in its Other. The unweighted studies (2008, and the 2007 and 2012 panels) are in the table view.",
            "Source : Études électorales québécoises et panels Durand, vote déclaré ; résultats officiels d'Élections Québec. Indice de dissimilarité : la moitié de la somme, sur les partis, de l'écart absolu entre la part déclarée et la part officielle, en points. C'est la part des répondants qui devraient changer de parti pour que l'enquête donne le résultat (0 : identiques). Un parti qu'une étude ne proposait pas compte dans ses Autres. Les études non pondérées (2008, et les panels de 2007 et 2012) sont dans la vue en tableau."))
```

If language explains little, weighting cannot be expected to do much
either, since it adjusts the sample to the census and not to the vote. The
index of dissimilarity lets us check: it is the share of respondents who
would have to change party for the survey to match the result. Weighting
brings 2012 and 2022 closer to the result, by
`r qz_num(di("qes2012", "unweighted") - di("qes2012", "weighted"))` and
`r qz_num(di("qes2022", "unweighted") - di("qes2022", "weighted"))` points,
and moves 2014 slightly away. Note, however, that the index stays between
`r qz_num(min(dissim$index[dissim$weighting == "weighted"]))` and
`r qz_num(max(dissim$index[dissim$weighting == "weighted"]))` points in the
weighted studies, and that two unweighted studies, the 2008 study and the
2007 panel, come closer than most of them (`r qz_num(di("qes2008", "unweighted"))`
and `r qz_num(di("qes2007_panel", "unweighted"))` points). A weight based on
census totals, then, does not by itself make a study's reported vote more
accurate.

## Before the vote: how far intentions move

```{r fig-crop, echo = FALSE}
crop_parties <- c("PLQ", "PQ", "ADQ", "QS")
d <- polls[polls$level %in% crop_parties, ]
d$party <- factor(d$level, levels = crop_parties)
d$x <- as.numeric(d$date)
# a centred three-month mean: the polls within 46 days of each poll
d <- do.call(rbind, lapply(split(d, d$party), function(g) {
  g$smooth <- vapply(g$x, function(t) mean(g$pct[abs(g$x - t) <= 46]), numeric(1))
  g
}))
election_date <- as.Date(h$election_date[h$study == "qes2008"][1])
election <- as.numeric(election_date)
months <- tr(c("Jan.", "Feb.", "Mar.", "Apr.", "May", "June", "July", "Aug.", "Sept.", "Oct.", "Nov.", "Dec."),
             c("janv.", "févr.", "mars", "avr.", "mai", "juin", "juil.", "août", "sept.", "oct.", "nov.", "déc."))
day_lab <- function(x) paste(as.integer(format(x, "%d")), months[as.integer(format(x, "%m"))], format(x, "%Y"))
month_lab <- function(x) paste(months[as.integer(format(x, "%m"))], format(x, "%Y"))
# months in full, for the prose
months_full <- tr(month.name, c("janvier", "février", "mars", "avril", "mai", "juin", "juillet", "août",
                                "septembre", "octobre", "novembre", "décembre"))
month_text <- function(x) paste(months_full[as.integer(format(x, "%m"))], format(x, "%Y"))
o <- official[official$year == 2008 & official$party %in% crop_parties, ]
o$party <- factor(o$party, levels = crop_parties)
last <- d[d$date == max(d$date), ]
last$label <- qz_party_label(last$party)
breaks <- as.Date(c("2007-07-01", "2008-01-01", "2008-07-01", "2009-01-01", "2009-07-01", "2010-01-01"))
pv <- function(date, p, col = "pct") polls[[col]][format(polls$date, "%Y-%m") == date & polls$level == p]
# the last poll before the election, and the range of each party's monthly intentions before it
before <- polls[polls$date < election_date & polls$level %in% crop_parties, ]
last_poll <- format(max(before$date), "%Y-%m")
n_polls <- length(unique(polls$group))
stopifnot(ov(2008, "PLQ") >= pv(last_poll, "PLQ", "lo"), ov(2008, "PLQ") <= pv(last_poll, "PLQ", "hi"),
          pv("2008-06", "PQ") > pv("2008-06", "PLQ"), pv("2008-05", "PLQ") > pv("2008-05", "PQ"),
          abs(gap("qes2008", "PLQ")) > abs(pv(last_poll, "PLQ") - ov(2008, "PLQ")))
# the polls' samples by region: the Québec City area was over-sampled by design
crop_reg <- qes_harmonize(studies = "qes_crop_2007_2010", targets = "region_cma3", quiet = TRUE)
quebec_cma <- 100 * mean(crop_reg$region_cma3 == "Quebec CMA", na.rm = TRUE)
stopifnot(quebec_cma > 15)

fig_crop <- function(mode) {
  k <- qz_tokens[[mode]]
  ggplot(d, aes(x, colour = party, group = party)) +
    geom_vline(xintercept = election, colour = k$axis, linewidth = 0.35) +
    annotate("text", x = election, y = 57, label = paste(tr("Election,", "Élection,"), day_lab(election_date)),
             hjust = -0.04, size = 3, colour = k$ink2, family = qz_font()) +
    geom_line(aes(y = pct), linewidth = 0.35, alpha = 0.45, show.legend = FALSE) +
    geom_line(aes(y = smooth), linewidth = 0.8, lineend = "round", linejoin = "round") +
    geom_point(data = o, aes(x = election, y = share_valid, fill = party, shape = "official"), size = 3.1,
               stroke = 0.8, colour = k$surface, inherit.aes = FALSE) +
    qz_end_labels(transform(last, x = as.numeric(date)), "x", "smooth", "label", "party", mode,
                  min_gap = 2.6, nudge_x = 60) +
    scale_party(mode, aesthetics = "colour", guide = guide_legend(order = 1)) +
    scale_party(mode, aesthetics = "fill") +
    scale_shape_manual(values = c(official = 23), labels = c(official = paste(tr("Official result,", "Résultat officiel,"), day_lab(election_date))),
                       guide = guide_legend(order = 2, override.aes = list(fill = k$ink2, colour = k$surface, size = 3))) +
    scale_x_continuous(breaks = as.numeric(breaks), labels = month_lab(breaks), expand = expansion(add = c(15, 110))) +
    scale_y_continuous(limits = c(0, 60), breaks = seq(0, 60, 10), labels = qz_pct_axis, expand = expansion(0)) +
    labs(x = NULL, y = tr("Vote intention (among those naming a party)", "Intention de vote (parmi ceux qui nomment un parti)")) +
    theme_qes(mode) +
    theme(legend.box = "horizontal", legend.box.just = "top")
}

tab <- polls[polls$level %in% qz_parties & polls$n > 0, ]
tab <- tab[order(tab$date, match(tab$level, qz_parties)), ]
# the CAQ (founded in November 2011) and the PCQ did not run in 2008 and
# did not exist or were not listed in these polls: "—", not a measured zero
ran <- tab$level == "other" | tab$level %in% official$party[official$year == 2008]
tab_crop <- setNames(data.frame(
  format(tab$date, "%Y-%m"), qz_party_label(tab$level),
  ifelse(ran, qz_cell(tab$pct, tab$lo, tab$hi, tab$n), "—"), ifelse(ran, tab$n, ""),
  row.names = NULL
), c(tr("Poll (month)", "Sondage (mois)"), tr("Party", "Parti"),
     tr("Intention, % [95% CI]", "Intention, % [IC à 95 %]"), "n"))
first_poll <- format(min(polls$date), "%Y-%m")
alt <- sprintf(tr(
  "Line chart of monthly CROP vote intentions from %s to %s for the PLQ, PQ, ADQ and QS, each as a three-month mean over the faint monthly values, with a line at the election of %s and the official results as diamonds. The ADQ falls from %s in the first poll to %s in the last poll before the election; the PLQ is at %s in that poll and took %s. Values in the table view.",
  "Graphique linéaire des intentions de vote mensuelles de CROP de %s à %s pour le PLQ, le PQ, l'ADQ et QS, chacune en moyenne sur trois mois par-dessus les valeurs mensuelles en pâle, avec une ligne à l'élection du %s et les résultats officiels en losanges. L'ADQ passe de %s au premier sondage à %s au dernier sondage avant l'élection ; le PLQ est à %s dans ce sondage et a obtenu %s. Valeurs dans la vue en tableau."),
  month_lab(min(polls$date)), month_lab(max(polls$date)), day_lab(election_date),
  qz_pct(pv(first_poll, "ADQ"), 0), qz_pct(pv(last_poll, "ADQ"), 0), qz_pct(pv(last_poll, "PLQ"), 0), qz_pct(ov(2008, "PLQ"), 0))
qz_figure(fig_crop, "crop",
  title = sprintf(tr("The ADQ fell from %s to %s before the 2008 election; the last poll had the PLQ at %s, and it took %s",
                     "L'ADQ est passée de %s à %s avant l'élection de 2008 ; le dernier sondage plaçait le PLQ à %s, et il a obtenu %s"),
                  qz_pct(pv(first_poll, "ADQ"), 0), qz_pct(pv(last_poll, "ADQ"), 0),
                  qz_pct(pv(last_poll, "PLQ"), 0), qz_pct(ov(2008, "PLQ"), 0)),
  subtitle = sprintf(tr("Monthly vote intentions in the CROP polls, %s to %s: three-month means over the monthly values, and the official result",
                        "Intentions de vote mensuelles des sondages CROP, de %s à %s : moyennes sur trois mois par-dessus les valeurs mensuelles, et résultat officiel"),
                     month_lab(min(polls$date)), month_lab(max(polls$date))),
  alt = alt, table = tab_crop, height = 4.2,
  note = sprintf(tr("Source: CROP polls, vote intention (with the undecided asked which party they lean towards, where the poll asked it), about %s respondents a month, among those who named a party; official results of Élections Québec (diamonds). Unweighted. By design, the polls over-sampled the Québec City area (%s of respondents, about twice its share of the population), so these shares give it too much weight. Bold lines: the mean of the polls within a month and a half of each poll; faint lines: each month. The 95%% confidence intervals of every poll are in the table view.",
                     "Source : sondages CROP, intention de vote (avec relance des indécis sur le parti vers lequel ils penchent, là où le sondage l'a demandé), environ %s répondants par mois, parmi ceux qui nomment un parti ; résultats officiels d'Élections Québec (losanges). Non pondéré. Par construction, les sondages surreprésentent la région de Québec (%s des répondants, environ le double de son poids dans la population) ; ces parts lui accordent donc trop de poids. Lignes foncées : la moyenne des sondages à un mois et demi ou moins de chaque sondage ; lignes pâles : chaque mois. Les intervalles de confiance à 95 %% de chaque sondage sont dans la vue en tableau."),
                 qz_num(round(stats::median(polls$n[polls$level == "PLQ"]) / 50) * 50, 0), qz_pct(quebec_cma, 0)))
```

A post-election study has one advantage over a poll: the vote it asks about
has already happened. Durand and Blais (2020) make the distinction between a
polling miss, when voters move after the last poll or the undecided break
unevenly, and a poll failure, when the method is at fault. The monthly CROP
polls show why the distinction matters. Between `r month_text(min(polls$date))`
and `r month_text(max(before$date))`, the Action démocratique du Québec
(ADQ), then the official opposition, went from
`r qz_pct(pv(first_poll, "ADQ"), 0)` to `r qz_pct(pv(last_poll, "ADQ"), 0)`
of vote intentions, and the PQ was back ahead of the PLQ in
`r month_text(as.Date("2008-06-15"))` (`r qz_pct(pv("2008-06", "PQ"), 0)`
against `r qz_pct(pv("2008-06", "PLQ"), 0)`). The last poll before the
election, in `r month_text(max(before$date))`, had the PLQ at
`r qz_pct(pv(last_poll, "PLQ"), 0)`; it took `r qz_pct(ov(2008, "PLQ"), 0)`,
within the poll's margin of error. The 2008 study, which asked respondents
how they voted after the election and is also unweighted, puts the PLQ at
`r qz_pct(rb("qes2008", "PLQ"), 0)`, further from the result than the last
poll. Since it was fielded after the vote, late movement cannot explain that
gap, small as it is. It comes from who answers and from what they say.

## In sum

The belief that surveys underestimate the Liberals holds for the Quebec
Election Studies from 2007 to 2014, but not after, and language, the usual
suspect, accounts for `r qz_num(min(closed))` to `r qz_num(max(closed))`
points of it. The PQ is over-reported in most studies, and the winner is
more often under-reported than over-reported. Our analysis is not without
limits. First, the reported vote is a recall, measured among respondents who
named a party. Second, a study of `r qz_int(n_typical)` voters (the median
of the six studies) has a margin of error of up to `r qz_num(moe_typical)`
points on each party before weighting, and close to
`r qz_num(moe_weighted, 0)` once the weights are taken into account, as the
intervals in the table views show. More importantly, the comparison with
official results cannot tell apart the two sources of error: voters who do
not take part in the survey, and respondents who do not report their vote
as cast. In Britain, the main cause of the 2015 polling error was
unrepresentative samples (Sturgis et al. 2018), as it was in Quebec in 1998
(Durand and Blais 1999); nothing here allows us to say whether it still is.
The open question is whether the CAQ's shortfall in 2022 is a one-off, or
whether its voters now play the part that Liberal voters long played: that
of an electorate less inclined to answer surveys.

## Further reading

- Atkeson, Lonna Rae. 1999. "'Sure, I Voted for the Winner!' Overreport of
  the Primary Vote for the Party Nominee in the National Election Studies."
  *Political Behavior* 21 (3): 197–215.
  [doi:10.1023/A:1022031432535](https://doi.org/10.1023/A:1022031432535)
- Bélanger, Éric, Philippe Mongrain, Thomas Gareau-Paquette, and
  Valérie-Anne Mahéo. 2025. "A Party that Went Viral? The Drivers of Support
  for the Parti Conservateur du Québec in the 2022 Election." *Canadian
  Journal of Political Science* 58 (2): 277–296.
  [doi:10.1017/S0008423924000829](https://doi.org/10.1017/S0008423924000829)
  (the 2022 study: the CAQ under-represented, the PCQ close to its result)
- Durand, Claire, and André Blais. 1999. "Why Did the Polls Go Wrong in the
  1998 Quebec Election? The Answer from Post-Election Polls." *Bulletin of
  Sociological Methodology* 62 (1): 43–47.
  [doi:10.1177/075910639906200105](https://doi.org/10.1177/075910639906200105)
- Durand, Claire, and André Blais. 2020. "Quebec 2018: A Failure of the
  Polls?" *Canadian Journal of Political Science* 53 (1): 133–150.
  [doi:10.1017/S0008423919000787](https://doi.org/10.1017/S0008423919000787)
- Sturgis, Patrick, Jouni Kuha, Nick Baker, Mario Callegaro, Stephen Fisher,
  Jane Green, Will Jennings, Benjamin E. Lauderdale, and Patten Smith. 2018.
  "An Assessment of the Causes of the Errors in the 2015 UK General Election
  Opinion Polls." *Journal of the Royal Statistical Society: Series A* 181
  (3): 757–781. [doi:10.1111/rssa.12329](https://doi.org/10.1111/rssa.12329)

On this site, [From two parties to four](realignment.html) follows the vote
of francophones and non-francophones since 1998, and
[Validation against official results](validation.html) runs these
comparisons for every study.

## About the data

- **Studies.** The six Quebec Election Studies (2007 to 2022) and the
  three Durand panels (2007, 2012, 2018), which asked respondents how they
  voted after the election. The 1998 polls interviewed francophones only
  and are left out. The `r n_polls` CROP polls, conducted almost every
  month from `r month_text(min(polls$date))` to
  `r month_text(max(polls$date))`, measure vote intention instead.
- **Shares.** Among respondents who named a party. Those who did not vote,
  did not know or refused to answer are left out. The Green Party, Option
  nationale and the smaller parties count as "Other", and so does a party
  a study did not list.
- **Official results.** Each party's share of valid votes, from Élections
  Québec.
- **Weights.** Each study's post-election weight, which adjusts the sample
  to the census on age, sex and language, and, depending on the study, on
  region or education. The 2008 study, the 2007 and 2012 panels and the
  CROP polls are shown unweighted (hollow dots in the charts). The 2008
  study's weights are calibrated on the vote itself, so they cannot be used
  to test the reported vote.
- **Language.** Francophones are respondents whose mother tongue is French.
  Where the mother tongue is missing or mixed, the language spoken at home
  is used instead.

The data on this page come from one call:

```{r show-harmonize, ref.label = "harmonize", eval = FALSE}
```
