---
title: "Are Quebec elections decided during the campaign?"
description: >
  The same respondents interviewed before and after the election: how
  many voted as they intended, where the undecided went, and which party
  won the campaign.
params:
  lang: "en"
---

*[Version française](fr-transitions.html)*

```{r setup, include = FALSE}
source("_setup.R", encoding = "UTF-8")
```

```{r harmonize, echo = FALSE}
h <- qes_harmonize(
  studies = c("qes1998", "qes2007_panel", "qes2012_panel", "qes2018_panel", "qes2022"),
  targets = "vote_choice",
  layout = "long", missing = "reasons", quiet = TRUE
)
```

```{r estimates, echo = FALSE}
# French typography on this page: a non-breaking space before "%"
qz_pct <- function(x, digits = 1) {
  out <- paste0(qz_num(x, digits), tr("%", " %"))
  out[is.na(x)] <- ""
  out
}
# small counts in words, and a capital for the start of a sentence
num_word <- function(n) tr(c("one", "two", "three", "four", "five", "six", "seven")[n],
                           c("un", "deux", "trois", "quatre", "cinq", "six", "sept")[n])
cap <- function(x) paste0(toupper(substr(x, 1, 1)), substring(x, 2))
# a party with its article ("the CAQ" / "la CAQ"), and after "of" ("of the
# CAQ" / "de la CAQ"), for the figure text
party_the <- function(p) unname(tr(
  c(PLQ = "the PLQ", PQ = "the PQ", ADQ = "the ADQ", QS = "QS", CAQ = "the CAQ", PCQ = "the PCQ"),
  c(PLQ = "le PLQ", PQ = "le PQ", ADQ = "l'ADQ", QS = "QS", CAQ = "la CAQ", PCQ = "le PCQ"))[p])
party_of <- function(p) unname(tr(
  c(PLQ = "of the PLQ", PQ = "of the PQ", ADQ = "of the ADQ", QS = "of QS", CAQ = "of the CAQ", PCQ = "of the PCQ"),
  c(PLQ = "du PLQ", PQ = "du PQ", ADQ = "de l'ADQ", QS = "de QS", CAQ = "de la CAQ", PCQ = "du PCQ"))[p])

panels <- c("qes1998", "qes2007_panel", "qes2012_panel", "qes2018_panel", "qes2022")
n_panels <- length(panels)
# before the election: the party intended (the undecided pushed toward the
# party they lean to), "none" (would not vote, none, would spoil) or
# "undecided" (still no answer after the push: don't know)
pre_cat <- function(x) {
  p <- qz_party_code(x$vote_choice)
  p[x$vote_choice %in% "Would not vote / none / would spoil"] <- "none"
  p[is.na(x$vote_choice) & x$vote_choice__na %in% c("dk", "dk_refused")] <- "undecided"
  p
}
# after the election: the party reported, or "abstained"
post_cat <- function(x) {
  p <- qz_party_code(x$vote_choice)
  p[is.na(x$vote_choice) & x$vote_choice__na %in% "not_voted"] <- "abstained"
  p
}
pre_levels <- c(qz_parties, "none", "undecided")
post_levels <- c(qz_parties, "abstained")
path_levels <- c("same", "switched", "undecided", "none")

# one row per respondent interviewed after the election, with the answer
# the same respondent gave before it
pairs <- lapply(panels, function(s) {
  pre <- h[h$study == s & h$wave_timing %in% "pre", ]
  post <- h[h$study == s & h$wave_timing %in% "post", ]
  post$before <- factor(pre_cat(pre)[match(post$qes_id, pre$qes_id)], levels = pre_levels)
  post$after <- factor(post_cat(post), levels = post_levels)
  b <- as.character(post$before)
  a <- as.character(post$after)
  post$stayed <- factor(ifelse(a == b, "same", "other"), levels = c("same", "other"))
  # among those who voted: voted as intended, switched from another party,
  # were undecided, or had said they would not vote
  post$path <- factor(ifelse(!a %in% qz_parties, NA,
                      ifelse(a == b, "same", ifelse(b %in% qz_parties, "switched", b))),
                      levels = path_levels)
  # the net gain of each party between intention and vote, in points
  for (p in setdiff(qz_parties, "other")) {
    post[[paste0("gain_", p)]] <- 100 * ((a %in% p) - (b %in% p))
  }
  d <- qz_design(post, s, "weight")
  subset(d, !is.na(before) & !is.na(after))
})
names(pairs) <- panels

moves <- do.call(rbind, lapply(panels, function(s) {
  d <- pairs[[s]]
  rbind(qz_tag(qz_share(d, "after", by = "before"), study = s, what = "row"),
        qz_tag(qz_share(subset(d, after != "abstained"), "stayed", by = "before"), study = s, what = "loyal"),
        qz_tag(qz_share(d, "path"), study = s, what = "path"),
        qz_tag(qz_share(subset(d, before %in% qz_parties), "after"), study = s, what = "named"))
}))
moves$year <- qz_year(moves$study)

gains <- do.call(rbind, lapply(panels, function(s) {
  ps <- setdiff(qz_parties, "other")
  ps <- ps[qz_party_status(s, ps) == "offered"]
  do.call(rbind, lapply(ps, function(p) qz_tag(qz_mean(pairs[[s]], paste0("gain_", p)), study = s, party = p)))
}))
gains$year <- qz_year(gains$study)
n_unweighted <- sum(!tapply(moves$weighted, moves$study, all))

# 1998: the recontact over-represents the undecided, non-disclosers and
# third-party supporters, so its unweighted net gains cannot be read. The
# weight the producers used for their published tables (ponder3, still
# under review, so not applied in the figures) shows how much this matters.
raw98 <- get_qes("qes1998", quiet = TRUE)
v98 <- pairs[["qes1998"]]$variables
v98$w_pi <- as.numeric(raw98$ponder3[v98$source_row])
d98 <- survey::svydesign(ids = ~1, weights = ~w_pi, data = v98)
gains98 <- do.call(rbind, lapply(c("PLQ", "PQ", "ADQ"), function(p) qz_tag(qz_mean(d98, paste0("gain_", p)), party = p)))
gw98 <- function(p, col = "mean") gains98[[col]][gains98$party == p]
adq98 <- c(unweighted = 100 * mean(v98$before %in% "ADQ"),
           weighted = 100 * stats::weighted.mean(v98$before %in% "ADQ", v98$w_pi))

before_lab <- c(PLQ = "PLQ", PQ = "PQ", ADQ = "ADQ", QS = "QS", CAQ = "CAQ", PCQ = "PCQ",
                other = tr("Other", "Autres"), none = tr("None / would not vote", "Aucun / ne voterait pas"),
                undecided = tr("Undecided", "Indécis"))
after_lab <- c(before_lab[qz_parties], abstained = tr("Did not vote", "N'a pas voté"))
loyal <- moves[moves$what == "loyal" & moves$level == "same", ]
lv <- function(s, p) loyal$pct[loyal$study == s & loyal$group == p]
lr <- function(p) range(loyal$pct[loyal$group == p & loyal$n >= qz_min_n])
mv <- function(s, b, a) moves$pct[moves$what == "row" & moves$study == s & moves$group == b & moves$level == a]
path <- moves[moves$what == "path", ]
pv <- function(s, l) path$pct[path$study == s & path$level == l]
pr <- function(l) range(path$pct[path$level == l])
gv <- function(s, p, col = "mean") gains[[col]][gains$study == s & gains$party == p]
official <- qz_official()
ov <- function(y, p) official$share_valid[official$year == y & official$party == p]
# parties above 10% of the valid votes in 2022
above <- 10
n_above <- sum(official$share_valid[official$year == 2022 & official$party != "other"] > above)
# the largest net gain and loss, without 1998 (see above)
gains_fig <- gains[gains$study != "qes1998", ]
best <- gains_fig[which.max(gains_fig$mean), ]
worst <- gains_fig[which.min(gains_fig$mean), ]
# abstention among those who had named a party, and the largest group of undecided
named_abst <- range(moves$pct[moves$what == "named" & moves$level == "abstained"])
und_n <- max(moves$n[moves$what == "row" & moves$group == "undecided"])
row_n <- function(s, g) moves$n[moves$what == "row" & moves$study == s & moves$group == g][1]
```

Since the 1988 federal election, Canadian election studies have shown
that opinion can move a great deal between the start of a campaign and
the vote (Johnston et al. 1992). It is often said that Quebec elections
are decided at the last minute, a claim that returns whenever the polls
miss the result, as they did in 1998 and 2018. In 1998 the polls had the
Parti Québécois (PQ) ahead, yet the Parti libéral du Québec (PLQ) took
more of the valid votes (`r qz_pct(ov(1998, "PLQ"), 1)` compared to
`r qz_pct(ov(1998, "PQ"), 1)`). In 2018 they forecast a close race
between the PLQ and the Coalition avenir Québec (CAQ), which won by
`r qz_num(ov(2018, "CAQ") - ov(2018, "PLQ"), 0)` points. Each time, part
of the explanation was sought among voters who changed their minds at the
end of the campaign and among non-disclosers, who would not reveal their
intention. Durand and Blais (1999) reject both a late Liberal rise and a
lower PQ turnout in 1998, and trace the error to sampling and to how
non-respondents behaved; for 2018, they conclude that last-minute shifts
and the vote of non-disclosers explain most of the gap (Durand and Blais
2020).

However, a cross-section tells us how the vote ended up, not who moved.
Panels can, because they interview the same people during the campaign
and after the vote. `r cap(num_word(n_panels))` Quebec election surveys
did so, from the 1998 CROP and CREATEC polls to the 2022 Quebec Election
Study, including the panels Claire Durand ran in 2007, 2012 and 2018. We
use them to ask a simple question: how many Quebec voters change their
minds during a campaign, and who gains from it? Between
`r qz_pct(pr("same")[1], 0)` and `r qz_pct(pr("same")[2], 0)` of those
who voted reported the party they had named during the campaign. The
undecided were a small group, from `r qz_pct(pr("undecided")[1], 0)` to
`r qz_pct(pr("undecided")[2], 0)` of voters, and voters who switched from
one party to another outnumbered them in every panel except 2018, where
the two groups were about the same size. On net, these moves rarely
changed a party's share by more than a few points. The largest gain is the
one the 2018 post-mortem pointed to: among the same respondents, the CAQ
gained `r qz_num(gv("qes2018_panel", "CAQ"), 1)` points between intention
and vote, and `r qz_pct(mv("qes2018_panel", "undecided", "CAQ"), 0)` of
that year's undecided voted CAQ. In short, a Quebec campaign moves many
voters but, on net, few votes.

## Most voters did what they said

```{r fig-matrix, echo = FALSE}
show <- c("qes2018_panel", "qes2022")
rows <- c("PLQ", "PQ", "QS", "CAQ", "PCQ", "other", "none", "undecided")
cols <- c("PLQ", "PQ", "QS", "CAQ", "PCQ", "other", "abstained")
after_short <- c(after_lab[setdiff(cols, "abstained")], abstained = tr("Did not\nvote", "N'a pas\nvoté"))
rn <- unique(moves[moves$what == "row" & moves$study %in% show, c("study", "group", "n")])
# the same rows and columns in both blocks; a row or column without values
# says why: not listed by the question, not an answer of the question, or
# fewer than 30 respondents
grid <- expand.grid(study = show, before = rows, after = cols, stringsAsFactors = FALSE)
grid <- merge(grid, moves[moves$what == "row", c("study", "group", "level", "pct")],
              by.x = c("study", "before", "after"), by.y = c("study", "group", "level"), all.x = TRUE)
grid$n <- rn$n[match(paste(grid$study, grid$before), paste(rn$study, rn$group))]
listed <- function(s, p) p %in% c("other", "none", "undecided", "abstained") | qz_party_status(s, p) == "offered"
grid$why <- ifelse(!listed(grid$study, grid$after), tr("n.l.", "n.p."),
             ifelse(!listed(grid$study, grid$before), tr("not listed", "non proposé"),
             ifelse(is.na(grid$n), tr("not an answer", "pas une réponse"),
             ifelse(grid$n < qz_min_n, paste0("n = ", grid$n), NA))))
grid$label <- ifelse(is.na(grid$why), qz_num(grid$pct, 0), "")
# a column the question did not list is "n.l." in each cell; a row without
# values carries one note, in its middle column, and nothing else
grid$label[!is.na(grid$why) & grid$why == tr("n.l.", "n.p.")] <- tr("n.l.", "n.p.")
empty_row <- paste(grid$study, grid$before) %in%
  paste(grid$study, grid$before)[!is.na(grid$why) & grid$why != tr("n.l.", "n.p.")]
grid$label[empty_row] <- ""
note_cell <- empty_row & grid$after == "CAQ"
grid$label[note_cell] <- grid$why[note_cell]
grid$key <- paste(grid$study, grid$before)
row_lab <- stats::setNames(ifelse(is.na(grid$n) | grid$n == 0, before_lab[grid$before],
                                  paste0(before_lab[grid$before], " (", grid$n, ")")), grid$key)
grid$y <- factor(grid$key, levels = rev(paste(rep(show, each = length(rows)), rows)))
grid$x <- factor(grid$after, levels = cols)
grid$panel <- factor(qz_study_label(grid$study), levels = qz_study_label(show))
d <- grid

fig_matrix <- function(mode) {
  k <- qz_tokens[[mode]]
  ramp <- grDevices::colorRampPalette(k$seq)(101)
  d$fillhex <- ifelse(is.na(d$why), ramp[pmin(101, pmax(1, round(d$pct) + 1))], k$wash)
  d$ink <- ifelse(is.na(d$why), qz_label_ink(d$fillhex, mode), k$ink2)
  ggplot(d, aes(x, y)) +
    # one tile layer, so the rows keep their order; cells without a value are grey
    qz_tile(mode, aes(fill = ifelse(is.na(why), pct, NA))) +
    geom_text(aes(label = label, colour = ink), size = 2.9, family = qz_font()) +
    facet_grid(panel ~ ., scales = "free_y", space = "free_y") +
    scale_fill_gradientn(colours = k$seq, limits = c(0, 100), breaks = c(0, 50, 100),
                         labels = qz_pct_axis, na.value = k$wash,
                         guide = guide_colourbar(barwidth = unit(120, "pt"), barheight = unit(7, "pt"))) +
    scale_colour_identity() +
    scale_x_discrete(position = "top", labels = after_short) +
    scale_y_discrete(labels = function(x) row_lab[x]) +
    labs(x = tr("Reported vote after the election", "Vote déclaré après l'élection"),
         y = tr("Intention during the campaign (n)", "Intention pendant la campagne (n)")) +
    theme_qes(mode, grid = "none") +
    theme(strip.text.y = element_text(angle = 0, hjust = 0, face = "bold", colour = k$ink),
          legend.position = "bottom", legend.justification = "left",
          legend.ticks = element_blank(),
          axis.text.x.top = element_text(lineheight = 0.9, vjust = 0),
          axis.title.x.top = element_text(margin = margin(b = 6)))
}

tab <- moves[moves$what == "row", ]
tab <- tab[order(tab$year, tab$study, match(tab$group, pre_levels), match(tab$level, post_levels)), ]
tab <- tab[tab$n > 0, ]
# a party that did not run, or that the question did not list, is "—" or
# "n.l.", not a measured zero
tab$status <- ifelse(tab$level %in% c("other", "abstained"), "offered", qz_party_status(tab$study, tab$level))
tab_matrix <- setNames(data.frame(
  tab$year, qz_study_label(tab$study), before_lab[tab$group], after_lab[tab$level],
  ifelse(tab$status == "not_run", "—", ifelse(tab$status == "not_listed", tr("n.l.", "n.p."),
                                                qz_cell(tab$pct, tab$lo, tab$hi, tab$n))),
  tab$n, qz_weight_text(tab$weighted), row.names = NULL
), c(tr("Election", "Élection"), tr("Study", "Étude"), tr("Intention", "Intention"),
     tr("Reported vote", "Vote déclaré"), tr("Row %, [95% CI]", "% de la rangée [IC à 95 %]"),
     tr("n (row)", "n (rangée)"), tr("Weighting", "Pondération")))
alt <- sprintf(tr(
  "Two heatmaps (2018 panel, QES 2022): rows are the vote intention during the campaign, with the number of respondents, columns the vote reported after the election, each cell the percentage of the row. The diagonal dominates: in 2022, %s of those who intended to vote CAQ reported a CAQ vote and %s of those who intended to vote QS a QS vote; %s of the undecided did not vote. Values in the table view.",
  "Deux cartes de chaleur (panel 2018, EEQ 2022) : les rangées sont l'intention de vote pendant la campagne, avec le nombre de répondants, les colonnes le vote déclaré après l'élection, chaque case le pourcentage de la rangée. La diagonale domine : en 2022, %s des personnes qui comptaient voter CAQ ont déclaré un vote CAQ et %s de celles qui comptaient voter QS, un vote QS ; %s des indécis n'ont pas voté. Valeurs dans la vue en tableau."),
  qz_pct(mv("qes2022", "CAQ", "CAQ"), 0), qz_pct(mv("qes2022", "QS", "QS"), 0),
  qz_pct(mv("qes2022", "undecided", "abstained"), 0))
qz_figure(fig_matrix, "matrix",
  title = sprintf(tr("Most voters did what they said; in 2018 the CAQ drew from every side, including %s of those who intended to vote PQ",
                     "La plupart des électeurs ont fait ce qu'ils avaient dit ; en 2018, la CAQ a puisé de tous les côtés, dont %s des personnes qui comptaient voter PQ"),
                  qz_pct(mv("qes2018_panel", "PQ", "CAQ"), 0)),
  subtitle = tr("From the intention during the campaign (rows) to the vote reported after the election (columns): the percentage of each row",
                "De l'intention pendant la campagne (rangées) au vote déclaré après l'élection (colonnes) : le pourcentage de chaque rangée"),
  alt = alt, table = tab_matrix, height = 6.2,
  note = tr("Source: qesR. Respondents interviewed during the campaign and after the election, in the 2018 Durand panel and the QES 2022, weighted with the post-election weight. Rows: the party intended, with the number of respondents; the undecided named no party, even when asked (where the panel asked) which party they leaned to. Each row includes those who did not vote. Grey cells have no value: the 2018 panel did not list the PCQ (n.l.), the 2022 campaign question had no 'would not vote' answer, and rows of fewer than 30 respondents are not drawn. Confidence intervals and the 1998, 2007 and 2012 panels are in the table view.",
            "Source : qesR. Répondants interrogés pendant la campagne et après l'élection, dans le panel Durand 2018 et l'EEQ 2022, pondérés avec la pondération postélectorale. Rangées : le parti prévu, avec le nombre de répondants ; les indécis n'ont nommé aucun parti, même quand on leur demandait (là où le panel le demandait) vers lequel ils penchaient. Chaque rangée inclut les personnes qui n'ont pas voté. Les cases grises n'ont pas de valeur : le panel 2018 ne proposait pas le PCQ (n.p.), la question de campagne de 2022 n'avait pas de réponse « ne voterait pas », et les rangées de moins de 30 répondants ne sont pas tracées. Les intervalles de confiance et les panels de 1998, 2007 et 2012 sont dans la vue en tableau."))
```

The diagonal dominates. In 2022, `r qz_pct(mv("qes2022", "CAQ", "CAQ"), 0)`
of those who intended to vote CAQ reported a CAQ vote (a share that counts
those who did not vote), and the flows between parties were small.
Québec solidaire (QS) kept the smallest share of its campaign supporters,
`r qz_pct(mv("qes2022", "QS", "QS"), 0)`:
`r qz_pct(mv("qes2022", "QS", "PLQ"), 0)` of them voted PLQ and
`r qz_pct(mv("qes2022", "QS", "PQ"), 0)` PQ. In 2018, the picture is quite
different. The CAQ drew from every side, with
`r qz_pct(mv("qes2018_panel", "PQ", "CAQ"), 0)` of those who intended to
vote PQ, `r qz_pct(mv("qes2018_panel", "QS", "CAQ"), 0)` of those who
intended to vote QS and `r qz_pct(mv("qes2018_panel", "none", "CAQ"), 0)`
of those who had said they would not vote. In other words, the winner of
2018 did not only hold its voters; it kept recruiting during the campaign.

Taken together, the share of voters who did not vote as they had said is
far from trivial: it ranges from `r qz_pct(100 - pr("same")[2], 0)` to
`r qz_pct(100 - pr("same")[1], 0)` across the `r num_word(n_panels)`
panels. The largest group among them switched from one party to another,
from `r qz_pct(pr("switched")[1], 0)` to `r qz_pct(pr("switched")[2], 0)`
of voters, while the undecided and those who had said they would not vote
made up the rest. In 2018 the two groups were about the same size
(`r qz_pct(pv("qes2018_panel", "switched"), 0)` switchers,
`r qz_pct(pv("qes2018_panel", "undecided"), 0)` undecided). That panel,
however, did not ask the follow-up question of some undecided respondents
interviewed by telephone, nor of those who said they would not vote, so
it counts more undecided than the others. The 2007 election, a three-way
race between the PLQ, the PQ and the Action démocratique du Québec (ADQ),
produced the most switchers, `r qz_pct(pv("qes2007_panel", "switched"), 0)`.

## Loyalty: voted as they intended

```{r fig-loyal, echo = FALSE}
loyal_parties <- c("PLQ", "PQ", "ADQ", "QS", "CAQ", "PCQ")
d <- loyal[loyal$group %in% loyal_parties & loyal$n >= qz_min_n, ]
d$party <- factor(d$group, levels = loyal_parties)
d$lab <- ifelse(d$study == "qes1998", tr("1998 polls", "Sondages de 1998"), qz_study_label(d$study))
d$key <- paste(d$group, d$study)
d$row <- factor(d$key, levels = rev(unique(d$key[order(d$party, qz_year(d$study))])))
row_lab <- stats::setNames(d$lab, d$key)
low <- d[d$study == "qes2007_panel" & d$group == "QS", ]
big <- range(d$pct[d$group %in% c("PLQ", "PQ", "CAQ")])
low$label <- sprintf(tr("%s: QS kept fewer than half", "%s : QS en a gardé moins de la moitié"), qz_pct(low$pct, 0))

fig_loyal <- function(mode) {
  k <- qz_tokens[[mode]]
  ggplot(d, aes(pct, row)) +
    geom_linerange(aes(xmin = lo, xmax = hi, colour = party), linewidth = 0.45, alpha = 0.6) +
    qz_dots(mode, d, aes(pct, row), colour_aes = "party", size = 2.6) +
    geom_text(data = low, aes(hi + 1.5, row, label = label), hjust = 0, size = 3, colour = k$ink,
              family = qz_font(), fontface = "bold") +
    facet_grid(party ~ ., scales = "free_y", space = "free_y", switch = "y") +
    scale_party(mode) +
    scale_y_discrete(labels = function(x) row_lab[x]) +
    scale_x_continuous(limits = c(30, 101), breaks = c(40, 60, 80, 100), labels = qz_pct_axis,
                       expand = expansion(0)) +
    labs(x = tr("Voted as intended (among those who voted; the axis starts at 30%)",
                "A voté comme prévu (parmi les votants ; l'axe commence à 30 %)"), y = NULL) +
    theme_qes(mode, grid = "x") +
    theme(strip.placement = "outside",
          strip.text.y.left = element_text(angle = 0, hjust = 1, vjust = 0.5, face = "bold", colour = k$ink,
                                           margin = margin(r = 8)),
          panel.spacing.y = unit(8, "pt"))
}

tab <- loyal[order(loyal$year, loyal$study, match(loyal$group, pre_levels)), ]
tab <- tab[tab$group %in% qz_parties & tab$n > 0, ]
tab_loyal <- setNames(data.frame(
  tab$year, qz_study_label(tab$study), before_lab[tab$group], qz_cell(tab$pct, tab$lo, tab$hi, tab$n), tab$n,
  qz_weight_text(tab$weighted), row.names = NULL
), c(tr("Election", "Élection"), tr("Study", "Étude"), tr("Intention", "Intention"),
     tr("Voted as intended, % [95% CI]", "A voté comme prévu, % [IC à 95 %]"), "n", tr("Weighting", "Pondération")))
alt <- sprintf(tr(
  "Dot chart grouped by party (PLQ, PQ, ADQ, QS, CAQ, PCQ), one row per panel study from 1998 to 2022: the share of those who intended to vote for the party during the campaign who reported voting for it, among those who voted, with confidence intervals. The PLQ, the PQ and the CAQ are between %s and %s; QS in the 2007 panel kept %s. In 2022 the CAQ kept %s and the PLQ %s. Values in the table view.",
  "Graphique à points groupé par parti (PLQ, PQ, ADQ, QS, CAQ, PCQ), une rangée par étude par panel de 1998 à 2022 : la part des personnes qui comptaient voter pour le parti pendant la campagne qui ont déclaré avoir voté pour lui, parmi les votants, avec intervalles de confiance. Le PLQ, le PQ et la CAQ sont entre %s et %s ; QS, dans le panel 2007, en a gardé %s. En 2022, la CAQ en a gardé %s et le PLQ %s. Valeurs dans la vue en tableau."),
  qz_pct(big[1], 0), qz_pct(big[2], 0), qz_pct(lv("qes2007_panel", "QS"), 0), qz_pct(lv("qes2022", "CAQ"), 0),
  qz_pct(lv("qes2022", "PLQ"), 0))
qz_figure(fig_loyal, "loyal",
  title = sprintf(tr("The PLQ, the PQ and the CAQ keep %s to %s of their campaign supporters; QS keeps fewer, down to %s in 2007",
                     "Le PLQ, le PQ et la CAQ gardent de %s à %s des personnes qui comptaient voter pour eux ; QS en garde moins, jusqu'à %s en 2007"),
                  qz_pct(big[1], 0), qz_pct(big[2], 0), qz_pct(lv("qes2007_panel", "QS"), 0)),
  subtitle = tr("Share of those who intended to vote for each party during the campaign who reported voting for it, among those who voted, by panel study, with 95% confidence intervals",
                "Part des personnes qui comptaient voter pour chaque parti pendant la campagne qui ont déclaré avoir voté pour lui, parmi les votants, selon l'étude par panel, avec intervalles de confiance à 95 %"),
  alt = alt, table = tab_loyal, height = 5.4,
  note = tr("Source: qesR. Respondents interviewed during the campaign and after the election who reported a vote. Lines: 95% confidence intervals. Filled dots are weighted with the post-election weight (2018 panel, QES 2022); hollow dots (1998, 2007 and 2012 panels) are not weighted here, because their weights are under review. The studies differ in design and mode, so comparisons across studies are best made within a party. Cells of fewer than 30 respondents are not drawn; the other parties are in the table view.",
            "Source : qesR. Répondants interrogés pendant la campagne et après l'élection qui ont déclaré un vote. Traits : intervalles de confiance à 95 %. Points pleins : pondérés avec la pondération postélectorale (panel 2018, EEQ 2022) ; points creux (panels 1998, 2007 et 2012) : non pondérés ici, leurs pondérations étant en révision. Les études diffèrent par leur devis et leur mode de collecte ; les comparaisons entre études se font donc de préférence à l'intérieur d'un parti. Les cellules de moins de 30 répondants ne sont pas tracées ; les autres partis sont dans la vue en tableau."))
```

Two patterns stand out. First, the PLQ is the most consistent party over
time: it keeps from `r qz_pct(lr("PLQ")[1], 0)` to
`r qz_pct(lr("PLQ")[2], 0)` of its campaign supporters at every election,
whereas the PQ ranges from `r qz_pct(lr("PQ")[1], 0)` to
`r qz_pct(lr("PQ")[2], 0)` and the CAQ from `r qz_pct(lr("CAQ")[1], 0)`
to `r qz_pct(lr("CAQ")[2], 0)`. These shares leave out those who did not
vote, which is why the CAQ keeps `r qz_pct(lv("qes2022", "CAQ"), 0)` here
in 2022, compared to `r qz_pct(mv("qes2022", "CAQ", "CAQ"), 0)` in the
matrix above. Second, the smaller parties keep the fewest. The ADQ kept
`r qz_pct(lv("qes1998", "ADQ"), 0)` in 1998, and QS kept
`r qz_pct(lv("qes2007_panel", "QS"), 0)` in 2007 and
`r qz_pct(lv("qes2012_panel", "QS"), 0)` in 2012. This is consistent with
a familiar account in which a voter who likes a small party during the
campaign returns to a party that can win on election day, although the
panels alone cannot tell strategic desertion from a simple change of
mind.

Loyalty is not a fixed trait of a party, however. The CAQ kept
`r qz_pct(lv("qes2012_panel", "CAQ"), 0)` of its campaign supporters in
2012, its first election, and `r qz_pct(lv("qes2018_panel", "CAQ"), 0)` in
2018, when it went on to win. The ADQ went from
`r qz_pct(lv("qes1998", "ADQ"), 0)` in 1998 to
`r qz_pct(lv("qes2007_panel", "ADQ"), 0)` in 2007, the year it became the
official opposition. These cases suggest, without establishing it, that
loyalty follows a party's momentum more than its size. The Parti
conservateur du Québec (PCQ) points the same way only if momentum, rather
than the chance of winning, is what matters: with
`r qz_pct(ov(2018, "PCQ"), 1)` of the vote in 2018, it kept
`r qz_pct(lv("qes2022", "PCQ"), 0)` of its campaign supporters in 2022 and
won no seat (on that surge, see Bélanger et al. 2025).

## Where the undecided went

```{r fig-undecided, echo = FALSE}
# "did not vote" first, at the baseline, then the parties in legend order
stack <- c("abstained", qz_parties)
d <- moves[moves$what == "row" & moves$group == "undecided" & moves$level %in% stack & moves$n >= qz_min_n, ]
d$level <- factor(d$level, levels = stack)
d <- d[order(d$study, d$level), ]
d$xmax <- ave(d$pct, d$study, FUN = cumsum)
d$xmin <- d$xmax - d$pct
d$row <- factor(paste0(qz_study_label(d$study), " (n = ", d$n, ")"),
                levels = rev(unique(paste0(qz_study_label(d$study), " (n = ", d$n, ")")[order(qz_year(d$study))])))
d$label <- ifelse(d$pct >= 8, qz_num(d$pct, 0), "")
abst <- moves$pct[moves$what == "row" & moves$group == "undecided" & moves$level == "abstained" & moves$n >= qz_min_n]

fig_undecided <- function(mode) {
  k <- qz_tokens[[mode]]
  fills <- c(abstained = k$surface, k$party)
  d$ink <- ifelse(d$level == "abstained", k$ink, qz_label_ink(fills[as.character(d$level)], mode))
  half <- 0.2
  ggplot(d) +
    geom_rect(data = d[d$level != "abstained", ],
              aes(xmin = xmin, xmax = xmax, ymin = as.numeric(row) - half, ymax = as.numeric(row) + half, fill = level),
              colour = k$surface, linewidth = 0.7) +
    # did not vote: an empty, outlined segment at the baseline
    geom_rect(data = d[d$level == "abstained", ],
              aes(xmin = xmin + 0.3, xmax = xmax - 0.3, ymin = as.numeric(row) - half + 0.02, ymax = as.numeric(row) + half - 0.02,
                  fill = level), colour = k$ink2, linewidth = 0.45) +
    geom_text(aes(x = (xmin + xmax) / 2, y = as.numeric(row), label = label, colour = ink),
              size = 2.8, family = qz_font()) +
    scale_fill_manual(values = fills, breaks = stack, labels = after_lab[stack],
                      guide = guide_legend(nrow = 1, override.aes = list(colour = c(k$ink2, rep(k$surface, length(qz_parties))),
                                                                         linewidth = c(0.45, rep(0.7, length(qz_parties)))))) +
    scale_colour_identity() +
    scale_y_continuous(breaks = seq_along(levels(d$row)), labels = levels(d$row), expand = expansion(add = 0.35)) +
    scale_x_continuous(limits = c(0, 100.5), breaks = c(0, 25, 50, 75, 100), labels = qz_pct_axis, expand = expansion(0)) +
    labs(x = tr("Share of the campaign's undecided", "Part des indécis de la campagne"), y = NULL) +
    theme_qes(mode, grid = "none") +
    theme(legend.key.width = unit(10, "pt"), legend.key.height = unit(10, "pt"))
}

tab <- moves[moves$what == "row" & moves$group == "undecided" & moves$n > 0, ]
tab <- tab[order(tab$year, tab$study, match(tab$level, post_levels)), ]
tab$status <- ifelse(tab$level %in% c("other", "abstained"), "offered", qz_party_status(tab$study, tab$level))
tab_undecided <- setNames(data.frame(
  tab$year, qz_study_label(tab$study), after_lab[tab$level],
  ifelse(tab$status == "not_run", "—", ifelse(tab$status == "not_listed", tr("n.l.", "n.p."),
                                                qz_cell(tab$pct, tab$lo, tab$hi, tab$n))),
  tab$n, qz_weight_text(tab$weighted), row.names = NULL
), c(tr("Election", "Élection"), tr("Study", "Étude"), tr("Reported vote", "Vote déclaré"),
     tr("Share of the undecided, % [95% CI]", "Part des indécis, % [IC à 95 %]"), "n", tr("Weighting", "Pondération")))
alt <- sprintf(tr(
  "100%% stacked horizontal bars, one per panel study from 1998 to 2022: what the respondents who were still undecided during the campaign reported after the election, with those who did not vote first. Between %s and %s did not vote; in 2018, %s voted CAQ; in 2022, %s did not vote and %s voted CAQ. Values in the table view.",
  "Barres horizontales empilées à 100 %%, une par étude par panel de 1998 à 2022 : ce qu'ont déclaré après l'élection les répondants encore indécis pendant la campagne, en commençant par ceux qui n'ont pas voté. Entre %s et %s n'ont pas voté ; en 2018, %s ont voté CAQ ; en 2022, %s n'ont pas voté et %s ont voté CAQ. Valeurs dans la vue en tableau."),
  qz_pct(min(abst), 0), qz_pct(max(abst), 0), qz_pct(mv("qes2018_panel", "undecided", "CAQ"), 0),
  qz_pct(mv("qes2022", "undecided", "abstained"), 0), qz_pct(mv("qes2022", "undecided", "CAQ"), 0))
qz_figure(fig_undecided, "undecided",
  title = sprintf(tr("%s to %s of the campaign's undecided did not vote; in 2018, %s of them voted CAQ, the winner",
                     "De %s à %s des indécis de la campagne n'ont pas voté ; en 2018, %s d'entre eux ont voté CAQ, le parti gagnant"),
                  qz_pct(min(abst), 0), qz_pct(max(abst), 0), qz_pct(mv("qes2018_panel", "undecided", "CAQ"), 0)),
  subtitle = tr("What the respondents still undecided during the campaign reported after the election, by panel study",
                "Ce qu'ont déclaré après l'élection les répondants encore indécis pendant la campagne, selon l'étude par panel"),
  alt = alt, table = tab_undecided, height = 3.2,
  note = sprintf(tr("Source: qesR. The undecided are respondents who named no party during the campaign, even when asked (where the panel asked) which party they leaned to, and who answered after the election. Weighted (2018 panel, QES 2022) or not weighted here, because their weights are under review (1998, 2007 and 2012 panels). Outlined segment: did not vote. Numbers: shares of 8%% or more. Each row rests on %s respondents or fewer, so the confidence intervals in the table view are wide. In the 2007 and 2018 panels, 'don't know' and 'refused' cannot be told apart and are both counted as undecided.",
            "Source : qesR. Les indécis sont les répondants qui n'ont nommé aucun parti pendant la campagne, même quand on leur demandait (là où le panel le demandait) vers lequel ils penchaient, et qui ont répondu après l'élection. Pondéré (panel 2018, EEQ 2022) ou non pondéré ici, les pondérations étant en révision (panels 1998, 2007 et 2012). Segment vide : n'a pas voté. Nombres : parts de 8 %% ou plus. Chaque rangée repose sur %s répondants ou moins : les intervalles de confiance de la vue en tableau sont donc larges. Dans les panels 2007 et 2018, « ne sait pas » et « refus » ne peuvent être distingués et sont tous deux comptés comme indécis."), und_n))
```

The undecided matter less for the result than for turnout. Between
`r qz_pct(min(abst), 0)` and `r qz_pct(max(abst), 0)` of them did not
vote, compared to `r qz_pct(named_abst[1], 0)` to
`r qz_pct(named_abst[2], 0)` of those who had named a party. Voters who
decide during the campaign are known to be the most open to its effects
(Fournier et al. 2004); in Quebec, a sizeable minority of the undecided
simply stay home. Those who did vote went in no single direction. In 1998 they
leaned to the PLQ (`r qz_pct(mv("qes1998", "undecided", "PLQ"), 0)`
versus `r qz_pct(mv("qes1998", "undecided", "PQ"), 0)` for the PQ), but
on only `r row_n("qes1998", "undecided")` respondents. In 2018 they broke
for the CAQ, and in 2022 they split across all the parties. The idea that
the undecided lean toward one party finds no steady support here.

## Who won the campaign

The matrix shows who moved; it does not show who came out ahead once the
moves in both directions are added up. We take, for each party, its
share of the reported vote minus its share of the intentions, among the
same respondents. A positive gap means the party won more voters during
the campaign than it lost.

```{r fig-gains, echo = FALSE}
# without 1998, whose unweighted net gains reflect how the recontact was
# drawn (see the text)
d <- gains_fig[gains_fig$n >= qz_min_n, ]
d$party <- factor(d$party, levels = qz_parties)
d$panel <- factor(qz_study_label(d$study), levels = qz_study_label(panels))
d$key <- paste(d$study, d$party)
d$row <- factor(d$key, levels = rev(d$key[order(match(d$study, panels), d$party)]))
row_lab <- stats::setNames(as.character(d$party), d$key)
d$label <- qz_pts(d$mean, 1)
lim <- max(abs(c(d$lo, d$hi))) + 2.5

fig_gains <- function(mode) {
  k <- qz_tokens[[mode]]
  ggplot(d, aes(mean, row)) +
    geom_vline(xintercept = 0, colour = k$axis, linewidth = 0.45) +
    geom_linerange(aes(xmin = lo, xmax = hi, colour = party), linewidth = 0.45, alpha = 0.6) +
    qz_dots(mode, d, aes(mean, row), colour_aes = "party", size = 2.6) +
    # each value on the side away from zero, past the end of its interval
    geom_text(aes(x = ifelse(mean >= 0, hi + 0.4, lo - 0.4), label = label, hjust = ifelse(mean >= 0, 0, 1)),
              size = 2.8, colour = k$ink2, family = qz_font()) +
    facet_grid(panel ~ ., scales = "free_y", space = "free_y", switch = "y") +
    scale_party(mode) +
    scale_y_discrete(labels = function(x) row_lab[x]) +
    scale_x_continuous(limits = c(-lim, lim), labels = function(x) qz_pts(x, 0), expand = expansion(0)) +
    labs(x = tr("Reported vote minus intention, among the same respondents (points)",
                "Vote déclaré moins intention, chez les mêmes répondants (points)"), y = NULL) +
    theme_qes(mode, grid = "x") +
    theme(strip.placement = "outside",
          strip.text.y.left = element_text(angle = 0, hjust = 1, vjust = 0.5, face = "bold", colour = k$ink,
                                           margin = margin(r = 8)),
          panel.spacing.y = unit(8, "pt"))
}

tab <- gains_fig[order(gains_fig$year, match(gains_fig$party, qz_parties)), ]
tab_gains <- setNames(data.frame(
  tab$year, qz_study_label(tab$study), tab$party,
  ifelse(tab$n < qz_min_n, tr("n < 30", "n < 30"),
         paste0(qz_num(tab$mean, 1), " [", qz_num(tab$lo, 1), tr(", ", " ; "), qz_num(tab$hi, 1), "]")),
  tab$n, qz_weight_text(tab$weighted), row.names = NULL
), c(tr("Election", "Élection"), tr("Study", "Étude"), tr("Party", "Parti"),
     tr("Net gain, points [95% CI]", "Gain net, points [IC à 95 %]"), "n", tr("Weighting", "Pondération")))
alt <- sprintf(tr(
  "Dot chart in %s panels, one per panel study from 2007 to 2022, one row per party: the party's share of the reported vote minus its share of the campaign intentions, among the same respondents, in points, with confidence intervals. Most gaps are within a few points of zero. The largest gain is that %s in %s, %s, the largest loss that %s in %s, %s; in 2022 the CAQ is at %s and the PQ at %s. Values in the table view.",
  "Graphique à points en %s panneaux, un par étude par panel de 2007 à 2022, une rangée par parti : la part du vote déclaré moins la part des intentions de campagne, chez les mêmes répondants, en points, avec intervalles de confiance. La plupart des écarts sont à quelques points de zéro. Le plus grand gain est celui %s en %s, %s, la plus grande perte celle %s en %s, %s ; en 2022, la CAQ est à %s et le PQ à %s. Valeurs dans la vue en tableau."),
  num_word(length(unique(d$study))), party_of(best$party), best$year, qz_pts(best$mean, 1),
  party_of(worst$party), worst$year, qz_pts(worst$mean, 1),
  qz_pts(gv("qes2022", "CAQ"), 1), qz_pts(gv("qes2022", "PQ"), 1))
qz_figure(fig_gains, "gains",
  title = sprintf(tr("On net, the campaign moved a few points at most: from %s for %s in %s to %s for %s in %s",
                     "En solde net, la campagne a déplacé quelques points tout au plus : de %s pour %s en %s à %s pour %s en %s"),
                  qz_pts(worst$mean, 1), party_the(worst$party), worst$year,
                  qz_pts(best$mean, 1), party_the(best$party), best$year),
  subtitle = tr("Each party's share of the reported vote minus its share of the campaign intentions, among the same respondents, with 95% confidence intervals",
                "Part du vote déclaré de chaque parti moins sa part des intentions de campagne, chez les mêmes répondants, avec intervalles de confiance à 95 %"),
  alt = alt, table = tab_gains, height = 5,
  note = tr("Source: qesR. All respondents interviewed during the campaign and after the election, including the undecided, those who said they would not vote and those who did not vote, so that the gains of the parties and of abstention add up. Filled dots: weighted with the post-election weight (2018 panel, QES 2022); hollow dots: not weighted here, because their weights are under review (2007 and 2012 panels). The 1998 polls are left out: their recontact over-represents the undecided, non-disclosers and third-party (ADQ) supporters, so their unweighted gains cannot be read (see the text). Parties that did not run, or that the question did not list, are left out.",
            "Source : qesR. Tous les répondants interrogés pendant la campagne et après l'élection, y compris les indécis, ceux qui disaient ne pas vouloir voter et ceux qui n'ont pas voté, de sorte que les gains des partis et de l'abstention s'additionnent. Points pleins : pondérés avec la pondération postélectorale (panel 2018, EEQ 2022) ; points creux : non pondérés ici, les pondérations étant en révision (panels 2007 et 2012). Les sondages de 1998 sont omis : leur recontact surreprésente les indécis, les discrets et les partisans des tiers partis (ADQ), de sorte que leurs gains non pondérés ne peuvent être lus (voir le texte). Les partis qui ne se présentaient pas, ou que la question ne proposait pas, sont omis."))
```

Most gaps are small, within a few points of zero. This is what one would
expect when most voters keep their intention and the remaining moves
partly cancel out. The two largest gains went to parties that went on to
form the government: the CAQ in 2018, with
`r qz_num(gv("qes2018_panel", "CAQ"), 1)` points (95% confidence interval:
`r qz_num(gv("qes2018_panel", "CAQ", "lo"), 1)` to
`r qz_num(gv("qes2018_panel", "CAQ", "hi"), 1)`), and the PQ in 2012, with
`r qz_num(gv("qes2012_panel", "PQ"), 1)` points
(`r qz_num(gv("qes2012_panel", "PQ", "lo"), 1)` to
`r qz_num(gv("qes2012_panel", "PQ", "hi"), 1)`). The 2018 gain fits
Durand and Blais's (2020) account. Winning the government did not require
winning the campaign. The PLQ in 2007 and the CAQ in 2022 formed
the government while losing ground during the campaign (by
`r qz_num(-gv("qes2007_panel", "PLQ"), 1)` and
`r qz_num(-gv("qes2022", "CAQ"), 1)` points), and the PQ gained
`r qz_num(gv("qes2022", "PQ"), 1)` points in 2022 without forming it.

We leave the 1998 polls out of the figure, because their net gains cannot
be read as they stand. The 1998 panel in qesR is made of two of the three
polls that Durand and Blais (1999) recontacted, those of CROP and
CREATEC, which interviewed francophones only. The recontact was drawn to
over-represent the undecided, non-disclosers and supporters of third
parties, and it is analysed here without weights (its weights are under
review). Unweighted, those who intended to vote ADQ make up
`r qz_pct(adq98["unweighted"], 0)` of the pairs; with the weight the
producers used for their own published tables, they make up
`r qz_pct(adq98["weighted"], 0)`. The unweighted numbers show the PLQ and
the PQ gaining about as much (`r qz_num(gv("qes1998", "PLQ"), 1)` and
`r qz_num(gv("qes1998", "PQ"), 1)` points), so there is no Liberal swing
at the PQ's expense, and the ADQ losing
`r qz_num(-gv("qes1998", "ADQ"), 1)` points. With the producers' weight,
the PLQ's gain is close to zero (`r qz_num(gw98("PLQ"), 1)` points) and
the PQ's loss is small (`r qz_num(-gw98("PQ"), 1)` points), both with
confidence intervals that include zero, while the ADQ's loss shrinks to
`r qz_num(-gw98("ADQ"), 1)` points. Either way, these numbers are
consistent with Durand and Blais (1999). This is hardly surprising: they
come from the same respondents, so they confirm the original analysis
rather than test it independently.

## What the numbers mean and their limits

Overall, we find mixed support for the view that Quebec elections are
decided at the last minute. On the one hand, from
`r qz_pct(100 - pr("same")[2], 0)` to `r qz_pct(100 - pr("same")[1], 0)`
of voters did not vote as they had said, and the CAQ of 2018 gained ground
during the campaign itself. On the other hand, most of these moves
cancel out, the undecided are few and abstain more often than the other
voters, and the net effect of the campaign on a party's share is usually
a few points. Two limits of these data call for caution.

First, the panels see only those who stayed. Respondents who dropped out
after the campaign are not in these tables, and if leaving the panel is
related to changing one's mind, the transitions lean toward stability.
Second, the vote after the election is a vote that respondents report
once the winner is known. Some voters may recall a vote closer to the
result than the one they cast, which would inflate the gains of the
winner, the CAQ in 2018 included. If recall simply leaned toward the
winner, though, the party that formed the government would gain every
time, and it lost ground in 2007 and 2022. The `r num_word(n_panels)`
studies also differ in design and mode, and `r num_word(n_unweighted)` of
them are analysed without weights, so comparisons across elections are
best read as orders of magnitude.

That most voters keep their intention across `r num_word(n_panels)`
elections and several survey designs is itself a finding. Of course, this
stability may not survive more fragmented elections, with
`r num_word(n_above)` parties above `r qz_pct(above, 0)` of the vote as in
2022; that is a question the next panels will have to answer.

## Going further

- Durand, Claire, and André Blais. 1999. "Why Did the Polls Go Wrong in
  the 1998 Quebec Election? The Answer from Post-Election Polls."
  *Bulletin of Sociological Methodology / Bulletin de Méthodologie
  Sociologique* 62 (1): 43–47.
  <https://doi.org/10.1177/075910639906200105>
- Durand, Claire, and André Blais. 2020. "Quebec 2018: A Failure of the
  Polls?" *Canadian Journal of Political Science* 53 (1): 133–150.
  <https://doi.org/10.1017/S0008423919000787>
- Fournier, Patrick, Richard Nadeau, André Blais, Elisabeth Gidengil, and
  Neil Nevitte. 2004. "Time-of-Voting Decision and Susceptibility to
  Campaign Effects." *Electoral Studies* 23 (4): 661–681.
  <https://doi.org/10.1016/j.electstud.2003.09.001>
- Johnston, Richard, André Blais, Henry E. Brady, and Jean Crête. 1992.
  *Letting the People Decide: Dynamics of a Canadian Election*. Montreal
  and Kingston: McGill-Queen's University Press.
  <https://doi.org/10.1515/9780773563643>
- On the PCQ voters of 2022, from the same Quebec Election Study:
  Bélanger, Éric, Philippe Mongrain, Thomas Gareau-Paquette, and
  Valérie-Anne Mahéo. 2025. "A Party that Went Viral? The Drivers of
  Support for the Parti Conservateur du Québec in the 2022 Election."
  *Canadian Journal of Political Science* 58 (2): 277–296.
  <https://doi.org/10.1017/S0008423924000829>

## About the data

- **Studies.** The studies that interviewed the same respondents during
  the campaign and after the election: the 1998 CROP and CREATEC polls
  (francophones only), the panels Claire Durand ran in 2007, 2012 and
  2018, and the 2022 Quebec Election Study. Only respondents of both
  waves are used. The 2022 study was an online panel, the 2018 panel
  mixed telephone and web, and the older studies were by telephone.
- **Intention.** The party named during the campaign. Respondents who
  hesitated were asked which party they leaned to, and that answer
  counts; those who still named no party are the undecided. In the 2018
  panel, some undecided respondents interviewed by telephone and those
  who said they would not vote were not asked the follow-up, so that
  panel counts more undecided. In the 2007 panel, a few respondents keep
  their first answer, where the producer's combined answer cannot be
  used.
- **Vote.** The party reported after the election, or did not vote.
- **Weights.** The post-election weight of each respondent in the 2018
  panel and the 2022 study. The 1998, 2007 and 2012 panels are not
  weighted here, because their weights are under review, and they are
  drawn hollow. The 1998 recontact also over-represents the undecided,
  non-disclosers and third-party (ADQ) supporters, so its net gains are
  left out of the last figure; the text compares them with and without
  the weight the producers used.
- **Attrition.** Respondents who left the panel after the campaign are
  not in these tables.
- **Code.** Every figure starts from the pooled `vote_choice`, in the
  long layout (one row per respondent and wave), matched within each
  study.

<details>
<summary>The call that builds the data</summary>

```{r show-harmonize, ref.label = "harmonize", eval = FALSE}
```

</details>
