---
title: "Validation against official results and the census"
params:
  lang: "en"
---

*[Version française](fr-validation.html)*

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 6)
# messages in the language of the page, whatever the locale of the build
options(qesR.lang = params$lang)
```

How close do the harmonized studies come to what is known about the
electorate? This page compares, for every study the harmonization
specification covers:

1. the **reported vote** with the official results of each general
   election, published by Élections Québec;
2. the **reported turnout** with the official turnout;
3. the **gender, age, mother tongue and education** of respondents with the
   census that precedes the study (Statistics Canada);
4. a few **relationships between answers** that any valid measure should
   show (construct validity).

The page is built when the website is built: the data files are downloaded
from their Dataverse deposits through qesR's cache and harmonized with
`qes_harmonize()`. The benchmarks are in `inst/extdata/validation/` of the
qesR source repository, each row with its source; the census margins also
ship with the package, the official results of Élections Québec and the
recorded report do not (their terms of use need Élections Québec's written
permission, which is pending). The same checks run every week on the pinned
files and fail when the weighted recall index (check V-L2 of the
specification) or a weighted census index rises more than 2 points above
its recorded value, when turnout over-reporting leaves 0 to 35 points, or
when a construct-validity direction (check V-L4) fails. An index that
falls, or rises by less, passes.

```{r packages, message = FALSE, warning = FALSE}
library(qesR)
library(ggplot2)
```

```{r run, message = FALSE}
tr <- function(en, fr) if (identical(params$lang, "fr")) fr else en
num <- function(x, digits = 1, sign = FALSE) {
  formatC(x, format = "f", digits = digits, flag = if (sign) "+" else "",
          decimal.mark = tr(".", ","))
}
# every study the specification covers, harmonized from the pinned files,
# against the benchmarks that ship with qesR
report <- qesR:::.qes_validation_run("all")
report <- qesR:::.qes_validation_gate(report, qesR:::.qes_validation_recorded())
studies <- qes_studies()
report$year <- studies$year[match(report$study, studies$study)]
report$weighting <- ifelse(report$weight == "none", tr("unweighted", "non pondéré"),
                           tr("weighted", "pondéré"))
# one row per study and check: the weighted value where the study has a
# reviewed weight, the unweighted one otherwise
best <- function(x) {
  x <- x[order(x$study, x$variable, x$weight == "none"), ]
  x[!duplicated(paste(x$study, x$variable)), ]
}
```

## 1. Reported vote and official results

For each study, the table gives the **dissimilarity index** between the
distribution of the reported vote and the official shares of valid votes:
half the sum of the absolute differences between the two distributions, in
points. It is the share of respondents who would have to change party for
the two to agree; 0 is a perfect match. Parties that the study's question
did not list are counted as "other party" on the official side, since that
is the only answer their voters could give.

```{r recall-table}
idx <- report[report$check == "recall" & is.na(report$level), ]
wide <- reshape(idx[, c("study", "year", "reference", "weighting", "n", "value")],
                idvar = c("study", "year", "reference"), timevar = "weighting",
                direction = "wide", drop = "n")
wide <- merge(wide, aggregate(n ~ study, idx, max), by = "study")
wide <- wide[order(wide$year, wide$study), ]
un <- wide[[paste0("value.", tr("unweighted", "non pondéré"))]]
we <- wide[[paste0("value.", tr("weighted", "pondéré"))]]
knitr::kable(
  data.frame(wide$study, wide$reference, wide$n, num(un),
             ifelse(is.na(we), "", num(we))),
  col.names = tr(c("Study", "Election", "N (reported a party)", "Index, unweighted", "Index, weighted"),
                 c("Étude", "Élection", "N (ont nommé un parti)", "Indice, non pondéré", "Indice, pondéré")),
  align = c("l", "l", "r", "r", "r")
)
```

The weighted index uses the weight the specification recommends for the
wave that asked the question; only `qes2012`, `qes2014`, `qes2018` and
`qes2022` have reviewed weights so far. The other studies' weights are not
documented well enough to be used (their rows are `needs_review`), so their
index is unweighted. `qes1998` interviewed francophones only, a population
the official results do not describe: it is shown for reference but not
checked.

```{r recall-plot, fig.height = 8, fig.alt = tr("For each study, the difference in points between the reported vote share of each party and its official share of valid votes; weighted where the study has a reviewed weight.", "Pour chaque étude, l'écart en points entre la part du vote déclaré de chaque parti et sa part officielle des votes valides ; pondéré lorsque l'étude a une pondération révisée.")}
# the weighted distribution where the study has a reviewed weight
shown <- best(idx[idx$study != "qes1998", ])
lv <- report[report$check == "recall" & !is.na(report$level) &
               paste(report$study, report$weight) %in% paste(shown$study, shown$weight), ]
lv$party <- ifelse(lv$level == "other", tr("Other", "Autre"), lv$level)
lv$label <- paste0(lv$study, " (", lv$weighting, ")")
lv$label <- factor(lv$label, unique(lv$label[order(lv$year, lv$study)]))
ggplot(lv, aes(x = value, y = party)) +
  geom_vline(xintercept = 0, colour = "grey50") +
  geom_segment(aes(x = 0, xend = value, yend = party), colour = "grey70") +
  geom_point(size = 2.2, colour = "#0468b9") +
  facet_wrap(~label, ncol = 3, scales = "free_y") +
  labs(
    x = tr("Reported minus official share (points)", "Part déclarée moins part officielle (points)"),
    y = NULL
  ) +
  theme_minimal(base_size = 11)
```

```{r recall-numbers, include = FALSE}
d <- function(study, level) {
  x <- lv$value[lv$study == study & lv$level == level]
  if (length(x)) x else NA_real_
}
```

## 2. Reported turnout

```{r turnout-table}
to <- best(report[report$check == "turnout", ])
to <- to[order(to$year, to$study), ]
# qes1998 (francophones only) is shown but not checked
checked <- to[!(to$status %in% "skipped"), ]
knitr::kable(
  data.frame(to$study, to$reference, to$weighting, to$n, num(to$estimate), num(to$benchmark),
             num(to$value, sign = TRUE)),
  col.names = tr(c("Study", "Election", "Weighting", "N", "Reported turnout (%)", "Official turnout (%)", "Difference (points)"),
                 c("Étude", "Élection", "Pondération", "N", "Participation déclarée (%)", "Participation officielle (%)", "Écart (points)")),
  align = c("l", "l", "l", "r", "r", "r", "r")
)
```

Official turnout is ballots cast over registered electors. Reported turnout
is the share who say they voted among respondents who answered yes or no;
those not registered or not eligible, and those who did not know or would
not say, are left out.

## 3. Respondents and the census

Each study is compared with the census that precedes its election (2006,
2011, 2016 or 2021). Statistics Canada publishes 18-and-over counts only
for gender and age in 2016 and 2021, so the nearest published age cut is
used elsewhere, and respondents are cut the same way when their age is
known:

| Margin | Census | Population compared | Table |
|---|---|---|---|
| Gender | 2016, 2021 | 18 and over | 98-10-0020-01 |
| Gender | 2006, 2011 | 20 and over, private households | 98-10-0384-01 |
| Age (six bands) | 2016, 2021 | 18 and over | 98-10-0020-01 |
| Age (five bands) | 2006, 2011 | 25 and over, private households | 98-10-0384-01 |
| Mother tongue | 2011, 2016, 2021 | 20 and over, outside institutions, one mother tongue | 98-10-0218-01 |
| Education | 2006 to 2021 | 25 and over, private households | 98-10-0384-01 |

The 2011 education and gender counts come from the National Household
Survey, which was voluntary. Mother tongue is compared among people who
report one mother tongue: the census's multiple answers and the surveys'
answers naming two languages are left out. Education is compared as
university (any university certificate, diploma or degree) or below.

```{r census-table}
cz <- report[report$check == "census" & is.na(report$level), ]
vars <- c(gender = tr("Gender", "Genre"), age_group6 = tr("Age", "Âge"),
          lang_mother = tr("Mother tongue", "Langue maternelle"),
          education = tr("Education", "Scolarité"))
cz$cell <- num(cz$value)
cw <- reshape(cz[, c("study", "year", "reference", "variable", "weighting", "cell")],
              idvar = c("study", "year", "reference", "weighting"), timevar = "variable",
              direction = "wide")
names(cw) <- sub("^cell\\.", "", names(cw))
for (v in names(vars)) if (is.null(cw[[v]])) cw[[v]] <- NA
cw <- cw[order(cw$year, cw$study, cw$weighting), ]
out <- cw[, c("study", "reference", "weighting", names(vars))]
out[is.na(out)] <- ""
knitr::kable(
  out, row.names = FALSE,
  col.names = c(tr(c("Study", "Census", "Weighting"), c("Étude", "Recensement", "Pondération")), unname(vars)),
  align = c("l", "l", "l", "r", "r", "r", "r")
)
```

Each cell is a dissimilarity index in points (0 is a perfect match); a
blank cell is a margin the study did not ask, or one the collected census
tables do not publish for that year (mother tongue in 2006).

```{r census-plot, fig.height = 7, fig.alt = tr("For the four studies with reviewed weights, the difference in points between the weighted share of each category of gender, age, mother tongue and education and its census share.", "Pour les quatre études dont la pondération est révisée, l'écart en points entre la part pondérée de chaque catégorie de genre, d'âge, de langue maternelle et de scolarité et sa part au recensement.")}
cl <- report[report$check == "census" & !is.na(report$level) & report$weight != "none", ]
cl$variable <- factor(vars[cl$variable], unname(vars))
cl$study <- factor(cl$study, unique(cl$study[order(cl$year)]))
ggplot(cl, aes(x = value, y = level, colour = study, shape = study)) +
  geom_vline(xintercept = 0, colour = "grey50") +
  geom_point(size = 2.2) +
  facet_wrap(~variable, ncol = 2, scales = "free_y") +
  labs(
    x = tr("Weighted survey share minus census share (points)",
           "Part pondérée de l'enquête moins part au recensement (points)"),
    y = NULL, colour = tr("Study", "Étude"), shape = tr("Study", "Étude")
  ) +
  theme_minimal(base_size = 11)
```

## 4. Relationships between answers

These checks do not compare with an outside source. They ask whether the
harmonized answers relate to one another the way decades of research say
they should; a mapping error (a party code swapped, a scale reversed)
would break them.

```{r construct-table}
k <- report[report$check == "construct", ]
k <- k[order(k$year, k$study, k$variable), ]
what <- c(interest_4pt = tr("Voters are more interested than nonvoters (mean, 1-4)",
                            "Les votants sont plus intéressés que les abstentionnistes (moyenne, 1-4)"),
          lr_self = tr("QS voters are to the left of CAQ voters (mean, 0-10)",
                       "Les électeurs de QS sont à gauche de ceux de la CAQ (moyenne, 0-10)"),
          sov_indep = tr("Yes to independence: PQ voters more than 40 points above PLQ voters (%)",
                         "Oui à l'indépendance : électeurs du PQ plus de 40 points au-dessus de ceux du PLQ (%)"),
          pid_prov = tr("Partisans who voted for their party: at least 55% (%)",
                        "Partisans qui ont voté pour leur parti : au moins 55 % (%)"))
knitr::kable(
  data.frame(k$study, what[k$variable], num(k$estimate),
             ifelse(is.na(k$benchmark), "", num(k$benchmark)),
             ifelse(k$status == "pass", tr("yes", "oui"), tr("no", "non"))),
  col.names = tr(c("Study", "Expected", "Value", "Compared with", "Holds"),
                 c("Étude", "Attendu", "Valeur", "Comparé à", "Vérifié")),
  align = c("l", "l", "r", "r", "l")
)
```

## Reading the differences

**Turnout is over-reported in every study**, here by
`r num(min(checked$value), 0)` to `r num(max(checked$value), 0)` points
(`qes1998`, francophones only, is left out). Two causes
add up and cannot be told apart with these data: some nonvoters say they
voted, and people who vote are more willing to answer election surveys in
the first place. The face-saving question formats of `qes2018` and
`qes2022` (several ways to say one did not vote) are meant to reduce the
first; weights calibrated on age, gender, education and language do not
correct either. The two rates also count different people: the official
rate is over registered electors, some of whom have moved or died, while
the surveys ask adults they reached. Any analysis of turnout from these
studies describes respondents, not the electorate.

**The reported vote is a recall, and it is not the vote.** It is asked
days to weeks after the election, of respondents whose participation is
itself selective. Recall is often said to drift toward the winner, but the
pattern here is not a simple winner's bonus: in `qes2012` the PQ, which won
by less than a point, is over-reported by `r num(d("qes2012", "PQ"))`
points and the PLQ under-reported by `r num(-d("qes2012", "PLQ"))`; in
`qes2022` the CAQ, which won, is under-reported by
`r num(-d("qes2022", "CAQ"))` points. Who answers matters at least as much
as how people remember: online panels tend to over-represent the politically
engaged and the educated, and the voters of some parties may be harder to
reach. Weighting on demographics does not remove this and can move the
index either way (compare the two columns of the first table).

**The question matters.** A party the question did not list could only be
reported as "other" (the Green Party in `qes2018` and `qes2022`, the
Conservative Party in `qes2018`). The two Durand panels of 2007 and 2012
ask the vote without reading the parties, by telephone; the other studies
read or show a list.
The grades of `qes_spec(view = "crosswalk")` record these differences.

**Samples are not the census population.** The census counts residents,
including people who are not citizens and cannot vote; `qes2022` sampled
citizens only. The census tables for mother tongue and education cover
private households, and their age cuts differ from 18 and over (section 3).
Three gaps stand out. Education: every study has more respondents with some
university than the census has holders of a university credential, partly
because the surveys ask the level reached (a year of university counts)
and the census the credential completed, and partly, most likely, because
people with more education are more willing to answer surveys; `qes2022`, weighted on education, is the only study
close to the census. Mother tongue in `qes2018`: English speakers are
over-represented and speakers of other languages under-represented; the
2018 weight corrects French against all other languages (its methodological
report, Table 14), not the split between English and other languages. Age
in `qes2018`: its weight corrects age in generational cells (16-18, 19-38,
39-58, 59-73, 74 and over: Table 12 of the same report), which cut across
the ten-year bands used here; within them the weighted sample still has too
few people aged 35 to 54 and too many aged 55 to 64.

**These benchmarks are coarse.** An index can hide errors that offset each
other: in 2012 the PQ and the PLQ finished less than a point apart, so a
study that swapped their codes would match the official results about as
well. That is why the harmonization is also checked against exact counts of
each answer in the original files and against the relationships of
section 4, and why each crosswalk row is reviewed against the
questionnaire.

## Sources

- Élections Québec, official results of the general elections of 1998 to
  2022, province-wide votes by party and ballots cast, from its archived
  results files (for example
  <https://donnees.electionsquebec.qc.ca/production/provincial/resultats/archives/gen2022-10-03/resultats.json>).
- Statistics Canada, census tables
  [98-10-0020-01](https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=9810002001)
  (age and gender, 2016 and 2021),
  [98-10-0218-01](https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=9810021801)
  (mother tongue by age, 2011 to 2021) and
  [98-10-0384-01](https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=9810038401)
  (highest certificate by age and gender, 2006 to 2021).

The benchmark tables are in `inst/extdata/validation/` of the source
repository, with the source of every row; the installed package
(`system.file("extdata", "validation", package = "qesR")`) has only the
census margins. The recorded results of the studies released under CC0
are in `validation_report.csv` there; those of
`qes2022`, whose licence (CC BY-NC 4.0) does not allow qesR to ship its
aggregates, are computed on this page from your copy of the file at build
time.
