---
title: "Validation against official results and the census"
params:
  lang: "en"
---

*[Version française](fr-validation.html)*

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE, fig.width = 7, fig.height = 6)
# messages in the language of the page, whatever the locale of the build
options(qesR.lang = params$lang)
library(qesR)
# numbers and labels in the page's language
tr <- function(en, fr) if (identical(params$lang, "fr")) fr else en
num <- function(x, digits = 1, sign = FALSE) {
  formatC(x, format = "f", digits = digits, flag = if (sign) "+" else "",
          decimal.mark = tr(".", ","))
}
# what the page downloads (the note below), from the site's data helpers
source("_data.R", encoding = "UTF-8")
dl <- qz_download(qz_studies)
```

::: {.qesr-download}
**This page downloads `r dl$n` studies** (about `r round(dl$mb)` MB) the first time it runs. `options(qesR.cache = "disk")` keeps them on disk for later sessions.
:::

How close do the harmonized studies come to what is known about the
electorate? This page compares, for every study the harmonization
specification covers:

1. the **reported vote** with the official results of each general
   election, published by Élections Québec;
2. the **reported turnout** with the official turnout;
3. the **gender, age, mother tongue and education** of respondents with the
   census that precedes the study (Statistics Canada);
4. a few **relationships between answers** that any valid measure should
   show (construct validity).

Every study is harmonized with `qes_harmonize()` from its original data
file. The same comparisons run automatically on the original files, so a
mapping error that moved these numbers would be caught.

```{r packages, message = FALSE, warning = FALSE}
library(qesR)
library(ggplot2)
```

```{r run, message = FALSE}
# every study the specification covers, harmonized from the original files,
# against the benchmarks that ship with qesR (internal functions, not part
# of qesR's interface)
report <- qesR:::.qes_validation_run("all")
report <- qesR:::.qes_validation_gate(report, qesR:::.qes_validation_recorded())
```

```{r run-tidy, include = FALSE}
studies <- qes_studies()
report$year <- studies$year[match(report$study, studies$study)]
report$weighting <- ifelse(report$weight == "none", tr("unweighted", "non pondéré"),
                           tr("weighted", "pondéré"))
# one row per study and check: the weighted value where the study has a
# usable weight, the unweighted one otherwise
best <- function(x) {
  x <- x[order(x$study, x$variable, x$weight == "none"), ]
  x[!duplicated(paste(x$study, x$variable)), ]
}
```

## 1. Reported vote and official results

For each study, the table gives the **dissimilarity index** between the
distribution of the reported vote and the official shares of valid votes:
half the sum of the absolute differences between the two distributions, in
points. It is the share of respondents who would have to change party for
the two to agree; 0 is a perfect match. Parties that the study's question
did not list are counted as "other party" on the official side, since that
is the only answer their voters could give.

```{r recall-table, echo = FALSE}
idx <- report[report$check == "recall" & is.na(report$level), ]
wide <- reshape(idx[, c("study", "year", "reference", "weighting", "n", "value")],
                idvar = c("study", "year", "reference"), timevar = "weighting",
                direction = "wide", drop = "n")
wide <- merge(wide, aggregate(n ~ study, idx, max), by = "study")
wide <- wide[order(wide$year, wide$study), ]
un <- wide[[paste0("value.", tr("unweighted", "non pondéré"))]]
we <- wide[[paste0("value.", tr("weighted", "pondéré"))]]
knitr::kable(
  data.frame(wide$study, wide$reference, wide$n, num(un),
             ifelse(is.na(we), "", num(we))),
  col.names = tr(c("Study", "Election", "N (reported a party)", "Index, unweighted", "Index, weighted"),
                 c("Étude", "Élection", "N (ont nommé un parti)", "Indice, non pondéré", "Indice, pondéré")),
  align = c("l", "l", "r", "r", "r")
)
```

The weighted index uses the weight the specification recommends for the
wave that asked the question; `qes2007`, `qes2012`, `qes2014`, `qes2018`,
`qes2022` and the 2018 panel have one. The weights of `qes2007_panel`,
`qes2012_panel`, the CROP polls and `qes1998` are not validated, and
`qes2008` has no recommended weight (both of its weights are calibrated on
the vote or on turnout), so their index is unweighted; the
[weights table](studies.html#weights) describes each study's weight. `qes1998` interviewed francophones only, a population
the official results do not describe: it is shown for reference but not
checked.

```{r recall-plot, echo = FALSE}
# the site's chart system (vignettes/articles/_theme.R): each figure drawn
# for the light and the dark theme, with its table view
qz_num <- function(x, digits = 1) num(x, digits)
source("_theme.R", encoding = "UTF-8", local = TRUE)
source("_data.R", encoding = "UTF-8", local = TRUE)
# the weighted distribution where the study has a usable weight
shown <- best(idx[idx$study != "qes1998", ])
lv <- report[report$check == "recall" & !is.na(report$level) &
               paste(report$study, report$weight) %in% paste(shown$study, shown$weight), ]
lv$weighted <- lv$weight != "none"
# one row per party (the same order in every study), one column per study
party_order <- c("PLQ", "PQ", "ADQ", "QS", "CAQ", "PCQ", "PVQ", "ON", "other")
lv$party <- factor(lv$level, levels = rev(intersect(party_order, lv$level)))
cols <- unique(lv[order(lv$year, lv$study), c("study", "weighted")])
col_lab <- paste0(qz_study_label(cols$study), ifelse(cols$weighted, "", tr("\nunweighted", "\nnon pondéré")))
lv$col <- factor(col_lab[match(lv$study, cols$study)], levels = col_lab)
# signed, rounded first so that a difference that rounds to zero is "0.0", not "-0.0"
signed <- function(x) { r <- round(x, 1); r[r == 0] <- 0; num(r, 1, sign = TRUE) }
lv$cell <- signed(lv$value)
lim <- max(abs(lv$value), na.rm = TRUE)
big <- lv[abs(lv$value) >= 5, ]
plq <- lv$value[lv$level == "PLQ" & lv$year >= 2007 & lv$year <= 2014]
caq22 <- lv$value[lv$level == "CAQ" & lv$study == "qes2022"]
fig_recall <- function(mode) {
  k <- qz_tokens[[mode]]
  ramp <- grDevices::colorRampPalette(c(k$no, k$wash, k$yes))(201)
  lv$ink <- qz_label_ink(ramp[round(100 * lv$value / lim) + 101], mode)
  ggplot(lv, aes(col, party)) +
    qz_tile(mode, aes(fill = value)) +
    # differences of 5 points or more are outlined
    geom_tile(data = big, fill = NA, colour = k$ink, linewidth = 0.6, width = 0.94, height = 0.9) +
    geom_text(aes(label = cell, colour = ink), size = 2.8, family = qz_font()) +
    scale_fill_gradient2(low = k$no, mid = k$wash, high = k$yes, midpoint = 0, limits = c(-lim, lim),
                         labels = function(x) num(x, 0),
                         guide = guide_colourbar(title.position = "top", barwidth = unit(140, "pt"), barheight = unit(7, "pt"))) +
    scale_colour_identity() +
    scale_x_discrete(position = "top") +
    scale_y_discrete(labels = function(x) qz_party_label(x)) +
    labs(x = NULL, y = NULL,
         fill = tr("Reported minus official share (points); outlined: 5 points or more",
                   "Part déclarée moins part officielle (points) ; encadré : 5 points ou plus")) +
    theme_qes(mode, grid = "none") +
    theme(legend.position = "bottom", legend.title = element_text(colour = k$ink2, size = 9.4),
          legend.ticks = element_blank(), axis.text.x.top = element_text(lineheight = 0.9, vjust = 0))
}
tab <- lv[order(lv$year, lv$study, match(lv$level, party_order)), ]
qz_figure(fig_recall, "recall",
  title = sprintf(tr("Surveys under-report the PLQ at every election from 2007 to 2014, by %s to %s points, and the CAQ by %s points in 2022",
                     "Les enquêtes sous-estiment le PLQ à chaque élection de 2007 à 2014, de %s à %s points, et la CAQ de %s points en 2022"),
                  num(min(abs(plq)), 0), num(max(abs(plq)), 0), num(abs(caq22), 0)),
  subtitle = tr("Reported minus official vote share, in points, by party (rows) and study (columns)",
                "Part du vote déclaré moins part officielle, en points, selon le parti (rangées) et l'étude (colonnes)"),
  alt = tr("Heatmap of the difference in points between each party's share of the reported vote and its official share of valid votes, one row per party and one column per study from 2007 to 2022; blue cells are over-reported, red cells under-reported, and differences of 5 points or more are outlined. The PLQ is under-reported in every study from 2007 to 2014 and the CAQ in 2022. Values in the table view.",
           "Carte de chaleur de l'écart en points entre la part de chaque parti dans le vote déclaré et sa part officielle des votes valides, une rangée par parti et une colonne par étude de 2007 à 2022 ; les cases bleues sont surestimées, les rouges sous-estimées, et les écarts de 5 points ou plus sont encadrés. Le PLQ est sous-estimé dans chaque étude de 2007 à 2014, et la CAQ en 2022. Valeurs dans la vue en tableau."),
  table = setNames(data.frame(qz_study_label(tab$study), tab$weighting, qz_party_label(tab$level), signed(tab$value),
                              row.names = NULL),
                   tr(c("Study", "Weighting", "Party", "Reported minus official (points)"),
                      c("Étude", "Pondération", "Parti", "Déclaré moins officiel (points)"))),
  note = tr("Weighted with the study's weight where it has a usable one; the 2008 study and the 2007 and 2012 Durand panels are unweighted. A blank cell: the party did not run, or the study counts it in Other. The 1998 polls, which interviewed francophones only, are left out.",
            "Pondéré avec la pondération de l'étude lorsqu'elle en a une utilisable ; l'étude de 2008 et les panels Durand de 2007 et 2012 sont non pondérés. Une case vide : le parti ne s'est pas présenté, ou l'étude le compte dans Autres. Les sondages de 1998, qui n'ont interrogé que des francophones, sont laissés de côté."),
  height = 4.6)
```

```{r recall-numbers, include = FALSE}
d <- function(study, level) {
  x <- lv$value[lv$study == study & lv$level == level]
  if (length(x)) x else NA_real_
}
```

## 2. Reported turnout

```{r turnout-table, echo = FALSE}
to <- best(report[report$check == "turnout", ])
to <- to[order(to$year, to$study), ]
# qes1998 (francophones only) is shown but not checked
checked <- to[!(to$status %in% "skipped"), ]
knitr::kable(
  data.frame(to$study, to$reference, to$weighting, to$n, num(to$estimate), num(to$benchmark),
             num(to$value, sign = TRUE)),
  col.names = tr(c("Study", "Election", "Weighting", "N", "Reported turnout (%)", "Official turnout (%)", "Difference (points)"),
                 c("Étude", "Élection", "Pondération", "N", "Participation déclarée (%)", "Participation officielle (%)", "Écart (points)")),
  align = c("l", "l", "l", "r", "r", "r", "r")
)
```

Official turnout is ballots cast over registered electors. Reported turnout
is the share who say they voted among respondents who answered yes or no;
those not registered or not eligible, and those who did not know or would
not say, are left out.

## 3. Respondents and the census

Each study is compared with the census that precedes its election (2006,
2011, 2016 or 2021). Gender and age are compared at 18 and over; for
mother tongue and education, whose tables have no 18-and-over cut, the
nearest published age cut is used, and respondents are cut the same way
when their age is known:

| Margin | Census | Population compared | Table |
|---|---|---|---|
| Gender, age (six bands) | 2016, 2021 | 18 and over | 98-10-0020-01 |
| Gender, age (six bands) | 2011 | 18 and over | Census Profile 98-316-XWE2011001 |
| Gender, age (six bands) | 2006 | 18 and over | 97-551-XCB2006009 |
| Mother tongue | 2011, 2016, 2021 | 20 and over, outside institutions, one mother tongue | 98-10-0218-01 |
| Education | 2006 to 2021 | 25 and over, private households | 98-10-0384-01 |

The 2011 education counts come from the National Household Survey, which
was voluntary. The 2006 table 97-551-XCB2006009 is no longer on Statistics
Canada's site; its file was retrieved from the Internet Archive. The one
2006 table found with mother tongue by age counts institutional residents
and many more multiple mother tongues, so 2006 has no mother-tongue
comparison. Mother tongue is compared among people who
report one mother tongue: the census's multiple answers and the surveys'
answers naming two languages are left out. Education is compared as
university (any university certificate, diploma or degree) or below.

```{r census-table, echo = FALSE}
cz <- report[report$check == "census" & is.na(report$level), ]
vars <- c(gender = tr("Gender", "Genre"), age_group6 = tr("Age", "Âge"),
          lang_mother = tr("Mother tongue", "Langue maternelle"),
          education = tr("Education", "Scolarité"))
cz$cell <- num(cz$value)
cw <- reshape(cz[, c("study", "year", "reference", "variable", "weighting", "cell")],
              idvar = c("study", "year", "reference", "weighting"), timevar = "variable",
              direction = "wide")
names(cw) <- sub("^cell\\.", "", names(cw))
for (v in names(vars)) if (is.null(cw[[v]])) cw[[v]] <- NA
cw <- cw[order(cw$year, cw$study, cw$weighting), ]
out <- cw[, c("study", "reference", "weighting", names(vars))]
out[is.na(out)] <- ""
knitr::kable(
  out, row.names = FALSE,
  col.names = c(tr(c("Study", "Census", "Weighting"), c("Étude", "Recensement", "Pondération")), unname(vars)),
  align = c("l", "l", "l", "r", "r", "r", "r")
)
```

Each cell is a dissimilarity index in points (0 is a perfect match); a
blank cell is a margin the study did not ask, or one the collected census
tables do not publish for that year (mother tongue in 2006).

```{r census-plot, echo = FALSE}
cl <- report[report$check == "census" & !is.na(report$level) & report$weight != "none", ]
# a two-category margin is one row (women; university): the other row mirrors it
cl <- cl[!cl$level %in% c("man", "below_university"), ]
# the story first: education and mother tongue, then age and gender
var_order <- c("education", "lang_mother", "age_group6", "gender")
cl$variable <- factor(vars[cl$variable], unname(vars[var_order]))
cl$study_lab <- factor(qz_study_label(cl$study), unique(qz_study_label(cl$study[order(cl$year)])))
level_lab <- c(man = tr("Men", "Hommes"), woman = tr("Women", "Femmes"),
               a18_24 = "18-24", a25_34 = "25-34", a35_44 = "35-44", a45_54 = "45-54",
               a55_64 = "55-64", a65_plus = tr("65 and over", "65 et plus"),
               french = tr("French", "Français"), english = tr("English", "Anglais"),
               other = tr("Other", "Autre"), below_university = tr("Below university", "Sous l'université"),
               university = tr("University", "Université"))
cl$level_label <- ifelse(cl$level %in% names(level_lab), level_lab[cl$level], cl$level)
cl$row <- factor(cl$level_label, levels = rev(unique(cl$level_label[order(cl$variable)])))
# a margin the study's weight is raked on is 0 by design: grey, "cal."
span <- stats::aggregate(value ~ study + variable, cl, function(v) max(abs(v)))
cl$calibrated <- paste(cl$study, cl$variable) %in% paste(span$study, span$variable)[span$value < 0.5]
cl$cell <- ifelse(cl$calibrated, tr("cal.", "cal."), signed(cl$value))
lim <- max(abs(cl$value[!cl$calibrated]), na.rm = TRUE)
uni <- cl$value[cl$level == "university"]
eng18 <- cl$value[cl$level == "english" & cl$study == "qes2018"]
fig_census <- function(mode) {
  k <- qz_tokens[[mode]]
  ramp <- grDevices::colorRampPalette(c(k$no, k$wash, k$yes))(201)
  cl$ink <- ifelse(cl$calibrated, k$ink2, qz_label_ink(ramp[round(100 * pmax(-1, pmin(1, cl$value / lim))) + 101], mode))
  ggplot(cl, aes(study_lab, row)) +
    qz_tile(mode, aes(fill = ifelse(calibrated, NA, value))) +
    geom_text(aes(label = cell, colour = ink), size = 2.7, family = qz_font()) +
    facet_grid(variable ~ ., scales = "free_y", space = "free_y") +
    scale_fill_gradient2(low = k$no, mid = k$wash, high = k$yes, midpoint = 0, limits = c(-lim, lim),
                         labels = function(x) num(x, 0), na.value = k$wash,
                         guide = guide_colourbar(title.position = "top", barwidth = unit(140, "pt"), barheight = unit(7, "pt"))) +
    scale_colour_identity() +
    scale_x_discrete(position = "top") +
    labs(x = NULL, y = NULL,
         fill = tr("Survey minus census (points); cal.: within half a point on every category",
                   "Enquête moins recensement (points) ; cal. : à moins d'un demi-point partout")) +
    theme_qes(mode, grid = "none") +
    theme(strip.text.y = element_text(angle = 0, hjust = 0, face = "bold", colour = k$ink),
          legend.position = "bottom", legend.title = element_text(colour = k$ink2, size = 9.4),
          legend.ticks = element_blank())
}
tab <- cl[order(cl$variable, cl$level, cl$year), ]
qz_figure(fig_census, "census",
  title = sprintf(tr("After weighting, university graduates are still over-represented by %s to %s points; in 2018, English speakers by %s",
                     "Après pondération, les diplômés universitaires restent surreprésentés de %s à %s points ; en 2018, les anglophones de %s"),
                  num(min(uni), 0), num(max(uni), 0), num(eng18, 0)),
  subtitle = tr("Weighted survey share minus census share, in points, by category and study; margins matched to within half a point, as raking targets are, are marked cal.",
                "Part pondérée de l'enquête moins part au recensement, en points, selon la catégorie et l'étude ; les marges atteintes à moins d'un demi-point, comme le sont les marges de calage, sont marquées cal."),
  alt = tr("Heatmap of the difference in points between the weighted share of each category of education, mother tongue, age and gender and its census share, one column per study with a usable weight; blue cells are over-represented, red cells under-represented, and grey cells marked cal. are margins the weight is raked on. University graduates are over-represented in every study. Values in the table view.",
           "Carte de chaleur de l'écart en points entre la part pondérée de chaque catégorie de scolarité, de langue maternelle, d'âge et de genre et sa part au recensement, une colonne par étude qui a une pondération utilisable ; les cases bleues sont surreprésentées, les rouges sous-représentées, et les cases grises marquées cal. sont des marges sur lesquelles la pondération est calée. Les diplômés universitaires sont surreprésentés dans chaque étude. Valeurs dans la vue en tableau."),
  table = setNames(data.frame(as.character(tab$variable), tab$level_label, as.character(tab$study_lab), signed(tab$value),
                              ifelse(tab$calibrated, tr("yes", "oui"), ""), row.names = NULL),
                   tr(c("Margin", "Category", "Study", "Survey minus census (points)", "Within half a point (cal.)"),
                      c("Marge", "Catégorie", "Étude", "Enquête moins recensement (points)", "À moins d'un demi-point (cal.)"))),
  note = tr("Studies with a usable weight, weighted; the census of the year before each study (Statistics Canada). Gender and education have two categories, so one row is shown (women; university): the other is its mirror. A margin is marked cal. when every category is within half a point of the census, as a margin the weight is raked on is by construction; a close match there says nothing about representativeness.",
            "Études qui ont une pondération utilisable, pondérées ; le recensement qui précède chaque étude (Statistique Canada). Le genre et la scolarité ont deux catégories : une seule rangée est montrée (femmes ; université), l'autre en est le miroir. Une marge est marquée cal. quand chaque catégorie est à moins d'un demi-point du recensement, comme l'est par construction une marge sur laquelle la pondération est calée ; un bon accord n'y dit rien de la représentativité."),
  height = 5.0)
```

## 4. Relationships between answers

These checks do not compare with an outside source. They ask whether the
harmonized answers relate to one another the way decades of research say
they should; a mapping error (a party code swapped, a scale reversed)
would break them.

```{r construct-table, echo = FALSE}
k <- report[report$check == "construct", ]
k <- k[order(k$year, k$study, k$variable), ]
what <- c(interest_4pt = tr("Voters are more interested than nonvoters (mean, 1-4)",
                            "Les votants sont plus intéressés que les abstentionnistes (moyenne, 1-4)"),
          lr_self = tr("QS voters are to the left of CAQ voters (mean, 0-10)",
                       "Les électeurs de QS sont à gauche de ceux de la CAQ (moyenne, 0-10)"),
          sov_indep = tr("Yes to independence: PQ voters more than 40 points above PLQ voters (%)",
                         "Oui à l'indépendance : électeurs du PQ plus de 40 points au-dessus de ceux du PLQ (%)"),
          pid_prov = tr("Partisans who voted for their party: at least 55% (%)",
                        "Partisans qui ont voté pour leur parti : au moins 55 % (%)"))
knitr::kable(
  data.frame(k$study, what[k$variable], num(k$estimate),
             ifelse(is.na(k$benchmark), "", num(k$benchmark)),
             ifelse(k$status == "pass", tr("yes", "oui"), tr("no", "non"))),
  col.names = tr(c("Study", "Expected", "Value", "Compared with", "Holds"),
                 c("Étude", "Attendu", "Valeur", "Comparé à", "Vérifié")),
  align = c("l", "l", "r", "r", "l")
)
```

## Reading the differences

**Turnout is over-reported in every study**, here by
`r num(min(checked$value), 0)` to `r num(max(checked$value), 0)` points
(`qes1998`, francophones only, is left out). Two causes
add up and cannot be told apart with these data: some nonvoters say they
voted, and people who vote are more willing to answer election surveys in
the first place. The face-saving question formats of `qes2018` and
`qes2022` (several ways to say one did not vote) are meant to reduce the
first; weights calibrated on age, gender, education and language do not
correct either. The two rates also count different people: the official
rate is over registered electors, some of whom have moved or died, while
the surveys ask adults they reached. Any analysis of turnout from these
studies describes respondents, not the electorate.

**The reported vote is a recall, and it is not the vote.** It is asked
days to weeks after the election, of respondents whose participation is
itself selective. Recall is often said to drift toward the winner, but the
pattern here is not a simple winner's bonus: in `qes2012` the PQ, which won
by less than a point, is over-reported by `r num(d("qes2012", "PQ"))`
points and the PLQ under-reported by `r num(-d("qes2012", "PLQ"))`; in
`qes2022` the CAQ, which won, is under-reported by
`r num(-d("qes2022", "CAQ"))` points. Who answers matters at least as much
as how people remember: online panels tend to over-represent the politically
engaged and the educated, and the voters of some parties may be harder to
reach. Weighting on demographics does not remove this and can move the
index either way (compare the two columns of the first table).

**The question matters.** A party the question did not list could only be
reported as "other" (the Green Party in `qes2018` and `qes2022`, the
Conservative Party in `qes2018`). The two Durand panels of 2007 and 2012
ask the vote without reading the parties, by telephone; the other studies
read or show a list.
The grades of `qes_spec(view = "crosswalk")` record these differences.

**Samples are not the census population.** The census counts residents,
including people who are not citizens and cannot vote; `qes2022` sampled
citizens only. The census tables for mother tongue and education cover
private households, and their age cuts differ from 18 and over (section 3).
Three gaps stand out. Education: every study has more respondents with some
university than the census has holders of a university credential, partly
because the surveys ask the level reached (a year of university counts)
and the census the credential completed, and partly, most likely, because
people with more education are more willing to answer surveys; `qes2022`, weighted on education, is the only study
close to the census. Mother tongue in `qes2018`: English speakers are
over-represented and speakers of other languages under-represented; the
2018 weight corrects French against all other languages (its methodological
report, Table 14), not the split between English and other languages. Age
in `qes2018`: its weight corrects age in generational cells (16-18, 19-38,
39-58, 59-73, 74 and over: Table 12 of the same report), which cut across
the ten-year bands used here; within them the weighted sample still has too
few people aged 35 to 54 and too many aged 55 to 64.

**These benchmarks are coarse.** An index can hide errors that offset each
other: in 2012 the PQ and the PLQ finished less than a point apart, so a
study that swapped their codes would match the official results about as
well. That is why the harmonization is also checked against exact counts of
each answer in the original files and against the relationships of
section 4, and why each mapping is checked against the questionnaire.

## Sources

- Élections Québec, official results of the general elections of 1998 to
  2022, province-wide votes by party and ballots cast, from its archived
  results files (for example
  <https://donnees.electionsquebec.qc.ca/production/provincial/resultats/archives/gen2022-10-03/resultats.json>).
- Statistics Canada, census tables
  [98-10-0020-01](https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=9810002001)
  (age and gender, 2016 and 2021),
  [98-10-0218-01](https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=9810021801)
  (mother tongue by age, 2011 to 2021) and
  [98-10-0384-01](https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=9810038401)
  (highest certificate by age and gender, 2006 to 2021); the 2011 Census
  Profile (catalogue 98-316-XWE2011001: age by sex); and the 2006 table
  97-551-XCB2006009 (age by sex; the Internet Archive's copy of Statistics
  Canada's file).

The benchmark tables, with the source of every row, are in the
[qesR source repository](https://github.com/ThomasGareau/qesR).
